### 2013-05-24 fluB.tex
Initial commit of manuscript.
Very rough sketch of manuscript. Design completely plagiarized from Trevor (because it's cool) ^_^.

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color} 
\usepackage{parskip}
\usepackage{float}

% text layout
\usepackage{geometry}
\geometry{textwidth=15cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace} 
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}

% comments
\usepackage{ulem}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}} 
\def\tbc#1{\textcolor{purple}{[#1]}}

%%% TITLE %%%
\title{\vspace{1.0cm} \LARGE \bf fluB}

\author[1]{Gytis Dudas}
\author[1]{Trevor Bedford}
\author[1]{Samantha Lycett}
\author[1,2]{Andrew Rambaut}

\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
\affil[2]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA.}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}

Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages (B/Victoria and B/Yamagata subtypes) on the basis of haemagglutinin inhibition (HI) assays. Frequent reassortment of the two subtypes has been noted in the past, but the effects of these reassortments have not been investigated in much detail.

\end{abstract}

\pagebreak

\section*{Introduction}
Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three species (influenza A, B and C) co-circulating in humans. Though lacking in antigenic diversity, influenza B viruses are increasingly being recognized as being an important economic burden \cite{paul-glezen2013}. In the 2012/2013 influenza season in Europe as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.

As members of \textit{Orthomyxoviridae}, all influenza viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains. The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000}) and thus the antigenic diversity of influenza B viruses is considered to be low. It has been suggested that influenza B viruses use a combination of reassortment and amino acid insertion-deletion to generate antigenic novelty.

Influenza B viruses are considered to be comprised of two antigenically distinct lineages - Victoria and Yamagata (referred to as Vic and Yam, respectively) which are named after the strains (B/Victoria/02/87 and B/Yamagata/16/88, respectively) with antigenically distinct haemagglutinin (HA) surface glycoproteins. The two HA lineages still co-circulate today and are thought to have shared a most recent common ancestor in xxxx. All other influenza B segments also show a split in xxxx (comment - our estimate \textit{circa} 1984, use a figure to show TMRCAs and rates for each segment?).

\section*{Methods}

\subsection*{Sequences and subtyping}
We compiled a dataset of 452 complete influenza B genomes from GISAID (comment - need acknowledgment tables). Each strain was subtyped (assigned to either Vic or Yam subtype) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS), where B/Norway/1/84 was considered as being representative of Victoria lineage.
Each strain was thus assigned 8 subtypes depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/02/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y). Seven of these where then used as discrete states in each of the trees (e.g. PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values) in BEAST (comment - reference BEAST). (comment - give figure with proportions of subtypes for each segment)

\subsection*{Building trees}
For each segment it was attempted to make 3 replicates with 200 million MCMC states (comment - only third replicates ran for that long) each under the HKY model of nucleotide substitutions, 3 codon partitions, robust counting (comment - reference robust counting)) under the GMRF Bayesian skyride demographic model (comment - reference skyride).
In addition we used "relaxed" tips (comment - \# of relaxed tips) for when the dates of isolation for strains were incomplete (i.e. when only year of isolation was available).
We inferred the ancestral subtypes of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS (comment - cite Philippe)). Because the posterior set of trees are labelled, we can infer intersubtype reassortments whenever a trait transition is observed (i.e. Y to V or V to Y). In addition, by reconstructing the subtypes of all other genomic segments jointly we can infer co-reassortment events when trait transitions occur on the same nodes in each tree.
We use a measure of persistence which is defined as the ratio of distances (in units of time) between a reassortant node and its most recent tip and compare it to the mean of distances between all nodes and their respective most recent tips. If this statistic is below 1 a reassortment event is presumed to be disadvantageous, if it is above 1 it is presumed it is advantageous.

\subsection*{Building a consensus set of reassortment events}
We built a consensus set of reassortment events by going through each tree in the posterior set of trees, identifying trait changes which are indicators of reassortment (with a cutoff of 1989 to avoid including the initial split between Vic and Yam as a reassortment event), taking the persistence value as the time between the date of reassortment (assumed to have occurred somewhere along the branch connecting nodes with opposite trait values) and the date of the most recent tip descended from that node. This was repeated for all nodes on the tree and the values averaged to give the average persistence time of nodes in the tree. We also noted which tips were descended from reassortant nodes.

We combined this list of reassortant clades across all posterior sets of segment trees based on descendant tips. There were a total of 3321 unique reassortant clades observed across all sets of posterior trees (comment - need a graph to show that most of the probability density is on a few well supported events or use different cutoffs?). By looking at only the nodes which occurred at least 20\% of the time in each tree and had a reassortment mapping that was found at least 50\% of the time, we recovered 29 "consensus" reassortment events.

\subsection*{Measure of diversity}
We inferred the diversity of each segment over time by taking the total distance between the most divergent branches on the tree at yearly intervals from 1986 to 2011.

\subsection*{Work in progress}
Currently running "null" traited trees where trait maps have been rearranged (keeping the same ratio of trait values).

\section*{Results}

\begin{figure}[h]
  \centering		
	\includegraphics[width=0.95\textwidth]{figures/InfB_diversityOverTime}
	\caption{\textbf{Provisional diversity over time plot (tip to tip rather than branch to branch).}}
\end{figure}

\bibliographystyle{plos}
\bibliography{fluB}
\end{document}

### 2013-05-26 fluB.tex
Trying to get this to work properly :D.
This should work. All the stuff that's been done so far going up on
github. Very provisional so far.
Adding line breaks after sentences.

@@
 The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000}) and thus the antigenic diversity of influenza B viruses is considered to be low. 
-It has been suggested that influenza B viruses use a combination of reassortment and amino acid insertion-deletion to generate antigenic novelty.
+It has been suggested that influenza B viruses use a combination of reassortment and amino acid insertion-deletion to generate antigenic novelty (comment - citation).
 

### 2013-05-27 fluB.tex
Changing "subtype" to "lineage", a few extra changes.

@@
 Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
-Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages (B/Victoria and B/Yamagata subtypes) on the basis of haemagglutinin inhibition (HI) assays. 
-Frequent reassortment of the two subtypes has been noted in the past, but the effects of these reassortments have not been investigated in much detail.
+Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages (referred to as B/Victoria and B/Yamagata subtypes (comment - is subtype okay to use here?)) on the basis of haemagglutinin inhibition (HI) assays.
+Frequent reassortment between the segments of these two lineages has been noted in the past (comment - cite Lindstrom), but the effects of these reassortments have not been investigated in much detail.
 
@@
 
-Influenza B viruses are considered to be comprised of two antigenically distinct lineages - Victoria and Yamagata (referred to as Vic and Yam, respectively) which are named after the strains (B/Victoria/02/87 and B/Yamagata/16/88, respectively) with antigenically distinct haemagglutinin (HA) surface glycoproteins.
+Influenza B viruses are considered to be comprised of two antigenically distinct lineages - Victoria and Yamagata (referred to as Vic and Yam, respectively) which are named after two strains: B/Victoria/02/87 and B/Yamagata/16/88, respectively, which possess antigenically distinct haemagglutinin (HA) surface glycoproteins.
 The two HA lineages still co-circulate today and are thought to have shared a most recent common ancestor in xxxx. 
@@
 We compiled a dataset of 452 complete influenza B genomes from GISAID (comment - need acknowledgment tables). 
-Each strain was subtyped (assigned to either Vic or Yam subtype) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS), where B/Norway/1/84 was considered as being representative of Victoria lineage.
-Each strain was thus assigned 8 subtypes depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/02/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).
+Each strain was assigned to a lineage (either Vic or Yam subtype) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS), where B/Norway/1/84 was considered as being representative of Victoria lineage.
+Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/02/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).
 Seven of these where then used as discrete states in each of the trees (e.g.
@@
 In addition we used "relaxed" tips (comment - \# of relaxed tips) for when the dates of isolation for strains were incomplete (i.e. when only year of isolation was available).
-We inferred the ancestral subtypes of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS (comment - cite Philippe)).
-Because the posterior set of trees are labelled, we can infer intersubtype reassortments whenever a trait transition is observed (i.e.
+We inferred the ancestral state of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS (comment - cite Philippe)).
+Because the posterior set of trees are labelled, we can infer interlineage reassortments whenever a trait transition is observed (i.e.
 Y to V or V to Y).
-In addition, by reconstructing the subtypes of all other genomic segments jointly we can infer co-reassortment events when trait transitions occur on the same nodes in each tree.
-We use a measure of persistence which is defined as the ratio of distances (in units of time) between a reassortant node and its most recent tip and compare it to the mean of distances between all nodes and their respective most recent tips.
+In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when trait transitions occur on the same nodes in each tree.
+We use a measure of persistence which is defined as the ratio of distances (in units of time) between a reassortant node and its most recent tip (the event is sampled uniformly over the branch displaying a trait value that is different from its parent node) and compare it to the mean of distances between all nodes and their respective most recent tips.
 If this statistic is below 1 a reassortment event is presumed to be disadvantageous, if it is above 1 it is presumed it is advantageous.
@@
 \subsection*{Building a consensus set of reassortment events}
-We built a consensus set of reassortment events by going through each tree in the posterior set of trees, identifying trait changes which are indicators of reassortment (with a cutoff of 1989 to avoid including the initial split between Vic and Yam as a reassortment event), taking the persistence value as the time between the date of reassortment (assumed to have occurred somewhere along the branch connecting nodes with opposite trait values) and the date of the most recent tip descended from that node.
+We built a consensus set of reassortment events by going through each tree in the posterior set of trees, identifying trait changes which are indicators of reassortment (with a cutoff of 1989 to avoid including the initial split between Vic and Yam as a reassortment event), taking the persistence value as the time between the date of reassortment and the date of the most recent tip descended from that node.
 This was repeated for all nodes on the tree and the values averaged to give the average persistence time of nodes in the tree. 
-We also noted which tips were descended from reassortant nodes.
+We also noted which tips were descended from reassortant nodes to identify the same events in different segments.
 
@@
 \section*{Results}
+Reassortment events that involve a basic polymerase (PB1 or PB2) or HA, but never all 3, have 95\% HPDs that overlap 1 with the exception of a stray PB2-NA-NS events which is near 1 and has very low posterior probability.
+(comment - need a good way of representing all consensus events)
 
@@
 	\includegraphics[width=0.95\textwidth]{figures/InfB_diversityOverTime}
-	\caption{\textbf{Provisional diversity over time plot (tip to tip rather than branch to branch).}}
+	\caption{\textbf{Provisional diversity over time plot (tip to tip rather than branch to branch) (comment - slightly unusual that NS has quite wide HPD intervals at a few places, even though posterior smears don't show a bimodal distribution).}}
 \end{figure}

### 2013-05-31 fluB.tex
Replacing references to subtypes with references to lineages.

@@
 We compiled a dataset of 452 complete influenza B genomes from GISAID (comment - need acknowledgment tables). 
-Each strain was assigned to a lineage (either Vic or Yam subtype) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS), where B/Norway/1/84 was considered as being representative of Victoria lineage.
+Each strain was assigned to a lineage (either Vic or Yam) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS), where B/Norway/1/84 was considered as being representative of Victoria lineage.
 Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/02/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).
@@
 PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values) in BEAST (comment - reference BEAST).
-(comment - give figure with proportions of subtypes for each segment)
+(comment - give figure with proportions of each lineage for each segment)
 

### 2013-07-23 fluB.tex
Very rough draft commit.
Committing what's been written so far in preparation for holiday
writing. Be warned - it's very ramble-y and vague in places. Also
missing most figures.

@@
 Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages (referred to as B/Victoria and B/Yamagata subtypes (comment - is subtype okay to use here?)) on the basis of haemagglutinin inhibition (HI) assays.
-Frequent reassortment between the segments of these two lineages has been noted in the past (comment - cite Lindstrom), but the effects of these reassortments have not been investigated in much detail.
+Frequent reassortment between the segments of these two lineages has been noted in the past (comment - cite Lindstrom et al 1999), but the effects of these reassortments have not been investigated in much detail.
 
@@
 Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three species (influenza A, B and C) co-circulating in humans.
-Though lacking in antigenic diversity, influenza B viruses are increasingly being recognized as being an important economic burden \cite{paul-glezen2013}.
-In the 2012/2013 influenza season in Europe as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.
+Most of these are attributed to influenza A viruses, which are considered to cause the most morbidity and mortality attributed to seasonal influenza.
+Though lacking the antigenic diversity that influenza A viruses possess, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
+Following the 2009 A/H1N1 pandemic influenza B viruses have circulated more widely than previously (note - reword this) and in the 2012/2013 influenza season in Europe as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.
 
-As members of \textit{Orthomyxoviridae}, all influenza viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
+Like all members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
 Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains.
-The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000}) and thus the antigenic diversity of influenza B viruses is considered to be low. 
-It has been suggested that influenza B viruses use a combination of reassortment and amino acid insertion-deletion to generate antigenic novelty (comment - citation).
+The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000}) and thus the antigenic diversity of influenza B viruses circulating in humans is considered to be low.
+It has been suggested that influenza B viruses use a combination of reassortment and amino acid insertion-deletions to generate antigenic diversity (comment - citation).
 
-Influenza B viruses are considered to be comprised of two antigenically distinct lineages - Victoria and Yamagata (referred to as Vic and Yam, respectively) which are named after two strains: B/Victoria/02/87 and B/Yamagata/16/88, respectively, which possess antigenically distinct haemagglutinin (HA) surface glycoproteins.
+Currently circulating influenza B viruses are comprised of two antigenically distinct lineages - Victoria and Yamagata (referred to as Vic and Yam, respectively) which are named after two strains: B/Victoria/02/87 and B/Yamagata/16/88, respectively, which possess antigenically distinct haemagglutinin (HA) surface glycoproteins.
 The two HA lineages still co-circulate today and are thought to have shared a most recent common ancestor in xxxx. 
-All other influenza B segments also show a split in xxxx (comment - our estimate \textit{circa} 1984, use a figure to show TMRCAs and rates for each segment?).
+All other influenza B segments also show a split in xxxx (comment - our estimate \textit{circa} 1984 figure \ref{TMRCAhpd}, use a figure to show TMRCAs and rates for each segment?), although the numbers of strains falling on either side of the split are different due to reassortments following the initial split of Yamagata and Victoria lineages in all segments.
+
+%\begin{figure}[h]
+%	\centering		
+%	\includegraphics[width=0.95\textwidth]{figures/InfBtmrcaHPD}
+%	\caption{\textbf{Provisional TMRCA plot.}}
+%	\label{TMRCAhpd}
+%\end{figure}
+
+Previously, through the use of time-stamped phylogenetic trees, the complete history of influenza B reassortment events has been reconstructed (comment - cite Chen \& Holmes 2008), and showed extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
+Another approach exists, however, whereby belonging to either the Victoria or Yamagata lineage in the tree of one segment can be used as a discrete trait in trees of other segments.
+Reassortment events which result in the replacement of one segment lineage by another show up as changes in the discrete trait along a branch and allow the inference of parameters related to the subtree descended from the node displaying a changed trait.
+Here, we use discrete traits to model inter-lineage reassortments between Victoria and Yamagata lineage segments in BEAST to look at lineage persistence times following reassortment events. 
+We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Victoria or the Yamagata lineage are selected against.
+In other segments (PA, NP, NA, MP and NS) a single lineage has been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
 
@@
 We compiled a dataset of 452 complete influenza B genomes from GISAID (comment - need acknowledgment tables). 
-Each strain was assigned to a lineage (either Vic or Yam) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS), where B/Norway/1/84 was considered as being representative of Victoria lineage.
+Each strain was assigned to a lineage (either Vic or Yam) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS), where B/Norway/1/84 (comment - B/Czechoslovakia/69/1990 could be used as well, since it is the only strain showing an entirely Vic genome) was considered as being representative of Victoria lineage.
 Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/02/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).
-Seven of these where then used as discrete states in each of the trees (e.g.
-PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values) in BEAST (comment - reference BEAST).
-(comment - give figure with proportions of each lineage for each segment)
+Seven of these where then used as discrete states in each of the trees (e.g. PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values) in BEAST (comment - reference BEAST).
 
@@
 For each segment it was attempted to make 3 replicates with 200 million MCMC states (comment - only third replicates ran for that long) each under the HKY model of nucleotide substitutions, 3 codon partitions, robust counting (comment - reference robust counting)) under the GMRF Bayesian skyride demographic model (comment - reference skyride).
-In addition we used "relaxed" tips (comment - \# of relaxed tips) for when the dates of isolation for strains were incomplete (i.e. when only year of isolation was available).
+In addition we used "relaxed" tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation).
 We inferred the ancestral state of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS (comment - cite Philippe)).
-Because the posterior set of trees are labelled, we can infer interlineage reassortments whenever a trait transition is observed (i.e.
-Y to V or V to Y).
-In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when trait transitions occur on the same nodes in each tree.
+Because the posterior set of trees have branches labelled with the inferred lineage in all other segments, we can detect inter-lineage reassortments whenever a trait transition is observed (i.e.
+Y to V or V to Y) between a node and its descendant in any segment.
+In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
+
 We use a measure of persistence which is defined as the ratio of distances (in units of time) between a reassortant node and its most recent tip (the event is sampled uniformly over the branch displaying a trait value that is different from its parent node) and compare it to the mean of distances between all nodes and their respective most recent tips.
-If this statistic is below 1 a reassortment event is presumed to be disadvantageous, if it is above 1 it is presumed it is advantageous.
+If this statistic is below 1 a reassortment event is presumed to be disadvantageous, if it is above 1 the event is presumed to be advantageous.
+
+Alternative - we start by allocating the tips in the tree to yearly bins starting from 1984 and ending in 2013.
+
+At each state of the MCMC chain we choose a random tip from each bin and traverse the tree towards the root until a reassortment event (i.e. a trait switch) is encountered.
+We log the traits which have transitioned at the reassortant node (i.e.
+the reassorting segment), the direction of transition (e.g.
+whether the parental node was V and the child node Y or vice versa) and the amount of time separating the tip and the node displaying a trait transition.
+The time taken to traverse back to a reassortment event is expected to be longer for advantageous reassortments than for non-advantageous reassortments.
 
 \subsection*{Building a consensus set of reassortment events}
-We built a consensus set of reassortment events by going through each tree in the posterior set of trees, identifying trait changes which are indicators of reassortment (with a cutoff of 1989 to avoid including the initial split between Vic and Yam as a reassortment event), taking the persistence value as the time between the date of reassortment and the date of the most recent tip descended from that node.
-This was repeated for all nodes on the tree and the values averaged to give the average persistence time of nodes in the tree. 
-We also noted which tips were descended from reassortant nodes to identify the same events in different segments.
+Maximum clade credibility trees were made after discarding the first 1000 trees as burnin. 
+Nodes with trait values different from the trait value of their parental node were classified as being reassortants.
+The persistence time, descendant tips and reassorting segments were logged for each reassortant node and aggregated across analyses based on descendant tips.
 
-We combined this list of reassortant clades across all posterior sets of segment trees based on descendant tips. 
-There were a total of 3321 unique reassortant clades observed across all sets of posterior trees (comment - need a graph to show that most of the probability density is on a few well supported events or use different cutoffs?).
-By looking at only the nodes which occurred at least 20\% of the time in each tree and had a reassortment mapping that was found at least 50\% of the time, we recovered 29 "consensus" reassortment events.
-
-\subsection*{Measure of diversity}
-We inferred the diversity of each segment over time by taking the total distance between the most divergent branches on the tree at yearly intervals from 1986 to 2011.
+\subsection*{Measures of diversity}
+We inferred the diversity of each segment over time by finding the oldest most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on diversity existing at each time period.
+In addition, we calculated mean pairwise diversity and mean pairwise TMRCA between branches inferred to be in virions possessing Victoria and Yamagata lineage PB1, PB2 and HA.
 
@@
 \section*{Results}
-Reassortment events that involve a basic polymerase (PB1 or PB2) or HA, but never all 3, have 95\% HPDs that overlap 1 with the exception of a stray PB2-NA-NS events which is near 1 and has very low posterior probability.
-(comment - need a good way of representing all consensus events)
+
+\subsection*{PA, NP, NA, MP and NS have periodically lost diversity}
+From TMRCA plots (see Figure \ref{tmrcaOT}) it is clear that diversity in PB1, PB2 and HA has been maintained since the initial split of Victoria and Yamagata lineages, whereas other segments (most notably NS) have repeatedly lost lineages.
+The plot shows the initial losses of diversity in PA, NP and MP segments in 1997 (x axis), indicative of the extinction of Victoria lineages of those segments and revealing that a reassortment had occured which meant that all sampled lineages circulating around 1997 shared a common ancestor around the year 1990 (y axis).
+By 2001, the Yamagata lineage of the NS segment had gone extinct as well, revealing that NS too had reassorted together with PA, NP and MP segment around the year 1990 (similar TMRCA along the y axis to previous reassortants).
+Another round of reassortment-related extinction took place since then in PA, MP and NA segments and another 3 in the NS segment.
+Unlike all other segments, however, all lineages of PB1, PB2 and HA segments have maintained diversity and lineages circulating in 2011 were still derived from the original split of the Victoria and Yamagata lineages.
+
+These differences in segment diversity are reflected in ratios of Victoria to Yamagata lineages in different segments: for the PB1, PB2 and HA segments the ratio of sequences belonging to either Victoria or Yamagata lineages are close to 0.5 (see Figure \ref{lineageRatios}, whereas all other segments are heavily skewed towards Yamagata lineage with the exception of NS segment which has more Victoria lineage sequences.
 
@@
 	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_diversityOverTime}
-	\caption{\textbf{Provisional diversity over time plot (tip to tip rather than branch to branch) (comment - slightly unusual that NS has quite wide HPD intervals at a few places, even though posterior smears don't show a bimodal distribution).}}
+	\includegraphics[width=0.95\textwidth]{figures/InfB_tmrcaOT_lines.png}
+	\caption{\textbf{TMRCA over time plot.}}
+	\label{tmrcaOT}
 \end{figure}
+
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/InfB_LineageRatios.png}
+	\caption{\textbf{Ratio of lineages in the dataset.}}
+	\label{lineageRatios}
+\end{figure}
+
+\subsection*{Diversity in PB1, PB2 and HA segments is not maintained independently}
+We find that the diversity preserved in PB1, PB2 and HA segments is not preserved independently in each segment.
+Plot XXXX (comment - make date-spectrum figure) shows the posterior distributions of inferred dates of reassortment.
+Following 1993 PB1, PB2 and HA reassortments are inferred to occur simultaneously at the same nodes.
+
+\subsection*{Numbers of reassortments and their persistence}
+
+(comment - this bit is for stratified REPU results)
+We find that on average, inter-lineage reassortments are common (number of trait switches observed: 116 after filtering, but some will be the same event, naive underestimate is 14.5 inter-lineage reassortments per segment) and favoured by natural selection (persistence statistic > 1, see Figure \ref{stratifiedREPUgraph} (comment - make figure prettier)), although this varies by segments involved and date of reassorment.
+
+In particular, reassortment events where same-lineage PB1, PB2 and HA segments reassort together have much better persistence than events where they reassort independently.
+Conversely, reassortments that produce mixed-lineage PB1-PB2-HA complexes have persistence statistic lower than an average node existing at the same time, with the exception of NS.
+We see reassortment events preserving the Victoria lineage PB1-PB2-HA complex occuring between 1990 and 1995 in PA, NP and MP (comment - if NS trees were better, there would be a PB1-PB2-HA preserving event in NS segment), coresponding to the origins of the reassortant lineage that eventually gave rise to B/Hong Kong/330/2001-like influenza B viruses, which were the first Victoria lineage HA-bearing viruses detected outside of Asia since 1990s (comment - cite paper).
+However, the persistence of segment lineages comprising these B/Hong Kong/330/2001-like reassortant viruses vary dramatically.
+Whereas the PA and MP lineages that were part of the reassortant B/Hong Kong/330/2001-like constellation had poor persistence, the NP lineage was "re-used" during the next inter-lineage reassortment occuring around 2000.
+
+The reassortment event occuring around the year 2000 (which produced B/Iowa/03/2002-like viruses), involved the PA, NA, MP and NS segments of Yamagata lineage reassorting with Victoria lineage PB1, PB2 and HA (the latter virus already contained a Yamagata lineage NP) with good persistence (over 2 times the persistence of all other nodes existing at the time).
+The greater persistence of PA, NA, MP and NS lineages is due to having left descendants close to the present (essentially becoming the "trunk" lineages of those segments).
+
+Additional reassortments occured in 2004 and 2006.
+In 2004 NP, MP and NS segments of Yamagata lineage reassorted with Victoria lineage PB1, PB2 and HA (with Yamagata lineage PA and NA).
+This event is one of the most interesting ones, since the NS segment of Victoria lineage PB1-PB2-HA complex, though ancestrally derived from Victoria lineage, was replaced by a Victoria lineage NS segment which had been evolving together with genome constellations derived from Yamagata lineage segments.
+Although the MP segment lineage reassorted in the 2004 event eventually became the "trunk" of all Victoria PB1-PB2-HA lineage bearing viruses, the NP(Yam) lineage was replaced by another one (also Yam) that was reassorted into the PB1-PB2-HA Victoria lineage-bearing background in 2006.
+
+From reassortment event persistence times it is clear that reassortment events that break up the PB1-PB2-HA complex have occured in the past but never persisted for long.
+We observe a sudden stop of PB1-PB2-HA complex-separating reassortment events after 1995 which are followed solely by PB1-PB2-HA complex-preserving reassortment events.
+
+(comment - this bit is for the tip descent method)
+On average it takes nearly 2 years longer for a random tip at any given point to trace its ancestry to a reassortant lineage where a PB1, PB2 and HA segments of the same lineage have reassorted together, although this varies by segment and time of reassortment.
+
+\subsection*{Differentation of segments along PB1, PB2 and HA lineages}
+(comment - this will be referring to 3 plots, which are mean pairwise tmrcas between branches under PB1, PB2 and HA traits, respectively)
+
+When comparing TMRCAs between branches of segment trees inferred to be under either Victoria or Yamagata lineage PB1, PB2 or HA, it is clear that diversity along the lineage axis has been preserved in PB1, PB2 and HA, whereas all other segments (PA, NP, NA, MP and NS) have repeatedly lost diversity, indicating reassortments with PB1, PB2 and HA segments.
+Interestingly, the diversity patterns between segments under inferred PB1, PB2 or HA lineages differ early on and this can be shown to be due to short lived reassortants with mixed-lineage PB1, PB2 and HA segments, namely the VVY (B/Bankok/163/1990-like), YVV (B/Nanchang/630/1994-like) and YVY (B/New York/24/1993-like) reassortants (comment - will require trees in supplementary info).
+Starting from about 1993, however, PB1, PB2 and HA segments of Victoria lineage have only reassorted as one with other segments of Yamagata lineage which is reflected in plots becoming identical, regardless of segment, post-1993.
+
+\section*{Discussion}
+
+\subsection*{Limitations of current study}
+The methods used in this paper have limited power to detect reassortments, since lineages have been fixed over time and thus all viruses eventually come to bear either Vic or Yam lineage PA, NP, NA, MP and NS segments.
+We can also only detect reassortment events where a lineage has switched from Vic to Yam or \textit{vice versa}, but not when a lineage has switched from a Yam lineage to another Yam lineage, which has occurred in the past (e.g.
+the 2004 Vic lineage PB1-PB2-HA complex reassortment with NP is only visible in the NP tree, since all viruses in our dataset from 1995 onwards, including Vic PB1-PB2-HA complex-bearing ones, had a Yamagata lineage-derived NP).
+
+\subsection*{Reassortment in influenza B is common and extensive}
+Extensive reassortment has long been established as a major pathway by which influenza B viruses generate diversity.
+Even focusing on a specific subset of reassortments (those that occur between two lineages and those where the previous lineage was of the opposite lineage) reveals 116 events and if we naively assume that all events are replicated across all trees it still means there have been over 14 inter-lineage reassortment events.
+We also see that reassortments occur across considerable phylogenetic distances and some segment lineages have been so successful as part of reassortant viruses that they out-competed their sister lineages (e.g.
+Yamagata lineages of PA, NP, MP and NA segments).
+(comment - could include diversity over time plots to show maximum phylogenetic distance attained prior to reassortment for each segment)
+
+\subsection*{Selection maintains the diversity in PB1-PB2-HA complexes of Victoria and Yamagata lineages}
+Rather surprisingly, 3 segments have maintained a ratio of Vic and Yam lineages very close to 0.5 (PB1, PB2 and HA) (see Figure \ref{lineageRatios}) and have preserved the diversity that has accumulated since the split of the two lineages (see Figure \ref{tmrcaOT}).
+Our results suggest that this is due to the 3 segments co-reassorting (comment - refer to posterior distributions of reassortant dates plot).
+We find evidence that this co-reassortment pattern between PB1, PB2 and HA is not due to skewed reassortment ((comment - add Marshall et al 2013 citation) have shown that reassortment efficiency is high in influenza A), but rather due to viruses bearing mixed-lineage PB1-PB2-HA complex having lower fitness than viruses with PB1-PB2-HA complexes derived entirely from Yam or Vic lineages (comment - refer to stratified REPU plot).
+Not only are events that split up pure-lineage PB1-PB2-HA complexes rare, but they also have poor persistence and trees have spent less than 10\% of the time in the mixed-lineage PB1-PB2-HA complex state, suggesting that they are selected against (comment - make plot of that).
+
+Thus our findings suggest that the link maintaining PB1, PB2 and HA lineages close to equilibrium is not due to biased reassortment, but a result of selection acting upon reassortants containing mixed-lineage PB1-PB2-HA segments.
+We find that by 1995 sufficient co-dependence had developed between PB1, PB2 and HA segments of Victoria and Yamagata lineages to prevent reassortant progeny containing mixed-lineage PB1-PB2-HA complexes from infecting sufficient numbers of individuals to cross the detection threshold.
+
+\subsection*{Segments vary in their abilities to reassort and success as part of reassortant viruses}
+A rather striking pattern in influenza B evolution is the fact that segments reassorted at the same time can have dramatically different fates.
+For example, PA, NP, MP and NS segments of Yamagata lineage were reassorted into Victoria lineage PB1-PB2-HA background in 1994, but of those 4 segments only the NP lineage persisted for a considerable amount of time, being replaced by another NP lineage in 2006, whereas PA was replaced in 2000 and MP together with NS were replaced by different lineages in 2004.
+This suggests that most non-PB1, PB2 or HA segments have the capacity to evolve somewhat independently of the rest of the segments present in the same virion.
+
+\subsection*{Interpreting our results}
+(comment - this bit is exceedingly unsubstantiated and ramble-y)
+Recently the antigenic drift of influenza B viruses has been characterised (comment - cite Bedford et al 2013) and Victoria lineage HA was found to drift nearly twice as fast as Yamagata lineage HA.
+A possible explanation for this would be differences in polymerase fidelity: if Victoria lineage polymerase introduced higher mutation rates, this would explain the higher antigenic drift rate seen in Vic lineage HA, but also would suggest that Vic lineage PB1-PB2-HA complex "borrows" segments from Yamagata lineage PB1-PB2-HA complex-bearing viruses and "breaks" them through going over something akin to each segment's error threshold, necessitating another round of reassortment.
+However, we find no evidence of differences in nucleotide, synonymous and non-synonymous rates of evolution between branches that were inferred to have been in the same virion as PB1, PB2 or HA segments of Vic or Yam lineages.
+Thus we have no evidence to suggest fundamental differences between Vic and Yam lineage PB1-PB2-HA complexes.
+
+Balancing selection would be expected, and does in the malaria parasite, preserve antigenic diversity, but it is unclear if maintenance of PB1-PB2-HA lineages is driven by balancing selection acting to preserve antigenic diversity of HA or the diversity of PB1 and PB2 segments.
+Recently (comment - cite Cobbin et al 2013) a similar link has been described in influenza A by XXXX, where the PB1 segment is reassorted more frequently than expected during vaccine strain creation.
+If influenza B virus PB1 has a similar gene-regulatory activity, then it would be reasonable to assume that HA and PB1 are associated through the necessity for HA to have appropriate levels of expression and PB2 hitchhikes together with PB1 due to some functional requirement as part of the polymerase complex.
+This does not explain why the Vic lineage of the PA segment was lost so readily, however, despite it being a part of the polymerase complex.
+
+\subsection*{Future evolution of influenza B viruses}
+From our current findings it is difficult to extropolate what reassortants will come to dominate the influenza B population in the far future.
+It seems likely that PB1, PB2 and HA segments of both Victoria and Yamagata lineages will either continue to co-exist in their co-adapted state or one lineage will go extinct.
+Until the 2004 and 2006 reassortment events it seemed like the NP segment would follow the PB1-PB2-HA complex in becoming the next segment to be recruited as part of a co-adapted segment complex.
+This method of gradually recruiting more segments to a co-adapted segment complex could have lead to the eventual speciation of influenza B into originally Vic and Yam lineage HA bearing species.
+As it stands now, however, the most diversity between Victoria and Yamgata PB1-PB2-HA lineage-bearing viruses has been built up in PA and NA segments (which shared a common ancestor around 2000) and it remains to be seen whether this diversity will too be lost following a reassortment event.
+
+Given the evolutionary history of the NP segment, it would be reasonable to focus sequencing efforts on complete influenza B genome sequencing.
+Our results indicate that successful reassortants are likely to come from the combination of Victoria lineage PB1-PB2-HA complex and PA, NP, NA and MP segments of Yamagata lineage (or NS segment of Vic lineage), though the number of Yamagata lineage segments can vary.
+It is also obvious that increased surveillance efforts will inevitably uncover a lot of reassortant viruses (possibly even viruses with mixed-lineage PB1-PB2-HA complexes) and the threshold at which these reassortants should be considered as being likely to take over the influenza B population would require further work.
+
+\subsection*{Testing the co-dependence between PB1, PB2 and HA segments}
+Our results suggest a simple test for the hypothesis that PB1, PB2 and HA segments of Victoria and Yamagata lineages have become co-dependent - if PB1, PB2 or HA segments from different lineages were combined artificially in a single virion, these should have lower fitness than virions containing wholly Victoria or wholly Yamagata lineage PB1-PB2-HA complex.
+We also expect that the fitness of artificially derived reassortants would decline with increasing date of strain isolation i.e.
+mixed-lineage PB1-PB2-HA containing reassortants with segments from older isolates having higher fitness than reassortants possessing segments of more recent isolates.
+
+\subsection*{Implications for influenza B surveillance and prevention}
+The findings presented here have several implications.
+On one hand, if the co-dependence between PB1, PB2 and HA segments is strong enough, influenza B viruses containing mixed-lineage PB1-PB2-HA complex detected by surveillance are unlikely to contribute to long-term evolution of influenza B viruses.
+On the other hand, the dependence of the HA segment on PB1 and PB2 segments (or vice versa) implies that more subtle disease control measures can be developed and used against influenza B viruses.
+Finally, the extent of inter-lineage reassortment argues for more extensive surveillance of influenza B viruses, as sequences of a handful of segments will not necessarily tell the whole story, especially since influenza B viruses have been known to evolve counter-intuitively (e.g.
+losing Victoria lineages of PA and NA segments despite their associations with the rest of the polymerase complex or role in antagonising HA activity, respectively).
+This study also highlights the gaps in our knowledge of the molecular biology of influenza B viruses.
+
+%\begin{figure}[h]
+%	\centering		
+%	\includegraphics[width=0.95\textwidth]{figures/InfB_TMRCAOverTime}
+%	\caption{\textbf{Provisional TMRCA over time plot.}}
+%\end{figure}
 

### 2013-07-23 fluB.tex
Minor changes.

@@
 Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages (referred to as B/Victoria and B/Yamagata subtypes (comment - is subtype okay to use here?)) on the basis of haemagglutinin inhibition (HI) assays.
-Frequent reassortment between the segments of these two lineages has been noted in the past (comment - cite Lindstrom et al 1999), but the effects of these reassortments have not been investigated in much detail.
+Frequent reassortment between the segments of these two lineages has been noted in the past \cite{lindstrom1999}, but the effects of these reassortments have not been investigated in much detail.
 
@@
 
-Previously, through the use of time-stamped phylogenetic trees, the complete history of influenza B reassortment events has been reconstructed (comment - cite Chen \& Holmes 2008), and showed extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
+Previously, through the use of time-stamped phylogenetic trees, the complete history of influenza B reassortment events has been reconstructed \cite{chen2008}, and showed extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
 Another approach exists, however, whereby belonging to either the Victoria or Yamagata lineage in the tree of one segment can be used as a discrete trait in trees of other segments.
@@
 
-Alternative - we start by allocating the tips in the tree to yearly bins starting from 1984 and ending in 2013.
-
+Alternative:
+We start by allocating the tips in the tree to yearly bins starting from 1984 and ending in 2013. 
 At each state of the MCMC chain we choose a random tip from each bin and traverse the tree towards the root until a reassortment event (i.e. a trait switch) is encountered.
@@
 
+(comment - need higher resolution for lineage ratios figure or switch to vector graphics)
 \begin{figure}[h]
@@
 Our results suggest that this is due to the 3 segments co-reassorting (comment - refer to posterior distributions of reassortant dates plot).
-We find evidence that this co-reassortment pattern between PB1, PB2 and HA is not due to skewed reassortment ((comment - add Marshall et al 2013 citation) have shown that reassortment efficiency is high in influenza A), but rather due to viruses bearing mixed-lineage PB1-PB2-HA complex having lower fitness than viruses with PB1-PB2-HA complexes derived entirely from Yam or Vic lineages (comment - refer to stratified REPU plot).
+We find evidence that this co-reassortment pattern between PB1, PB2 and HA is not due to skewed reassortment (Marhsall \textit{et al.} \cite{marshall2013} have shown that reassortment efficiency is high in influenza A), but rather due to viruses bearing mixed-lineage PB1-PB2-HA complex having lower fitness than viruses with PB1-PB2-HA complexes derived entirely from Yam or Vic lineages (comment - refer to stratified REPU plot).
 Not only are events that split up pure-lineage PB1-PB2-HA complexes rare, but they also have poor persistence and trees have spent less than 10\% of the time in the mixed-lineage PB1-PB2-HA complex state, suggesting that they are selected against (comment - make plot of that).
@@
 Balancing selection would be expected, and does in the malaria parasite, preserve antigenic diversity, but it is unclear if maintenance of PB1-PB2-HA lineages is driven by balancing selection acting to preserve antigenic diversity of HA or the diversity of PB1 and PB2 segments.
-Recently (comment - cite Cobbin et al 2013) a similar link has been described in influenza A by XXXX, where the PB1 segment is reassorted more frequently than expected during vaccine strain creation.
+Recently a similar link has been described in influenza A by Cobbin \textit{et al.} \cite{cobbin2013}, where the PB1 segment is reassorted more frequently than expected during vaccine strain creation.
 If influenza B virus PB1 has a similar gene-regulatory activity, then it would be reasonable to assume that HA and PB1 are associated through the necessity for HA to have appropriate levels of expression and PB2 hitchhikes together with PB1 due to some functional requirement as part of the polymerase complex.
@@
 On one hand, if the co-dependence between PB1, PB2 and HA segments is strong enough, influenza B viruses containing mixed-lineage PB1-PB2-HA complex detected by surveillance are unlikely to contribute to long-term evolution of influenza B viruses.
-On the other hand, the dependence of the HA segment on PB1 and PB2 segments (or vice versa) implies that more subtle disease control measures can be developed and used against influenza B viruses.
-Finally, the extent of inter-lineage reassortment argues for more extensive surveillance of influenza B viruses, as sequences of a handful of segments will not necessarily tell the whole story, especially since influenza B viruses have been known to evolve counter-intuitively (e.g.
-losing Victoria lineages of PA and NA segments despite their associations with the rest of the polymerase complex or role in antagonising HA activity, respectively).
-This study also highlights the gaps in our knowledge of the molecular biology of influenza B viruses.
+On the other hand, the dependence of the HA segment on PB1 and PB2 segments (or \textit{vice versa}) implies that there is a possibility that more subtle disease control measures could be developed and used against influenza B viruses, or, if similar co-adapted segment complexes were discovered in influenza A, seasonal influenza in general.
+In addition, the extent of inter-lineage reassortment argues for more intensive surveillance of influenza B viruses, as sequences of a handful of segments will not necessarily tell the whole story, especially since influenza B viruses have been known to evolve counter-intuitively (\textit{e.g.} losing Victoria lineages of PA and NA segments despite their associations with the rest of the polymerase complex or role in antagonising HA activity, respectively).
+Finally, this study also highlights the gaps in our knowledge about the molecular biology of influenza B viruses.
 
@@
 
-\bibliographystyle{plos}
+\bibliographystyle{pnas}
 \bibliography{fluB}

### 2013-07-24 fluB.tex
Minor changes to text, new figure.
Adding a figure showing the total time spent in "pure" vs "mixed"
PB1-PB2-HA.

@@
 We find evidence that this co-reassortment pattern between PB1, PB2 and HA is not due to skewed reassortment (Marhsall \textit{et al.} \cite{marshall2013} have shown that reassortment efficiency is high in influenza A), but rather due to viruses bearing mixed-lineage PB1-PB2-HA complex having lower fitness than viruses with PB1-PB2-HA complexes derived entirely from Yam or Vic lineages (comment - refer to stratified REPU plot).
-Not only are events that split up pure-lineage PB1-PB2-HA complexes rare, but they also have poor persistence and trees have spent less than 10\% of the time in the mixed-lineage PB1-PB2-HA complex state, suggesting that they are selected against (comment - make plot of that).
+Not only are events that split up pure-lineage PB1-PB2-HA complexes rare, but they also have poor persistence and trees have spent less than 10\% of the time in the mixed-lineage PB1-PB2-HA complex state (Figure \ref{stateTime}), suggesting that they are selected against (comment - make plot of that).
+
+(comment - PB1, PB2 and HA are missing extra years in the mixed-lineage bar due to invisible reassortments e.g.
+VVY is invisible in HA tree, YVV in PB1 and YVY in PB2; solution: remove PB1, PB2 and HA bars or make it clear in the text)
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/InfB_stateTime.png}
+	\caption{\textbf{Total time spent in "pure" vs "mixed" PB1-PB2-HA}}
+	\label{stateTime}
+\end{figure}
 
@@
 For example, PA, NP, MP and NS segments of Yamagata lineage were reassorted into Victoria lineage PB1-PB2-HA background in 1994, but of those 4 segments only the NP lineage persisted for a considerable amount of time, being replaced by another NP lineage in 2006, whereas PA was replaced in 2000 and MP together with NS were replaced by different lineages in 2004.
-This suggests that most non-PB1, PB2 or HA segments have the capacity to evolve somewhat independently of the rest of the segments present in the same virion.
+This suggests that most non-PB1, PB2 or HA segments have the capacity to evolve somewhat independently of the rest of the genome.
 
@@
 On the other hand, the dependence of the HA segment on PB1 and PB2 segments (or \textit{vice versa}) implies that there is a possibility that more subtle disease control measures could be developed and used against influenza B viruses, or, if similar co-adapted segment complexes were discovered in influenza A, seasonal influenza in general.
-In addition, the extent of inter-lineage reassortment argues for more intensive surveillance of influenza B viruses, as sequences of a handful of segments will not necessarily tell the whole story, especially since influenza B viruses have been known to evolve counter-intuitively (\textit{e.g.} losing Victoria lineages of PA and NA segments despite their associations with the rest of the polymerase complex or role in antagonising HA activity, respectively).
+In addition, the extent of inter-lineage reassortment argues for more intensive surveillance of influenza B viruses, as sequences of a handful of segments will not necessarily tell the whole story, especially since influenza B viruses have been known to evolve counter-intuitively (\textit{e.g.} losing Victoria lineages of PA and NA segments despite associations with the rest of the polymerase complex or antagonising HA activity, respectively).
 Finally, this study also highlights the gaps in our knowledge about the molecular biology of influenza B viruses.

### 2013-07-24 fluB.tex
Minor comment.

@@
 Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
-Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages (referred to as B/Victoria and B/Yamagata subtypes (comment - is subtype okay to use here?)) on the basis of haemagglutinin inhibition (HI) assays.
+Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages (referred to as B/Victoria and B/Yamagata subtypes (comment - is subtype okay to use here?
+- \tbc{Unfortunately, `subtype' is not kosher to refer to Vic and Yam.
+It's annoying.
+The most common term in the literature is `lineage'.
+I suspect you could also get away with `virus', as in the `B/Vic virus'.})) on the basis of haemagglutinin inhibition (HI) assays.
 Frequent reassortment between the segments of these two lineages has been noted in the past \cite{lindstrom1999}, but the effects of these reassortments have not been investigated in much detail.

### 2013-07-24 fluB.tex
Minor revisions and comments.

@@
 \author[1]{Gytis Dudas}
-\author[1]{Trevor Bedford}
+\author[2]{Trevor Bedford}
 \author[1]{Samantha Lycett}
-\author[1,2]{Andrew Rambaut}
+\author[1,3]{Andrew Rambaut}
 
 \affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
-\affil[2]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA.}
+\affil[2]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA.}
+\affil[3]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA.}
 
@@
 \section*{Introduction}
-Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three species (influenza A, B and C) co-circulating in humans.
-Most of these are attributed to influenza A viruses, which are considered to cause the most morbidity and mortality attributed to seasonal influenza.
-Though lacking the antigenic diversity that influenza A viruses possess, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
-Following the 2009 A/H1N1 pandemic influenza B viruses have circulated more widely than previously (note - reword this) and in the 2012/2013 influenza season in Europe as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.
+Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three virus types (influenza A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality.
+Though lacking the degree of antigenic diversity that influenza A viruses possess, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
+Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.
 
@@
 Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains.
-The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000}) and thus the antigenic diversity of influenza B viruses circulating in humans is considered to be low.
-It has been suggested that influenza B viruses use a combination of reassortment and amino acid insertion-deletions to generate antigenic diversity (comment - citation).
+The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000}), thus limiting the virus's available antigenic reservoir. 
+It has been suggested that influenza B viruses use a combination of reassortment and point mutations to generate antigenic diversity (comment - citation).
 
-Currently circulating influenza B viruses are comprised of two antigenically distinct lineages - Victoria and Yamagata (referred to as Vic and Yam, respectively) which are named after two strains: B/Victoria/02/87 and B/Yamagata/16/88, respectively, which possess antigenically distinct haemagglutinin (HA) surface glycoproteins.
+Currently circulating influenza B viruses are comprised of two antigenically distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after two strains, B/Victoria/02/87 and B/Yamagata/16/88, which possess antigenically distinct haemagglutinin (HA) surface glycoproteins.
 The two HA lineages still co-circulate today and are thought to have shared a most recent common ancestor in xxxx. 
@@
 
-Previously, through the use of time-stamped phylogenetic trees, the complete history of influenza B reassortment events has been reconstructed \cite{chen2008}, and showed extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
+Previously, through the use of time-stamped phylogenetic trees, the reassortment history of major clades of influenza B has been reconstructed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
 Another approach exists, however, whereby belonging to either the Victoria or Yamagata lineage in the tree of one segment can be used as a discrete trait in trees of other segments.
@@
 In other segments (PA, NP, NA, MP and NS) a single lineage has been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
+
+\tbc{I think a cartoon displaying a reassortment graph for 2 segments (Vic and Yam lineages with segments A and B) plus the decomposed discrete-trait trees (A tree with B traits and B tree with A traits) would really help convey what you've done.
+Even if it doesn't get used in the paper, it would be great for a talk.}
 
@@
 \subsection*{Building trees}
+\tbc{I generally prefer to describe the model (HKY, etc...) then say that posterior samples were drawn using MCMC and describe sampling regime.}
 For each segment it was attempted to make 3 replicates with 200 million MCMC states (comment - only third replicates ran for that long) each under the HKY model of nucleotide substitutions, 3 codon partitions, robust counting (comment - reference robust counting)) under the GMRF Bayesian skyride demographic model (comment - reference skyride).
-In addition we used "relaxed" tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation).
+In addition we used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) \tbc{CITE}.
 We inferred the ancestral state of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS (comment - cite Philippe)).
@@
 \subsection*{PA, NP, NA, MP and NS have periodically lost diversity}
-From TMRCA plots (see Figure \ref{tmrcaOT}) it is clear that diversity in PB1, PB2 and HA has been maintained since the initial split of Victoria and Yamagata lineages, whereas other segments (most notably NS) have repeatedly lost lineages.
+We find that diversity in PB1, PB2 and HA has been maintained since the initial split of the Victoria and Yamagata lineages, whereas other segments (most notably NS) have undergone repeated lineage replacement (Figure~\ref{tmrcaOT}).
 The plot shows the initial losses of diversity in PA, NP and MP segments in 1997 (x axis), indicative of the extinction of Victoria lineages of those segments and revealing that a reassortment had occured which meant that all sampled lineages circulating around 1997 shared a common ancestor around the year 1990 (y axis).
 By 2001, the Yamagata lineage of the NS segment had gone extinct as well, revealing that NS too had reassorted together with PA, NP and MP segment around the year 1990 (similar TMRCA along the y axis to previous reassortants).
+\tbc{Here, you say things like `the Yamagata lineage of NS' and `the extinction of Victoria lineages'; the Vic/Yam distinction can't be seen in this figure.
+If you want to discuss these details you should provide a figure that makes them apparent.
+I'm not immediately sure how to display this.
+Or wait to discuss the Vic/Yam stratification until you have a figure that supports it.}
 Another round of reassortment-related extinction took place since then in PA, MP and NA segments and another 3 in the NS segment.
@@
 These differences in segment diversity are reflected in ratios of Victoria to Yamagata lineages in different segments: for the PB1, PB2 and HA segments the ratio of sequences belonging to either Victoria or Yamagata lineages are close to 0.5 (see Figure \ref{lineageRatios}, whereas all other segments are heavily skewed towards Yamagata lineage with the exception of NS segment which has more Victoria lineage sequences.
+\tbc{This would be more convincing if you showed a trend through time.}
 

### 2013-07-25 fluB.tex
Adding 1st round of changes suggested by Trevor.

@@
 Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
-Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages (referred to as B/Victoria and B/Yamagata subtypes (comment - is subtype okay to use here?
-- \tbc{Unfortunately, `subtype' is not kosher to refer to Vic and Yam.
-It's annoying.
-The most common term in the literature is `lineage'.
-I suspect you could also get away with `virus', as in the `B/Vic virus'.})) on the basis of haemagglutinin inhibition (HI) assays.
+Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages on the basis of haemagglutinin inhibition (HI) assays.
 Frequent reassortment between the segments of these two lineages has been noted in the past \cite{lindstrom1999}, but the effects of these reassortments have not been investigated in much detail.
@@
 \tbc{I generally prefer to describe the model (HKY, etc...) then say that posterior samples were drawn using MCMC and describe sampling regime.}
-For each segment it was attempted to make 3 replicates with 200 million MCMC states (comment - only third replicates ran for that long) each under the HKY model of nucleotide substitutions, 3 codon partitions, robust counting (comment - reference robust counting)) under the GMRF Bayesian skyride demographic model (comment - reference skyride).
+
+Temporal phylogenies were recovered for each segment using the HKY model of nucleotide substitution, 3 codon partitions, robust counting (comment - reference robust counting) and GMRF Bayesian skyride (comment - reference skyride) as the demographic model over the course of an MCMC chain (3 parallel runs of 200 million states, sampling every 20000 steps).
 In addition we used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) \tbc{CITE}.

### 2013-08-16 fluB.tex
Major rewrite of most text.
Added two extra figures ("between diversity" and a sketch of a track
plot) and moved one figure within text.
Rewrite is mostly going for a slightly more minimalist and purely
diversity-centered design.
I think the case is fairly convincing as it is, but there is still room
for using persistence statistics to say something about other kinds of
reassortments.

@@
 The two HA lineages still co-circulate today and are thought to have shared a most recent common ancestor in xxxx. 
-All other influenza B segments also show a split in xxxx (comment - our estimate \textit{circa} 1984 figure \ref{TMRCAhpd}, use a figure to show TMRCAs and rates for each segment?), although the numbers of strains falling on either side of the split are different due to reassortments following the initial split of Yamagata and Victoria lineages in all segments.
+All other influenza B segments also show a split in xxxx (comment - our estimate \textit{circa} 1984 figure \ref{TMRCAhpd}, use a figure to show TMRCAs and rates for each segment?), although the numbers of strains falling on either side of the split are different due to reassortments following the initial split of Yam and Vict lineages in all segments and fixation of one or the other lineage in some segments.
 
@@
 Another approach exists, however, whereby belonging to either the Victoria or Yamagata lineage in the tree of one segment can be used as a discrete trait in trees of other segments.
-Reassortment events which result in the replacement of one segment lineage by another show up as changes in the discrete trait along a branch and allow the inference of parameters related to the subtree descended from the node displaying a changed trait.
+Reassortment events which result in the replacement of one segment lineage by another show up as changes in the discrete trait along a branch and allow the inference of parameters related to the subtree descended from the node displaying a changed trait (see Figure \ref{methodFig}).
 Here, we use discrete traits to model inter-lineage reassortments between Victoria and Yamagata lineage segments in BEAST to look at lineage persistence times following reassortment events. 
-We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Victoria or the Yamagata lineage are selected against.
+We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage are selected against and, to the best of our knowledge, have not been sampled since 1996.
 In other segments (PA, NP, NA, MP and NS) a single lineage has been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
 
-\tbc{I think a cartoon displaying a reassortment graph for 2 segments (Vic and Yam lineages with segments A and B) plus the decomposed discrete-trait trees (A tree with B traits and B tree with A traits) would really help convey what you've done.
-Even if it doesn't get used in the paper, it would be great for a talk.}
+\begin{figure}[h]
+ \centering		
+	\includegraphics[width=0.45\textwidth]{figures/TreeFigure2}
+	\caption{\textbf{Schematic depiction of method.}}
+	\label{methodFig}
+\end{figure}
 
@@
 We inferred the diversity of each segment over time by finding the oldest most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on diversity existing at each time period.
-In addition, we calculated mean pairwise diversity and mean pairwise TMRCA between branches inferred to be in virions possessing Victoria and Yamagata lineage PB1, PB2 and HA.
+In addition, we calculated mean pairwise diversity and mean pairwise TMRCA between branches inferred to be in virions possessing Vic and Yam lineage PB1, PB2 and HA.
 
@@
 
-\subsection*{PA, NP, NA, MP and NS have periodically lost diversity}
-We find that diversity in PB1, PB2 and HA has been maintained since the initial split of the Victoria and Yamagata lineages, whereas other segments (most notably NS) have undergone repeated lineage replacement (Figure~\ref{tmrcaOT}).
-The plot shows the initial losses of diversity in PA, NP and MP segments in 1997 (x axis), indicative of the extinction of Victoria lineages of those segments and revealing that a reassortment had occured which meant that all sampled lineages circulating around 1997 shared a common ancestor around the year 1990 (y axis).
-By 2001, the Yamagata lineage of the NS segment had gone extinct as well, revealing that NS too had reassorted together with PA, NP and MP segment around the year 1990 (similar TMRCA along the y axis to previous reassortants).
-\tbc{Here, you say things like `the Yamagata lineage of NS' and `the extinction of Victoria lineages'; the Vic/Yam distinction can't be seen in this figure.
-If you want to discuss these details you should provide a figure that makes them apparent.
-I'm not immediately sure how to display this.
-Or wait to discuss the Vic/Yam stratification until you have a figure that supports it.}
-Another round of reassortment-related extinction took place since then in PA, MP and NA segments and another 3 in the NS segment.
-Unlike all other segments, however, all lineages of PB1, PB2 and HA segments have maintained diversity and lineages circulating in 2011 were still derived from the original split of the Victoria and Yamagata lineages.
+\subsection*{PA, NP, NA, MP and NS segments have periodically lost diversity}
+The split of Vic and Yam lineages can be found in all segments.
+Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
+Some viruses with mixed-lineage genomes have become common ancestors to many later strains (i.e.
+the "trunk" of the phylogenetic tree) and the segment lineages involved so successful, as to become fixed within the influenza B population circulating today.
+We find that Vic lineage PA, NP, NA and MP segments have gone extinct (i.e.
+no sequences from those lineages have been sampled in recent years see Figure \ref{lineageRatiosOverTime}), replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
+Similarly, Vic lineage NS has become fixed in the viral population as well.
+By looking at the oldest most recent common ancestor of all lineages existing within the tree of each segment at yearly timeslices (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby sampled sequences trace their ancestry to increasingly more recent common ancestors.
 
-These differences in segment diversity are reflected in ratios of Victoria to Yamagata lineages in different segments: for the PB1, PB2 and HA segments the ratio of sequences belonging to either Victoria or Yamagata lineages are close to 0.5 (see Figure \ref{lineageRatios}, whereas all other segments are heavily skewed towards Yamagata lineage with the exception of NS segment which has more Victoria lineage sequences.
-\tbc{This would be more convincing if you showed a trend through time.}
+\subsection*{PB1, PB2 and HA segments continue to circulate as two lineages and have not reassorted recently}
+Unlike PA, NP, NA, MP and NS segments, which have periodically lost diversity, we find that PB1, PB2 and HA segments have continued to accumulate diversity since the initial split of Vic and Yam lineages (see Figure \ref{tmrcaOT}).
+In addition, there is evidence to suggest that the numbers of sequences that belong to either Vic and Yam lineages of PB1, PB2 and HA segments have been sampled at a ratio close to 0.5 (see Figure \ref{lineageRatiosOverTime}), presumably due to the action of some form of balancing selection, possibly to maintain antigenic diversity.
+Our findings suggest that the diversity in PB1, PB2 and HA segments is not maintained independently, but is very likely a result of co-adaptation between these segments.
+By measuring the mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in diversity, which indicate that an inter-lineage reassortment event has taken place.
+In example \ref{ARGexample}, if, over time slices, we measured mean pairwise distance between red and blue branches in tree 1, after they have been reciprocally labelled with each tip's position (with respect to some pre-defined bifurcation) in tree 2, we would find a drop in this measure, since the distance between tip 6 and tip 5 is now shorter and does not go through the root of the tree.
+Thus, by measuring pairwise distances between differently labelled branches we get a sense of how panmictic different segments are with respect to each other.
+If we repeat this procedure for all trees and focus on PB1, PB2 and HA lineage labels, we find that mean pairwise diversity between branches labelled as having for example, a PB1 segment of Vic or Yam lineage in PB2 and HA trees (and other combinations of the 3 trees and their labels), show small drops in diversity initially, but from the 1997 time slice onwards all branches labelled as Vic or Yam have a common ancestor at the root of the tree.
+We also find evidence of co-reassortment in PB1, PB2 and HA segments: in Figure \ref{betweenDiversity} all three graphs become identical after 1997, indicating that the same PB1, PB2 and HA labels have been assigned to the same branches in trees of all other segments.
+Our findings therefore suggest that a single configuration of Yam and Vic lineages of PB1, PB2 and HA segments has been preserved since the split of Vic and Yam lineages and have not reassorted with respect to each other, but rather have co-reassorted as a single unit, from 1997 onwards.
+
+\subsection*{Victoria and Yamagata lineages of PB1-PB2-HA are co-adapted segment complexes}
+We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
+Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic or entirely Yam PB1, PB2 and HA segments (i.e.
+where a branch was labelled as having Vic PB1, Vic PB2 and Vic HA) compared to the sum of branch lengths where one or more of the labels is of another lineage (e.g.
+Vic PB1, Yam PB2 and Yam HA).
+Together with Figure \ref{betweenDiversity} it suggests that some inter-lineage reassortment has occured between PB1, PB2 and HA segments soon after the split of Vic and Yam lineages, but the resulting viruses were not fit and did not become the trunk of the tree.
+Viruses bearing mixed-lineage PB1-PB2-HA complexes had the following constellations of PB1, PB2 and HA: VVY (B/Bankok/163/1990-like), YVV (B/Nanchang/630/1994-like) and YVY (B/New York/24/1993-like).
+We do not show results for other lineage labels as none of them show this pattern of reciprocally preserved diversity with any other segment.
+
+\subsection*{Other segments evolve with considerable independence}
+The reconstructed evolution of Victoria and Yamagata lineages of all segments reveals a considerable degree of autonomy.
+For example, of all the Yamagata lineage segments that were reassorted into Victoria lineage PB1-PB2-HA background giving rise to B/Hong Kong/330/2001-like viruses, most segments went extinct following later reassortments, save for NP, which continued circulating until, at least, 2011.
+A more dramatic example can be found in the NS segment, where the Vic lineage had been fixed in the population after reassorting into an otherwise entirely Yam-derived genome constellation and then reassorted back into a Vic lineage PB1-PB2-HA background, replacing the NS lineage that was present there.
 
@@
 	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_LineageRatios.png}
+	\includegraphics[width=0.55\textwidth]{figures/InfB_LineageRatiosOverTime.png}
 	\caption{\textbf{Ratio of lineages in the dataset.}}
@@
 
-\subsection*{Diversity in PB1, PB2 and HA segments is not maintained independently}
-We find that the diversity preserved in PB1, PB2 and HA segments is not preserved independently in each segment.
-Plot XXXX (comment - make date-spectrum figure) shows the posterior distributions of inferred dates of reassortment.
-Following 1993 PB1, PB2 and HA reassortments are inferred to occur simultaneously at the same nodes.
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/InfB_betweenDiversity.png}
+	\caption{\textbf{Ratio of lineages in the dataset.}}
+	\label{betweenDiversity}
+\end{figure}
 
-\subsection*{Numbers of reassortments and their persistence}
-
-(comment - this bit is for stratified REPU results)
-We find that on average, inter-lineage reassortments are common (number of trait switches observed: 116 after filtering, but some will be the same event, naive underestimate is 14.5 inter-lineage reassortments per segment) and favoured by natural selection (persistence statistic > 1, see Figure \ref{stratifiedREPUgraph} (comment - make figure prettier)), although this varies by segments involved and date of reassorment.
-
-In particular, reassortment events where same-lineage PB1, PB2 and HA segments reassort together have much better persistence than events where they reassort independently.
-Conversely, reassortments that produce mixed-lineage PB1-PB2-HA complexes have persistence statistic lower than an average node existing at the same time, with the exception of NS.
-We see reassortment events preserving the Victoria lineage PB1-PB2-HA complex occuring between 1990 and 1995 in PA, NP and MP (comment - if NS trees were better, there would be a PB1-PB2-HA preserving event in NS segment), coresponding to the origins of the reassortant lineage that eventually gave rise to B/Hong Kong/330/2001-like influenza B viruses, which were the first Victoria lineage HA-bearing viruses detected outside of Asia since 1990s (comment - cite paper).
-However, the persistence of segment lineages comprising these B/Hong Kong/330/2001-like reassortant viruses vary dramatically.
-Whereas the PA and MP lineages that were part of the reassortant B/Hong Kong/330/2001-like constellation had poor persistence, the NP lineage was "re-used" during the next inter-lineage reassortment occuring around 2000.
-
-The reassortment event occuring around the year 2000 (which produced B/Iowa/03/2002-like viruses), involved the PA, NA, MP and NS segments of Yamagata lineage reassorting with Victoria lineage PB1, PB2 and HA (the latter virus already contained a Yamagata lineage NP) with good persistence (over 2 times the persistence of all other nodes existing at the time).
-The greater persistence of PA, NA, MP and NS lineages is due to having left descendants close to the present (essentially becoming the "trunk" lineages of those segments).
-
-Additional reassortments occured in 2004 and 2006.
-In 2004 NP, MP and NS segments of Yamagata lineage reassorted with Victoria lineage PB1, PB2 and HA (with Yamagata lineage PA and NA).
-This event is one of the most interesting ones, since the NS segment of Victoria lineage PB1-PB2-HA complex, though ancestrally derived from Victoria lineage, was replaced by a Victoria lineage NS segment which had been evolving together with genome constellations derived from Yamagata lineage segments.
-Although the MP segment lineage reassorted in the 2004 event eventually became the "trunk" of all Victoria PB1-PB2-HA lineage bearing viruses, the NP(Yam) lineage was replaced by another one (also Yam) that was reassorted into the PB1-PB2-HA Victoria lineage-bearing background in 2006.
-
-From reassortment event persistence times it is clear that reassortment events that break up the PB1-PB2-HA complex have occured in the past but never persisted for long.
-We observe a sudden stop of PB1-PB2-HA complex-separating reassortment events after 1995 which are followed solely by PB1-PB2-HA complex-preserving reassortment events.
-
-(comment - this bit is for the tip descent method)
-On average it takes nearly 2 years longer for a random tip at any given point to trace its ancestry to a reassortant lineage where a PB1, PB2 and HA segments of the same lineage have reassorted together, although this varies by segment and time of reassortment.
-
-\subsection*{Differentation of segments along PB1, PB2 and HA lineages}
-(comment - this will be referring to 3 plots, which are mean pairwise tmrcas between branches under PB1, PB2 and HA traits, respectively)
-
-When comparing TMRCAs between branches of segment trees inferred to be under either Victoria or Yamagata lineage PB1, PB2 or HA, it is clear that diversity along the lineage axis has been preserved in PB1, PB2 and HA, whereas all other segments (PA, NP, NA, MP and NS) have repeatedly lost diversity, indicating reassortments with PB1, PB2 and HA segments.
-Interestingly, the diversity patterns between segments under inferred PB1, PB2 or HA lineages differ early on and this can be shown to be due to short lived reassortants with mixed-lineage PB1, PB2 and HA segments, namely the VVY (B/Bankok/163/1990-like), YVV (B/Nanchang/630/1994-like) and YVY (B/New York/24/1993-like) reassortants (comment - will require trees in supplementary info).
-Starting from about 1993, however, PB1, PB2 and HA segments of Victoria lineage have only reassorted as one with other segments of Yamagata lineage which is reflected in plots becoming identical, regardless of segment, post-1993.
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/InfB_stateTime.png}
+	\caption{\textbf{Total time spent in "pure" vs "mixed" PB1-PB2-HA}}
+	\label{stateTime}
+\end{figure}
 
@@
 the 2004 Vic lineage PB1-PB2-HA complex reassortment with NP is only visible in the NP tree, since all viruses in our dataset from 1995 onwards, including Vic PB1-PB2-HA complex-bearing ones, had a Yamagata lineage-derived NP).
+
+\subsection*{PB1-PB2-HA co-adaptation has evolved over time}
+Our findings suggest that PB1, PB2 and HA segments of Vic and Yam lineages represent co-adapted co-reassorting segment complexes.
+We argue that PB1-PB2-HA complexes have become co-dependent on each other to the extent that any reassortant viruses possessing mixed-lineage PB1-PB2-HA are unfit and possibly never get sampled or, if sampled, never persist for long enough to be fixed in the influenza B population.
+Supporting this, we detect at least 3 genome constellations having mixed-lineage PB1-PB2-HA sampled early in the first decade of the split between Vic and Yam lineages, when, presumably, Vic and Yam lineages of PB1, PB2 and HA segments were still similar at the protein sequence level to produce viable, though unfit, influenza B viruses.
+We interpret lack of reassortment within the PB1-PB2-HA complex across the lineage boundary in recent times as evidence of a considerable degree of co-dependence within the Vic and Yam PB1-PB2-HA lineages, whereby mixed-lineage reassortants are inviable or unsampled by surveillance due to being outcompeted by viruses with pure-lineage PB1-PB2-HA complexes.
+
+\subsection*{Potential causes of co-dependence between PB1-PB2-HA}
+Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, e.g. PB1-PB2-PA and HA-NA.
+Though it would be easy to explain co-adaptation between the aforementioned segments by referring to their functional roles (e.g.
+PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities), our findings suggest a more subtle relationship between PB1, PB2 and HA segments.
+It is clear that PB1-PB2-HA segments do not preferentially reassort together, because there have been at least 3 sampled mixed-lineage PB1-PB2-HA complex constellations, which did not become fixed in the population and recent evidence suggests, that, at least in influenza A, reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
+In the absence of clear functional explanations of why PB1, PB2 and HA should be co-adapted, we draw attention to a recent influenza A study, which found that during the process of vaccine creation, whereby recent human influenza A isolates are reassorted with an egg-adapted strain in an attempt to reassort HA-NA of the recent isolate into the otherwise entirely egg-adapted genome background.
+Cobbin et al.
+\cite{cobbin2013} found that the reassortant viruses following this procedure, together with the HA-NA segments the PB1 segment gets co-reassorted more frequently than expected by chance.
+Cobbin et al. suggest a gene regulatory role for the PB1 segment, as its presence or absence has dramatic effects on HA concentration on the surface of virions and total virion production.
+Without more, and more recent, reassortment events we cannot offer more than speculation to the nature of the relationship between PB1, PB2 and HA segments of Vic and Yam lineages.
+We do however see several potential causes for the phenomenon we observe.
+It would be possible that balancing selection is preserving the diversity in one segment, the others hitchhiking along.
+A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (NA of Vic lineage having gone extinct), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient diversity.
+Alternatively, either differences in gene expression or fidelity between Vic and Yam PB1 or PB2 could cause an apparent co-adaptation between PB1, PB2 and HA segments.
+Finally, it is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply become co-adapted to each other, without any one segment being the driver of diversity preservation.
+
+\subsection*{The future of influenza B viruses}
+The co-adaptation between Vic PB1-PB2-HA and Yam PB1-PB2-HA complexes is a potential insight into things to come.
+We see two potential alternative paths of evolution.
+One, where more segments will be recruited into these complexes, as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex.
+For some time it seemed like the NP segment would be recruited into the PB1-PB2-HA complex, but recent reassortment events suggest that all Yam NP lineages reassorted into Vic lineage PB1-PB2-HA background are being outcompeted by a Yam NP lineage reassorted most recently (represented by viruses with the B/Brisbane/33/2008-like genome constellations).
+Another alternative is continued reassortment of all non-PB1-PB2-HA segments between Vic and Yam PB1-PB2-HA complexes.
+Finally, it is not impossible that one PB1-PB2-HA complex lineage will go extinct in the future, by chance or by being outcompeted by the other lineage, thus leaving a single-lineage influenza B population.
+Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether such two-lineage dynamics have not occurred in the distant past.
+In the former two scenarios (speciation and co-circulation) we have expectations about the long term trend of reassortment.
+If the influenza B population is on its way to speciation we expect that over time inter-lineage reassortment frequency should decrease, as more and more segment lineages get recruited into co-adapted segment complexes.
+On the other hand, if influenza B viruses continue co-circulating as two PB1-PB2-HA lineages exchanging non-PB1-PB2-HA segments, reassortment frequency is expected to be constant.
+Seeing how rarely influenza B viruses reassort across the Vic-Yam lineage boundary, the timescale for finding out which of the aforementioned scenarios, if any, are under way, is impractical at best.
+
+\subsection*{Implications for influenza B virus control and prevention}
+It is our hope that understanding the mechanism(s) underlying the links between Vic and Yam lineage PB1-PB2-HA complexes might suggest a more varied approach to both influenza B surveillance and control.
+For one, our findings suggest that viruses bearing mixed-lineage PB1-PB2-HA complexes, if unveiled by surveillance, are unlikely to contribute to future influenza B evolution.
+Similarly, in the absence of sequence data, it should be possible to predict the lineage of PB1, PB2 and HA segments of influenza B viruses by sequencing only one of those segments.
+It remains to be seen whether equivalent or similar links exist between segments of influenza A viruses, which, by far, present a much greater threat to human health worldwide than influenza B viruses.
+In addition, by discovering the underlying mechanism by which PB1, PB2 and HA are co-adapted might reveal potential cracks in the armour of influenza viruses, allowing for the creation of anti-virals to be used with currently deployed oseltamivir and zanamivir in a multi-drug approach similar to the one used against bacteria.
+
+\subsection*{Predictions}
+We see two potential future directions of research.
+Using previously developed plasmid systems it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments.
+We make two predictions for such experiments.
+One, viruses with mixed-lineage PB1-PB2-HA complexes will have reduced fitness compared to pure-lineage PB1-PB2-HA complexes.
+Two, if the relationship between Vic and Yam lineage PB1, PB2 and HA segments has indeed evolved over time, viruses with mixed-lineage PB1-PB2-HA complexes isolated earlier will perform better than viruses with PB1-PB2-HA isolated more recently.
+However, there is potential for pleiotropic effects of other segments to interfere with fitness measurements.
+A second approach is to use sequence data to look for linkage disequilibrium between polymorphisms on PB1, PB2 and HA segments of Vic and Yam lineages.
+Polymorphisms originating close in time on different segments of Vic and Yam lineages could be used to infer the sites potentially responsible for the observed co-adaptation of PB1, PB2 and HA segments.
+This could then pave way for wet-lab experiments where, for example, Vic PB1, PB2 or HA segments would be manipulated at potential sites to get Yam-like segments, which after being reassorted into Yam genome constellations should give reassortants of similar fitness to viruses with entirely Yam-derived PB1-PB2-HA complexes.
+
+\section*{Conclusions}
+In this study we have investigated the patterns of reassortment amongst two influenza B virus segment lineages and found consistent co-reassortment of 3 segments: PB1, PB2 and HA.
+Our findings reveal that Vic and Yam PB1, PB2 and HA segments co-associate amongst themselves with segments of their own lineage, thus resulting in an entirely Vic-lineage-derived PB1-PB2-HA segment complex and an entirely Yam-lineage-derived PB1-PB2-HA segment complex.
+Though we detect strains with mixed-lineage PB1-PB2-HA complexes early after the split between Vic and Yam lineages, we find that they have performed poorly and have comprised only about 10\% of the total evolution undergone by influenza B viruses sampled in this study.
+We argue that this is evidence of a selectively maintained relationship between PB1-PB2-HA complex of Vic and Yam lineages and not due to biased reassortment.
+In addition, the close timing of these mixed-lineage PB1-PB2-HA reassortants close to the split of Vic and Yam lineages suggests that the relationship between each lineage's PB1-PB2-HA complexes have developed over time and in their current state are undetected by surveillance due to either poor fitness or inviability of reassortants with mixed-lineage PB1-PB2-HA complexes.
+
+(comment - END of the newer version)
 
@@
 VVY is invisible in HA tree, YVV in PB1 and YVY in PB2; solution: remove PB1, PB2 and HA bars or make it clear in the text)
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_stateTime.png}
-	\caption{\textbf{Total time spent in "pure" vs "mixed" PB1-PB2-HA}}
-	\label{stateTime}
-\end{figure}
 

### 2013-08-28 fluB.tex
Adding multiple files, minor adjustments to text.
Added poster version of the manuscript, acknowledgement tables (.tsv and
.xls formats) and added a few references to the manuscript in
preparation for ploughing through the manuscript once I'm back.

@@
 Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains.
-The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000}), thus limiting the virus's available antigenic reservoir. 
-It has been suggested that influenza B viruses use a combination of reassortment and point mutations to generate antigenic diversity (comment - citation).
+The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), thus limiting the virus's available antigenic reservoir. 
+It has been suggested that influenza B viruses use a combination of reassortment and point mutations to generate antigenic diversity \cite{nerome1998,mccullers1999}.
 
@@
 \subsection*{Building trees}
-\tbc{I generally prefer to describe the model (HKY, etc...) then say that posterior samples were drawn using MCMC and describe sampling regime.}
-
 Temporal phylogenies were recovered for each segment using the HKY model of nucleotide substitution, 3 codon partitions, robust counting (comment - reference robust counting) and GMRF Bayesian skyride (comment - reference skyride) as the demographic model over the course of an MCMC chain (3 parallel runs of 200 million states, sampling every 20000 steps).
@@
 We inferred the diversity of each segment over time by finding the oldest most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on diversity existing at each time period.
-In addition, we calculated mean pairwise diversity and mean pairwise TMRCA between branches inferred to be in virions possessing Vic and Yam lineage PB1, PB2 and HA.
+In addition, we calculated mean pairwise TMRCA between branches inferred to be in virions possessing Vic and Yam lineage PB1, PB2 and HA.
 
@@
 
-(comment - END of the newer version)
-
-\subsection*{Reassortment in influenza B is common and extensive}
-Extensive reassortment has long been established as a major pathway by which influenza B viruses generate diversity.
-Even focusing on a specific subset of reassortments (those that occur between two lineages and those where the previous lineage was of the opposite lineage) reveals 116 events and if we naively assume that all events are replicated across all trees it still means there have been over 14 inter-lineage reassortment events.
-We also see that reassortments occur across considerable phylogenetic distances and some segment lineages have been so successful as part of reassortant viruses that they out-competed their sister lineages (e.g.
-Yamagata lineages of PA, NP, MP and NA segments).
-(comment - could include diversity over time plots to show maximum phylogenetic distance attained prior to reassortment for each segment)
-
-\subsection*{Selection maintains the diversity in PB1-PB2-HA complexes of Victoria and Yamagata lineages}
-Rather surprisingly, 3 segments have maintained a ratio of Vic and Yam lineages very close to 0.5 (PB1, PB2 and HA) (see Figure \ref{lineageRatios}) and have preserved the diversity that has accumulated since the split of the two lineages (see Figure \ref{tmrcaOT}).
-Our results suggest that this is due to the 3 segments co-reassorting (comment - refer to posterior distributions of reassortant dates plot).
-We find evidence that this co-reassortment pattern between PB1, PB2 and HA is not due to skewed reassortment (Marhsall \textit{et al.} \cite{marshall2013} have shown that reassortment efficiency is high in influenza A), but rather due to viruses bearing mixed-lineage PB1-PB2-HA complex having lower fitness than viruses with PB1-PB2-HA complexes derived entirely from Yam or Vic lineages (comment - refer to stratified REPU plot).
-Not only are events that split up pure-lineage PB1-PB2-HA complexes rare, but they also have poor persistence and trees have spent less than 10\% of the time in the mixed-lineage PB1-PB2-HA complex state (Figure \ref{stateTime}), suggesting that they are selected against (comment - make plot of that).
-
-(comment - PB1, PB2 and HA are missing extra years in the mixed-lineage bar due to invisible reassortments e.g.
-VVY is invisible in HA tree, YVV in PB1 and YVY in PB2; solution: remove PB1, PB2 and HA bars or make it clear in the text)
-
-Thus our findings suggest that the link maintaining PB1, PB2 and HA lineages close to equilibrium is not due to biased reassortment, but a result of selection acting upon reassortants containing mixed-lineage PB1-PB2-HA segments.
-We find that by 1995 sufficient co-dependence had developed between PB1, PB2 and HA segments of Victoria and Yamagata lineages to prevent reassortant progeny containing mixed-lineage PB1-PB2-HA complexes from infecting sufficient numbers of individuals to cross the detection threshold.
-
-\subsection*{Segments vary in their abilities to reassort and success as part of reassortant viruses}
-A rather striking pattern in influenza B evolution is the fact that segments reassorted at the same time can have dramatically different fates.
-For example, PA, NP, MP and NS segments of Yamagata lineage were reassorted into Victoria lineage PB1-PB2-HA background in 1994, but of those 4 segments only the NP lineage persisted for a considerable amount of time, being replaced by another NP lineage in 2006, whereas PA was replaced in 2000 and MP together with NS were replaced by different lineages in 2004.
-This suggests that most non-PB1, PB2 or HA segments have the capacity to evolve somewhat independently of the rest of the genome.
-
-\subsection*{Interpreting our results}
-(comment - this bit is exceedingly unsubstantiated and ramble-y)
-Recently the antigenic drift of influenza B viruses has been characterised (comment - cite Bedford et al 2013) and Victoria lineage HA was found to drift nearly twice as fast as Yamagata lineage HA.
-A possible explanation for this would be differences in polymerase fidelity: if Victoria lineage polymerase introduced higher mutation rates, this would explain the higher antigenic drift rate seen in Vic lineage HA, but also would suggest that Vic lineage PB1-PB2-HA complex "borrows" segments from Yamagata lineage PB1-PB2-HA complex-bearing viruses and "breaks" them through going over something akin to each segment's error threshold, necessitating another round of reassortment.
-However, we find no evidence of differences in nucleotide, synonymous and non-synonymous rates of evolution between branches that were inferred to have been in the same virion as PB1, PB2 or HA segments of Vic or Yam lineages.
-Thus we have no evidence to suggest fundamental differences between Vic and Yam lineage PB1-PB2-HA complexes.
-
-Balancing selection would be expected, and does in the malaria parasite, preserve antigenic diversity, but it is unclear if maintenance of PB1-PB2-HA lineages is driven by balancing selection acting to preserve antigenic diversity of HA or the diversity of PB1 and PB2 segments.
-Recently a similar link has been described in influenza A by Cobbin \textit{et al.} \cite{cobbin2013}, where the PB1 segment is reassorted more frequently than expected during vaccine strain creation.
-If influenza B virus PB1 has a similar gene-regulatory activity, then it would be reasonable to assume that HA and PB1 are associated through the necessity for HA to have appropriate levels of expression and PB2 hitchhikes together with PB1 due to some functional requirement as part of the polymerase complex.
-This does not explain why the Vic lineage of the PA segment was lost so readily, however, despite it being a part of the polymerase complex.
-
-\subsection*{Future evolution of influenza B viruses}
-From our current findings it is difficult to extropolate what reassortants will come to dominate the influenza B population in the far future.
-It seems likely that PB1, PB2 and HA segments of both Victoria and Yamagata lineages will either continue to co-exist in their co-adapted state or one lineage will go extinct.
-Until the 2004 and 2006 reassortment events it seemed like the NP segment would follow the PB1-PB2-HA complex in becoming the next segment to be recruited as part of a co-adapted segment complex.
-This method of gradually recruiting more segments to a co-adapted segment complex could have lead to the eventual speciation of influenza B into originally Vic and Yam lineage HA bearing species.
-As it stands now, however, the most diversity between Victoria and Yamgata PB1-PB2-HA lineage-bearing viruses has been built up in PA and NA segments (which shared a common ancestor around 2000) and it remains to be seen whether this diversity will too be lost following a reassortment event.
-
-Given the evolutionary history of the NP segment, it would be reasonable to focus sequencing efforts on complete influenza B genome sequencing.
-Our results indicate that successful reassortants are likely to come from the combination of Victoria lineage PB1-PB2-HA complex and PA, NP, NA and MP segments of Yamagata lineage (or NS segment of Vic lineage), though the number of Yamagata lineage segments can vary.
-It is also obvious that increased surveillance efforts will inevitably uncover a lot of reassortant viruses (possibly even viruses with mixed-lineage PB1-PB2-HA complexes) and the threshold at which these reassortants should be considered as being likely to take over the influenza B population would require further work.
-
-\subsection*{Testing the co-dependence between PB1, PB2 and HA segments}
-Our results suggest a simple test for the hypothesis that PB1, PB2 and HA segments of Victoria and Yamagata lineages have become co-dependent - if PB1, PB2 or HA segments from different lineages were combined artificially in a single virion, these should have lower fitness than virions containing wholly Victoria or wholly Yamagata lineage PB1-PB2-HA complex.
-We also expect that the fitness of artificially derived reassortants would decline with increasing date of strain isolation i.e.
-mixed-lineage PB1-PB2-HA containing reassortants with segments from older isolates having higher fitness than reassortants possessing segments of more recent isolates.
-
-\subsection*{Implications for influenza B surveillance and prevention}
-The findings presented here have several implications.
-On one hand, if the co-dependence between PB1, PB2 and HA segments is strong enough, influenza B viruses containing mixed-lineage PB1-PB2-HA complex detected by surveillance are unlikely to contribute to long-term evolution of influenza B viruses.
-On the other hand, the dependence of the HA segment on PB1 and PB2 segments (or \textit{vice versa}) implies that there is a possibility that more subtle disease control measures could be developed and used against influenza B viruses, or, if similar co-adapted segment complexes were discovered in influenza A, seasonal influenza in general.
-In addition, the extent of inter-lineage reassortment argues for more intensive surveillance of influenza B viruses, as sequences of a handful of segments will not necessarily tell the whole story, especially since influenza B viruses have been known to evolve counter-intuitively (\textit{e.g.} losing Victoria lineages of PA and NA segments despite associations with the rest of the polymerase complex or antagonising HA activity, respectively).
-Finally, this study also highlights the gaps in our knowledge about the molecular biology of influenza B viruses.
-
 %\begin{figure}[h]

### 2013-09-27 fluB.tex
Major update.
Adding all the references, extra methods, changed some figures, wrote a
bit more text.

@@
 \renewcommand\Affilfont{\small \normalfont}
+\makeatletter
+\renewcommand\AB@affilsepx{, \protect\Affilfont}
+\makeatother
 
@@
 Currently circulating influenza B viruses are comprised of two antigenically distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after two strains, B/Victoria/02/87 and B/Yamagata/16/88, which possess antigenically distinct haemagglutinin (HA) surface glycoproteins.
-The two HA lineages still co-circulate today and are thought to have shared a most recent common ancestor in xxxx. 
-All other influenza B segments also show a split in xxxx (comment - our estimate \textit{circa} 1984 figure \ref{TMRCAhpd}, use a figure to show TMRCAs and rates for each segment?), although the numbers of strains falling on either side of the split are different due to reassortments following the initial split of Yam and Vict lineages in all segments and fixation of one or the other lineage in some segments.
+The two HA lineages still co-circulate today and are thought to have emerged around 1983 \cite{rota1990}. 
+All other influenza B segments also show a split around that time (our combined estimate across all segments place it \textit{circa} 1982.95 (95\% highest posterior density 1981.53--1984.21) figure \ref{TMRCAhpd}), although the numbers of strains falling on either side of the split are different due to reassortments following the initial split of Yam and Vic lineages in all segments and fixation of one or the other lineage in some segments.
 
@@
 Previously, through the use of time-stamped phylogenetic trees, the reassortment history of major clades of influenza B has been reconstructed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
+
 Another approach exists, however, whereby belonging to either the Victoria or Yamagata lineage in the tree of one segment can be used as a discrete trait in trees of other segments.
@@
 Here, we use discrete traits to model inter-lineage reassortments between Victoria and Yamagata lineage segments in BEAST to look at lineage persistence times following reassortment events.
+
 We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage are selected against and, to the best of our knowledge, have not been sampled since 1996.
@@
 	\includegraphics[width=0.45\textwidth]{figures/TreeFigure2}
-	\caption{\textbf{Schematic depiction of method.}}
+	\caption{\textbf{Schematic depiction of method.}
+We begin by assigning sequences falling on either side of a specified bifurcation within each segment tree to different lineages.
+We then transfer lineage labels from one tree to the same tips in another tree.
+Label switches thus indicate reassortment events that combine lineages falling on different sides of the bifurcation specified earlier.
+In the example above tip number 6 is determined to be such a reassortant.}
 	\label{methodFig}
@@
 \subsection*{Sequences and subtyping}
-We compiled a dataset of 452 complete influenza B genomes from GISAID (comment - need acknowledgment tables). 
-Each strain was assigned to a lineage (either Vic or Yam) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS), where B/Norway/1/84 (comment - B/Czechoslovakia/69/1990 could be used as well, since it is the only strain showing an entirely Vic genome) was considered as being representative of Victoria lineage.
+We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID}. 
+Each strain was assigned to a lineage (either Vic or Yam) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
+
 Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/02/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).
-Seven of these where then used as discrete states in each of the trees (e.g. PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values) in BEAST (comment - reference BEAST).
+Seven of these where then used as discrete states in each of the trees (e.g. PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values) in BEAST \cite{drummond2012}.
 
 \subsection*{Building trees}
-Temporal phylogenies were recovered for each segment using the HKY model of nucleotide substitution, 3 codon partitions, robust counting (comment - reference robust counting) and GMRF Bayesian skyride (comment - reference skyride) as the demographic model over the course of an MCMC chain (3 parallel runs of 200 million states, sampling every 20000 steps).
-In addition we used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) \tbc{CITE}.
-We inferred the ancestral state of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS (comment - cite Philippe)).
+Temporal phylogenies were recovered for each segment using the HKY model of nucleotide substitution, 3 codon partitions, robust counting \cite{obrien2009} and GMRF Bayesian skyride \cite{minin2008} as the demographic model over the course of an MCMC chain (3 parallel runs of 200 million states, sampling every 20000 steps).
+In addition we used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip dates would change and be estimated over the course of the MCMC chain (available in BEAST v1.7.4 \cite{drummond2012} via XML file editing as a date precision parameter).
+
+We inferred the ancestral state of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS \cite{lemey2009}).
 Because the posterior set of trees have branches labelled with the inferred lineage in all other segments, we can detect inter-lineage reassortments whenever a trait transition is observed (i.e.
-Y to V or V to Y) between a node and its descendant in any segment.
+Y to V or V to Y, see Figure \ref{methodFig}) between a node and its descendant in any segment.
 In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
 
-We use a measure of persistence which is defined as the ratio of distances (in units of time) between a reassortant node and its most recent tip (the event is sampled uniformly over the branch displaying a trait value that is different from its parent node) and compare it to the mean of distances between all nodes and their respective most recent tips.
-If this statistic is below 1 a reassortment event is presumed to be disadvantageous, if it is above 1 the event is presumed to be advantageous.
+\subsection*{Measures of diversity}
+We inferred the diversity of each segment over time by finding the oldest most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on the maximum amount of diversity existing at each time period.
+In addition, we calculated mean pairwise TMRCA between branches inferred to be in virions possessing Vic and Yam lineage PB1, PB2 and HA.
+This gave us an averaged measure of how much a particular segment reassorts with respect to PB1, PB2 and HA segments.
+If Vic and Yam lineages of PB1, PB2 and HA segments were to be considered as being separate populations this measure would be equivalent to `between population' diversity.
 
-Alternative:
-We start by allocating the tips in the tree to yearly bins starting from 1984 and ending in 2013. 
-At each state of the MCMC chain we choose a random tip from each bin and traverse the tree towards the root until a reassortment event (i.e. a trait switch) is encountered.
-We log the traits which have transitioned at the reassortant node (i.e.
-the reassorting segment), the direction of transition (e.g.
-whether the parental node was V and the child node Y or vice versa) and the amount of time separating the tip and the node displaying a trait transition.
-The time taken to traverse back to a reassortment event is expected to be longer for advantageous reassortments than for non-advantageous reassortments.
+We also calculated the total amount of evolutionary time each segment has spent in virions with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments by taking the sum of branch lengths in each tree under different combinations of the PB1, PB2 and HA labels.
+This gives a measure of how successful, over long periods of time, each particular PB1-PB2-HA constellation has been.
 
-\subsection*{Building a consensus set of reassortment events}
-Maximum clade credibility trees were made after discarding the first 1000 trees as burnin. 
-Nodes with trait values different from the trait value of their parental node were classified as being reassortants.
-The persistence time, descendant tips and reassorting segments were logged for each reassortant node and aggregated across analyses based on descendant tips.
+\subsection*{Tree to tree similarities}
+In order to get a measure of tree to tree similarity we first extracted 1000 (sampling frequency roughly equal to 1 tree/180000 MCMC states) trees from the posterior distributions of trees of each segment.
+Pairwise comparison between each pair of segments was performed by comparing trees at each MCMC state.
 
-\subsection*{Measures of diversity}
-We inferred the diversity of each segment over time by finding the oldest most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on diversity existing at each time period.
-In addition, we calculated mean pairwise TMRCA between branches inferred to be in virions possessing Vic and Yam lineage PB1, PB2 and HA.
+We express the similarity by referring to 3 descriptive statistics (sum, mean and standard deviation) of the distribution of TMRCA differences ($\Delta$TMRCA) between pairs of tips in the two trees.
+These parameters are estimated for each pairing of trees at each MCMC state and provides us with 95\% highest posterior density intervals for those parameters, thus giving us the ability to test specific hypotheses regarding similarities between the trees of different segments.
+Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and therefore insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
+If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as $\frac{2\times [A:B]}{[A:A]+[B:B]}$, where [A:B] is $\Delta$TMRCA of a pair of tips between tree A and tree B and [A:A] is $\Delta$TMRCA between the same pair of tips between tree A and a replicate of tree A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
 
-\subsection*{Work in progress}
-Currently running "null" traited trees where trait maps have been rearranged (keeping the same ratio of trait values).
+We also attempted to calculate subtree prune and regraft (SPR) distances between trees using rSPR \cite{whidden2009,whidden2010,whidden2013}.
+However, because the time taken to calculate the SPR distances depends on the SPR distances themselves, we find that SPR distance calculations over posterior sets of trees are impractical for all but the most similar trees.
+
+\subsection*{Identifying linked sites in PB1, PB2 and HA segments}
+We attempted to identify amino acid sites that exhibit linkage disequilibrium (LD) between PB1, PB2 and HA segments.
+We chose the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic for biallelic loci, but also quantifies LD when there are more than two alleles per locus.
+
+We also looked at amino acid sites which are different and fixed between sequences that were labelled as Victoria or Yamagata earlier and which possess entirely Victoria-derived or entirely Yamagata-derived PB1, PB2 and HA segments.
+For the latter analysis we split our PB1, PB2 and HA amino acid alignments into temporal 5-year bins ranging from 1984 to 2013. 
+By scanning Victoria and Yamagata alignments at each 5-year bin we can identify sites which are different and fixed between the two lineages and how the sites have evolved over the past 30 years.
+This was done to control for sites in PB1, PB2 and HA proteins discovered by the LD analysis which, in addition to being more sensitive, might also be linked due to co-evolving sub-lineages within Vic and Yam lineages.
 
@@
 Some viruses with mixed-lineage genomes have become common ancestors to many later strains (i.e.
-the "trunk" of the phylogenetic tree) and the segment lineages involved so successful, as to become fixed within the influenza B population circulating today.
+the `trunk' of the phylogenetic tree) and the segment lineages involved so successful, as to become fixed within the influenza B population circulating today.
+
 We find that Vic lineage PA, NP, NA and MP segments have gone extinct (i.e.
@@
 In addition, there is evidence to suggest that the numbers of sequences that belong to either Vic and Yam lineages of PB1, PB2 and HA segments have been sampled at a ratio close to 0.5 (see Figure \ref{lineageRatiosOverTime}), presumably due to the action of some form of balancing selection, possibly to maintain antigenic diversity.
-Our findings suggest that the diversity in PB1, PB2 and HA segments is not maintained independently, but is very likely a result of co-adaptation between these segments.
+
 By measuring the mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in diversity, which indicate that an inter-lineage reassortment event has taken place.
@@
 Thus, by measuring pairwise distances between differently labelled branches we get a sense of how panmictic different segments are with respect to each other.
+
 If we repeat this procedure for all trees and focus on PB1, PB2 and HA lineage labels, we find that mean pairwise diversity between branches labelled as having for example, a PB1 segment of Vic or Yam lineage in PB2 and HA trees (and other combinations of the 3 trees and their labels), show small drops in diversity initially, but from the 1997 time slice onwards all branches labelled as Vic or Yam have a common ancestor at the root of the tree.
@@
 
+\begin{figure}
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/InfB_betweenDiversity.png}
+	\caption{\textbf{Mean pairwise TMRCA between Vic and Yam branches under PB1, PB2 and HA label sets.}
+PB1, PB2 and HA segment labels indicate that these segments show reciprocal preservation of diversity, which dates back to the split of Vic and Yam lineages.
+All other segments show increasingly more recent TMRCAs between branches labelled as Vic and Yam in PB1, PB2 and HA label sets.}
+	\label{betweenDiversity}
+\end{figure}
+
 \subsection*{Victoria and Yamagata lineages of PB1-PB2-HA are co-adapted segment complexes}
-We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
-Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic or entirely Yam PB1, PB2 and HA segments (i.e.
-where a branch was labelled as having Vic PB1, Vic PB2 and Vic HA) compared to the sum of branch lengths where one or more of the labels is of another lineage (e.g.
-Vic PB1, Yam PB2 and Yam HA).
-Together with Figure \ref{betweenDiversity} it suggests that some inter-lineage reassortment has occured between PB1, PB2 and HA segments soon after the split of Vic and Yam lineages, but the resulting viruses were not fit and did not become the trunk of the tree.
+We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary recently, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
+Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
+
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/InfB_stateTime.png}
+	\caption{\textbf{Amount of evolutionary time spent under different PB1-PB2-HA constellations.}
+All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.
+(comment - based on 1 set of posteriors (comment - give numbers, HPDs)}
+	\label{stateTime}
+\end{figure}
+
+Together with Figure \ref{betweenDiversity} it suggests that some inter-lineage reassortment has occured between PB1, PB2 and HA segments soon after the split of Vic and Yam lineages, but the resulting viruses were not fit and did not become the `trunk' of the tree.
 Viruses bearing mixed-lineage PB1-PB2-HA complexes had the following constellations of PB1, PB2 and HA: VVY (B/Bankok/163/1990-like), YVV (B/Nanchang/630/1994-like) and YVY (B/New York/24/1993-like).
@@
 The reconstructed evolution of Victoria and Yamagata lineages of all segments reveals a considerable degree of autonomy.
-For example, of all the Yamagata lineage segments that were reassorted into Victoria lineage PB1-PB2-HA background giving rise to B/Hong Kong/330/2001-like viruses, most segments went extinct following later reassortments, save for NP, which continued circulating until, at least, 2011.
+For example, of all the Yamagata lineage segments that were reassorted into Victoria lineage PB1-PB2-HA background giving rise to B/Hong Kong/330/2001-like viruses, most segments went extinct following later reassortments, save for NP, which continued circulating until, at the latest, 2011.
 A more dramatic example can be found in the NS segment, where the Vic lineage had been fixed in the population after reassorting into an otherwise entirely Yam-derived genome constellation and then reassorted back into a Vic lineage PB1-PB2-HA background, replacing the NS lineage that was present there.
@@
 	\includegraphics[width=0.95\textwidth]{figures/InfB_tmrcaOT_lines.png}
-	\caption{\textbf{TMRCA over time plot.}}
+	\caption{\textbf{Oldest TMRCA of all surviving lineages over time.}
+PA, NP, NA, MP and NS segments of influenza B viruses show periodic losses of diversity, indicating lineage turn over.
+PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.}
 	\label{tmrcaOT}
@@
 	\centering	
-	\includegraphics[width=0.55\textwidth]{figures/InfB_LineageRatiosOverTime.png}
+	\includegraphics[width=0.75\textwidth]	{figures/InfB_LineageRatiosOverTime.png}
 	\caption{\textbf{Ratio of lineages in the dataset.}}
-	\label{lineageRatios}
+	\label{lineageRatiosOverTime}
 \end{figure}
 
-\begin{figure}[h]
+\begin{table}[h]
 	\centering
-	\includegraphics[width=0.95\textwidth]{figures/InfB_betweenDiversity.png}
-	\caption{\textbf{Ratio of lineages in the dataset.}}
-	\label{betweenDiversity}
-\end{figure}
+	\label{stateTimeTable}
+	\caption{Amount of evolutionary time spent under different PB1-PB2-HA constellations.
+All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.}
+	\makebox[\textwidth][c]{
+	\begin{tabular}{ c | c c c c } 
+	\hline
+	Segment						&	PB1				&	PB2				&	PA				&	HA					\\ \hline	
+	Victoria mean 95\% HPDs		&	\textbf{251.21}	&	\textbf{245.13}	&	\textbf{230.51} 	&	\textbf{245.46} 		\\
+								&	234.43--270.60	&	 226.25--264.23	&	212.94--247.24	&	228.94--262.99 		\\
+	Yamagata mean 95\% HPDs		&	\textbf{228.66} 	&	\textbf{229.65} 	&	\textbf{208.38} 	&	\textbf{192.82} 	 	\\
+								&	212.94--244.51	&	211.80--247.79	&	193.28--224.65	&	179.30--207.69		\\
+	Mixed-lineage mean 95\% HPDs	&	\textbf{30.06} 		&	\textbf{25.06} 		&	\textbf{30.90} 		&	\textbf{14.45} 		 	\\
+								&	26.46--34.16		&	20.54--29.69		&	25.22--36.13		&	10.14--18.80			\\
+	\hline
+	Segment							&	NP				&	NA				&	MP				&	NS				\\ \hline	
+	Victoria mean 95\% HPDs			&	\textbf{220.60} 	&	\textbf{226.73} 	&	\textbf{202.38} 	&	\textbf{222.23}	\\
+									&	201.65--241.03	&	210.91--242.12	&	176.77--227.67	&	193.31--249.61 	\\
+	Yamagata mean 95\% HPDs		 	&	\textbf{204.18} 	&	\textbf{192.27} 	&	\textbf{184.13} 	&	\textbf{210.93} 	\\
+									&	188.68--221.73	&	179.28--205.74	&	165.97--204.10	&	190.02--231.61	\\
+	Mixed-lineage mean 95\% HPDs 		&	\textbf{43.66}		&	\textbf{34.05} 		&	\textbf{39.41} 		&	\textbf{49.36} 		\\
+									&	34.67--50.86		&	27.50--40.90		&	29.80---48.90	&	33.49--62.62		\\
 
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_stateTime.png}
-	\caption{\textbf{Total time spent in "pure" vs "mixed" PB1-PB2-HA}}
-	\label{stateTime}
-\end{figure}
+	\end{tabular}	
+	}
+\end{table}
+
+\subsection*{Co-adaptation candidate sites do not fall within intuitive regions of PB1, PB2 and HA proteins}
+We find that amongst the pairs of sites exhibiting the highest $\chi^{2}_{df}$ statistics none fall within the regions previously described as forming the contacts between PB1 and PB2 proteins of the influenza A polymerase.
+Linkage disequilibrium analysis of the HA segment shows that as many as 11-12\% of the amino acid sites in HA exhibit high linkage disequilibrium with respect to PB1 and PB2 segments (compared to 6-7\% in PB1 and PB2).
+This probably reflects the independent accumulation of advantageous amino acid substitutions within Victoria and Yamagata lineages and sub-clades.
+In fact all amino acid sites in the HA segment exhibiting high linkage disequilibrium with sites on PB1 and PB2 segments are found within the HA1 region of HA.
 
@@
 We can also only detect reassortment events where a lineage has switched from Vic to Yam or \textit{vice versa}, but not when a lineage has switched from a Yam lineage to another Yam lineage, which has occurred in the past (e.g.
-the 2004 Vic lineage PB1-PB2-HA complex reassortment with NP is only visible in the NP tree, since all viruses in our dataset from 1995 onwards, including Vic PB1-PB2-HA complex-bearing ones, had a Yamagata lineage-derived NP).
+the reassortment events giving rise to B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like viruses, which involved a Vic lineage PB1-PB2-HA complex reassorting with NP of Yamagata lineage is only visible in the NP tree, since all viruses in our dataset from 1995 onwards, including Vic PB1-PB2-HA complex-bearing ones, already had a Yamagata lineage-derived NP).
+However, this study focuses on PB1, PB2 and HA segments, all of which have survived as two lineages up to this day and thus are more informative for inferring reassortments.
 
@@
 Our findings suggest that PB1, PB2 and HA segments of Vic and Yam lineages represent co-adapted co-reassorting segment complexes.
-We argue that PB1-PB2-HA complexes have become co-dependent on each other to the extent that any reassortant viruses possessing mixed-lineage PB1-PB2-HA are unfit and possibly never get sampled or, if sampled, never persist for long enough to be fixed in the influenza B population.
-Supporting this, we detect at least 3 genome constellations having mixed-lineage PB1-PB2-HA sampled early in the first decade of the split between Vic and Yam lineages, when, presumably, Vic and Yam lineages of PB1, PB2 and HA segments were still similar at the protein sequence level to produce viable, though unfit, influenza B viruses.
-We interpret lack of reassortment within the PB1-PB2-HA complex across the lineage boundary in recent times as evidence of a considerable degree of co-dependence within the Vic and Yam PB1-PB2-HA lineages, whereby mixed-lineage reassortants are inviable or unsampled by surveillance due to being outcompeted by viruses with pure-lineage PB1-PB2-HA complexes.
+Previous studies have suggested similarities between the evolution of PB1, PB2 and HA segments of influenza B viruses \cite{hiromoto2000,lindstrom1999}, based on topology similarities between phylogenetic trees of the three segments, albeit the trees used in those studies contained limited numbers of sequences sampled somewhat close in time to the split of Victoria and Yamagata lineages.
+Despite the apparent similarities between PB1, PB2 and HA trees, we have detected at least 3 genome constellations having mixed-lineage PB1-PB2-HA sampled early in the first decade of the split between Vic and Yam lineages, when, presumably, Vic and Yam lineages of PB1, PB2 and HA segments were still similar at the amino acid level to produce viable, though presumably unfit, influenza B viruses.
+
+Our results also suggested that all segments have spent very little evolutionary time in the same virions as the 3 mixed-lineage PB1-PB2-HA constellations, compared to either entirely Yam PB1-PB2-HA or entirely Vic PB1-PB2-HA bearing viruses.
+We interpret lack of reassortment within the PB1-PB2-HA complex across the lineage boundary in recent times as evidence of a considerable degree of co-dependence within the Vic and Yam PB1-PB2-HA lineages, whereby recent reassortants bearing mixed-lineage PB1-PB2-HA complexes are inviable or unsampled by surveillance due to being outcompeted by viruses with pure-lineage PB1-PB2-HA complexes.
 
 \subsection*{Potential causes of co-dependence between PB1-PB2-HA}
-Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, e.g. PB1-PB2-PA and HA-NA.
+Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, e.g.
+PB1-PB2-PA and HA-NA \cite{mccullers2004}.
 Though it would be easy to explain co-adaptation between the aforementioned segments by referring to their functional roles (e.g.
-PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities), our findings suggest a more subtle relationship between PB1, PB2 and HA segments.
-It is clear that PB1-PB2-HA segments do not preferentially reassort together, because there have been at least 3 sampled mixed-lineage PB1-PB2-HA complex constellations, which did not become fixed in the population and recent evidence suggests, that, at least in influenza A, reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
-In the absence of clear functional explanations of why PB1, PB2 and HA should be co-adapted, we draw attention to a recent influenza A study, which found that during the process of vaccine creation, whereby recent human influenza A isolates are reassorted with an egg-adapted strain in an attempt to reassort HA-NA of the recent isolate into the otherwise entirely egg-adapted genome background.
-Cobbin et al.
-\cite{cobbin2013} found that the reassortant viruses following this procedure, together with the HA-NA segments the PB1 segment gets co-reassorted more frequently than expected by chance.
-Cobbin et al. suggest a gene regulatory role for the PB1 segment, as its presence or absence has dramatic effects on HA concentration on the surface of virions and total virion production.
-Without more, and more recent, reassortment events we cannot offer more than speculation to the nature of the relationship between PB1, PB2 and HA segments of Vic and Yam lineages.
-We do however see several potential causes for the phenomenon we observe.
-It would be possible that balancing selection is preserving the diversity in one segment, the others hitchhiking along.
-A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (NA of Vic lineage having gone extinct), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient diversity.
-Alternatively, either differences in gene expression or fidelity between Vic and Yam PB1 or PB2 could cause an apparent co-adaptation between PB1, PB2 and HA segments.
-Finally, it is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply become co-adapted to each other, without any one segment being the driver of diversity preservation.
+PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities), our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.
+
+It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together, because there have been at least 3 sampled mixed-lineage PB1-PB2-HA complex constellations, which did not become fixed in the population and recent evidence suggests that, at least in influenza A, reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
+In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we suggest several alternatives.
+
+Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \cite{bergeron2010,fulvini2011}.
+Early influenza A vaccines were derived by infecting chicken eggs with an egg-adapted strain and a seasonal human isolate in the presence of antisera raised against the HA and NA proteins of the egg-adapted strain, thus selecting for reassortants which were egg-adapted (\textit{i.e.} had internal segments derived from the egg-adapted strain) but possessed the HA and NA segments of the seasonal strain.
+However, in addition to producing reassortants with HA-NA derived from the seasonal strain, as intended, the second most frequent class of reassortants produced using this methods were those with PB1, HA and NA derived from seasonal strains \cite{bergeron2010,fulvini2011}.
+
+Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \cite{cobbin2013}.
+Additional evidence for a relationship between PB1 and HA segments in influenza A viruses is given by previous influenza pandemics, which were caused by avian-human influenza A virus reassortants.
+It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments (and in the case of A/H2N2, NA as well) derived from avian influenza A viruses \cite{kawaoka1989}.
+
+Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
+A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
+Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments', which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
+It is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
+This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in hybrids of \textit{Saccharomyces cerevisiae} and \textit{Saccharomyces bayanus} \cite{lee2008}.
 
 \subsection*{The future of influenza B viruses}
-The co-adaptation between Vic PB1-PB2-HA and Yam PB1-PB2-HA complexes is a potential insight into things to come.
-We see two potential alternative paths of evolution.
-One, where more segments will be recruited into these complexes, as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex.
-For some time it seemed like the NP segment would be recruited into the PB1-PB2-HA complex, but recent reassortment events suggest that all Yam NP lineages reassorted into Vic lineage PB1-PB2-HA background are being outcompeted by a Yam NP lineage reassorted most recently (represented by viruses with the B/Brisbane/33/2008-like genome constellations).
-Another alternative is continued reassortment of all non-PB1-PB2-HA segments between Vic and Yam PB1-PB2-HA complexes.
-Finally, it is not impossible that one PB1-PB2-HA complex lineage will go extinct in the future, by chance or by being outcompeted by the other lineage, thus leaving a single-lineage influenza B population.
-Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether such two-lineage dynamics have not occurred in the distant past.
-In the former two scenarios (speciation and co-circulation) we have expectations about the long term trend of reassortment.
+We see two potential paths of evolution for influenza B viruses.
+More segments could be recruited into co-adapted segment complexes (PB1, PB2 and HA segments having started this first), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
+
+This may seem to be the case in Figure \ref{tmrcaOT} where the NP segment, exhibiting the fourth oldest TMRCA by 2011, looks like it might begin co-evolving with the PB1-PB2-HA complexes. 
+However, the continuous measure of diversity between Vic and Yam lineages of PB1, PB2 and HA segments in Figure \ref{betweenDiversity} show that the NP segment is skewed towards very recent TMRCAs, suggesting that a large number of recent isolates are products of recent reassortment between the NP segment and PB1-PB2-HA complexes (in fact, this is due to viruses possessing B/Brisbane/33/2008-like genome constellations, which comprise a large part of recently isolated influenza B virus strains).
+The other alternative is continued reassortment of all non-PB1-PB2-HA segments between Vic and Yam PB1-PB2-HA complexes, or even extinction of one or the other lineage of the PB1-PB2-HA complex.
+
+Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether such two-lineage dynamics have not occurred in the history of influenza B viruses.
 If the influenza B population is on its way to speciation we expect that over time inter-lineage reassortment frequency should decrease, as more and more segment lineages get recruited into co-adapted segment complexes.
-On the other hand, if influenza B viruses continue co-circulating as two PB1-PB2-HA lineages exchanging non-PB1-PB2-HA segments, reassortment frequency is expected to be constant.
+
+On the other hand, influenza B viruses might continue co-circulating as two PB1-PB2-HA lineages exchanging non-PB1-PB2-HA segments, with reassortment frequency being roughly constant.
 Seeing how rarely influenza B viruses reassort across the Vic-Yam lineage boundary, the timescale for finding out which of the aforementioned scenarios, if any, are under way, is impractical at best.
+
+\subsection*{Predictions}
+Using previously developed plasmid systems \cite{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
+We predict that artificially produced viruses with mixed-lineage PB1-PB2-HA complexes will have reduced fitness when compared to viruses with pure-lineage (\textit{i.e.} entirely Vic or entirely Yam) PB1-PB2-HA complexes.
+In addition, we expect the relationship between Vic and Yam lineage PB1, PB2 and HA segments to be dependent on date of segment isolation, as viruses with mixed-lineage PB1-PB2-HA complexes isolated earlier should perform better than viruses with PB1-PB2-HA segments isolated more recently.
+
+However, we also see that epistatic effects might interfere with fitness measurements, if for example non-PB1-PB2-HA segments are also temporally mismatched.
+Ideally, the co-adaptation would be easier to understand by referring to the structures of PB1 and PB2 proteins, as the link between these would be intuitive.
+We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins that fall within the regions that form contacts within the influenza B polymerase heterotrimer, suggesting more subtle roles for sites we have identified.
 
@@
 Similarly, in the absence of sequence data, it should be possible to predict the lineage of PB1, PB2 and HA segments of influenza B viruses by sequencing only one of those segments.
-It remains to be seen whether equivalent or similar links exist between segments of influenza A viruses, which, by far, present a much greater threat to human health worldwide than influenza B viruses.
-In addition, by discovering the underlying mechanism by which PB1, PB2 and HA are co-adapted might reveal potential cracks in the armour of influenza viruses, allowing for the creation of anti-virals to be used with currently deployed oseltamivir and zanamivir in a multi-drug approach similar to the one used against bacteria.
 
-\subsection*{Predictions}
-We see two potential future directions of research.
-Using previously developed plasmid systems it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments.
-We make two predictions for such experiments.
-One, viruses with mixed-lineage PB1-PB2-HA complexes will have reduced fitness compared to pure-lineage PB1-PB2-HA complexes.
-Two, if the relationship between Vic and Yam lineage PB1, PB2 and HA segments has indeed evolved over time, viruses with mixed-lineage PB1-PB2-HA complexes isolated earlier will perform better than viruses with PB1-PB2-HA isolated more recently.
-However, there is potential for pleiotropic effects of other segments to interfere with fitness measurements.
-A second approach is to use sequence data to look for linkage disequilibrium between polymorphisms on PB1, PB2 and HA segments of Vic and Yam lineages.
-Polymorphisms originating close in time on different segments of Vic and Yam lineages could be used to infer the sites potentially responsible for the observed co-adaptation of PB1, PB2 and HA segments.
-This could then pave way for wet-lab experiments where, for example, Vic PB1, PB2 or HA segments would be manipulated at potential sites to get Yam-like segments, which after being reassorted into Yam genome constellations should give reassortants of similar fitness to viruses with entirely Yam-derived PB1-PB2-HA complexes.
+It remains to be seen whether equivalent or similar links exist between segments of influenza A viruses, which present a much greater threat to human health worldwide than influenza B viruses.
+By discovering the underlying mechanism by which PB1, PB2 and HA are co-adapted might reveal potential targets for new classes of antiviral drugs.
 
 \section*{Conclusions}
-In this study we have investigated the patterns of reassortment amongst two influenza B virus segment lineages and found consistent co-reassortment of 3 segments: PB1, PB2 and HA.
-Our findings reveal that Vic and Yam PB1, PB2 and HA segments co-associate amongst themselves with segments of their own lineage, thus resulting in an entirely Vic-lineage-derived PB1-PB2-HA segment complex and an entirely Yam-lineage-derived PB1-PB2-HA segment complex.
-Though we detect strains with mixed-lineage PB1-PB2-HA complexes early after the split between Vic and Yam lineages, we find that they have performed poorly and have comprised only about 10\% of the total evolution undergone by influenza B viruses sampled in this study.
-We argue that this is evidence of a selectively maintained relationship between PB1-PB2-HA complex of Vic and Yam lineages and not due to biased reassortment.
-In addition, the close timing of these mixed-lineage PB1-PB2-HA reassortants close to the split of Vic and Yam lineages suggests that the relationship between each lineage's PB1-PB2-HA complexes have developed over time and in their current state are undetected by surveillance due to either poor fitness or inviability of reassortants with mixed-lineage PB1-PB2-HA complexes.
+In this study we have investigated the patterns of reassortment amongst two influenza B virus segment lineages and found preserved diversity and consistent co-reassortment of 3 segments: PB1, PB2 and HA.
+Our findings reveal that Vic and Yam PB1, PB2 and HA segments co-associate amongst themselves with segments of their own lineage, thus resulting in an entirely Vic-lineage-derived PB1-PB2-HA segment complex and an entirely Yam-lineage-derived PB1-PB2-HA segment complex that co-circulate within the influenza B virus population, occasionally exchanging non-PB1-PB2-HA segments.
+Though we detect strains with mixed-lineage PB1-PB2-HA complexes early after the split between Vic and Yam lineages, we find that they have performed poorly and have circulated for short amounts of evolutionary time when compared to viruses with `pure' lineage PB1-PB2-HA complexes.
+
+We argue that this is evidence for a selectively maintained relationship between PB1-PB2-HA complex of Vic and Yam lineages and not due to biased reassortment.
+In addition, the close timing of these mixed-lineage PB1-PB2-HA reassortants close to the split of Vic and Yam lineages suggests that the relationship between each lineage's PB1-PB2-HA complexes have developed over time and in their current state reassortant viruses with mixed-lineage PB1-PB2-HA complexes are undetected by surveillance due to either poor fitness or inviability.
 

### 2013-10-04 fluB.tex
Adding bits on LD.
Adding provisional LD plots, which still require some work to look
swish and some discussion of sites at high LD.

@@
 We attempted to identify amino acid sites that exhibit linkage disequilibrium (LD) between PB1, PB2 and HA segments.
-We chose the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic for biallelic loci, but also quantifies LD when there are more than two alleles per locus.
-
-We also looked at amino acid sites which are different and fixed between sequences that were labelled as Victoria or Yamagata earlier and which possess entirely Victoria-derived or entirely Yamagata-derived PB1, PB2 and HA segments.
-For the latter analysis we split our PB1, PB2 and HA amino acid alignments into temporal 5-year bins ranging from 1984 to 2013. 
-By scanning Victoria and Yamagata alignments at each 5-year bin we can identify sites which are different and fixed between the two lineages and how the sites have evolved over the past 30 years.
-This was done to control for sites in PB1, PB2 and HA proteins discovered by the LD analysis which, in addition to being more sensitive, might also be linked due to co-evolving sub-lineages within Vic and Yam lineages.
+We chose the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
 
@@
 PB1, PB2 and HA segment labels indicate that these segments show reciprocal preservation of diversity, which dates back to the split of Vic and Yam lineages.
-All other segments show increasingly more recent TMRCAs between branches labelled as Vic and Yam in PB1, PB2 and HA label sets.}
+All other segments show increasingly more recent TMRCAs between branches labelled as Vic and Yam in PB1, PB2 and HA label sets.
+All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
 	\label{betweenDiversity}
@@
 All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.
-(comment - based on 1 set of posteriors (comment - give numbers, HPDs)}
+(comment - based on 1 set of posteriors (comment - give numbers, HPDs)
+All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
 	\label{stateTime}
@@
 PA, NP, NA, MP and NS segments of influenza B viruses show periodic losses of diversity, indicating lineage turn over.
-PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.}
+PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
+All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
 	\label{tmrcaOT}
@@
 
-\subsection*{Co-adaptation candidate sites do not fall within intuitive regions of PB1, PB2 and HA proteins}
-We find that amongst the pairs of sites exhibiting the highest $\chi^{2}_{df}$ statistics none fall within the regions previously described as forming the contacts between PB1 and PB2 proteins of the influenza A polymerase.
+\subsection*{Amino acid sites with high LD between PB1, PB2 and HA}
+We collected sites which exhibit high LD ($\chi^{2}_{df}\ge$0.5).
+By performing hierarchical clustering we identify clusters of sites which exhibit similar LD patterns.
+We identify 4 main classes of sites within the clusters:
+\begin{enumerate}
+	\item Sites that are associated with the split of Vic-Yam lineages in PB1 and PB2 segments.
+	These sites form the largest clusters in our heatmaps and encompass all three segments (PB1, PB2 and HA).
+	PB1 sites 176, 508, PB2 sites 115, 301, 383, 397, 468 and HA sites 91, 163, 270 and 277 are in complete or nearly complete linkage disequilibrium ($\chi^{2}_{df}>$0.97) and are the sites that are fixed and different between Vic and Yam lineages of PB1, PB2 and HA segments.
+	PB1 sites 211, 213 and 214, PB2 sites 382 and 483 and HA sites 96 and 190 are in reduced (but still high) linkage disequilibrium with sites exhibiting high LD due to a combination of homoplastic amino acid substitutions and amino acid sites that have undergone substitutions some time after the Vic-Yam split.
+	\item Sites associated with the B/Beijing/76/1998 genome, a presumed member of a divergent and poorly sampled lineage.
+	The B/Beijing/76/1998 genome possesses substitutions at PB1 site 547, PB2 sites 26, 55, 75, 238, 371 and 743 and HA sites 100 and 263.
+	\item Sites associated with the B/Ibaraki/2/1985 genome, another presumed member of a poorly sampled and divergent lineage.
+	These sites are: 170, 736 on PB1, site 76 on PB2 and sites 265, 289, 302, 317, 328 on HA.
+	\item Sites which do not show complete PB1-PB2-HA associations and are represented by mostly single isolates with `concerted' substitutions on pairs of segments (PB1-PB2, PB1-HA or PB2-HA), but not all three.
+\end{enumerate}
+
 Linkage disequilibrium analysis of the HA segment shows that as many as 11-12\% of the amino acid sites in HA exhibit high linkage disequilibrium with respect to PB1 and PB2 segments (compared to 6-7\% in PB1 and PB2).
-This probably reflects the independent accumulation of advantageous amino acid substitutions within Victoria and Yamagata lineages and sub-clades.
-In fact all amino acid sites in the HA segment exhibiting high linkage disequilibrium with sites on PB1 and PB2 segments are found within the HA1 region of HA.
+This probably reflects the independent accumulation of advantageous amino acid substitutions within individual lineages, as the HA would be expected to accumulate many non-synonymous substitutions via antigenic drift.
+In fact all amino acid sites in the HA segment exhibiting high linkage disequilibrium are found within the HA1 region of HA.
+
+(comment - will require a lot of fiddling to get the heatmaps displayed in a clear way)
+\begin{figure}[h]
+	\centering	
+	\includegraphics[width=0.65\textwidth]	{figures/Chi_PB1_PB2.png}
+	\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and PB2 segments based on $\chi^{2}_{df}$ values.}}
+	\label{ChiPB1PB2}
+\end{figure}
+
+\begin{figure}[h]
+	\centering	
+	\includegraphics[width=0.65\textwidth]	{figures/Chi_PB1_HA.png}
+	\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and HA segments based on $\chi^{2}_{df}$ values.}}
+	\label{ChiPB1HA}
+\end{figure}
+
+\begin{figure}[h]
+	\centering	
+	\includegraphics[width=0.65\textwidth]	{figures/Chi_PB2_HA.png}
+	\caption{\textbf{Hierarchical clustering of amino acid sites on PB2 and HA segments based on $\chi^{2}_{df}$ values.}}
+	\label{ChiPB2HA}
+\end{figure}
 
@@
 We interpret lack of reassortment within the PB1-PB2-HA complex across the lineage boundary in recent times as evidence of a considerable degree of co-dependence within the Vic and Yam PB1-PB2-HA lineages, whereby recent reassortants bearing mixed-lineage PB1-PB2-HA complexes are inviable or unsampled by surveillance due to being outcompeted by viruses with pure-lineage PB1-PB2-HA complexes.
+
+We have identified amino acid sites which exhibit a high degree of linkage disequilibrium between PB1, PB2 and HA segments.
+Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.
+
+None of the sites we have identified on PB1 and PB2 proteins fall within the previously described regions that form the contacts between the two polymerase subunits \cite{sugiyama2009} (comment - citation pertains to influenza A, but all high LD sites are far away from those regions anyway).
+
+(comment - PB1 sites 211, 213 and 214 within nuclear localization signal? PB1 site 176 kind of nearby too)
+(comment - PB1 site 508 is 28 amino acids away from super conserved RdRp motif IV - significant?)
+
+(comment - PB2 sites 382, 383, 397, 468 and 483 fall within cap-binding region of PB2?)
 
@@
 It is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
-This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in hybrids of \textit{Saccharomyces cerevisiae} and \textit{Saccharomyces bayanus} \cite{lee2008}.
+This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \cite{lee2008}.
 

### 2013-10-18 fluB.tex
Adding average LD measures between proteins.
Adding average LD measures between proteins, some text that goes with
it and updated heatmaps by focusing on more common alleles.

@@
 
+(comment - could try normalizing at each tip-tip comparison)
 We express the similarity by referring to 3 descriptive statistics (sum, mean and standard deviation) of the distribution of TMRCA differences ($\Delta$TMRCA) between pairs of tips in the two trees.
@@
 Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and therefore insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
-If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as $\frac{2\times [A:B]}{[A:A]+[B:B]}$, where [A:B] is $\Delta$TMRCA of a pair of tips between tree A and tree B and [A:A] is $\Delta$TMRCA between the same pair of tips between tree A and a replicate of tree A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
+
+If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as $\frac{[A:A']+[B:B']}{2\times [A:B]}$, where [A:B] is $\Delta$TMRCA of a pair of tips between tree A and tree B and [A:A'] is $\Delta$TMRCA between the same pair of tips between tree A and a replicate of tree A (A'), to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
 
 We also attempted to calculate subtree prune and regraft (SPR) distances between trees using rSPR \cite{whidden2009,whidden2010,whidden2013}.
-However, because the time taken to calculate the SPR distances depends on the SPR distances themselves, we find that SPR distance calculations over posterior sets of trees are impractical for all but the most similar trees.
+However, because the time taken to calculate the SPR distances depends on the SPR distances themselves, we find that exact SPR distance calculations over posterior sets of trees are impractical for all but the most similar trees.
+We thus calculated approximate SPR distances for all trees, using the normalization procedure described above.
 
-\subsection*{Identifying linked sites in PB1, PB2 and HA segments}
-We attempted to identify amino acid sites that exhibit linkage disequilibrium (LD) between PB1, PB2 and HA segments.
+\subsection*{Estimating linkage disequilibrium across the influenza B genome}
+We attempted to estimate linkage disequilibrium (LD) between amino acid sites across the influenza B proteome.
 We chose the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
+LD was estimated only at loci where each `allele' was represented by at least two isolates and gaps were ignored and not considered as polymorphisms.
+Ignoring gaps in estimating LD was a practical decision based on some sequences missing nucleotides at the beginning and end of sequences.
+However, we did notice at least one polymorphic indel within the HA protein, which is associated with the Vic-Yam split and not with sequence quality.
+In addition, some influenza B genome constellations have not reassorted for prolonged periods of time and have acquired amino acid substitutions on several segments, resulting in linkage disequilibrium.
+These single-pair associations between sites result in high LD and come from poorly sampled lineages, thus we choose to only focus on alleles that have been sampled at least twice.
+
+We also estimated average pairwise LD between proteins by calculating the proportion of site pairs that exhibit LD $\geq$ 0.1.
 
@@
 We identify 4 main classes of sites within the clusters:
+(comment - re-identify sequences/clades responsible for pairs of sites with high LD)
 \begin{enumerate}
@@
 	PB1 sites 176, 508, PB2 sites 115, 301, 383, 397, 468 and HA sites 91, 163, 270 and 277 are in complete or nearly complete linkage disequilibrium ($\chi^{2}_{df}>$0.97) and are the sites that are fixed and different between Vic and Yam lineages of PB1, PB2 and HA segments.
-	PB1 sites 211, 213 and 214, PB2 sites 382 and 483 and HA sites 96 and 190 are in reduced (but still high) linkage disequilibrium with sites exhibiting high LD due to a combination of homoplastic amino acid substitutions and amino acid sites that have undergone substitutions some time after the Vic-Yam split.
+	PB1 sites 211, 213 and 214, PB2 sites 382 and 483 and HA sites 96 and 190 are in reduced (but still high) linkage disequilibrium with sites exhibiting high LD due to a combination of homoplastic amino acid substitutions and amino acid sites that have undergone substitutions in either the Vic or the Yam part of the tree some time after the Vic-Yam split.
 	\item Sites associated with the B/Beijing/76/1998 genome, a presumed member of a divergent and poorly sampled lineage.
@@
 	\item Sites which do not show complete PB1-PB2-HA associations and are represented by mostly single isolates with `concerted' substitutions on pairs of segments (PB1-PB2, PB1-HA or PB2-HA), but not all three.
+	We have attempted to minimize the number of these sites by focusing only on alleles that were observed in at least two isolates.
 \end{enumerate}
 
-Linkage disequilibrium analysis of the HA segment shows that as many as 11-12\% of the amino acid sites in HA exhibit high linkage disequilibrium with respect to PB1 and PB2 segments (compared to 6-7\% in PB1 and PB2).
-This probably reflects the independent accumulation of advantageous amino acid substitutions within individual lineages, as the HA would be expected to accumulate many non-synonymous substitutions via antigenic drift.
-In fact all amino acid sites in the HA segment exhibiting high linkage disequilibrium are found within the HA1 region of HA.
+\begin{figure}[h]
+	\centering	
+	\includegraphics[width=0.75\textwidth]	{figures/InfB_segmentLD.png}
+	\caption{\textbf{LD comparison between influenza B proteins.}
+Comparison of linkage disequilibrium between amino acid sites on influenza B proteins.
+Numbers on the heatmap are proportion of pairs of polymorphic sites (where each allele is represented by at least 2 isolates) that exhibit LD $\geq$ 0.1.}
+	\label{segmentLD}
+\end{figure}
 
@@
 Our findings suggest that PB1, PB2 and HA segments of Vic and Yam lineages represent co-adapted co-reassorting segment complexes.
-Previous studies have suggested similarities between the evolution of PB1, PB2 and HA segments of influenza B viruses \cite{hiromoto2000,lindstrom1999}, based on topology similarities between phylogenetic trees of the three segments, albeit the trees used in those studies contained limited numbers of sequences sampled somewhat close in time to the split of Victoria and Yamagata lineages.
+Previous studies have suggested similarities between the evolution of PB1, PB2 and HA segments of influenza B viruses \cite{hiromoto2000,lindstrom1999}, based on topology similarities between phylogenetic trees of the three segments, albeit the trees used contained limited numbers of sequences sampled somewhat close in time to the split of Victoria and Yamagata lineages.
 Despite the apparent similarities between PB1, PB2 and HA trees, we have detected at least 3 genome constellations having mixed-lineage PB1-PB2-HA sampled early in the first decade of the split between Vic and Yam lineages, when, presumably, Vic and Yam lineages of PB1, PB2 and HA segments were still similar at the amino acid level to produce viable, though presumably unfit, influenza B viruses.
@@
 
-None of the sites we have identified on PB1 and PB2 proteins fall within the previously described regions that form the contacts between the two polymerase subunits \cite{sugiyama2009} (comment - citation pertains to influenza A, but all high LD sites are far away from those regions anyway).
+None of the sites we have identified on PB1 and PB2 proteins fall within the previously described regions that form the contacts between the two polymerase subunits \cite{sugiyama2009}.
 
-(comment - PB1 sites 211, 213 and 214 within nuclear localization signal? PB1 site 176 kind of nearby too)
-(comment - PB1 site 508 is 28 amino acids away from super conserved RdRp motif IV - significant?)
-
-(comment - PB2 sites 382, 383, 397, 468 and 483 fall within cap-binding region of PB2?)
+Sites 211, 213 and 214 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
+Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211, 213 and 214 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
+Previous research \cite{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
+Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 positions 211 and 214 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at positions 211 and 214 and threonine (T) at position 213.
+It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
+Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \cite{fodor2004}.
+Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \cite{guilligay2008}.
+Again, we cannot determine whether this finding is significant or not.
 

### 2013-10-24 fluB.tex
Adding extra provisional plots.
Adding plots of pairwise tree-to-tree comparisons: normalized SPR and
mean delta TMRCA.
Also adding MDS plot with segments, where distances are proportion of
site pairs that show LD>0.5 (i.e. of the included heatmap).

@@
 
+\subsection*{All segments reassort equally but some are more equal than others}
+(comment - normalized approximate SPR distance plot suggests that all segments reassort to the same extent, but the mean \Delta TMRCA plot suggests that some segment pairs occassionally have TMRCAs similar to themselves i.e.
+only short-evolutionary-distance reassortments are observed)
+(comment - also Sam expected that MP and NA would come out the way they did)
+
 \subsection*{Potential causes of co-dependence between PB1-PB2-HA}

### 2013-12-02 fluB.tex
Revisions.

--- 2013-10-18 fluB.tex
@@
 %%% TITLE %%%
-\title{\vspace{1.0cm} \LARGE \bf fluB}
+\title{\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex}
 
@@
 \affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
-\affil[2]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA.}
-\affil[3]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA.}
+\affil[2]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
+\affil[3]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
 
@@
 \section*{Introduction}
-Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three virus types (influenza A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality.
+Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality.
 Though lacking the degree of antigenic diversity that influenza A viruses possess, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
@@
 Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains.
-The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), thus limiting the virus's available antigenic reservoir. 
-It has been suggested that influenza B viruses use a combination of reassortment and point mutations to generate antigenic diversity \cite{nerome1998,mccullers1999}.
+The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), thus limiting the virus's pandemic potential. 
+Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2013}.
 
-Currently circulating influenza B viruses are comprised of two antigenically distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after two strains, B/Victoria/02/87 and B/Yamagata/16/88, which possess antigenically distinct haemagglutinin (HA) surface glycoproteins.
-The two HA lineages still co-circulate today and are thought to have emerged around 1983 \cite{rota1990}. 
+Currently circulating influenza B viruses are comprised of two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/02/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}.
+These two lineages possess antigenically distinct HA surface glycoproteins allowing them to co-circulate in the human population.
 All other influenza B segments also show a split around that time (our combined estimate across all segments place it \textit{circa} 1982.95 (95\% highest posterior density 1981.53--1984.21) figure \ref{TMRCAhpd}), although the numbers of strains falling on either side of the split are different due to reassortments following the initial split of Yam and Vic lineages in all segments and fixation of one or the other lineage in some segments.
+\tbc{Save this for results.}
 
-%\begin{figure}[h]
-%	\centering		
-%	\includegraphics[width=0.95\textwidth]{figures/InfBtmrcaHPD}
-%	\caption{\textbf{Provisional TMRCA plot.}}
-%	\label{TMRCAhpd}
-%\end{figure}
+Previously, through the use of time-stamped phylogenetic trees, the reassortment history of major clades of influenza B has been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
+\tbc{This analysis was limited because...}
+However, another approach exists that provides a significantly cleaner look at reassortment patterns in influenza B.
+In this approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used as a discrete trait in trees of other segments.
+Reassortment events which result in the replacement of one segment lineage by another show up as changes in the discrete trait along a branch and allow the inference of parameters related to the subtree descended from the node displaying a changed trait (Figure \ref{methodFig}).
+Here, we use discrete traits to model inter-lineage reassortments between Victoria and Yamagata lineage segments to look at lineage persistence times following reassortment events.
 
-Previously, through the use of time-stamped phylogenetic trees, the reassortment history of major clades of influenza B has been reconstructed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
-
-Another approach exists, however, whereby belonging to either the Victoria or Yamagata lineage in the tree of one segment can be used as a discrete trait in trees of other segments.
-Reassortment events which result in the replacement of one segment lineage by another show up as changes in the discrete trait along a branch and allow the inference of parameters related to the subtree descended from the node displaying a changed trait (see Figure \ref{methodFig}).
-Here, we use discrete traits to model inter-lineage reassortments between Victoria and Yamagata lineage segments in BEAST to look at lineage persistence times following reassortment events.
-
-We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage are selected against and, to the best of our knowledge, have not been sampled since 1996.
+We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage are selected against and, to the best of our knowledge, have not been sampled since 1996.
 In other segments (PA, NP, NA, MP and NS) a single lineage has been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
@@
 Label switches thus indicate reassortment events that combine lineages falling on different sides of the bifurcation specified earlier.
-In the example above tip number 6 is determined to be such a reassortant.}
+In the example above tip number 6 is determined to be such a reassortant. \tbc{Give this A and B panels.}}
 	\label{methodFig}
@@
 \subsection*{Sequences and subtyping}
-We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID}. 
+
+\tbc{Be careful with the word `subtyping'.}
+
+We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} \tbc{dating from 1984(?) to 2011(?)}. 
 Each strain was assigned to a lineage (either Vic or Yam) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
@@
 Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/02/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).
-Seven of these where then used as discrete states in each of the trees (e.g. PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values) in BEAST \cite{drummond2012}.
 
 \subsection*{Building trees}
-Temporal phylogenies were recovered for each segment using the HKY model of nucleotide substitution, 3 codon partitions, robust counting \cite{obrien2009} and GMRF Bayesian skyride \cite{minin2008} as the demographic model over the course of an MCMC chain (3 parallel runs of 200 million states, sampling every 20000 steps).
-In addition we used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip dates would change and be estimated over the course of the MCMC chain (available in BEAST v1.7.4 \cite{drummond2012} via XML file editing as a date precision parameter).
+Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
+Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate HKY models for each of the 3 codon partitions, and additionally included robust counting \cite{obrien2009} to report realized substitutions.
+We modeled the demographic process using a GMRF Bayesian skyride model \cite{minin2008}.
+We used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
+We ran 3 independent MCMC chains, each with 200 million states, sampled every 20000 steps.
+Samples were combined across runs \tbc{?} to give \tbc{XXX} total samples for each segment and convergence was assessed by \tbc{...}.
+
+\tbc{This belongs in this section:} Seven of these were then used as discrete states in each of the trees (e.g. PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values).
 
@@
 
+\tbc{I think that for these first two sections, you just need figures \ref{tmrcaOT} and \ref{lineageRatiosOverTime}.
+Figure \ref{betweenDiversity} seems redundant, and the multiple `perspectives' is confusing to be the first thing shown.}
+
 \subsection*{PA, NP, NA, MP and NS segments have periodically lost diversity}
-The split of Vic and Yam lineages can be found in all segments.
+
+The split of Vic and Yam lineages can be seen in all segments.
 Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
@@
 \subsection*{Victoria and Yamagata lineages of PB1-PB2-HA are co-adapted segment complexes}
+
+\tbc{Did you decide to shelve the REPU statistics?
+At least in theory, this should be much more convincing that PB1-PB2-HA are held together by selection rather than `mutation' (i.e.
+due to packaging signals).
+REPU should be a stronger argument than figure \ref{betweenDiversity}.
+....
+I see.
+You can't do something like REPU if you don't observe any transient reassortants in the first place.
+This suggests either that selection is very strong or mutation is very rare.
+But later on you mention that `there have been at least 3 sampled mixed-lineage PB1-PB2-HA complex constellations'.
+More description of these constellations would be helpful, when they appear and how long they last.
+Also report how long the average lineage persists.}
+
+\tbc{Can you reject the hypothesis that PB1 and PB2 stuck to HA stochastically?}
+
 We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary recently, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
@@
 \subsection*{Other segments evolve with considerable independence}
+
+\tbc{Are the observed rates of extinction / co-option fit with the effective population of the flu B population?
+I.e.
+we generally get 10 years back to a common ancestor within a lineage / segment.
+Can we think of the extinction of segments as an aspect of there existing a loosely panmictic flu B population for everything but PB1, PB2 and HA?}
+
 The reconstructed evolution of Victoria and Yamagata lineages of all segments reveals a considerable degree of autonomy.
@@
 PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
-All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs). \tbc{I think this would be more intuitive if the y-axis (years) were reversed with 1980 at the top.}}
 	\label{tmrcaOT}
@@
 	\caption{Amount of evolutionary time spent under different PB1-PB2-HA constellations.
-All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.}
+All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes. \tbc{Seems better in figure form.}}
 	\makebox[\textwidth][c]{
@@
 
+\subsection*{Railroad schematic plot of influenza B reassortment history}
+
+\tbc{Title is approximate.}
+
+\subsection*{Linkage disequilibrium analysis confirms phylogenetic findings}
+
+\tbc{The 8x8 MDS analysis suggests that PB1-PB2-HA are the only segments with substantial coevolution.}
+
 \subsection*{Amino acid sites with high LD between PB1, PB2 and HA}
+
+\tbc{Look for signal in the sites analysis for antigenic sites in HA.
+These sites can be seen here: https://github.com/trvrb/stem/tree/master/data.
+Although this analysis will be difficult if LD varies with site-specific substitution rate....}
+
 We collected sites which exhibit high LD ($\chi^{2}_{df}\ge$0.5).
@@
 
+\subsection*{Rehash streamlined version of results for people who don't read results sections.}
+
+\tbc{Title is approximate.}
+
 \subsection*{Limitations of current study}
@@
 Again, we cannot determine whether this finding is significant or not.
+
+\subsection*{All segments reassort equally but some are more equal than others}
+(comment - normalized approximate SPR distance plot suggests that all segments reassort to the same extent, but the mean $\Delta$ TMRCA plot suggests that some segment pairs occassionally have TMRCAs similar to themselves i.e.
+only short-evolutionary-distance reassortments are observed)
+(comment - also Sam expected that MP and NA would come out the way they did)
 

### 2013-12-03 fluB.tex
A few text changes... much more to go.

@@
 Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains.
-The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), thus limiting the virus's pandemic potential. 
+The only known persistent host of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), thus limiting the virus's potential to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir.
 Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2013}.
@@
 \tbc{This analysis was limited because...}
-However, another approach exists that provides a significantly cleaner look at reassortment patterns in influenza B.
-In this approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used as a discrete trait in trees of other segments.
+Here, we introduce another approach that reveals a evolutionarily intriguing pattern of reassortment in influenza B.
+In this approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual viruses in the tree of the other segments.
 Reassortment events which result in the replacement of one segment lineage by another show up as changes in the discrete trait along a branch and allow the inference of parameters related to the subtree descended from the node displaying a changed trait (Figure \ref{methodFig}).
@@
 
-We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage are selected against and, to the best of our knowledge, have not been sampled since 1996.
+We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as distinct Victoria (Vic) and Yamagata (Yam) lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage are selected against and, to the best of our knowledge, have not been sampled since 1996.
 In other segments (PA, NP, NA, MP and NS) a single lineage has been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.

### 2013-12-04 fluB.tex
Some changes to text, update to figures.
Text will require a lot of revision.
There's two new figures (SPR and delta TMRCA plots) and a new colour
scheme, which has more contrast and doesn't have faint yellow as a
colour.
Also adding an alternative version of TMRCA over time as per Trevor's
suggestion.

@@
 These two lineages possess antigenically distinct HA surface glycoproteins allowing them to co-circulate in the human population.
-All other influenza B segments also show a split around that time (our combined estimate across all segments place it \textit{circa} 1982.95 (95\% highest posterior density 1981.53--1984.21) figure \ref{TMRCAhpd}), although the numbers of strains falling on either side of the split are different due to reassortments following the initial split of Yam and Vic lineages in all segments and fixation of one or the other lineage in some segments.
-\tbc{Save this for results.}
+All other influenza B segments also show a split into two lineages around that time.
 
-Previously, through the use of time-stamped phylogenetic trees, the reassortment history of major clades of influenza B has been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
-\tbc{This analysis was limited because...}
+Previously, through the combined use of time-stamped and maximum likelihood phylogenetic trees evolutionary rate, selective pressures and reassortment history of major clades of influenza B has been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
+This analysis was limited because it used limited and different numbers of sequences to infer trees of each segment and relied on maximum likelihood phylogenies to infer the history of reassortments.
 Here, we introduce another approach that reveals a evolutionarily intriguing pattern of reassortment in influenza B.
@@
 
-\subsection*{Sequences and subtyping}
+\subsection*{Sequences and lineage assignment}
 
-\tbc{Be careful with the word `subtyping'.}
-
-We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} \tbc{dating from 1984(?) to 2011(?)}. 
+We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012. 
 Each strain was assigned to a lineage (either Vic or Yam) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
@@
 We used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
-We ran 3 independent MCMC chains, each with 200 million states, sampled every 20000 steps.
-Samples were combined across runs \tbc{?} to give \tbc{XXX} total samples for each segment and convergence was assessed by \tbc{...}.
+We ran 3 independent MCMC chains, each with 200 million states, sampled every 20000 steps and discarded the first 10\% of the MCMC states as burn-in.
+Samples were combined across runs to give 27000 total samples for each segment and convergence was assessed by Tracer.
 
-\tbc{This belongs in this section:} Seven of these were then used as discrete states in each of the trees (e.g. PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values).
-
+Every sequence was assigned 7 discrete traits corresponding to the lineages of all other segments with which a strain was isolated (e.g.
+PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values).
 We inferred the ancestral state of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS \cite{lemey2009}).
@@
 \subsection*{Measures of diversity}
-We inferred the diversity of each segment over time by finding the oldest most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on the maximum amount of diversity existing at each time period.
+We inferred the diversity of each segment over time by finding the most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on the maximum amount of diversity existing at each time period.
 In addition, we calculated mean pairwise TMRCA between branches inferred to be in virions possessing Vic and Yam lineage PB1, PB2 and HA.
@@
 \subsection*{Tree to tree similarities}
-In order to get a measure of tree to tree similarity we first extracted 1000 (sampling frequency roughly equal to 1 tree/180000 MCMC states) trees from the posterior distributions of trees of each segment.
+In order to get a measure of tree to tree similarity we first extracted 900 trees from the posterior distributions of trees for each segment.
+These were combined across 3 independent MCMC chains to give a total of 2700 trees.
 Pairwise comparison between each pair of segments was performed by comparing trees at each MCMC state.
 
-(comment - could try normalizing at each tip-tip comparison)
-We express the similarity by referring to 3 descriptive statistics (sum, mean and standard deviation) of the distribution of TMRCA differences ($\Delta$TMRCA) between pairs of tips in the two trees.
+We express the similarity of two trees by referring to the mean of the distribution of TMRCA differences ($\Delta$TMRCA) between pairs of tips in the two trees.
 These parameters are estimated for each pairing of trees at each MCMC state and provides us with 95\% highest posterior density intervals for those parameters, thus giving us the ability to test specific hypotheses regarding similarities between the trees of different segments.
@@
 Ignoring gaps in estimating LD was a practical decision based on some sequences missing nucleotides at the beginning and end of sequences.
-However, we did notice at least one polymorphic indel within the HA protein, which is associated with the Vic-Yam split and not with sequence quality.
+However, there is at least one polymorphic indel within the HA protein, which is associated with the Vic-Yam split and not with sequence quality.
 In addition, some influenza B genome constellations have not reassorted for prolonged periods of time and have acquired amino acid substitutions on several segments, resulting in linkage disequilibrium.
-These single-pair associations between sites result in high LD and come from poorly sampled lineages, thus we choose to only focus on alleles that have been sampled at least twice.
+These single site pairs result in high LD and come from poorly sampled lineages, thus we choose to only focus on alleles that have been sampled at least twice.
 
-We also estimated average pairwise LD between proteins by calculating the proportion of site pairs that exhibit LD $\geq$ 0.1.
+We also calculated mean LD for all pairs of segments to quantify associations between different segments.
+
+\subsection*{Confirming findings with a larger dataset using lineage assignment}
+The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates were downloaded from GISAID.
+Neighbour-joining trees of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/02/87 or B/Yamagata/16/88 sequences.
+Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted.
 
@@
 Similarly, Vic lineage NS has become fixed in the viral population as well.
-By looking at the oldest most recent common ancestor of all lineages existing within the tree of each segment at yearly timeslices (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby sampled sequences trace their ancestry to increasingly more recent common ancestors.
+By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.
 
@@
 By measuring the mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in diversity, which indicate that an inter-lineage reassortment event has taken place.
-In example \ref{ARGexample}, if, over time slices, we measured mean pairwise distance between red and blue branches in tree 1, after they have been reciprocally labelled with each tip's position (with respect to some pre-defined bifurcation) in tree 2, we would find a drop in this measure, since the distance between tip 6 and tip 5 is now shorter and does not go through the root of the tree.
+In example \ref{methodFig}, if, over time slices, we measured mean pairwise distance between red and blue branches in tree 1, after they have been reciprocally labelled with position of each tip in tree 2 (with respect to some pre-defined bifurcation), we would find a drop in this measure, since the distance between tip 6 and tip 5 is now shorter and does not go through the root of the tree.
 Thus, by measuring pairwise distances between differently labelled branches we get a sense of how panmictic different segments are with respect to each other.
@@
 
-\begin{figure}
+\begin{figure}[h]
 	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_betweenDiversity.png}
-	\caption{\textbf{Mean pairwise TMRCA between Vic and Yam branches under PB1, PB2 and HA label sets.}
-PB1, PB2 and HA segment labels indicate that these segments show reciprocal preservation of diversity, which dates back to the split of Vic and Yam lineages.
-All other segments show increasingly more recent TMRCAs between branches labelled as Vic and Yam in PB1, PB2 and HA label sets.
+	\includegraphics[width=0.95\textwidth]{figures/InfB_stateTime.png}
+	\caption{\textbf{Amount of evolutionary time spent under different PB1-PB2-HA constellations.}
+All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.
+(comment - based on 1 set of posteriors (comment - give numbers, HPDs)
 All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
-	\label{betweenDiversity}
+	\label{stateTime}
 \end{figure}
@@
 
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_stateTime.png}
-	\caption{\textbf{Amount of evolutionary time spent under different PB1-PB2-HA constellations.}
-All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.
-(comment - based on 1 set of posteriors (comment - give numbers, HPDs)
-All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
-	\label{stateTime}
-\end{figure}
-
-Together with Figure \ref{betweenDiversity} it suggests that some inter-lineage reassortment has occured between PB1, PB2 and HA segments soon after the split of Vic and Yam lineages, but the resulting viruses were not fit and did not become the `trunk' of the tree.
-Viruses bearing mixed-lineage PB1-PB2-HA complexes had the following constellations of PB1, PB2 and HA: VVY (B/Bankok/163/1990-like), YVV (B/Nanchang/630/1994-like) and YVY (B/New York/24/1993-like).
-We do not show results for other lineage labels as none of them show this pattern of reciprocally preserved diversity with any other segment.
+\subsection*{PB1-PB2-HA reassortants are rare and transient}
+Figures \ref{betweenDiversity} and \ref{stateTime} suggests that some inter-lineage reassortment has occured between PB1, PB2 and HA segments soon after the split of Vic and Yam lineages, but the resulting viruses were not fit and did not become the `trunk' of the tree.
+From complete genome data we have identified 3 classes of PB1-PB2-HA reassortants with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like), YVV (B/Nanchang/630/1994-like) and YVY (B/New York/24/1993-like).
+An additional class of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like) were found when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
+These reassortants are represented by 16 sequences and were isolated in New Zealand and Australia between June 9 and November 12 in 2005.
+In all cases PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
 
@@
 PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
-All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs). \tbc{I think this would be more intuitive if the y-axis (years) were reversed with 1980 at the top.}}
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs). \tbc{I think this would be more intuitive if the y-axis (years) were reversed with 1980 at the top.}}
 	\label{tmrcaOT}
@@
 
-\begin{table}[h]
+\begin{figure}
 	\centering		
-	\label{stateTimeTable}
-	\caption{Amount of evolutionary time spent under different PB1-PB2-HA constellations.
-All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes. \tbc{Seems better in figure form.}}
-	\makebox[\textwidth][c]{
-	\begin{tabular}{ c | c c c c } 
-	\hline
-	Segment						&	PB1				&	PB2				&	PA				&	HA					\\ \hline	
-	Victoria mean 95\% HPDs		&	\textbf{251.21}	&	\textbf{245.13}	&	\textbf{230.51} 	&	\textbf{245.46} 		\\
-								&	234.43--270.60	&	 226.25--264.23	&	212.94--247.24	&	228.94--262.99 		\\
-	Yamagata mean 95\% HPDs		&	\textbf{228.66} 	&	\textbf{229.65} 	&	\textbf{208.38} 	&	\textbf{192.82} 	 	\\
-								&	212.94--244.51	&	211.80--247.79	&	193.28--224.65	&	179.30--207.69		\\
-	Mixed-lineage mean 95\% HPDs	&	\textbf{30.06} 		&	\textbf{25.06} 		&	\textbf{30.90} 		&	\textbf{14.45} 		 	\\
-								&	26.46--34.16		&	20.54--29.69		&	25.22--36.13		&	10.14--18.80			\\
-	\hline
-	Segment							&	NP				&	NA				&	MP				&	NS				\\ \hline	
-	Victoria mean 95\% HPDs			&	\textbf{220.60} 	&	\textbf{226.73} 	&	\textbf{202.38} 	&	\textbf{222.23}	\\
-									&	201.65--241.03	&	210.91--242.12	&	176.77--227.67	&	193.31--249.61 	\\
-	Yamagata mean 95\% HPDs		 	&	\textbf{204.18} 	&	\textbf{192.27} 	&	\textbf{184.13} 	&	\textbf{210.93} 	\\
-									&	188.68--221.73	&	179.28--205.74	&	165.97--204.10	&	190.02--231.61	\\
-	Mixed-lineage mean 95\% HPDs 		&	\textbf{43.66}		&	\textbf{34.05} 		&	\textbf{39.41} 		&	\textbf{49.36} 		\\
-									&	34.67--50.86		&	27.50--40.90		&	29.80---48.90	&	33.49--62.62		\\
+	\includegraphics[width=0.95\textwidth]{figures/InfB_betweenDiversity.png}
+	\caption{\textbf{Mean pairwise TMRCA between Vic and Yam branches under PB1, PB2 and HA label sets.}
+PB1, PB2 and HA segment labels indicate that these segments show reciprocal preservation of diversity, which dates back to the split of Vic and Yam lineages.
+All other segments show increasingly more recent TMRCAs between branches labelled as Vic and Yam in PB1, PB2 and HA label sets.
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+	\label{betweenDiversity}
+\end{figure}
 
-	\end{tabular}	
-	}
-\end{table}
+\subsection*{Reassortment rates are not significantly different between PB1, PB2 and HA}
+
+\begin{figure}
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/InfB_normalizedApproxSPR.png}
+	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+	\label{SPRdistances}
+\end{figure}
+
+\begin{figure}
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/InfB_normalizedMuDeltaTMRCA.png}
+	\caption{\textbf{Normalized mean $\Delta$TMRCA statistics between pairs of segments.}
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+	\label{deltaTMRCA}
+\end{figure}
 
@@
 
-We collected sites which exhibit high LD ($\chi^{2}_{df}\ge$0.5).
-By performing hierarchical clustering we identify clusters of sites which exhibit similar LD patterns.
-We identify 4 main classes of sites within the clusters:
-(comment - re-identify sequences/clades responsible for pairs of sites with high LD)
-\begin{enumerate}
-	\item Sites that are associated with the split of Vic-Yam lineages in PB1 and PB2 segments.
-	These sites form the largest clusters in our heatmaps and encompass all three segments (PB1, PB2 and HA).
-	PB1 sites 176, 508, PB2 sites 115, 301, 383, 397, 468 and HA sites 91, 163, 270 and 277 are in complete or nearly complete linkage disequilibrium ($\chi^{2}_{df}>$0.97) and are the sites that are fixed and different between Vic and Yam lineages of PB1, PB2 and HA segments.
-	PB1 sites 211, 213 and 214, PB2 sites 382 and 483 and HA sites 96 and 190 are in reduced (but still high) linkage disequilibrium with sites exhibiting high LD due to a combination of homoplastic amino acid substitutions and amino acid sites that have undergone substitutions in either the Vic or the Yam part of the tree some time after the Vic-Yam split.
-	\item Sites associated with the B/Beijing/76/1998 genome, a presumed member of a divergent and poorly sampled lineage.
-	The B/Beijing/76/1998 genome possesses substitutions at PB1 site 547, PB2 sites 26, 55, 75, 238, 371 and 743 and HA sites 100 and 263.
-	\item Sites associated with the B/Ibaraki/2/1985 genome, another presumed member of a poorly sampled and divergent lineage.
-	These sites are: 170, 736 on PB1, site 76 on PB2 and sites 265, 289, 302, 317, 328 on HA.
-	\item Sites which do not show complete PB1-PB2-HA associations and are represented by mostly single isolates with `concerted' substitutions on pairs of segments (PB1-PB2, PB1-HA or PB2-HA), but not all three.
-	We have attempted to minimize the number of these sites by focusing only on alleles that were observed in at least two isolates.
-\end{enumerate}
+We perform hierarchical clustering to identify clusters of sites which exhibit similar LD patterns across pairs of segments.
 

### 2013-12-09 fluB.tex
Minor text additions.
TODO soon-ish:
LD MDS plots, MDS vs LD distances plots.

@@
 
-\subsection*{Confirming findings with a larger dataset using lineage assignment}
+\subsection*{Confirming findings with a larger dataset and lineage assignment to sequences}
 The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates were downloaded from GISAID.
@@
 We also find evidence of co-reassortment in PB1, PB2 and HA segments: in Figure \ref{betweenDiversity} all three graphs become identical after 1997, indicating that the same PB1, PB2 and HA labels have been assigned to the same branches in trees of all other segments.
-Our findings therefore suggest that a single configuration of Yam and Vic lineages of PB1, PB2 and HA segments has been preserved since the split of Vic and Yam lineages and have not reassorted with respect to each other, but rather have co-reassorted as a single unit, from 1997 onwards.
 
@@
 \subsection*{Linkage disequilibrium analysis confirms phylogenetic findings}
+Linkage disequilibrium analyses (Figure \ref{segmentLD}) confirm our findings from phylogenetic analyses that PB1, PB2 and HA segments are linked.
+We observe much higher mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
+In some cases this value exceed the estimated LD of amino acid sites within the same protein (e.g. PB1 and HA).
+Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD.
+These sites define the split between Vic and Yam lineages within those segments.
+In addition, there are sites on PB1, PB2 and HA proteins which also show high (albeit smaller) LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
 
@@
 Comparison of linkage disequilibrium between amino acid sites on influenza B proteins.
-Numbers on the heatmap are proportion of pairs of polymorphic sites (where each allele is represented by at least 2 isolates) that exhibit LD $\geq$ 0.1.}
+Numbers on the heatmap are mean $\chi^{2}_{df}$ between pairs of proteins.}
 	\label{segmentLD}
@@
 \subsection*{All segments reassort equally but some are more equal than others}
-(comment - normalized approximate SPR distance plot suggests that all segments reassort to the same extent, but the mean $\Delta$ TMRCA plot suggests that some segment pairs occassionally have TMRCAs similar to themselves i.e.
-only short-evolutionary-distance reassortments are observed)
-(comment - also Sam expected that MP and NA would come out the way they did)
+The 95\% HPD intervals of normalized approximate SPR distances between pairs of segments (Figure \ref{SPRdistances}) overlap, suggesting there is no evidence of differences in reassortment rate between all segments.
+However, SPR distances would only approximate the true number of reassortments.
+Together with an approximation of SPR distance, this might be an indication of lack of power to detect differences in reassortment rate.
 
@@
 
-It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together, because there have been at least 3 sampled mixed-lineage PB1-PB2-HA complex constellations, which did not become fixed in the population and recent evidence suggests that, at least in influenza A, reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
+It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations (which did not become fixed in the population), our estimates of the number of reassortments do not differ significantly between all segments (Figure \ref{SPRdistances} and recent evidence suggests that, at least in influenza A, reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
 In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we suggest several alternatives.
@@
 It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments (and in the case of A/H2N2, NA as well) derived from avian influenza A viruses \cite{kawaoka1989}.
+However, there have also been reassortant influenza viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \cite{gregory2002} and H1N1/09 in 2009 \cite{smith2009}.
 

### 2013-12-13 fluB.tex
Minor text changes, beginning section on SPR and dTMRCA results.

@@
 
-Previously, through the combined use of time-stamped and maximum likelihood phylogenetic trees evolutionary rate, selective pressures and reassortment history of major clades of influenza B has been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
+Previously, through the combined use of time-stamped and maximum likelihood phylogenetic trees evolutionary rate, selective pressures and reassortment history of major clades of influenza B have been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
 This analysis was limited because it used limited and different numbers of sequences to infer trees of each segment and relied on maximum likelihood phylogenies to infer the history of reassortments.
@@
 In this approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual viruses in the tree of the other segments.
-Reassortment events which result in the replacement of one segment lineage by another show up as changes in the discrete trait along a branch and allow the inference of parameters related to the subtree descended from the node displaying a changed trait (Figure \ref{methodFig}).
-Here, we use discrete traits to model inter-lineage reassortments between Victoria and Yamagata lineage segments to look at lineage persistence times following reassortment events.
+Reassortment events which result in the replacement of one segment lineage by another show up as changes in the discrete trait along a branch (Figure \ref{methodFig}).
+Here, we use discrete traits to reconstruct major reassortment events in the recent history of influenza B viruses and to partition diversity across segments of influenza B virus strains to quantify reassortment dynamics over time.
 
-We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as distinct Victoria (Vic) and Yamagata (Yam) lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage are selected against and, to the best of our knowledge, have not been sampled since 1996.
+We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as distinct Victoria (Vic) and Yamagata (Yam) lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and once isolated belong to transient lineages.
 In other segments (PA, NP, NA, MP and NS) a single lineage has been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
@@
 Because the posterior set of trees have branches labelled with the inferred lineage in all other segments, we can detect inter-lineage reassortments whenever a trait transition is observed (i.e.
-Y to V or V to Y, see Figure \ref{methodFig}) between a node and its descendant in any segment.
+Yam to Vic or Vic to Yam, see Figure \ref{methodFig}) between a node and its descendant in any segment.
 In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
@@
 Neighbour-joining trees of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/02/87 or B/Yamagata/16/88 sequences.
-Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted.
+Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.
 
@@
 
-By measuring the mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in diversity, which indicate that an inter-lineage reassortment event has taken place.
-In example \ref{methodFig}, if, over time slices, we measured mean pairwise distance between red and blue branches in tree 1, after they have been reciprocally labelled with position of each tip in tree 2 (with respect to some pre-defined bifurcation), we would find a drop in this measure, since the distance between tip 6 and tip 5 is now shorter and does not go through the root of the tree.
-Thus, by measuring pairwise distances between differently labelled branches we get a sense of how panmictic different segments are with respect to each other.
+By measuring the mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in diversity, which indicate that an inter-lineage reassortment event has taken place and thus is a qualitative measure of panmixis between pairs of segments.
 
-If we repeat this procedure for all trees and focus on PB1, PB2 and HA lineage labels, we find that mean pairwise diversity between branches labelled as having for example, a PB1 segment of Vic or Yam lineage in PB2 and HA trees (and other combinations of the 3 trees and their labels), show small drops in diversity initially, but from the 1997 time slice onwards all branches labelled as Vic or Yam have a common ancestor at the root of the tree.
+Here we focus on PB1, PB2 and HA lineage labels and find that mean pairwise diversity between branches labelled as having for example, a PB1 segment of Vic or Yam lineage in PB2 and HA trees (and other combinations of the 3 trees and their labels), show small drops in diversity initially, but from the 1997 time slice onwards all branches labelled as Vic or Yam have a common ancestor at the root of the tree.
 We also find evidence of co-reassortment in PB1, PB2 and HA segments: in Figure \ref{betweenDiversity} all three graphs become identical after 1997, indicating that the same PB1, PB2 and HA labels have been assigned to the same branches in trees of all other segments.
@@
 All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.
-(comment - based on 1 set of posteriors (comment - give numbers, HPDs)
 All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
@@
 PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
-All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs). \tbc{I think this would be more intuitive if the y-axis (years) were reversed with 1980 at the top.}}
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
 	\label{tmrcaOT}
@@
 
-\subsection*{Reassortment rates are not significantly different between PB1, PB2 and HA}
-
+\subsection*{Reassortment rates are not significantly different between PB1, PB2 and HA, but reassortment distances can be}
+Estimates of normalized SPR distances between pairs of segments (Figure \ref{SPRdistances}) show that there is no evidence of differences in reassortment rates between all segments, based on overlapping 95\% highest posterior density (HPD) intervals.
 \begin{figure}
@@
 
+Unlike normalized SPR distances normalized $\Delta$TMRCA estimates between some segments (Figure \ref{deltaTMRCA}) indicate that reassortment distance (as quantified by $\Delta$TMRCA) between some segments, namely PB1-PB2-HA (but also NA-M1 and PA-M1-NS1), are much more similar to $\Delta$TMRCA estimates between replicate trees.
 \begin{figure}
@@
 \subsection*{Amino acid sites with high LD between PB1, PB2 and HA}
-
-\tbc{Look for signal in the sites analysis for antigenic sites in HA.
-These sites can be seen here: https://github.com/trvrb/stem/tree/master/data.
-Although this analysis will be difficult if LD varies with site-specific substitution rate....}
 

### 2013-12-20 fluB.tex
Include railroad track plot in text.
Using the railroad track plot and elaborating on some lineages with
reference to the plot.
Also adding more information with regards to mixed-lineage reassortant
lineages (number of sequences, dates of first and last isolate).

@@
 
-The split of Vic and Yam lineages can be seen in all segments.
+The split of Vic and Yam lineages can be seen in all segments \cite{chen2008}.
 Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
@@
 Unlike PA, NP, NA, MP and NS segments, which have periodically lost diversity, we find that PB1, PB2 and HA segments have continued to accumulate diversity since the initial split of Vic and Yam lineages (see Figure \ref{tmrcaOT}).
-In addition, there is evidence to suggest that the numbers of sequences that belong to either Vic and Yam lineages of PB1, PB2 and HA segments have been sampled at a ratio close to 0.5 (see Figure \ref{lineageRatiosOverTime}), presumably due to the action of some form of balancing selection, possibly to maintain antigenic diversity.
+In addition, there is evidence to suggest that the numbers of sequences that belong to either Vic and Yam lineages of PB1, PB2 and HA segments have been sampled at a ratio close to 0.5 (see Figure \ref{lineageRatiosOverTime}), presumably due to the action of some form of balancing selection.
 
@@
 
-Here we focus on PB1, PB2 and HA lineage labels and find that mean pairwise diversity between branches labelled as having for example, a PB1 segment of Vic or Yam lineage in PB2 and HA trees (and other combinations of the 3 trees and their labels), show small drops in diversity initially, but from the 1997 time slice onwards all branches labelled as Vic or Yam have a common ancestor at the root of the tree.
-We also find evidence of co-reassortment in PB1, PB2 and HA segments: in Figure \ref{betweenDiversity} all three graphs become identical after 1997, indicating that the same PB1, PB2 and HA labels have been assigned to the same branches in trees of all other segments.
+In Figure \ref{betweenDiversity} we focus on PB1, PB2 and HA lineage labels, which have both Vic and Yam lineages up to recent times.
+We find that there have been some losses of diversity, indicative of label mixing i.e. reassortment, in the early years of the Vic--Yam split (1986--1996) between HA and PB1--PB2 segments.
+After 1997, however, mean pairwise between label diversity increases in PB1, PB2 and HA segments, whereby Vic and Yam labels share a common ancestor only at the root.
+Our data thus suggest that since 1997 there has been no reassortments between Vic and Yam lineages of PB1, PB2 and HA segments.
+In addition mean pairwise between label diversity measurements after 1997 become identical regardless of chosen label, which suggests that PB1, PB2 and HA lineage labels are assigned identically in all trees investigated.
 
@@
 
-\subsection*{Victoria and Yamagata lineages of PB1-PB2-HA are co-adapted segment complexes}
+\subsection*{Recent PB1-PB2-HA constellations are entirely derived from Victoria or Yamagata lineage segments}
 
@@
 Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
+Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations.
 
-\subsection*{PB1-PB2-HA reassortants are rare and transient}
-Figures \ref{betweenDiversity} and \ref{stateTime} suggests that some inter-lineage reassortment has occured between PB1, PB2 and HA segments soon after the split of Vic and Yam lineages, but the resulting viruses were not fit and did not become the `trunk' of the tree.
-From complete genome data we have identified 3 classes of PB1-PB2-HA reassortants with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like), YVV (B/Nanchang/630/1994-like) and YVY (B/New York/24/1993-like).
-An additional class of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like) were found when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
-These reassortants are represented by 16 sequences and were isolated in New Zealand and Australia between June 9 and November 12 in 2005.
+We have identified 3 classes of mixed-lineage PB1-PB2-HA reassortants from our data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5th Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8th Jan 1993 -- 1994).
+An additional class of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was found when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
 In all cases PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
-
-\subsection*{Other segments evolve with considerable independence}
-
-\tbc{Are the observed rates of extinction / co-option fit with the effective population of the flu B population?
-I.e.
-we generally get 10 years back to a common ancestor within a lineage / segment.
-Can we think of the extinction of segments as an aspect of there existing a loosely panmictic flu B population for everything but PB1, PB2 and HA?}
-
-The reconstructed evolution of Victoria and Yamagata lineages of all segments reveals a considerable degree of autonomy.
-For example, of all the Yamagata lineage segments that were reassorted into Victoria lineage PB1-PB2-HA background giving rise to B/Hong Kong/330/2001-like viruses, most segments went extinct following later reassortments, save for NP, which continued circulating until, at the latest, 2011.
-A more dramatic example can be found in the NS segment, where the Vic lineage had been fixed in the population after reassorting into an otherwise entirely Yam-derived genome constellation and then reassorted back into a Vic lineage PB1-PB2-HA background, replacing the NS lineage that was present there.
+In particular reassortment events combining PB1 and PB2 segments of different lineages, e.g.
+B/Nanchang/630/1994-like and B/New York/24/1993-like isolates, exhibit poor sampling and short circulation times.
+In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past.
 
@@
 
+\subsection*{Railroad schematic plot of influenza B reassortment history}
+
+\tbc{Are the observed rates of extinction / co-option fit with the effective population of the flu B population?
+I.e.
+we generally get 10 years back to a common ancestor within a lineage / segment.
+Can we think of the extinction of segments as an aspect of there existing a loosely panmictic flu B population for everything but PB1, PB2 and HA?}
+
+We reconstructed reassortment events detectable by discrete traits.
+Figure \ref{railroadPlot} focuses only on reassortments that have occured after 1990.
+We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV).
+Figure \ref{railroadPlot} shows that reassorting segments appear to evolve with a considerable degree of autonomy.
+For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to B/Iowa/03/2002-like isolates.
+A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has been derived from Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
+
+\begin{figure}
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/RailroadPlotDated.pdf}
+	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
+Victoria and Yamagata lineage genomes are represented by 8 (red or blue, respectively) parallel lines.
+Inter-lineage reassortment events are indicated by lines entering a different genome.
+The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the parent node and mean TMRCA of reassortant clades).
+Lineage extinction dates are not shown accurately.}
+	\label{railroadPlot}
+\end{figure}
+
 \subsection*{Reassortment rates are not significantly different between PB1, PB2 and HA, but reassortment distances can be}
 Estimates of normalized SPR distances between pairs of segments (Figure \ref{SPRdistances}) show that there is no evidence of differences in reassortment rates between all segments, based on overlapping 95\% highest posterior density (HPD) intervals.
+
 \begin{figure}
@@
 Unlike normalized SPR distances normalized $\Delta$TMRCA estimates between some segments (Figure \ref{deltaTMRCA}) indicate that reassortment distance (as quantified by $\Delta$TMRCA) between some segments, namely PB1-PB2-HA (but also NA-M1 and PA-M1-NS1), are much more similar to $\Delta$TMRCA estimates between replicate trees.
+
 \begin{figure}
@@
 \end{figure}
-
-\subsection*{Railroad schematic plot of influenza B reassortment history}
-
-\tbc{Title is approximate.}
 

### 2013-12-21 fluB.tex
Adding recap in discussion.
Rehashing results, adding letters to both panels in methodFig.

@@
 Label switches thus indicate reassortment events that combine lineages falling on different sides of the bifurcation specified earlier.
-In the example above tip number 6 is determined to be such a reassortant. \tbc{Give this A and B panels.}}
+In the example above tip number 6 is determined to be such a reassortant.}
 	\label{methodFig}
@@
 
-\subsection*{Reassortment rates are not significantly different between PB1, PB2 and HA, but reassortment distances can be}
+\subsection*{Numbers of reassortments are not significantly different between all segments, but reassortment distances can be}
 Estimates of normalized SPR distances between pairs of segments (Figure \ref{SPRdistances}) show that there is no evidence of differences in reassortment rates between all segments, based on overlapping 95\% highest posterior density (HPD) intervals.
@@
 \subsection*{Rehash streamlined version of results for people who don't read results sections.}
-
+In this paper we have shown that PB1, PB2 and HA segments of influenza B viruses are the only ones that have maintained both a Vic and a Yam lineage (Figure \ref{LineageRatiosOverTime}).
+Evidence suggests that this is a result of lack of reassortment between Vic and Yam lineages of these segments (Figure \ref{betweenDiversity}) which have evolved co-assorting amino acid sites detectable as high linkage disequilibrium (\ref{segmentLD}).
+We propose that this pattern of non-reassortment is due to the action of selection and not simply biased reassortment.
+Figures \ref{SPRdistance} and \ref{deltaTMRCA} show that despite all pairs of segments having the same reassortment frequency, PB1, PB2 and HA segments exhibit much lower reassortment `distances'.
+In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were much more similar at the nucleotide and amino acid levels.
+The most recent lineage of influenza B viruses with mixed-lineage PB1-PB2-HA complexes (B/Waikato/6/2005-like) only circulated for 5 months before going extinct.
+We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands.
+If both PB1-PB2-HA complex lineages continue circulating in humans we expect more segments to be recruited to the two complexes, resulting in eventual sympatric speciation of influenza B viruses. 
 \tbc{Title is approximate.}

### 2013-12-23 fluB.tex
Adding more text here and there.
Mostly focus on deltaTMRCA and SPR distance discussion and adding more
references to B/Waikato/6/2005-like isolates (with different lineage
HA).

@@
 \subsection*{Limitations of current study}
-The methods used in this paper have limited power to detect reassortments, since lineages have been fixed over time and thus all viruses eventually come to bear either Vic or Yam lineage PA, NP, NA, MP and NS segments.
+The methods used in this paper have limited power to detect reassortments, since lineages have been fixed over time and thus all viruses eventually come to bear either Vic or Yam lineage PA, NP, NA, MP and NS segments with the exception of PB1, PB2 and HA segments.
 We can also only detect reassortment events where a lineage has switched from Vic to Yam or \textit{vice versa}, but not when a lineage has switched from a Yam lineage to another Yam lineage, which has occurred in the past (e.g.
-the reassortment events giving rise to B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like viruses, which involved a Vic lineage PB1-PB2-HA complex reassorting with NP of Yamagata lineage is only visible in the NP tree, since all viruses in our dataset from 1995 onwards, including Vic PB1-PB2-HA complex-bearing ones, already had a Yamagata lineage-derived NP).
+the reassortment events giving rise to B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like viruses in Figure \ref{railroadPlot}, which involved a Vic lineage PB1-PB2-HA complex reassorting with NP of Yamagata lineage is only visible in the NP tree, since all viruses in our dataset from 1995 onwards, including Vic PB1-PB2-HA complex-bearing ones, already had a Yamagata lineage-derived NP).
 However, this study focuses on PB1, PB2 and HA segments, all of which have survived as two lineages up to this day and thus are more informative for inferring reassortments.
@@
 Previous studies have suggested similarities between the evolution of PB1, PB2 and HA segments of influenza B viruses \cite{hiromoto2000,lindstrom1999}, based on topology similarities between phylogenetic trees of the three segments, albeit the trees used contained limited numbers of sequences sampled somewhat close in time to the split of Victoria and Yamagata lineages.
-Despite the apparent similarities between PB1, PB2 and HA trees, we have detected at least 3 genome constellations having mixed-lineage PB1-PB2-HA sampled early in the first decade of the split between Vic and Yam lineages, when, presumably, Vic and Yam lineages of PB1, PB2 and HA segments were still similar at the amino acid level to produce viable, though presumably unfit, influenza B viruses.
+Despite the apparent similarities between PB1, PB2 and HA trees, we have detected at least 4 genome constellations having mixed-lineage PB1-PB2-HA, 3 of which were sampled early in the first decade of the split between Vic and Yam lineages, when, presumably, Vic and Yam lineages of PB1, PB2 and HA segments were still similar at the amino acid level to produce viable, though presumably unfit, influenza B viruses.
+The fourth genome constellation having mixed-lineage PB1-PB2-HA complex was isolated in Australia and New Zealand in 2005 over a 5 month period and have not been detected since.
 
-Our results also suggested that all segments have spent very little evolutionary time in the same virions as the 3 mixed-lineage PB1-PB2-HA constellations, compared to either entirely Yam PB1-PB2-HA or entirely Vic PB1-PB2-HA bearing viruses.
-We interpret lack of reassortment within the PB1-PB2-HA complex across the lineage boundary in recent times as evidence of a considerable degree of co-dependence within the Vic and Yam PB1-PB2-HA lineages, whereby recent reassortants bearing mixed-lineage PB1-PB2-HA complexes are inviable or unsampled by surveillance due to being outcompeted by viruses with pure-lineage PB1-PB2-HA complexes.
+Our results also suggested that all segments have spent very little evolutionary time with the 3 mixed-lineage PB1-PB2-HA constellations, compared to either entirely Yam PB1-PB2-HA or entirely Vic PB1-PB2-HA bearing viruses.
+We interpret lack of reassortment within the PB1-PB2-HA complex across the Vic--Yam lineage boundary in recent times as evidence of a considerable degree of co-dependence within the Vic and Yam PB1-PB2-HA lineages, whereby recent reassortants bearing mixed-lineage PB1-PB2-HA complexes are inviable or unsampled by surveillance due to being outcompeted by viruses with pure-lineage PB1-PB2-HA complexes.
 
 We have identified amino acid sites which exhibit a high degree of linkage disequilibrium between PB1, PB2 and HA segments.
-Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.
+Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three proteins at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.
 
@@
 \subsection*{All segments reassort equally but some are more equal than others}
-The 95\% HPD intervals of normalized approximate SPR distances between pairs of segments (Figure \ref{SPRdistances}) overlap, suggesting there is no evidence of differences in reassortment rate between all segments.
-However, SPR distances would only approximate the true number of reassortments.
-Together with an approximation of SPR distance, this might be an indication of lack of power to detect differences in reassortment rate.
+The 95\% HPD intervals of normalized approximate SPR distances between pairs of segments (Figure \ref{SPRdistances}) overlap, suggesting there is no evidence of differences in the numbers of reassortments (or reassortment rate given as number of SPR moves per tree, data not shown) between all segments.
+There is reason to suspect that our SPR distance analysis simply lacks power.
+We used an approximation of the true SPR distances and SPR distances themselves could only approximate (and underestimate) the actual numbers of reassortments.
+However, exact SPR distances could be estimated for a limited number of segment pairs (PB1--PB2 and PB1--HA) and analyses indicate that after normalization exact and approximate SPR distances are not significantly different (supplementary figure of exact-approximate SPR distance differences and/or correlations of the two?).
+
+Significant differences between pairs of segments can be found when considering `reassortment distance'.
+Our estimates suggest that tips in PB1, PB2 and HA trees have similar TMRCAs (Figure \ref{deltaTMRCA}), suggesting that reassortments between the three rarely bring together lineages that have been evolving separately for prolonged periods of time.
+We cannot determine whether this effect is specifically due to non-reassortment of Vic and Yam PB1-PB2-HA lineages or a generic property of these three segments.
+In addition, we see evidence of this in MP and NA trees (comment - get references for MP-NA association in fluA, otherwise put down as personal communication or commonly held suspicion).
 
@@
 In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we suggest several alternatives.
+
+It is perhaps easiest to explain co-dependence between PB1 and PB2 segments based on their functions as part of the influenza RNA-dependent RNA polymerase (RdRp) heterotrimer.
+Indeed, trees of PB1 and PB2 segments exhibit high similarity in tip--tip TMRCAs (Figure \ref{deltaTMRCA}), suggesting highly similar trees.
+In addition, PB1--PB2 reassortants are the rarest and least persistent amongst mixed-lineage PB1-PB2-HA strains and have not been isolated since 1996.
+
+Explaining the co-dependence between PB1+2 and HA segments is more difficult.
 
@@
 Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
-A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
+A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct in 2002), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
 Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments', which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
@@
 
+Though it is possible that PB1 and PB2 segments are co-adapted and the HA segment is simply hitchhiking along stochastically, we find it unlikely given that B/Waikato/6/2005-like strains, the most recent isolates with a mixed-lineage PB1-PB2-HA complex, circulated for a limited amount of time within a limited geographic area, suggesting that this particular lineage was outcompeted by influenza B strains with `pure' PB1-PB2-HA complexes.
+
 \subsection*{The future of influenza B viruses}
 We see two potential paths of evolution for influenza B viruses.
-More segments could be recruited into co-adapted segment complexes (PB1, PB2 and HA segments having started this first), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
+More segments could be recruited into co-adapted segment complexes (PB1, PB2 and HA segments having started this first, akin to genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
 
 This may seem to be the case in Figure \ref{tmrcaOT} where the NP segment, exhibiting the fourth oldest TMRCA by 2011, looks like it might begin co-evolving with the PB1-PB2-HA complexes. 
-However, the continuous measure of diversity between Vic and Yam lineages of PB1, PB2 and HA segments in Figure \ref{betweenDiversity} show that the NP segment is skewed towards very recent TMRCAs, suggesting that a large number of recent isolates are products of recent reassortment between the NP segment and PB1-PB2-HA complexes (in fact, this is due to viruses possessing B/Brisbane/33/2008-like genome constellations, which comprise a large part of recently isolated influenza B virus strains).
+However, the continuous measure of diversity between Vic and Yam lineages of PB1, PB2 and HA segments in Figure \ref{betweenDiversity} show that the NP segment is skewed towards very recent TMRCAs, suggesting that a large number of recent isolates are products of recent reassortment between the NP segment and PB1-PB2-HA complexes.
+In fact, this is due to viruses possessing B/Brisbane/33/2008-like genome constellations, which comprise a large part of recently isolated influenza B virus strains (Figure \ref{railroadPlot}).
 The other alternative is continued reassortment of all non-PB1-PB2-HA segments between Vic and Yam PB1-PB2-HA complexes, or even extinction of one or the other lineage of the PB1-PB2-HA complex.
@@
 Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether such two-lineage dynamics have not occurred in the history of influenza B viruses.
-If the influenza B population is on its way to speciation we expect that over time inter-lineage reassortment frequency should decrease, as more and more segment lineages get recruited into co-adapted segment complexes.
+If the influenza B population is on its way to speciation we expect that over time inter-lineage reassortment frequency should decrease, as more and more segment lineages would be recruited into co-adapted segment complexes.
 
@@
 In addition, we expect the relationship between Vic and Yam lineage PB1, PB2 and HA segments to be dependent on date of segment isolation, as viruses with mixed-lineage PB1-PB2-HA complexes isolated earlier should perform better than viruses with PB1-PB2-HA segments isolated more recently.
+
+Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1+2-HA reassortants.
 
@@
 In this study we have investigated the patterns of reassortment amongst two influenza B virus segment lineages and found preserved diversity and consistent co-reassortment of 3 segments: PB1, PB2 and HA.
-Our findings reveal that Vic and Yam PB1, PB2 and HA segments co-associate amongst themselves with segments of their own lineage, thus resulting in an entirely Vic-lineage-derived PB1-PB2-HA segment complex and an entirely Yam-lineage-derived PB1-PB2-HA segment complex that co-circulate within the influenza B virus population, occasionally exchanging non-PB1-PB2-HA segments.
-Though we detect strains with mixed-lineage PB1-PB2-HA complexes early after the split between Vic and Yam lineages, we find that they have performed poorly and have circulated for short amounts of evolutionary time when compared to viruses with `pure' lineage PB1-PB2-HA complexes.
+Our findings reveal that Vic and Yam lineages of PB1, PB2 and HA segments co-associate amongst themselves with segments of their own lineage, thus resulting in an entirely Vic-lineage-derived PB1-PB2-HA segment complex and an entirely Yam-lineage-derived PB1-PB2-HA segment complex that co-circulate within the influenza B virus population, occasionally exchanging non-PB1-PB2-HA segments.
+Though we detect strains with mixed-lineage PB1-PB2-HA complexes, we find that they have performed poorly and have circulated for short amounts of evolutionary time when compared to viruses with `pure' lineage PB1-PB2-HA complexes.
 
 We argue that this is evidence for a selectively maintained relationship between PB1-PB2-HA complex of Vic and Yam lineages and not due to biased reassortment.
-In addition, the close timing of these mixed-lineage PB1-PB2-HA reassortants close to the split of Vic and Yam lineages suggests that the relationship between each lineage's PB1-PB2-HA complexes have developed over time and in their current state reassortant viruses with mixed-lineage PB1-PB2-HA complexes are undetected by surveillance due to either poor fitness or inviability.
+Given sufficient time we expect the two PB1-PB2-HA complexes to recruit the rest of the genomic segments into co-adapted and co-reassorting segment complexes eventually resulting in sympatric speciation of influenza B viruses.
 

### 2014-01-05 fluB.tex
Moving and adding text, adding MDS figure.
Adding text to discussion, moving paragraphs around and adding MDS
figure.

@@
 
-\subsection*{Numbers of reassortments are not significantly different between all segments, but reassortment distances can be}
-Estimates of normalized SPR distances between pairs of segments (Figure \ref{SPRdistances}) show that there is no evidence of differences in reassortment rates between all segments, based on overlapping 95\% highest posterior density (HPD) intervals.
+\subsection*{All segments reassort equally but some are more equal than others}
+The 95\% HPD intervals of normalized approximate SPR distances between pairs of segments (Figure \ref{SPRdistances}) overlap, suggesting there is no evidence of differences in the numbers of reassortments (or reassortment rate given as number of SPR moves per tree, data not shown) between all segments.
+There is reason to suspect that our SPR distance analysis simply lacks power.
+We used an approximation of the true SPR distances and SPR distances themselves could only approximate (and underestimate) the actual numbers of reassortments.
+However, exact SPR distances could be estimated for a limited number of segment pairs (PB1-PB2 and PB1-HA) and analyses indicate that after normalization exact and approximate SPR distances are not significantly different (supplementary figure of exact-approximate SPR distance differences and/or correlations of the two?).
+
+Significant differences between pairs of segments can be found when considering `reassortment distance'.
+Our estimates suggest that tips in PB1, PB2 and HA trees have similar TMRCAs (Figure \ref{deltaTMRCA}), suggesting that reassortments between the three rarely bring together lineages that have been evolving separately for prolonged periods of time.
+We cannot determine whether this effect is specifically due to non-reassortment of Vic and Yam PB1-PB2-HA lineages or a generic property of these three segments.
+In addition, we see evidence of this in MP and NA trees (comment - get references for MP-NA association in fluA, otherwise put down as personal communication or commonly held suspicion).
 
@@
 \end{figure}
-
-Unlike normalized SPR distances normalized $\Delta$TMRCA estimates between some segments (Figure \ref{deltaTMRCA}) indicate that reassortment distance (as quantified by $\Delta$TMRCA) between some segments, namely PB1-PB2-HA (but also NA-M1 and PA-M1-NS1), are much more similar to $\Delta$TMRCA estimates between replicate trees.
 
@@
 
-\subsection*{Amino acid sites with high LD between PB1, PB2 and HA}
-
-We perform hierarchical clustering to identify clusters of sites which exhibit similar LD patterns across pairs of segments.
-
 \begin{figure}[h]
@@
 (comment - will require a lot of fiddling to get the heatmaps displayed in a clear way)
+(comment - move heatmaps to supplementary information, if anyone's interested in specific sites they can look it up)
 \begin{figure}[h]
@@
 \subsection*{Rehash streamlined version of results for people who don't read results sections.}
-In this paper we have shown that PB1, PB2 and HA segments of influenza B viruses are the only ones that have maintained both a Vic and a Yam lineage (Figure \ref{LineageRatiosOverTime}).
-Evidence suggests that this is a result of lack of reassortment between Vic and Yam lineages of these segments (Figure \ref{betweenDiversity}) which have evolved co-assorting amino acid sites detectable as high linkage disequilibrium (\ref{segmentLD}).
+In this paper we have shown that PB1, PB2 and HA segments of influenza B viruses are the only ones that have maintained both a Vic and a Yam lineage (Figure \ref{lineageRatiosOverTime}).
+Evidence suggests that this is a result of prolonged lack of reassortment between Vic and Yam lineages of these segments (Figure \ref{betweenDiversity}) which have evolved co-assorting amino acid sites detectable as high linkage disequilibrium (Figure \ref{segmentLD}).
 We propose that this pattern of non-reassortment is due to the action of selection and not simply biased reassortment.
-Figures \ref{SPRdistance} and \ref{deltaTMRCA} show that despite all pairs of segments having the same reassortment frequency, PB1, PB2 and HA segments exhibit much lower reassortment `distances'.
+Figures \ref{SPRdistances} and \ref{deltaTMRCA} show that despite all pairs of segments having the same reassortment frequency, PB1, PB2 and HA segments exhibit much lower reassortment `distances'.
+Reassortment `distances' represent the temporal separation of reassorting lineages and suggests that reassortments between PB1, PB2 and HA segments tend to combine lineages that have relatively recently been part of the same genome (i.e.
+that have shared a recent common virion).
 In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were much more similar at the nucleotide and amino acid levels.
 The most recent lineage of influenza B viruses with mixed-lineage PB1-PB2-HA complexes (B/Waikato/6/2005-like) only circulated for 5 months before going extinct.
-We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands.
-If both PB1-PB2-HA complex lineages continue circulating in humans we expect more segments to be recruited to the two complexes, resulting in eventual sympatric speciation of influenza B viruses. 
+We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands i.e. small numbers of genes that resist being homogenized through gene flow.
+Thus, if both PB1-PB2-HA complex lineages continue circulating in humans we expect more segments to be recruited to the two complexes, resulting in eventual sympatric speciation of influenza B viruses.
 \tbc{Title is approximate.}
 
-\subsection*{Limitations of current study}
-The methods used in this paper have limited power to detect reassortments, since lineages have been fixed over time and thus all viruses eventually come to bear either Vic or Yam lineage PA, NP, NA, MP and NS segments with the exception of PB1, PB2 and HA segments.
-We can also only detect reassortment events where a lineage has switched from Vic to Yam or \textit{vice versa}, but not when a lineage has switched from a Yam lineage to another Yam lineage, which has occurred in the past (e.g.
-the reassortment events giving rise to B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like viruses in Figure \ref{railroadPlot}, which involved a Vic lineage PB1-PB2-HA complex reassorting with NP of Yamagata lineage is only visible in the NP tree, since all viruses in our dataset from 1995 onwards, including Vic PB1-PB2-HA complex-bearing ones, already had a Yamagata lineage-derived NP).
-However, this study focuses on PB1, PB2 and HA segments, all of which have survived as two lineages up to this day and thus are more informative for inferring reassortments.
-
+(move the following section to supplementary information? I imagine reading through a list of sites is a boring thing to do) 
 \subsection*{PB1-PB2-HA co-adaptation has evolved over time}
 Our findings suggest that PB1, PB2 and HA segments of Vic and Yam lineages represent co-adapted co-reassorting segment complexes.
-Previous studies have suggested similarities between the evolution of PB1, PB2 and HA segments of influenza B viruses \cite{hiromoto2000,lindstrom1999}, based on topology similarities between phylogenetic trees of the three segments, albeit the trees used contained limited numbers of sequences sampled somewhat close in time to the split of Victoria and Yamagata lineages.
+Previous studies have suggested similarities between the evolution of PB1, PB2 and HA segments of influenza B viruses \cite{hiromoto2000,lindstrom1999}, based on topology similarities between phylogenetic trees of the three segments, albeit the trees used in those studies contained limited numbers of sequences sampled somewhat close in time to the split of Victoria and Yamagata lineages.
 Despite the apparent similarities between PB1, PB2 and HA trees, we have detected at least 4 genome constellations having mixed-lineage PB1-PB2-HA, 3 of which were sampled early in the first decade of the split between Vic and Yam lineages, when, presumably, Vic and Yam lineages of PB1, PB2 and HA segments were still similar at the amino acid level to produce viable, though presumably unfit, influenza B viruses.
@@
 
-Our results also suggested that all segments have spent very little evolutionary time with the 3 mixed-lineage PB1-PB2-HA constellations, compared to either entirely Yam PB1-PB2-HA or entirely Vic PB1-PB2-HA bearing viruses.
+Linkage disequilibrium analyses suggest that co-adaptation between PB1, PB2 and HA proteins is an ongoing process.
+In addition to putative `speciation' amino acid sites that exhibit nearly complete linkage disequilibrium there exist a few co-associating amino acid polymorphisms ($\chi^{2}_{df}$ $\geq$ 0.5).
+These represent new non-synonymous mutations that have recently arisen within the Vic or the Yam PB1-PB2-HA background which might contribute to reassortment isolation between PB1, PB2 and HA segments.
+
+Our results also suggested that all segments have spent very little evolutionary time with the 3 mixed-lineage PB1-PB2-HA constellations we detect in our dataset, compared to either entirely Yam PB1-PB2-HA or entirely Vic PB1-PB2-HA bearing viruses.
 We interpret lack of reassortment within the PB1-PB2-HA complex across the Vic--Yam lineage boundary in recent times as evidence of a considerable degree of co-dependence within the Vic and Yam PB1-PB2-HA lineages, whereby recent reassortants bearing mixed-lineage PB1-PB2-HA complexes are inviable or unsampled by surveillance due to being outcompeted by viruses with pure-lineage PB1-PB2-HA complexes.
@@
 Again, we cannot determine whether this finding is significant or not.
-
-\subsection*{All segments reassort equally but some are more equal than others}
-The 95\% HPD intervals of normalized approximate SPR distances between pairs of segments (Figure \ref{SPRdistances}) overlap, suggesting there is no evidence of differences in the numbers of reassortments (or reassortment rate given as number of SPR moves per tree, data not shown) between all segments.
-There is reason to suspect that our SPR distance analysis simply lacks power.
-We used an approximation of the true SPR distances and SPR distances themselves could only approximate (and underestimate) the actual numbers of reassortments.
-However, exact SPR distances could be estimated for a limited number of segment pairs (PB1--PB2 and PB1--HA) and analyses indicate that after normalization exact and approximate SPR distances are not significantly different (supplementary figure of exact-approximate SPR distance differences and/or correlations of the two?).
-
-Significant differences between pairs of segments can be found when considering `reassortment distance'.
-Our estimates suggest that tips in PB1, PB2 and HA trees have similar TMRCAs (Figure \ref{deltaTMRCA}), suggesting that reassortments between the three rarely bring together lineages that have been evolving separately for prolonged periods of time.
-We cannot determine whether this effect is specifically due to non-reassortment of Vic and Yam PB1-PB2-HA lineages or a generic property of these three segments.
-In addition, we see evidence of this in MP and NA trees (comment - get references for MP-NA association in fluA, otherwise put down as personal communication or commonly held suspicion).
 
@@
 Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
-A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct in 2002), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
+A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct in 2002 (Figures \ref{tmrcaOT} and \ref{railroadPlot})), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
 Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments', which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
@@
 \subsection*{The future of influenza B viruses}
-We see two potential paths of evolution for influenza B viruses.
+We see three potential paths of evolution for influenza B viruses.
 More segments could be recruited into co-adapted segment complexes (PB1, PB2 and HA segments having started this first, akin to genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
 
-This may seem to be the case in Figure \ref{tmrcaOT} where the NP segment, exhibiting the fourth oldest TMRCA by 2011, looks like it might begin co-evolving with the PB1-PB2-HA complexes. 
+Sympatric speciation in other systems usually requires strong barriers to introgression, e.g.
+infertility of F1 hybrids can lead to the evolution of prezygotic reproductive isolation otherwise known as reinforcement or Wallace effect.
+Many questions remain unaddressed in the case of influenza B viruses, however.
+
+If influenza B viruses are undergoing sympatric speciation, does co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (e.g.
+unique packaging signals) or is co-infection so rare that speciation occurs via mutation-driven co-evolution?
+Given that PB1 and PB2 segments are functionally linked, is it reasonable to expect that the next segment to be recruited to the PB1-PB2-HA segment complexes will be functionally linked to either PB1/PB2 or HA?
+In Figure \ref{tmrcaOT} where the NP segment exhibits the fourth oldest TMRCA by 2011 looks like it might begin co-evolving with the PB1-PB2-HA complexes. 
 However, the continuous measure of diversity between Vic and Yam lineages of PB1, PB2 and HA segments in Figure \ref{betweenDiversity} show that the NP segment is skewed towards very recent TMRCAs, suggesting that a large number of recent isolates are products of recent reassortment between the NP segment and PB1-PB2-HA complexes.
 In fact, this is due to viruses possessing B/Brisbane/33/2008-like genome constellations, which comprise a large part of recently isolated influenza B virus strains (Figure \ref{railroadPlot}).
-The other alternative is continued reassortment of all non-PB1-PB2-HA segments between Vic and Yam PB1-PB2-HA complexes, or even extinction of one or the other lineage of the PB1-PB2-HA complex.
 
-Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether such two-lineage dynamics have not occurred in the history of influenza B viruses.
-If the influenza B population is on its way to speciation we expect that over time inter-lineage reassortment frequency should decrease, as more and more segment lineages would be recruited into co-adapted segment complexes.
+Are the reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background entering a Victoria PB1-PB2-HA background via reassortment indicative of rare events followed by selective sweeps?
+If so, are the sweeps products of sufficiently strong selection which might prevent sympatric speciation of other segments?
 
-On the other hand, influenza B viruses might continue co-circulating as two PB1-PB2-HA lineages exchanging non-PB1-PB2-HA segments, with reassortment frequency being roughly constant.
-Seeing how rarely influenza B viruses reassort across the Vic-Yam lineage boundary, the timescale for finding out which of the aforementioned scenarios, if any, are under way, is impractical at best.
+It's not unfeasible for selective sweeps following reassortments to breakdown developing co-adaptation of segments, especially if they are not functionally linked and have themselves been reassorted into a new PB1-PB2-HA background recently.
+This is the second scenario we might expect to occur, whereby relatively frequent reassortments occur between Vic and Yam lineages and are followed by selective sweeps.
+In this case influenza B viruses would undergo periodic genome homogenization with the exception of PB1, PB2 and HA segments.
+However, this model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.
+
+Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether dynamics similar to Victoria and Yamagata lineages have not occurred in influenza B virus genomes before and left no trace through extinction.
+We find it unlikely that either Victoria or Yamagata lineage PB1-PB2-HA complexes will go extinct stochastically in the near future, as they have co-circulated for prolonged periods at a ratio close to 0.5, suggesting the action of balancing selection (Figure \ref{lineageRatiosOverTime}).
+Depletion of susceptible individuals through influenza pandemics or vaccination seem unlikely as well.
+Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines do not produce lifelong immunity.
 
@@
 
-Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1+2-HA reassortants.
+Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
 
@@
 It is our hope that understanding the mechanism(s) underlying the links between Vic and Yam lineage PB1-PB2-HA complexes might suggest a more varied approach to both influenza B surveillance and control.
-For one, our findings suggest that viruses bearing mixed-lineage PB1-PB2-HA complexes, if unveiled by surveillance, are unlikely to contribute to future influenza B evolution.
+For one, our findings suggest that viruses bearing mixed-lineage PB1-PB2-HA complexes, if unveiled by surveillance, are unlikely to contribute to future influenza B evolution (as seen with B/Waikato/6/2005-like viruses).
 Similarly, in the absence of sequence data, it should be possible to predict the lineage of PB1, PB2 and HA segments of influenza B viruses by sequencing only one of those segments.

### 2014-01-22 fluB.tex
Deleting bits that will go into supp materials.
Revising text, minor changes to some figures.
Will be adding supplementary information shortly.

@@
 
-We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as distinct Victoria (Vic) and Yamagata (Yam) lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and once isolated belong to transient lineages.
+We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as distinct Victoria (Vic) and Yamagata (Yam) lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
 In other segments (PA, NP, NA, MP and NS) a single lineage has been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
+This has occured through repeated reassortments and subsequent fixation of reassortant viruses within the influenza B population.
 
@@
 	\caption{\textbf{Schematic depiction of method.}
-We begin by assigning sequences falling on either side of a specified bifurcation within each segment tree to different lineages.
-We then transfer lineage labels from one tree to the same tips in another tree.
+A) We begin by assigning sequences falling on either side of a specified bifurcation within each segment tree to different lineages.
+B) We then transfer lineage labels from one tree to the same tips in another tree.
 Label switches thus indicate reassortment events that combine lineages falling on different sides of the bifurcation specified earlier.
-In the example above tip number 6 is determined to be such a reassortant.}
+C) In the example above tip number 6 is determined to be such a reassortant.}
 	\label{methodFig}
@@
 These parameters are estimated for each pairing of trees at each MCMC state and provides us with 95\% highest posterior density intervals for those parameters, thus giving us the ability to test specific hypotheses regarding similarities between the trees of different segments.
-Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and therefore insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
+Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
 
@@
 
-We also calculated mean LD for all pairs of segments to quantify associations between different segments.
+We also calculated mean LD for all pairs of segments to quantify whole-segment associations.
 
-\subsection*{Confirming findings with a larger dataset and lineage assignment to sequences}
-The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates were downloaded from GISAID.
+\subsection*{Confirming findings with a larger dataset through lineage assignment}
+The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates sampled 1984-2013 were downloaded from GISAID.
 Neighbour-joining trees of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/02/87 or B/Yamagata/16/88 sequences.
@@
 \section*{Results}
-
-\tbc{I think that for these first two sections, you just need figures \ref{tmrcaOT} and \ref{lineageRatiosOverTime}.
-Figure \ref{betweenDiversity} seems redundant, and the multiple `perspectives' is confusing to be the first thing shown.}
 
@@
 
-We find that Vic lineage PA, NP, NA and MP segments have gone extinct (i.e.
-no sequences from those lineages have been sampled in recent years see Figure \ref{lineageRatiosOverTime}), replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
-Similarly, Vic lineage NS has become fixed in the viral population as well.
+We find that Vic lineage PA, NP, NA and MP segments have gone extinct (\textit{i.e.} no Vic lineage sequences of those segments have been sampled in recent years see Figure \ref{lineageRatiosOverTime}), replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
+Similarly, Vic lineage NS has become fixed in the influenza B virus population as well.
 By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.
 
-\subsection*{PB1, PB2 and HA segments continue to circulate as two lineages and have not reassorted recently}
-Unlike PA, NP, NA, MP and NS segments, which have periodically lost diversity, we find that PB1, PB2 and HA segments have continued to accumulate diversity since the initial split of Vic and Yam lineages (see Figure \ref{tmrcaOT}).
-In addition, there is evidence to suggest that the numbers of sequences that belong to either Vic and Yam lineages of PB1, PB2 and HA segments have been sampled at a ratio close to 0.5 (see Figure \ref{lineageRatiosOverTime}), presumably due to the action of some form of balancing selection.
+\subsection*{PB1, PB2 and HA segments continue to circulate as two lineages}
+We find that PB1, PB2 and HA segments of influenza B have not undergone lineage turnover, unlike PA, NP, NA, MP and NS segments.
+This is shown in Figure \ref{tmrcaOT}, where PB1, PB2 and HA lineages are shown to be derived from a consistently old TMRCA, indicating continued accumulation of diversity since the split of Vic and Yam lineages.
+We suggest that the maintenance of both Vic and Yam lineages in PB1, PB2 and HA segments might be caused by the action of some form of balancing selection.
+In Figure \ref{lineageRatiosOverTime} the ratios at which Vic and Yam lineages of PB1, PB2 and HA segments have been isolated appear to be close to 0.5 over long periods of time.
+On a year-to-year basis the ratios can fluctuate dramatically and suggesting frequency-dependent domination of one or the other lineage within a given year.
 
-By measuring the mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in diversity, which indicate that an inter-lineage reassortment event has taken place and thus is a qualitative measure of panmixis between pairs of segments.
+\subsection*{PB1, PB2 and HA segments do not reassort across the Victoria-Yamagata boundary}
+By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in diversity, which indicate that an inter-lineage reassortment event has taken place.
+Because we are taking the mean of pairwise comparisons, this method gives a qualitative measure of panmixis between Victoria and Yamagata lineages in two trees.
 
-In Figure \ref{betweenDiversity} we focus on PB1, PB2 and HA lineage labels, which have both Vic and Yam lineages up to recent times.
-We find that there have been some losses of diversity, indicative of label mixing i.e. reassortment, in the early years of the Vic--Yam split (1986--1996) between HA and PB1--PB2 segments.
-After 1997, however, mean pairwise between label diversity increases in PB1, PB2 and HA segments, whereby Vic and Yam labels share a common ancestor only at the root.
-Our data thus suggest that since 1997 there has been no reassortments between Vic and Yam lineages of PB1, PB2 and HA segments.
-In addition mean pairwise between label diversity measurements after 1997 become identical regardless of chosen label, which suggests that PB1, PB2 and HA lineage labels are assigned identically in all trees investigated.
+We focus only on PB1, PB2 and HA lineage labels (traits), since all other segments have fixed either the Vic or the Yam lineage.
+Figure \ref{betweenDiversity} shows that all trees have reassorted with respect to Victoria and Yamagata lineages of PB1, PB2 and HA.
+However, we also see that PB1, PB2 and HA traits show reciprocal preservation of diversity after 1997.
+This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments (i.e.
+PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees).
+In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1 and PB2 segments on one hand and HA on the other.
+These reductions in diversity represent small clades with PB1 and PB2 segments from one lineage and the HA segment from another, which go extinct by 1997.
+
+In addition, the assignment of Vic or Yam lineages of PB1, PB2 and HA segments to branches of other segment trees is very similar and often identical.
+This results in PB1, PB2 and HA lineage labels switching between Vic and Yam simultaneously in all trees after 1997, suggesting co-reassortment of Vic and Yam lineages of PB1, PB2 and HA segments.
 
@@
 \subsection*{Recent PB1-PB2-HA constellations are entirely derived from Victoria or Yamagata lineage segments}
-
-\tbc{Did you decide to shelve the REPU statistics?
-At least in theory, this should be much more convincing that PB1-PB2-HA are held together by selection rather than `mutation' (i.e.
-due to packaging signals).
-REPU should be a stronger argument than figure \ref{betweenDiversity}.
-....
-I see.
-You can't do something like REPU if you don't observe any transient reassortants in the first place.
-This suggests either that selection is very strong or mutation is very rare.
-But later on you mention that `there have been at least 3 sampled mixed-lineage PB1-PB2-HA complex constellations'.
-More description of these constellations would be helpful, when they appear and how long they last.
-Also report how long the average lineage persists.}
-
 \tbc{Can you reject the hypothesis that PB1 and PB2 stuck to HA stochastically?}
 
-We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary recently, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
+We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary since 1997, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
 Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
-Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations.
+Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations.
 
-We have identified 3 classes of mixed-lineage PB1-PB2-HA reassortants from our data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5th Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8th Jan 1993 -- 1994).
-An additional class of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was found when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
+We have identified 3 classes of mixed-lineage PB1-PB2-HA reassortants from our data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
+An additional class of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
 In all cases PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
-In particular reassortment events combining PB1 and PB2 segments of different lineages, e.g.
-B/Nanchang/630/1994-like and B/New York/24/1993-like isolates, exhibit poor sampling and short circulation times.
-In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past.
+In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each having two sequences), exhibit poor sampling and short circulation times (up to two years in the case of B/Nanchang/630/1994-like viruses).
+In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past (up to 4 years for B/Bankok/163/1990-like, although B/Waikato/6/2005-like only circulated for 5 months).
 
@@
 
-\subsection*{All segments reassort equally but some are more equal than others}
-The 95\% HPD intervals of normalized approximate SPR distances between pairs of segments (Figure \ref{SPRdistances}) overlap, suggesting there is no evidence of differences in the numbers of reassortments (or reassortment rate given as number of SPR moves per tree, data not shown) between all segments.
-There is reason to suspect that our SPR distance analysis simply lacks power.
-We used an approximation of the true SPR distances and SPR distances themselves could only approximate (and underestimate) the actual numbers of reassortments.
-However, exact SPR distances could be estimated for a limited number of segment pairs (PB1-PB2 and PB1-HA) and analyses indicate that after normalization exact and approximate SPR distances are not significantly different (supplementary figure of exact-approximate SPR distance differences and/or correlations of the two?).
+\subsection*{PB1, PB2 and HA trees experience reassortment `distance' effects}
+Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of trees after normalization (see Methods).
+If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to see small-scale similarities between phylogenetic trees of those segments.
+In our case we expect SPR distances (which are proportional to the numbers of reassortment that have taken place between trees) to reflect the raw reassortment rate.
+In Figure \ref{SPR distances} the 95\% HPD intervals of normalized approximate SPR distances between pairs of segments overlap, suggesting there is no evidence of differences in the numbers of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, data not shown) between all segments.
+We note that our SPR distance analysis might simply lack power.
+SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
+In addition, we use approximate SPR distances, rather than exact SPR distances.
+However, we do estimate exact SPR distances for a limited number of segment pairs (PB1-PB2 and PB1-HA) and find that after normalization exact and approximate SPR distances are not significantly different.
 
-Significant differences between pairs of segments can be found when considering `reassortment distance'.
-Our estimates suggest that tips in PB1, PB2 and HA trees have similar TMRCAs (Figure \ref{deltaTMRCA}), suggesting that reassortments between the three rarely bring together lineages that have been evolving separately for prolonged periods of time.
-We cannot determine whether this effect is specifically due to non-reassortment of Vic and Yam PB1-PB2-HA lineages or a generic property of these three segments.
-In addition, we see evidence of this in MP and NA trees (comment - get references for MP-NA association in fluA, otherwise put down as personal communication or commonly held suspicion).
+Although we do not find evidence of differences in numbers of raw reassortments between all pairs of trees, we find support for a reassortment `distance' effect.
+Figure \ref{deltaTMRCA} shows normalized mean $\Delta$TMRCA values for all pairs of trees.
+Most segment pairs show very low values for this statistic, indicating differences in TMRCAs.
+PB1, PB2 and HA trees, on the other hand, exhibit TMRCAs between themselves which are, on average, much more similar to the replicates of those trees.
+This suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves with lineages with which they have not recently shared the same virion.
+In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees.
+Previously, associations between MP and NA segments have been suspected in influenza A viruses (Samantha Lycett, personal communication).
 
@@
 We observe much higher mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
-In some cases this value exceed the estimated LD of amino acid sites within the same protein (e.g. PB1 and HA).
+In some cases this value exceed the estimated LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA).
 Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD.
-These sites define the split between Vic and Yam lineages within those segments.
+These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
 In addition, there are sites on PB1, PB2 and HA proteins which also show high (albeit smaller) LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
@@
 
-(comment - will require a lot of fiddling to get the heatmaps displayed in a clear way)
-(comment - move heatmaps to supplementary information, if anyone's interested in specific sites they can look it up)
-\begin{figure}[h]
-	\centering	
-	\includegraphics[width=0.65\textwidth]	{figures/Chi_PB1_PB2.png}
-	\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and PB2 segments based on $\chi^{2}_{df}$ values.}}
-	\label{ChiPB1PB2}
-\end{figure}
-
-\begin{figure}[h]
-	\centering	
-	\includegraphics[width=0.65\textwidth]	{figures/Chi_PB1_HA.png}
-	\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and HA segments based on $\chi^{2}_{df}$ values.}}
-	\label{ChiPB1HA}
-\end{figure}
-
-\begin{figure}[h]
-	\centering	
-	\includegraphics[width=0.65\textwidth]	{figures/Chi_PB2_HA.png}
-	\caption{\textbf{Hierarchical clustering of amino acid sites on PB2 and HA segments based on $\chi^{2}_{df}$ values.}}
-	\label{ChiPB2HA}
-\end{figure}
-
 \section*{Discussion}
@@
 Figures \ref{SPRdistances} and \ref{deltaTMRCA} show that despite all pairs of segments having the same reassortment frequency, PB1, PB2 and HA segments exhibit much lower reassortment `distances'.
-Reassortment `distances' represent the temporal separation of reassorting lineages and suggests that reassortments between PB1, PB2 and HA segments tend to combine lineages that have relatively recently been part of the same genome (i.e.
-that have shared a recent common virion).
+Reassortment `distances' represent the temporal separation of reassorting lineages and suggests that reassortments between PB1, PB2 and HA segments tend to combine lineages that have relatively recently been part of the same genome (\textit{i.e.} that have shared a recent common virion).
 In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were much more similar at the nucleotide and amino acid levels.
 The most recent lineage of influenza B viruses with mixed-lineage PB1-PB2-HA complexes (B/Waikato/6/2005-like) only circulated for 5 months before going extinct.
-We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands i.e. small numbers of genes that resist being homogenized through gene flow.
+We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands \textit{i.e.} small numbers of genes that resist being homogenized through gene flow.
 Thus, if both PB1-PB2-HA complex lineages continue circulating in humans we expect more segments to be recruited to the two complexes, resulting in eventual sympatric speciation of influenza B viruses.
@@
 
-(move the following section to supplementary information? I imagine reading through a list of sites is a boring thing to do) 
-\subsection*{PB1-PB2-HA co-adaptation has evolved over time}
-Our findings suggest that PB1, PB2 and HA segments of Vic and Yam lineages represent co-adapted co-reassorting segment complexes.
-Previous studies have suggested similarities between the evolution of PB1, PB2 and HA segments of influenza B viruses \cite{hiromoto2000,lindstrom1999}, based on topology similarities between phylogenetic trees of the three segments, albeit the trees used in those studies contained limited numbers of sequences sampled somewhat close in time to the split of Victoria and Yamagata lineages.
-Despite the apparent similarities between PB1, PB2 and HA trees, we have detected at least 4 genome constellations having mixed-lineage PB1-PB2-HA, 3 of which were sampled early in the first decade of the split between Vic and Yam lineages, when, presumably, Vic and Yam lineages of PB1, PB2 and HA segments were still similar at the amino acid level to produce viable, though presumably unfit, influenza B viruses.
-The fourth genome constellation having mixed-lineage PB1-PB2-HA complex was isolated in Australia and New Zealand in 2005 over a 5 month period and have not been detected since.
+\subsection*{Potential causes of co-dependence between PB1-PB2-HA}
+Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, \textit{e.g.} PB1-PB2-PA and HA-NA \cite{mccullers2004}.
+Though it would be easy to explain co-adaptation between the aforementioned segments by referring to their functional roles (\textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities), our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.
 
-Linkage disequilibrium analyses suggest that co-adaptation between PB1, PB2 and HA proteins is an ongoing process.
-In addition to putative `speciation' amino acid sites that exhibit nearly complete linkage disequilibrium there exist a few co-associating amino acid polymorphisms ($\chi^{2}_{df}$ $\geq$ 0.5).
-These represent new non-synonymous mutations that have recently arisen within the Vic or the Yam PB1-PB2-HA background which might contribute to reassortment isolation between PB1, PB2 and HA segments.
-
-Our results also suggested that all segments have spent very little evolutionary time with the 3 mixed-lineage PB1-PB2-HA constellations we detect in our dataset, compared to either entirely Yam PB1-PB2-HA or entirely Vic PB1-PB2-HA bearing viruses.
-We interpret lack of reassortment within the PB1-PB2-HA complex across the Vic--Yam lineage boundary in recent times as evidence of a considerable degree of co-dependence within the Vic and Yam PB1-PB2-HA lineages, whereby recent reassortants bearing mixed-lineage PB1-PB2-HA complexes are inviable or unsampled by surveillance due to being outcompeted by viruses with pure-lineage PB1-PB2-HA complexes.
-
-We have identified amino acid sites which exhibit a high degree of linkage disequilibrium between PB1, PB2 and HA segments.
-Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three proteins at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.
-
-None of the sites we have identified on PB1 and PB2 proteins fall within the previously described regions that form the contacts between the two polymerase subunits \cite{sugiyama2009}.
-
-Sites 211, 213 and 214 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
-Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211, 213 and 214 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
-Previous research \cite{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
-Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 positions 211 and 214 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at positions 211 and 214 and threonine (T) at position 213.
-It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
-Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \cite{fodor2004}.
-Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \cite{guilligay2008}.
-Again, we cannot determine whether this finding is significant or not.
-
-\subsection*{Potential causes of co-dependence between PB1-PB2-HA}
-Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, e.g.
-PB1-PB2-PA and HA-NA \cite{mccullers2004}.
-Though it would be easy to explain co-adaptation between the aforementioned segments by referring to their functional roles (e.g.
-PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities), our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.
-
-It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations (which did not become fixed in the population), our estimates of the number of reassortments do not differ significantly between all segments (Figure \ref{SPRdistances} and recent evidence suggests that, at least in influenza A, reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
+It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations (which did not become fixed in the population) and our estimates of the number of reassortments do not differ significantly between all segments.
+Figure \ref{SPRdistances} and recent evidence suggests that, at least in influenza A, reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
 In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we suggest several alternatives.
@@
 It is perhaps easiest to explain co-dependence between PB1 and PB2 segments based on their functions as part of the influenza RNA-dependent RNA polymerase (RdRp) heterotrimer.
-Indeed, trees of PB1 and PB2 segments exhibit high similarity in tip--tip TMRCAs (Figure \ref{deltaTMRCA}), suggesting highly similar trees.
+Indeed, trees of PB1 and PB2 segments exhibit high similarity in tip--tip TMRCAs (Figure \ref{deltaTMRCA}), suggesting highly similar TMRCAs.
 In addition, PB1--PB2 reassortants are the rarest and least persistent amongst mixed-lineage PB1-PB2-HA strains and have not been isolated since 1996.
@@
 Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \cite{bergeron2010,fulvini2011}.
-Early influenza A vaccines were derived by infecting chicken eggs with an egg-adapted strain and a seasonal human isolate in the presence of antisera raised against the HA and NA proteins of the egg-adapted strain, thus selecting for reassortants which were egg-adapted (\textit{i.e.} had internal segments derived from the egg-adapted strain) but possessed the HA and NA segments of the seasonal strain.
+Early influenza A vaccines used to be derived by infecting chicken eggs with an egg-adapted strain and a seasonal human isolate in the presence of antisera raised against the HA and NA proteins of the egg-adapted strain, thus selecting for reassortants which were egg-adapted (\textit{i.e.} had internal segments derived from the egg-adapted strain) but possessed the HA and NA segments of the seasonal strain.
 However, in addition to producing reassortants with HA-NA derived from the seasonal strain, as intended, the second most frequent class of reassortants produced using this methods were those with PB1, HA and NA derived from seasonal strains \cite{bergeron2010,fulvini2011}.
@@
 Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments', which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
+This is easiest to interpret as frequency-dependent balancing selection acting to preserve antigenic diversity.
+If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
 It is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
@@
 
-Though it is possible that PB1 and PB2 segments are co-adapted and the HA segment is simply hitchhiking along stochastically, we find it unlikely given that B/Waikato/6/2005-like strains, the most recent isolates with a mixed-lineage PB1-PB2-HA complex, circulated for a limited amount of time within a limited geographic area, suggesting that this particular lineage was outcompeted by influenza B strains with `pure' PB1-PB2-HA complexes.
+Finally it is possible that PB1 and PB2 segments are co-adapted whereas the HA segment is simply hitchhiking along stochastically. 
+We find it unlikely given that B/Waikato/6/2005-like strains, the most recent isolates with a mixed-lineage PB1-PB2-HA complex, circulated for a limited amount of time within a limited geographic area, suggesting that this particular lineage was outcompeted by influenza B strains with `pure' PB1-PB2-HA complexes.
 
@@
 We see three potential paths of evolution for influenza B viruses.
-More segments could be recruited into co-adapted segment complexes (PB1, PB2 and HA segments having started this first, akin to genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
+More segments could be recruited into the two currently circulating co-adapted segment complexes (PB1, PB2 and HA segments being the genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
 
-Sympatric speciation in other systems usually requires strong barriers to introgression, e.g.
-infertility of F1 hybrids can lead to the evolution of prezygotic reproductive isolation otherwise known as reinforcement or Wallace effect.
-Many questions remain unaddressed in the case of influenza B viruses, however.
+Sympatric speciation in other systems usually requires strong barriers to introgression, \textit{e.g.} infertility of F1 hybrids can lead to the evolution of prezygotic reproductive isolation otherwise known as reinforcement or Wallace effect.
+To what extent this would apply to influenza B viruses remains unknown.
 
-If influenza B viruses are undergoing sympatric speciation, does co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (e.g.
-unique packaging signals) or is co-infection so rare that speciation occurs via mutation-driven co-evolution?
+If influenza B viruses are undergoing sympatric speciation, does co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or is co-infection so rare that speciation occurs via mutation-driven co-evolution?
 Given that PB1 and PB2 segments are functionally linked, is it reasonable to expect that the next segment to be recruited to the PB1-PB2-HA segment complexes will be functionally linked to either PB1/PB2 or HA?
-In Figure \ref{tmrcaOT} where the NP segment exhibits the fourth oldest TMRCA by 2011 looks like it might begin co-evolving with the PB1-PB2-HA complexes. 
-However, the continuous measure of diversity between Vic and Yam lineages of PB1, PB2 and HA segments in Figure \ref{betweenDiversity} show that the NP segment is skewed towards very recent TMRCAs, suggesting that a large number of recent isolates are products of recent reassortment between the NP segment and PB1-PB2-HA complexes.
-In fact, this is due to viruses possessing B/Brisbane/33/2008-like genome constellations, which comprise a large part of recently isolated influenza B virus strains (Figure \ref{railroadPlot}).
 
-Are the reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background entering a Victoria PB1-PB2-HA background via reassortment indicative of rare events followed by selective sweeps?
-If so, are the sweeps products of sufficiently strong selection which might prevent sympatric speciation of other segments?
+Are the reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background entering a Victoria PB1-PB2-HA background via reassortment indicative of rare events followed by selective sweeps or stochastic fixation?
 

### 2014-01-22 fluB_supp.tex
Adding supplementary materials.
Moving hierarchical clustering figures and stuff from main text to
supplementary info, adding additional figures of tree-to-tree
comparisons.
Removing outdated figures.

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color} 
\usepackage{parskip}
\usepackage{float}

% text layout
\usepackage{geometry}
\geometry{textwidth=15cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace} 
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

% comments
\usepackage{ulem}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}} 
\def\tbc#1{\textcolor{purple}{[#1]}}

\begin{document}

\section*{Linkage disequilibrium data}

\begin{figure}[h]
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB1_PB2.png}
\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and PB2 segments based on $\chi^{2}_{df}$ values.}}
\label{ChiPB1PB2}
\end{figure}

\begin{figure}[h]
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB1_HA.png}
\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and HA segments based on $\chi^{2}_{df}$ values.}}
\label{ChiPB1HA}
\end{figure}

\begin{figure}[h]
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB2_HA.png}
\caption{\textbf{Hierarchical clustering of amino acid sites on PB2 and HA segments based on $\chi^{2}_{df}$ values.}}
\label{ChiPB2HA}
\end{figure}

We have identified amino acid sites which exhibit a high degree of linkage disequilibrium between PB1, PB2 and HA segments.
Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three proteins at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.

None of the sites we have identified on PB1 and PB2 proteins fall within the previously described regions that form the contacts between the two polymerase subunits \cite{sugiyama2009}.

Sites 211, 213 and 214 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211, 213 and 214 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
Previous research \cite{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 positions 211 and 214 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at positions 211 and 214 and threonine (T) at position 213.
It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \cite{fodor2004}.
Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \cite{guilligay2008}.
Again, we cannot determine whether this finding is significant or not.

\begin{figure}[h]
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_trees.png}
\caption{\textbf{}}
\label{}
\end{figure}

\begin{figure}[h]
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_replicates.png}
\caption{\textbf{}}
\label{}
\end{figure}

\begin{figure}[h]
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
\caption{\textbf{}}
\label{}
\end{figure}

\begin{figure}[h]
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
\caption{\textbf{}}
\label{}
\end{figure}

\begin{figure}[h]
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_normRErate.png}
\caption{\textbf{}}
\label{}
\end{figure}

\begin{figure}[h]
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_trees.png}
\caption{\textbf{}}
\label{}
\end{figure}

\begin{figure}[h]
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_replicates.png}
\caption{\textbf{}}
\label{}
\end{figure}

\bibliographystyle{pnas}
\bibliography{fluB}
\end{document}

### 2014-01-23 fluB_supp.tex
Adding/moving figures to supp figures.

@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_trees.png}
-\caption{\textbf{}}
+\caption{\textbf{Approximate SPR distances between all pairs of trees of segments.}}
 \label{}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_replicates.png}
-\caption{\textbf{}}
+\caption{\textbf{Approximate SPR distances between replicate trees of each segment.}}
 \label{}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
-\caption{\textbf{}}
+\caption{\textbf{Mean $\Delta$TMRCA between all pairs of trees of segments.}}
 \label{}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
-\caption{\textbf{}}
+\caption{\textbf{Mean $\Delta$TMRCA between replicate trees of each segment.}}
 \label{}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_normRErate.png}
-\caption{\textbf{}}
+\caption{\textbf{Normalized reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
 \label{}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_trees.png}
-\caption{\textbf{}}
+\caption{\textbf{Reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
 \label{}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_replicates.png}
-\caption{\textbf{}}
+\caption{\textbf{Reassortment rate (approximate number of SPR moves per total time in the tree) between replicate trees of each segment.}}
 \label{}

### 2014-01-30 fluB.tex
Adding text revisions suggested by @rambaut
Text revisions, moving MDS plots to supplementary info and using them
in supplementary info.

@@
 \author[2]{Trevor Bedford}
-\author[1]{Samantha Lycett}
-\author[1,3]{Andrew Rambaut}
+\author[1,3]{Samantha Lycett}
+\author[1,4]{Andrew Rambaut}
 
@@
 \affil[2]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
-\affil[3]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
+\affil[3]{Institute of Biodiversity Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK}
+\affil[4]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
 
@@
 Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality.
-Though lacking the degree of antigenic diversity that influenza A viruses possess, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
+Although the known antigenic diversity of influenza B viruses is less than that of influenza A viruses, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
 Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.
@@
 Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains.
-The only known persistent host of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), thus limiting the virus's potential to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir.
+Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), and are thought to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir.
 Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2013}.
 
-Currently circulating influenza B viruses are comprised of two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/02/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}.
+Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/02/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}.
 These two lineages possess antigenically distinct HA surface glycoproteins allowing them to co-circulate in the human population.
@@
 
-Previously, through the combined use of time-stamped and maximum likelihood phylogenetic trees evolutionary rate, selective pressures and reassortment history of major clades of influenza B have been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
-This analysis was limited because it used limited and different numbers of sequences to infer trees of each segment and relied on maximum likelihood phylogenies to infer the history of reassortments.
-Here, we introduce another approach that reveals a evolutionarily intriguing pattern of reassortment in influenza B.
+Previously, through phylogenetic analyses the evolutionary rate, selective pressures and reassortment history of major clades of influenza B have been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
+Here, we expand upon the methods used previously and introduce another approach that reveals a evolutionarily intriguing pattern of reassortment in influenza B.
 In this approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual viruses in the tree of the other segments.
-Reassortment events which result in the replacement of one segment lineage by another show up as changes in the discrete trait along a branch (Figure \ref{methodFig}).
-Here, we use discrete traits to reconstruct major reassortment events in the recent history of influenza B viruses and to partition diversity across segments of influenza B virus strains to quantify reassortment dynamics over time.
+By modelling these labels on a phylogenetic tree, reassortment events which result in the replacement of one segment lineage by another show up as label changes along a branch (Figure \ref{methodFig}).
+We use this method to reconstruct major reassortment events in the recent history of influenza B viruses and to partition diversity across segments of influenza B virus strains to quantify reassortment dynamics over time.
 
@@
 
-We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012. 
-Each strain was assigned to a lineage (either Vic or Yam) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
+We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information).
+The segments of each strain were assigned to a lineage (either Vic or Yam) by making maximum likelihood trees of each segment and identifying whether the isolate was more closely related to B/Victoria/02/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
 
@@
 Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
-Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate HKY models for each of the 3 codon partitions, and additionally included robust counting \cite{obrien2009} to report realized substitutions.
-We modeled the demographic process using a GMRF Bayesian skyride model \cite{minin2008}.
-We used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
+Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate HKY models for each of the 3 codon partitions, and additionally used a recently described method for reporting estimates of realized synonymous and non-synonymous substitutions \cite{obrien2009}.
+We used a flexible demographic model (Bayesian skyride \cite{minin2008}).
+We accounted for incomplete sampling dates for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
 We ran 3 independent MCMC chains, each with 200 million states, sampled every 20000 steps and discarded the first 10\% of the MCMC states as burn-in.
-Samples were combined across runs to give 27000 total samples for each segment and convergence was assessed by Tracer.
+After assessing convergence of all 3 MCMC chains by Tracer, we combined all analyses to give a total of 27000 samples from the posterior distribution of trees.
 
@@
 \subsection*{Measures of diversity}
-We inferred the diversity of each segment over time by finding the most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on the maximum amount of diversity existing at each time period.
-In addition, we calculated mean pairwise TMRCA between branches inferred to be in virions possessing Vic and Yam lineage PB1, PB2 and HA.
-This gave us an averaged measure of how much a particular segment reassorts with respect to PB1, PB2 and HA segments.
+We inferred the diversity of each segment over time by plotting the most recent common ancestor of all branches existing at yearly intervals, which places an upper bound on the maximum amount of diversity existing at each time period.
+In addition, we calculated mean pairwise TMRCA between branches labelled as Vic and Yam for PB1, PB2 and HA.
+This gave us a measure of how much a particular segment reassorts with respect to PB1, PB2 and HA segments.
 If Vic and Yam lineages of PB1, PB2 and HA segments were to be considered as being separate populations this measure would be equivalent to `between population' diversity.
 
-We also calculated the total amount of evolutionary time each segment has spent in virions with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments by taking the sum of branch lengths in each tree under different combinations of the PB1, PB2 and HA labels.
+We also calculated the total amount of evolutionary time spent by each segment with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments.
+We do this by summing the branch lengths in each tree under 3 different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
 This gives a measure of how successful, over long periods of time, each particular PB1-PB2-HA constellation has been.
@@
 \subsection*{Tree to tree similarities}
-In order to get a measure of tree to tree similarity we first extracted 900 trees from the posterior distributions of trees for each segment.
-These were combined across 3 independent MCMC chains to give a total of 2700 trees.
+We subsampled our combined posterior distribution of trees to give a total of 2700 trees.
 Pairwise comparison between each pair of segments was performed by comparing trees at each MCMC state.
@@
 
-If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as $\frac{[A:A']+[B:B']}{2\times [A:B]}$, where [A:B] is $\Delta$TMRCA of a pair of tips between tree A and tree B and [A:A'] is $\Delta$TMRCA between the same pair of tips between tree A and a replicate of tree A (A'), to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
+If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as $\frac{[A\_{i}:A\_{i}']+[B\_{i}:B\_{i}']}{2\times [A\_{i}:B\_{i}]}$, where [A\_{i}:B\_{i}] is $\Delta$TMRCA of a pair of tips between \textit{i}th tree derived from alignment A and \textit{i}th tree derived from alignment B and [A\_{i}:A\_{i}'] is $\Delta$TMRCA between the same pair of tips between \textit{i}th tree derived from alignment A and \textit{i}th tree derived from an independent run of alignment A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
 
-We also attempted to calculate subtree prune and regraft (SPR) distances between trees using rSPR \cite{whidden2009,whidden2010,whidden2013}.
-However, because the time taken to calculate the SPR distances depends on the SPR distances themselves, we find that exact SPR distance calculations over posterior sets of trees are impractical for all but the most similar trees.
-We thus calculated approximate SPR distances for all trees, using the normalization procedure described above.
+Subtree prune and regraft (SPR) distances between phylogenetic trees are an approximate measure of the numbers of reassortment or recombination events.
+Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
+We calculated approximate SPR distances \cite{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
+Approximate SPR distances were normalized using the procedure described above.
 
 \subsection*{Estimating linkage disequilibrium across the influenza B genome}
-We attempted to estimate linkage disequilibrium (LD) between amino acid sites across the influenza B proteome.
-We chose the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
-LD was estimated only at loci where each `allele' was represented by at least two isolates and gaps were ignored and not considered as polymorphisms.
-Ignoring gaps in estimating LD was a practical decision based on some sequences missing nucleotides at the beginning and end of sequences.
-However, there is at least one polymorphic indel within the HA protein, which is associated with the Vic-Yam split and not with sequence quality.
-In addition, some influenza B genome constellations have not reassorted for prolonged periods of time and have acquired amino acid substitutions on several segments, resulting in linkage disequilibrium.
-These single site pairs result in high LD and come from poorly sampled lineages, thus we choose to only focus on alleles that have been sampled at least twice.
+We attempted to estimate linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
+We use the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
+LD was estimated only at loci where each `allele' was represented by at least two isolates.
+We ignored gaps in the alignment and did not consider them as polymorphisms.
+However, there is at least one polymorphic indel within the HA protein, which is associated with the Vic-Yam split and not with incomplete sequences.
 
-We also calculated mean LD for all pairs of segments to quantify whole-segment associations.
+We also calculated mean LD for all pairs of segments to quantify the overall association between them.
 
 \subsection*{Confirming findings with a larger dataset through lineage assignment}
-The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates sampled 1984-2013 were downloaded from GISAID.
-Neighbour-joining trees of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/02/87 or B/Yamagata/16/88 sequences.
+The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates sampled 1984-2013 were compiled from GISAID.
+Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/02/87 or B/Yamagata/16/88 sequences.
 Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.
@@
 Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
-Some viruses with mixed-lineage genomes have become common ancestors to many later strains (i.e.
-the `trunk' of the phylogenetic tree) and the segment lineages involved so successful, as to become fixed within the influenza B population circulating today.
+Some viruses with mixed-lineage genomes have become fixed in the influenza B virus population \textit{i.e.} became the `trunk' of the phylogenetic tree of a segment.
 
 We find that Vic lineage PA, NP, NA and MP segments have gone extinct (\textit{i.e.} no Vic lineage sequences of those segments have been sampled in recent years see Figure \ref{lineageRatiosOverTime}), replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
-Similarly, Vic lineage NS has become fixed in the influenza B virus population as well.
+Conversely, Vic lineage NS has become fixed in the influenza B virus population as well.
 By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.
@@
 In Figure \ref{lineageRatiosOverTime} the ratios at which Vic and Yam lineages of PB1, PB2 and HA segments have been isolated appear to be close to 0.5 over long periods of time.
-On a year-to-year basis the ratios can fluctuate dramatically and suggesting frequency-dependent domination of one or the other lineage within a given year.
+On a year-to-year basis the ratios can fluctuate dramatically suggesting frequency-dependent domination of one or the other lineage within a given year.
 
@@
 
-We focus only on PB1, PB2 and HA lineage labels (traits), since all other segments have fixed either the Vic or the Yam lineage.
-Figure \ref{betweenDiversity} shows that all trees have reassorted with respect to Victoria and Yamagata lineages of PB1, PB2 and HA.
-However, we also see that PB1, PB2 and HA traits show reciprocal preservation of diversity after 1997.
-This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments (i.e.
-PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees).
+We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
+Increasing mean between label TMRCAs for all segments in Figure \ref{betweenDiversity} indicate that every segment has reassorted with respect to Victoria and Yamagata lineages of PB1, PB2 and HA segments.
+However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
+This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments (\textit{i.e.} PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees).
 In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1 and PB2 segments on one hand and HA on the other.
@@
 Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
-Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations.
+Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).
 
-We have identified 3 classes of mixed-lineage PB1-PB2-HA reassortants from our data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
-An additional class of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
-In all cases PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
+We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
+An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
+In all cases, PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
 In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each having two sequences), exhibit poor sampling and short circulation times (up to two years in the case of B/Nanchang/630/1994-like viruses).
@@
 
-We reconstructed reassortment events detectable by discrete traits.
+We reconstructed reassortment events that were detected by using lineage labels.
 Figure \ref{railroadPlot} focuses only on reassortments that have occured after 1990.
@@
 In our case we expect SPR distances (which are proportional to the numbers of reassortment that have taken place between trees) to reflect the raw reassortment rate.
-In Figure \ref{SPR distances} the 95\% HPD intervals of normalized approximate SPR distances between pairs of segments overlap, suggesting there is no evidence of differences in the numbers of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, data not shown) between all segments.
+In Figure \ref{SPRdistances} the 95\% HPD intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, data not shown) between all segments.
 We note that our SPR distance analysis might simply lack power.
@@
 PB1, PB2 and HA trees, on the other hand, exhibit TMRCAs between themselves which are, on average, much more similar to the replicates of those trees.
-This suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves with lineages with which they have not recently shared the same virion.
+This suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, \textit{i.e.} there is isolation by distance.
 In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees.
@@
 \subsection*{Linkage disequilibrium analysis confirms phylogenetic findings}
-Linkage disequilibrium analyses (Figure \ref{segmentLD}) confirm our findings from phylogenetic analyses that PB1, PB2 and HA segments are linked.
-We observe much higher mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
-In some cases this value exceed the estimated LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA).
-Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD.
+Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.
+In Figure \ref{segmentLD} their signature is greater mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
+In some cases this value exceeds the estimated LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA).
+Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD (see Supplementary information).
 These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
@@
 
-\tbc{The 8x8 MDS analysis suggests that PB1-PB2-HA are the only segments with substantial coevolution.}
-
 \begin{figure}[h]
 	\centering	
-	\includegraphics[width=0.75\textwidth]	{figures/InfB_segmentLD.png}
+	\includegraphics[width=0.75\textwidth]	{figures/InfB_aaMeanLD.png}
 	\caption{\textbf{LD comparison between influenza B proteins.}

### 2014-01-30 fluB_supp.tex
Adding text revisions suggested by @rambaut
Text revisions, moving MDS plots to supplementary info and using them
in supplementary info.

@@
 
+\begin{figure}[h]
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_locs_aa.png}
+\caption{\textbf{2-dimensional MDS embedding of segment based on mean amino acid linkage disequilibrium ($\chi\_{df}^{2}$).}}
+\label{}
+\end{figure}
+
+\begin{figure}[h]
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_corrs_aa.png}
+\caption{\textbf{Correlation between mean amino acid linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
+\label{}
+\end{figure}
+
+\begin{figure}[h]
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_locs_nt.png}
+\caption{\textbf{2-dimensional MDS embedding of segment based on mean nucleotide linkage disequilibrium ($\chi\_{df}^{2}$)}}
+\label{}
+\end{figure}
+
+\begin{figure}[h]
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_corrs_nt.png}
+\caption{\textbf{Correlation between mean nucleotide linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
+\label{}
+\end{figure}
+
 \bibliographystyle{pnas}

### 2014-01-30 fluB_supp.tex
Add title to supplement.

@@
 
+%%% TITLE %%%
+\title{\vspace{1.0cm} \huge \bf Supplemental information:\\ \LARGE Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex}
+
+\author[1]{Gytis Dudas}
+\author[2]{Trevor Bedford}
+\author[1,3]{Samantha Lycett}
+\author[1,4]{Andrew Rambaut}
+
+\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
+\affil[2]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
+\affil[3]{Institute of Biodiversity Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK}
+\affil[4]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
+
+\date{\today}
+
 \begin{document}
+\maketitle
 

### 2014-01-31 fluB.tex
Manuscript revisions.

@@
 Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
-Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages on the basis of haemagglutinin inhibition (HI) assays.
-Frequent reassortment between the segments of these two lineages has been noted in the past \cite{lindstrom1999}, but the effects of these reassortments have not been investigated in much detail.
+Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages. 
+Frequent reassortment between the segments of these two lineages has been noted in the past, but the effects of these reassortments have not been investigated in much detail.
+\tbc{Expand abstract to broadly outline results and give a 1 sentence conclusion.}
 
@@
 \section*{Introduction}
-Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality.
-Although the known antigenic diversity of influenza B viruses is less than that of influenza A viruses, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
+Seasonal influenza causes between 250,000 and 500,000 deaths annually and is comprised of three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \cite{flufactsheet}.
+However, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
 Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.
-
-Like all members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
-Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains.
-Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), and are thought to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir.
+Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
+Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of circulating human strains and with non-human influenza A strains.
+Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis), and are thought to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir.
 Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2013}.
 
-Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/02/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}.
-These two lineages possess antigenically distinct HA surface glycoproteins allowing them to co-circulate in the human population.
-All other influenza B segments also show a split into two lineages around that time.
+Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/2/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}.
+These two lineages now possess antigenically distinct HA surface glycoproteins \tbc{CITE} allowing them to co-circulate in the human population.
+Phylogenetic analysis of evolutionary rate, selective pressures and reassortment history of major clades of influenza B has shown extensive and often complicated patterns of reassortment between all segments of influenza B viruses between and within the Vic and Yam lineages \cite{chen2008}.
+Here, we extend previous methods to reveal an evolutionarily intriguing pattern of reassortment in influenza B.
+In our approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual viruses in the tree of the other segments.
+By modelling the transition between labels on a phylogenetic tree, reassortment events which result in the replacement of one segment's lineage by another show up as label changes along a branch (Figure \ref{methodFig}).
+We use this method to reconstruct major reassortment events and quantify reassortment dynamics over time in influenza B viruses.
 
-Previously, through phylogenetic analyses the evolutionary rate, selective pressures and reassortment history of major clades of influenza B have been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
-Here, we expand upon the methods used previously and introduce another approach that reveals a evolutionarily intriguing pattern of reassortment in influenza B.
-In this approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual viruses in the tree of the other segments.
-By modelling these labels on a phylogenetic tree, reassortment events which result in the replacement of one segment lineage by another show up as label changes along a branch (Figure \ref{methodFig}).
-We use this method to reconstruct major reassortment events in the recent history of influenza B viruses and to partition diversity across segments of influenza B virus strains to quantify reassortment dynamics over time.
-
-We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as distinct Victoria (Vic) and Yamagata (Yam) lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
-In other segments (PA, NP, NA, MP and NS) a single lineage has been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
+We show that despite extensive reassortment, three of the eight segments, PB1, PB2 and HA, still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
+In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
 This has occured through repeated reassortments and subsequent fixation of reassortant viruses within the influenza B population.
@@
 	\includegraphics[width=0.45\textwidth]{figures/TreeFigure2}
-	\caption{\textbf{Schematic depiction of method.}
-A) We begin by assigning sequences falling on either side of a specified bifurcation within each segment tree to different lineages.
+	\caption{\textbf{Schematic analysis of reassortment patterns.}
+	A) We begin by assigning sequences falling on either side of a specified bifurcation within each segment tree to different lineages, in this case, the Victoria and Yamagata bifurcation that occurred in the early 1980s.
 	B) We then transfer lineage labels from one tree to the same tips in another tree.
-Label switches thus indicate reassortment events that combine lineages falling on different sides of the bifurcation specified earlier.
-C) In the example above tip number 6 is determined to be such a reassortant.}
+	Label switches thus indicate reassortment events that combine lineages falling on different sides of the Vic/Yam bifurcation.
+	C) A reassortment graph depiction shows that tip number 6 is determined to be a reassortant.
+	\tbc{Relabel as "Segment 1" and "Segment 2" instead of "Tree 1" and "Tree 2".}
+	}
 	\label{methodFig}
@@
 We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information).
-The segments of each strain were assigned to a lineage (either Vic or Yam) by making maximum likelihood trees of each segment and identifying whether the isolate was more closely related to B/Victoria/02/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
+The segments of each strain were assigned to a lineage (either Vic or Yam) by making maximum likelihood trees of each segment \tbc{using what software?} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
 
-Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/02/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).
+Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).
 
@@
 Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
-Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate HKY models for each of the 3 codon partitions, and additionally used a recently described method for reporting estimates of realized synonymous and non-synonymous substitutions \cite{obrien2009}.
-We used a flexible demographic model (Bayesian skyride \cite{minin2008}).
+Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \cite{obrien2009}.
+We used a flexible Bayesian skyride demographic model \cite{minin2008}.
 We accounted for incomplete sampling dates for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
-We ran 3 independent MCMC chains, each with 200 million states, sampled every 20000 steps and discarded the first 10\% of the MCMC states as burn-in.
-After assessing convergence of all 3 MCMC chains by Tracer, we combined all analyses to give a total of 27000 samples from the posterior distribution of trees.
+We ran 3 independent MCMC chains, each with 200 million states, sampled every 20,000 steps and discarded the first 10\% of the MCMC states as burn-in.
+After assessing convergence of all 3 MCMC chains by visual inspection using Tracer \tbc{CITE}, we combined samples across chains to give a total of 27,000 samples from the posterior distribution of trees.
+\tbc{The MCC tree for each segment should be shown in supplemental information.  I would imagine one figure with eight panels.}
 
@@
 PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values).
-We inferred the ancestral state of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS \cite{lemey2009}).
-Because the posterior set of trees have branches labelled with the inferred lineage in all other segments, we can detect inter-lineage reassortments whenever a trait transition is observed (i.e.
-Yam to Vic or Vic to Yam, see Figure \ref{methodFig}) between a node and its descendant in any segment.
+We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete traits using an asymmetric transition matrix \tbc{CITE asymmetric model} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates \cite{lemey2009}).
+Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing trait transitions, i.e.
+Yam to Vic or Vic to Yam (see Figure \ref{methodFig}).
 In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
@@
 The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates sampled 1984-2013 were compiled from GISAID.
-Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/02/87 or B/Yamagata/16/88 sequences.
+Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences.
 Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.

### 2014-01-31 fluB.tex
Continued revisions.

@@
 \subsection*{Measures of diversity}
-We inferred the diversity of each segment over time by plotting the most recent common ancestor of all branches existing at yearly intervals, which places an upper bound on the maximum amount of diversity existing at each time period.
+We inferred the diversity of each segment at a single point in time by estimating the date of the most recent common ancestor of all branches at yearly intervals, which places an upper bound on the maximum amount of diversity existing at each time period.
 In addition, we calculated mean pairwise TMRCA between branches labelled as Vic and Yam for PB1, PB2 and HA.
@@
 We subsampled our combined posterior distribution of trees to give a total of 2700 trees.
-Pairwise comparison between each pair of segments was performed by comparing trees at each MCMC state.
+Pairwise comparison between each pair of segments was performed by comparing individual pairs of sampled trees.
 
@@
 
-If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as $\frac{[A\_{i}:A\_{i}']+[B\_{i}:B\_{i}']}{2\times [A\_{i}:B\_{i}]}$, where [A\_{i}:B\_{i}] is $\Delta$TMRCA of a pair of tips between \textit{i}th tree derived from alignment A and \textit{i}th tree derived from alignment B and [A\_{i}:A\_{i}'] is $\Delta$TMRCA between the same pair of tips between \textit{i}th tree derived from alignment A and \textit{i}th tree derived from an independent run of alignment A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
+If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as 
+\begin{equation}
+\frac{[A_{i}:A_{i}']+[B_{i}:B_{i}']}{2\times [A_{i}:B_{i}]},
+\frac{[A_{i}:A_{i}']+[B_{i}:B_{i}']}{2\times [A_{i}:B_{i}]},
+\end{equation}
+where $f(A_i,B_j)$ is the $\Delta$TMRCA of a pair of tips between the \textit{i}th tree derived from alignment A and the \textit{j}th tree derived from alignment B and $f(A_i,A_j)$ is the $\Delta$TMRCA between the same pair of tips between \textit{i}th tree derived from alignment A and \textit{j}th tree derived from an independent run of alignment A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
 

### 2014-01-31 fluB.tex
Adding text revisions suggested by @trvrb
Changes to text and method figure and adding references.
Tempted to call PB1-PB2-HA a genomic speciation raft rather than island,
seems catchy and more accurate as an analogy.

@@
 Frequent reassortment between the segments of these two lineages has been noted in the past, but the effects of these reassortments have not been investigated in much detail.
-\tbc{Expand abstract to broadly outline results and give a 1 sentence conclusion.}
+We investigate inter-lineage reassortments by modelling the lineages of all genomic segments on phylogenetic trees.
+Our analyses indicate that of the 8 segments of influenza B viruses only PB1, PB2 and HA segments still have a Victoria and a Yamagata lineage and that currently circulating strains possess PB1, PB2 and HA segments derived entirely from one or the other lineage.
+We argue that this is due to selection against reassortant viruses with mixed lineage PB1, PB2 and HA segments.
+Given sufficient time we expect influenza B viruses to undergo sympatric speciation, if more segments get recruited by the already reassortment-isolated PB1-PB2-HA speciation island.
 
@@
 Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/2/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}.
-These two lineages now possess antigenically distinct HA surface glycoproteins \tbc{CITE} allowing them to co-circulate in the human population.
+These two lineages now possess antigenically distinct HA surface glycoproteins \cite{kanegae1990,rota1990,nerome1998,nakagawa2002,ansaldi2003} allowing them to co-circulate in the human population.
 Phylogenetic analysis of evolutionary rate, selective pressures and reassortment history of major clades of influenza B has shown extensive and often complicated patterns of reassortment between all segments of influenza B viruses between and within the Vic and Yam lineages \cite{chen2008}.
@@
 	C) A reassortment graph depiction shows that tip number 6 is determined to be a reassortant.
-	\tbc{Relabel as "Segment 1" and "Segment 2" instead of "Tree 1" and "Tree 2".}
 	}
@@
 We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information).
-The segments of each strain were assigned to a lineage (either Vic or Yam) by making maximum likelihood trees of each segment \tbc{using what software?} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
+The segments of each strain were assigned to a lineage (either Vic or Yam) by making maximum likelihood trees of each segment using PhyML \cite{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
 
@@
 We ran 3 independent MCMC chains, each with 200 million states, sampled every 20,000 steps and discarded the first 10\% of the MCMC states as burn-in.
-After assessing convergence of all 3 MCMC chains by visual inspection using Tracer \tbc{CITE}, we combined samples across chains to give a total of 27,000 samples from the posterior distribution of trees.
+After assessing convergence of all 3 MCMC chains by visual inspection using Tracer \cite{tracer}, we combined samples across chains to give a total of 27,000 samples from the posterior distribution of trees.
 \tbc{The MCC tree for each segment should be shown in supplemental information.  I would imagine one figure with eight panels.}
 
-Every sequence was assigned 7 discrete traits corresponding to the lineages of all other segments with which a strain was isolated (e.g.
-PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values).
-We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete traits using an asymmetric transition matrix \tbc{CITE asymmetric model} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates \cite{lemey2009}).
-Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing trait transitions, i.e.
-Yam to Vic or Vic to Yam (see Figure \ref{methodFig}).
+Every sequence was assigned 7 discrete traits corresponding to the lineages of all other segments with which a strain was isolated (\textit{e.g.} PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values).
+We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete traits using an asymmetric transition matrix \cite{lemey2009} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates.
+Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing trait transitions, \textit{i.e.} Yam to Vic or Vic to Yam (see Figure \ref{methodFig}).
 In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
@@
 \subsection*{Recent PB1-PB2-HA constellations are entirely derived from Victoria or Yamagata lineage segments}
-\tbc{Can you reject the hypothesis that PB1 and PB2 stuck to HA stochastically?}
 
@@
 \subsection*{Railroad schematic plot of influenza B reassortment history}
-
-\tbc{Are the observed rates of extinction / co-option fit with the effective population of the flu B population?
-I.e.
-we generally get 10 years back to a common ancestor within a lineage / segment.
-Can we think of the extinction of segments as an aspect of there existing a loosely panmictic flu B population for everything but PB1, PB2 and HA?}
 
@@
 
-\subsection*{Rehash streamlined version of results for people who don't read results sections.}
+\subsection*{Evidence of incipient sympatric speciation}
 In this paper we have shown that PB1, PB2 and HA segments of influenza B viruses are the only ones that have maintained both a Vic and a Yam lineage (Figure \ref{lineageRatiosOverTime}).
@@
 We propose that this pattern of non-reassortment is due to the action of selection and not simply biased reassortment.
-Figures \ref{SPRdistances} and \ref{deltaTMRCA} show that despite all pairs of segments having the same reassortment frequency, PB1, PB2 and HA segments exhibit much lower reassortment `distances'.
+Figures \ref{SPRdistances} and \ref{deltaTMRCA} show that despite all pairs of segments having the same reassortment frequency, suggesting there is no bias in segment packaging, PB1, PB2 and HA segments exhibit much lower reassortment `distances'.
 Reassortment `distances' represent the temporal separation of reassorting lineages and suggests that reassortments between PB1, PB2 and HA segments tend to combine lineages that have relatively recently been part of the same genome (\textit{i.e.} that have shared a recent common virion).
@@
 Thus, if both PB1-PB2-HA complex lineages continue circulating in humans we expect more segments to be recruited to the two complexes, resulting in eventual sympatric speciation of influenza B viruses.
-\tbc{Title is approximate.}
 

### 2014-01-31 fluB_supp.tex
Manuscript revisions.
Using approximate SPR validation figures in supp.
Including figures showing that normalized approximate and exact SPR
distances are not different.

@@
 
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thepage}{S\arabic{page}}
+
 \section*{Linkage disequilibrium data}
 
-\begin{figure}[h]
+\begin{figure}
 \centering  
@@
 
-\begin{figure}[h]
+\begin{figure}
 \centering  
@@
 
-\begin{figure}[h]
+\begin{figure}
 \centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_locs_aa.png}
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist.png}
+\caption{\textbf{Difference}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist2.png}
+\caption{\textbf{Ratio}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist.png}
+\caption{\textbf{Difference}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist2.png}
+\caption{\textbf{Ratio}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist.png}
+\caption{\textbf{Difference}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
+\caption{\textbf{Ratio}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
+\caption{\textbf{Correlation}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
+\caption{\textbf{Correlation}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB2-HA_corr.png}
+\caption{\textbf{Correlation}}
+\label{}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_aaLD_MDS.png}
 \caption{\textbf{2-dimensional MDS embedding of segment based on mean amino acid linkage disequilibrium ($\chi\_{df}^{2}$).}}
@@
 
-\begin{figure}[h]
+\begin{figure}
 \centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_corrs_aa.png}
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_aaLD_MatrixMDScorr.png}
 \caption{\textbf{Correlation between mean amino acid linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
@@
 
-\begin{figure}[h]
+\begin{figure}
 \centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_locs_nt.png}
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_ntLD_MDS.png}
 \caption{\textbf{2-dimensional MDS embedding of segment based on mean nucleotide linkage disequilibrium ($\chi\_{df}^{2}$)}}
@@
 
-\begin{figure}[h]
+\begin{figure}
 \centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_corrs_nt.png}
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_ntLD_MatrixMDScorr.png}
 \caption{\textbf{Correlation between mean nucleotide linkage disequilibrium and MDS embedding distance for all pairs of segments.}}

### 2014-01-31 fluB.tex
More revisions.

@@
 Frequent reassortment between the segments of these two lineages has been noted in the past, but the effects of these reassortments have not been investigated in much detail.
-We investigate inter-lineage reassortments by modelling the lineages of all genomic segments on phylogenetic trees.
-Our analyses indicate that of the 8 segments of influenza B viruses only PB1, PB2 and HA segments still have a Victoria and a Yamagata lineage and that currently circulating strains possess PB1, PB2 and HA segments derived entirely from one or the other lineage.
-We argue that this is due to selection against reassortant viruses with mixed lineage PB1, PB2 and HA segments.
-Given sufficient time we expect influenza B viruses to undergo sympatric speciation, if more segments get recruited by the already reassortment-isolated PB1-PB2-HA speciation island.
+We investigate inter-lineage reassortments by comparing phylogenetic trees across genomic segments.
+Our analyses indicate that of the 8 segments of influenza B viruses only PB1, PB2 and HA segments maintained separate Victoria and Yamagata lineages and that currently circulating strains possess PB1, PB2 and HA segments derived entirely from one or the other lineage; other segments have repeatedly reassorted between lineages thereby reducing genetic diversity.
+We argue that this difference between segments is due to selection against reassortant viruses with mixed lineage PB1, PB2 and HA segments.
+Given sufficient time we expect influenza B viruses to undergo sympatric speciation, if more segments get recruited by the already reassortment-isolated PB1-PB2-HA gene complex.
 
@@
 
-We express the similarity of two trees by referring to the mean of the distribution of TMRCA differences ($\Delta$TMRCA) between pairs of tips in the two trees.
+We express the similarity of two trees by comparing to the mean of the distribution of TMRCA differences between pairs of tips in the two trees.
+In comparing TMRCAs of trees A and B we calculate
+\begin{equation}
+\Delta(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
+\end{equation}
+where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $g(A_{ij},B_{ij})$ is the difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree A and the \textit{i}th posterior sample of tree B.
+\tbc{Please correct if I misinterpreted what you were doing.}
+Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
+
 These parameters are estimated for each pairing of trees at each MCMC state and provides us with 95\% highest posterior density intervals for those parameters, thus giving us the ability to test specific hypotheses regarding similarities between the trees of different segments.
 Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
-
-If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as 
-\begin{equation}
-\frac{[A_{i}:A_{i}']+[B_{i}:B_{i}']}{2\times [A_{i}:B_{i}]},
-\frac{[A_{i}:A_{i}']+[B_{i}:B_{i}']}{2\times [A_{i}:B_{i}]},
-\end{equation}
-where $f(A_i,B_j)$ is the $\Delta$TMRCA of a pair of tips between the \textit{i}th tree derived from alignment A and the \textit{j}th tree derived from alignment B and $f(A_i,A_j)$ is the $\Delta$TMRCA between the same pair of tips between \textit{i}th tree derived from alignment A and \textit{j}th tree derived from an independent run of alignment A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
 
@@
 \subsection*{Estimating linkage disequilibrium across the influenza B genome}
-We attempted to estimate linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
-We use the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
-LD was estimated only at loci where each `allele' was represented by at least two isolates.
+We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
+We used the $\chi^{2}_{df}$ \tbc{I couldn't download the article PDF, but is the $df$ really necessary?} statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
+LD was estimated only at loci where each nucleotide allele was represented by at least two isolates.
 We ignored gaps in the alignment and did not consider them as polymorphisms.
-However, there is at least one polymorphic indel within the HA protein, which is associated with the Vic-Yam split and not with incomplete sequences.
-
 We also calculated mean LD for all pairs of segments to quantify the overall association between them.
@@
 \subsection*{Confirming findings with a larger dataset through lineage assignment}
-The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates sampled 1984-2013 were compiled from GISAID.
+The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates \tbc{`same' is confusing here, these are different isolates from those discussed previously?} sampled 1984-2013 were compiled from GISAID.
 Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences.
@@
 
-The split of Vic and Yam lineages can be seen in all segments \cite{chen2008}.
+The split into Vic and Yam lineages can be seen in all segments \cite{chen2008} \tbc{also reference supplemental tree figure}.
 Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
@@
 
-We find that Vic lineage PA, NP, NA and MP segments have gone extinct (\textit{i.e.} no Vic lineage sequences of those segments have been sampled in recent years see Figure \ref{lineageRatiosOverTime}), replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
+We find that Vic lineage PA, NP, NA and MP segments have gone extinct (\textit{i.e.} no Vic lineage sequences of those segments have been sampled in recent years see Figure \ref{lineageRatiosOverTime}) \tbc{figures should not be referenced out of order, either make this figure 2 or remove this reference, or push this section back}, replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
 Conversely, Vic lineage NS has become fixed in the influenza B virus population as well.
-By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.
+By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}) \tbc{again, restructure to fix out of order figure reference}, it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.
+
+\tbc{This reads more like a beginning Discussion paragraph. These first Results paragraphs should be closer to the data.  Just describe what you show in figures.}
 

### 2014-02-02 fluB.tex
Changing figure order, moving results around.
Also changing LD results from heatmap to pairwise comparisons (has the
side effect of making the LD results a bit more dramatic).
Hopefully resolves #5

@@
 Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.
+
 Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
@@
 Phylogenetic analysis of evolutionary rate, selective pressures and reassortment history of major clades of influenza B has shown extensive and often complicated patterns of reassortment between all segments of influenza B viruses between and within the Vic and Yam lineages \cite{chen2008}.
+
 Here, we extend previous methods to reveal an evolutionarily intriguing pattern of reassortment in influenza B.
@@
 We accounted for incomplete sampling dates for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
+
 We ran 3 independent MCMC chains, each with 200 million states, sampled every 20,000 steps and discarded the first 10\% of the MCMC states as burn-in.
@@
 
-We express the similarity of two trees by comparing to the mean of the distribution of TMRCA differences between pairs of tips in the two trees.
+We express the similarity of two trees by comparing the means of the distribution of TMRCA differences between pairs of tips in the two trees.
 In comparing TMRCAs of trees A and B we calculate
@@
 where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $g(A_{ij},B_{ij})$ is the difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree A and the \textit{i}th posterior sample of tree B.
-\tbc{Please correct if I misinterpreted what you were doing.}
 Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
@@
 We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
-We used the $\chi^{2}_{df}$ \tbc{I couldn't download the article PDF, but is the $df$ really necessary?} statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
+We used the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
 LD was estimated only at loci where each nucleotide allele was represented by at least two isolates.
@@
 \subsection*{Confirming findings with a larger dataset through lineage assignment}
-The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates \tbc{`same' is confusing here, these are different isolates from those discussed previously?} sampled 1984-2013 were compiled from GISAID.
-Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences.
+We downloaded all available sequences of influenza B isolates sampled 1984--2013 from GISAID for which PB1, PB2 and HA segments were sequenced.
+This comprised a dataset of 1433 isolates in total. 
+Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences, as before.
 Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.
@@
 
-We find that Vic lineage PA, NP, NA and MP segments have gone extinct (\textit{i.e.} no Vic lineage sequences of those segments have been sampled in recent years see Figure \ref{lineageRatiosOverTime}) \tbc{figures should not be referenced out of order, either make this figure 2 or remove this reference, or push this section back}, replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
-Conversely, Vic lineage NS has become fixed in the influenza B virus population as well.
-By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}) \tbc{again, restructure to fix out of order figure reference}, it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.
+By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.
 
-\tbc{This reads more like a beginning Discussion paragraph. These first Results paragraphs should be closer to the data.  Just describe what you show in figures.}
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.65\textwidth]{figures/InfB_tmrcaOT_lines.png}
+	\caption{\textbf{Oldest TMRCA of all surviving lineages over time.}
+PA, NP, NA, MP and NS segments of influenza B viruses show periodic losses of diversity, indicating lineage turn over.
+PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+	\label{tmrcaOT}
+\end{figure}
 
-\subsection*{PB1, PB2 and HA segments continue to circulate as two lineages}
-We find that PB1, PB2 and HA segments of influenza B have not undergone lineage turnover, unlike PA, NP, NA, MP and NS segments.
-This is shown in Figure \ref{tmrcaOT}, where PB1, PB2 and HA lineages are shown to be derived from a consistently old TMRCA, indicating continued accumulation of diversity since the split of Vic and Yam lineages.
-We suggest that the maintenance of both Vic and Yam lineages in PB1, PB2 and HA segments might be caused by the action of some form of balancing selection.
-In Figure \ref{lineageRatiosOverTime} the ratios at which Vic and Yam lineages of PB1, PB2 and HA segments have been isolated appear to be close to 0.5 over long periods of time.
-On a year-to-year basis the ratios can fluctuate dramatically suggesting frequency-dependent domination of one or the other lineage within a given year.
+Whereas PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, we do not find evidence of broad-scale lineage turnover for PB1, PB2 and HA segments.
+Figure \ref{tmrcaOT} shows that lineages of PB1, PB2 and HA segments are consistently derived from an old TMRCA, indicative of continued accumulation of diversity since the split of those segments into Vic and Yam lineages.
 
@@
 Increasing mean between label TMRCAs for all segments in Figure \ref{betweenDiversity} indicate that every segment has reassorted with respect to Victoria and Yamagata lineages of PB1, PB2 and HA segments.
+
+\begin{figure}
+	\centering		
+	\includegraphics[width=0.75\textwidth]{figures/InfB_betweenDiversity.png}
+	\caption{\textbf{Mean pairwise TMRCA between Vic and Yam branches under PB1, PB2 and HA label sets.}
+PB1, PB2 and HA segment labels indicate that these segments show reciprocal preservation of diversity, which dates back to the split of Vic and Yam lineages.
+All other segments show increasingly more recent TMRCAs between branches labelled as Vic and Yam in PB1, PB2 and HA label sets.
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+	\label{betweenDiversity}
+\end{figure}
+
+\begin{figure}[h]
+	\centering	
+	\includegraphics[width=0.65\textwidth]	{figures/InfB_LineageRatiosOverTime.png}
+	\caption{\textbf{Ratio of lineages in the dataset.}}
+	\label{lineageRatiosOverTime}
+\end{figure}
+
 However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
@@
 
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_stateTime.png}
-	\caption{\textbf{Amount of evolutionary time spent under different PB1-PB2-HA constellations.}
-All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.
-All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
-	\label{stateTime}
-\end{figure}
-
-\subsection*{Recent PB1-PB2-HA constellations are entirely derived from Victoria or Yamagata lineage segments}
-
-We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary since 1997, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
-Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
-Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).
-
-We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
-An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
-In all cases, PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
-In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each having two sequences), exhibit poor sampling and short circulation times (up to two years in the case of B/Nanchang/630/1994-like viruses).
-In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past (up to 4 years for B/Bankok/163/1990-like, although B/Waikato/6/2005-like only circulated for 5 months).
-
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_tmrcaOT_lines.png}
-	\caption{\textbf{Oldest TMRCA of all surviving lineages over time.}
-PA, NP, NA, MP and NS segments of influenza B viruses show periodic losses of diversity, indicating lineage turn over.
-PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
-All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
-	\label{tmrcaOT}
-\end{figure}
-
-(comment - need higher resolution for lineage ratios figure or switch to vector graphics)
-\begin{figure}[h]
-	\centering	
-	\includegraphics[width=0.75\textwidth]	{figures/InfB_LineageRatiosOverTime.png}
-	\caption{\textbf{Ratio of lineages in the dataset.}}
-	\label{lineageRatiosOverTime}
-\end{figure}
-
-\begin{figure}
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_betweenDiversity.png}
-	\caption{\textbf{Mean pairwise TMRCA between Vic and Yam branches under PB1, PB2 and HA label sets.}
-PB1, PB2 and HA segment labels indicate that these segments show reciprocal preservation of diversity, which dates back to the split of Vic and Yam lineages.
-All other segments show increasingly more recent TMRCAs between branches labelled as Vic and Yam in PB1, PB2 and HA label sets.
-All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
-	\label{betweenDiversity}
-\end{figure}
+By plotting the ratios of sequences assigned to either Vic or Yam lineage in each segment (Figure \ref{lineageRatiosOverTime}) it is evident that losses of diversity seen in Figures \ref{tmrcaOT} and \ref{betweenDiversity} are related to fixations of either the Vic or the Yam lineage in PA, NP, NA, MP and NS segments.
+These losses of diversity correspond to extinctions of Vic lineage in PA, NP, NA and MP segments and Yam lineage in NS segment.
+Similarly, the lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA segments is reflected in Figure \ref{lineageRatiosOverTime} where the two lineages have been isolated at a ratio close to 0.5 over long periods of time.
+On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically suggesting frequency-dependent domination of one or the other lineage within a given season.
 
@@
 We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV).
-Figure \ref{railroadPlot} shows that reassorting segments appear to evolve with a considerable degree of autonomy.
-For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to B/Iowa/03/2002-like isolates.
-A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has been derived from Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
 
-\begin{figure}
+\begin{figure}[h]
 	\centering		
@@
 
+Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
+For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to B/Iowa/03/2002-like isolates.
+A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has been derived from Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
+
+\subsection*{Recent PB1-PB2-HA constellations are entirely derived from Victoria or Yamagata lineage segments}
+
+We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary since 1997 (Figure \ref{betweenDiversity}), form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
+Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
+Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).
+
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.65\textwidth]{figures/InfB_stateTime.png}
+	\caption{\textbf{Amount of evolutionary time spent under different PB1-PB2-HA constellations.}
+All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.
+All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+	\label{stateTime}
+\end{figure}
+
+We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
+An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
+
+In all cases, PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
+In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each having two sequences), exhibit poor sampling and short circulation times (up to two years in the case of B/Nanchang/630/1994-like viruses).
+In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past (up to 4 years for B/Bankok/163/1990-like, although B/Waikato/6/2005-like only circulated for 5 months).
+
 \subsection*{PB1, PB2 and HA trees experience reassortment `distance' effects}
@@
 In our case we expect SPR distances (which are proportional to the numbers of reassortment that have taken place between trees) to reflect the raw reassortment rate.
+
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedApproxSPR.png}
+	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+	\label{SPRdistances}
+\end{figure}
+
 In Figure \ref{SPRdistances} the 95\% HPD intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, data not shown) between all segments.
@@
 
-\begin{figure}
+\begin{figure}[h]
 	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_normalizedApproxSPR.png}
-	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
-All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
-	\label{SPRdistances}
-\end{figure}
-
-\begin{figure}
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/InfB_normalizedMuDeltaTMRCA.png}
+	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedMuDeltaTMRCA.png}
 	\caption{\textbf{Normalized mean $\Delta$TMRCA statistics between pairs of segments.}
@@
 In some cases this value exceeds the estimated LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA).
+
+\begin{figure}[h]
+	\centering	
+	\includegraphics[width=0.65\textwidth]	{figures/InfB_aaMeanLD.png}
+	\caption{\textbf{LD comparison between influenza B proteins.}
+Pairwise comparisons of linkage disequilibrium between amino acid sites on influenza B proteins.}
+	\label{segmentLD}
+\end{figure}
+
 Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD (see Supplementary information).
@@
 In addition, there are sites on PB1, PB2 and HA proteins which also show high (albeit smaller) LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
-
-\begin{figure}[h]
-	\centering	
-	\includegraphics[width=0.75\textwidth]	{figures/InfB_aaMeanLD.png}
-	\caption{\textbf{LD comparison between influenza B proteins.}
-Comparison of linkage disequilibrium between amino acid sites on influenza B proteins.
-Numbers on the heatmap are mean $\chi^{2}_{df}$ between pairs of proteins.}
-	\label{segmentLD}
-\end{figure}
 
@@
 Figures \ref{SPRdistances} and \ref{deltaTMRCA} show that despite all pairs of segments having the same reassortment frequency, suggesting there is no bias in segment packaging, PB1, PB2 and HA segments exhibit much lower reassortment `distances'.
+
 Reassortment `distances' represent the temporal separation of reassorting lineages and suggests that reassortments between PB1, PB2 and HA segments tend to combine lineages that have relatively recently been part of the same genome (\textit{i.e.} that have shared a recent common virion).
@@
 Explaining the co-dependence between PB1+2 and HA segments is more difficult.
-
 Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \cite{bergeron2010,fulvini2011}.
@@
 Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
-A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct in 2002 (Figures \ref{tmrcaOT} and \ref{railroadPlot})), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
+A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct in 2002 (Figures \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot})), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
+
 Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments', which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
@@
 
-%\begin{figure}[h]
-%	\centering		
-%	\includegraphics[width=0.95\textwidth]{figures/InfB_TMRCAOverTime}
-%	\caption{\textbf{Provisional TMRCA over time plot.}}
-%\end{figure}
-
 \bibliographystyle{pnas}

### 2014-02-03 fluB.tex
Some edits and comments (prefixed by '%%AR').

@@
 
-\subsection*{Sequences and lineage assignment}
+%% AR - I think there are too many subheading in this section...
+%\subsection*{Sequences and lineage assignment}
 
@@
 
-\subsection*{Building trees}
+%\subsection*{Building trees}
 Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
@@
 
-\subsection*{Estimating linkage disequilibrium across the influenza B genome}
+\subsection*{Linkage disequilibrium across the influenza B genome}
 We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
@@
 
-\subsection*{Confirming findings with a larger dataset through lineage assignment}
+%% AR - Merge this section with the start of the methods (just before 'Measures of Diversity').
+%\subsection*{Confirming findings with a larger dataset through lineage assignment}
 We downloaded all available sequences of influenza B isolates sampled 1984--2013 from GISAID for which PB1, PB2 and HA segments were sequenced.
@@
 
-\subsection*{PA, NP, NA, MP and NS segments have periodically lost diversity}
+%% AR - suggest removal of some of the subheadings...
+%\subsection*{PA, NP, NA, MP and NS segments have periodically lost diversity}
 
-The split into Vic and Yam lineages can be seen in all segments \cite{chen2008} \tbc{also reference supplemental tree figure}.
-Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
+The differentiation into Vic and Yam lineages can be seen in all segments \cite{chen2008} \tbc{also reference supplemental tree figure}.
+Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
 Some viruses with mixed-lineage genomes have become fixed in the influenza B virus population \textit{i.e.} became the `trunk' of the phylogenetic tree of a segment.
 
-By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.
+Looking at the most recent common ancestor of all lineages existing at yearly time slices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity with sequences sampled more recently tracing their ancestry to increasingly more recent common ancestors.
 
@@
 
-Whereas PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, we do not find evidence of broad-scale lineage turnover for PB1, PB2 and HA segments.
+Whereas PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, we see no evidence of broad-scale lineage turnover for PB1, PB2 and HA segments.
 Figure \ref{tmrcaOT} shows that lineages of PB1, PB2 and HA segments are consistently derived from an old TMRCA, indicative of continued accumulation of diversity since the split of those segments into Vic and Yam lineages.
@@
 
-By plotting the ratios of sequences assigned to either Vic or Yam lineage in each segment (Figure \ref{lineageRatiosOverTime}) it is evident that losses of diversity seen in Figures \ref{tmrcaOT} and \ref{betweenDiversity} are related to fixations of either the Vic or the Yam lineage in PA, NP, NA, MP and NS segments.
+By plotting the ratios of sequences assigned to either Vic or Yam lineage in each segment (Figure \ref{lineageRatiosOverTime}) it is evident that losses of diversity seen in Figures \ref{tmrcaOT} and \ref{betweenDiversity} are related to the repeated fixation of either the Vic or the Yam lineage in PA, NP, NA, MP and NS segments.
 These losses of diversity correspond to extinctions of Vic lineage in PA, NP, NA and MP segments and Yam lineage in NS segment.
@@
 
-\subsection*{Railroad schematic plot of influenza B reassortment history}
+%% AR - perhaps this section should go in the discussion as it a summary of the reconstructions?
+\subsection*{Schematic plot of influenza B reassortment history}
 
@@
 
-\subsection*{Potential causes of co-dependence between PB1-PB2-HA}
+\subsection*{Causes of co-dependence between PB1-PB2-HA}
 Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, \textit{e.g.} PB1-PB2-PA and HA-NA \cite{mccullers2004}.
-Though it would be easy to explain co-adaptation between the aforementioned segments by referring to their functional roles (\textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities), our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.
+Though it would be easy to explain co-adaptation between these segments by referring to their functional roles (\textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities), our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.
 
@@
 Figure \ref{SPRdistances} and recent evidence suggests that, at least in influenza A, reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
-In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we suggest several alternatives.
+In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we offer several alternatives.
 
@@
 
-Explaining the co-dependence between PB1+2 and HA segments is more difficult.
+Explaining the co-dependence of PB1+2 and HA segments is more difficult.
 Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \cite{bergeron2010,fulvini2011}.
@@
 If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
+%%AR - I am not sure I understand the following. I guess this is simply a null model that what we observe is simply chance (HA/P1&2 ressortants failed to fix). Why does this need a name?
 It is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
@@
 
+%%AR - this is the partner hypothesis for the one on line 384. Put this up there. Also what would stop one or other of the PB1+PB2 complexes from fixing? 
 Finally it is possible that PB1 and PB2 segments are co-adapted whereas the HA segment is simply hitchhiking along stochastically. 
@@
 
+%%AR - rather than predictions, I think this would be better described as a possible experiment with different outcomes supporting one hypothesis or the other.
 \subsection*{Predictions}
@@
 
-\subsection*{Implications for influenza B virus control and prevention}
-It is our hope that understanding the mechanism(s) underlying the links between Vic and Yam lineage PB1-PB2-HA complexes might suggest a more varied approach to both influenza B surveillance and control.
-For one, our findings suggest that viruses bearing mixed-lineage PB1-PB2-HA complexes, if unveiled by surveillance, are unlikely to contribute to future influenza B evolution (as seen with B/Waikato/6/2005-like viruses).
-Similarly, in the absence of sequence data, it should be possible to predict the lineage of PB1, PB2 and HA segments of influenza B viruses by sequencing only one of those segments.
-
-It remains to be seen whether equivalent or similar links exist between segments of influenza A viruses, which present a much greater threat to human health worldwide than influenza B viruses.
-By discovering the underlying mechanism by which PB1, PB2 and HA are co-adapted might reveal potential targets for new classes of antiviral drugs.
+%%AR I would be very wary of this. Don't go out on a limb unnecessarily. Perhaps if there is experimental evidence to support it.
+%\subsection*{Implications for influenza B virus control and prevention}
+%It is our hope that understanding the mechanism(s) underlying the links between Vic and Yam lineage PB1-PB2-HA complexes might suggest a more varied approach to both influenza B surveillance and control.
+%For one, our findings suggest that viruses bearing mixed-lineage PB1-PB2-HA complexes, if unveiled by surveillance, are unlikely to contribute to future influenza B evolution (as seen with B/Waikato/6/2005-like viruses).
+%Similarly, in the absence of sequence data, it should be possible to predict the lineage of PB1, PB2 and HA segments of influenza B viruses by sequencing only one of those segments.
+%
+%It remains to be seen whether equivalent or similar links exist between segments of influenza A viruses, which present a much greater threat to human health worldwide than influenza B viruses.
+%By discovering the underlying mechanism by which PB1, PB2 and HA are co-adapted might reveal potential targets for new classes of antiviral drugs.
 

### 2014-02-03 fluB.tex
Addressing suggestions by @rambaut, other changes
Addressing Andrew's points, also minor text alterations.
Mostly removing excessive brackets, adding clearer references to, and
excuses for not using, the larger (1433) dataset.
Added acknowledgment section, so everyone can put their funding in
there.

@@
 
-Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
+Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments, a process known as reassortment. 
 Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of circulating human strains and with non-human influenza A strains.
@@
 \section*{Methods}
+We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in Supplementary information).
+The longest protein coding region of each segment was extracted and used for all further analyses.
+%% GD - the following bit is meant to explain why we use M1 and NS1 alignments to make trees, but refer to them as MP and NS respectively in the text and figures, but I also realized that those labels have been used very inconsistently in the figures.
+I'll correct this when I re-make the figures to be slightly better resolution and also fix some minor typeface issues I've noticed.
+"M1" and "NS1" will only be used in the LD figure, since that is relevant if measuring amino acid LD, whereas whenever trees of M1 or NS1 are used we can be reasonably sure that the history of the MP and NS segment is accurately captured by M1 and NS1 regions.
+We thus assume that homologous recombination has not taken place and that the evolutionary history of the whole segment can be inferred from the longest coding sequence in the segment.
+To date there has been little evidence of homologous recombination in influenza viruses \cite{chare2003,boni2008,han2010}.
+The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \cite{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
 
-%% AR - I think there are too many subheading in this section...
-%\subsection*{Sequences and lineage assignment}
+Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y.
 
-We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information).
-The segments of each strain were assigned to a lineage (either Vic or Yam) by making maximum likelihood trees of each segment using PhyML \cite{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
+We also downloaded all available sequences of influenza B isolates sampled 1984--2013 from GISAID for which PB1, PB2 and HA segments were sequenced.
+This comprised a dataset of 1433 isolates in total which became available only after the primary analyses were performed using the smaller dataset and had too many sequences to analyze using the methods described later.
+Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences, as before.
+Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.
 
-Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).
-
-%\subsection*{Building trees}
 Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
@@
 
-Every sequence was assigned 7 discrete traits corresponding to the lineages of all other segments with which a strain was isolated (\textit{e.g.} PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values).
-We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete traits using an asymmetric transition matrix \cite{lemey2009} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates.
-Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing trait transitions, \textit{i.e.} Yam to Vic or Vic to Yam (see Figure \ref{methodFig}).
+Every sequence was assigned 7 discrete traits in BEAUti corresponding to the lineages of all other segments with which a strain was isolated \textit{e.g.} PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values.
+We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete states using an asymmetric transition matrix \cite{lemey2009} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates.
+Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing state transitions, \textit{i.e.} Yam to Vic or Vic to Yam (see Figure \ref{methodFig}).
 In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
@@
 We inferred the diversity of each segment at a single point in time by estimating the date of the most recent common ancestor of all branches at yearly intervals, which places an upper bound on the maximum amount of diversity existing at each time period.
-In addition, we calculated mean pairwise TMRCA between branches labelled as Vic and Yam for PB1, PB2 and HA.
+In addition, we calculated mean pairwise time of most recent common ancestor (TMRCA) between branches labelled as Vic and Yam for PB1, PB2 and HA.
 This gave us a measure of how much a particular segment reassorts with respect to PB1, PB2 and HA segments.
@@
 We subsampled our combined posterior distribution of trees to give a total of 2700 trees.
-Pairwise comparison between each pair of segments was performed by comparing individual pairs of sampled trees.
+Pairwise comparisons between each pair of segments were performed by comparing individual pairs of sampled trees.
 
@@
 We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
-We used the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
+For quantifying LD we used the $\chi^{2}_{df}$ statistic \cite{zhao2005}, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
 LD was estimated only at loci where each nucleotide allele was represented by at least two isolates.
@@
 We also calculated mean LD for all pairs of segments to quantify the overall association between them.
-
-%% AR - Merge this section with the start of the methods (just before 'Measures of Diversity').
-%\subsection*{Confirming findings with a larger dataset through lineage assignment}
-We downloaded all available sequences of influenza B isolates sampled 1984--2013 from GISAID for which PB1, PB2 and HA segments were sequenced.
-This comprised a dataset of 1433 isolates in total. 
-Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences, as before.
-Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.
+Linkage disequilibrium was not estimated for the larger PB1, PB2 and HA dataset, as it would be impossible to put these measurements into a genome-wide perspective of linkage disequilibrium using just the data at hand.
 
 \section*{Results}
-
-%% AR - suggest removal of some of the subheadings...
-%\subsection*{PA, NP, NA, MP and NS segments have periodically lost diversity}
-
 The differentiation into Vic and Yam lineages can be seen in all segments \cite{chen2008} \tbc{also reference supplemental tree figure}.
 Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
-Some viruses with mixed-lineage genomes have become fixed in the influenza B virus population \textit{i.e.} became the `trunk' of the phylogenetic tree of a segment.
+Some segment lineages associated with mixed-lineage genomes have become fixed in the influenza B virus population \textit{i.e.} became the `trunk' of the phylogenetic tree of a segment.
 
@@
 \subsection*{PB1, PB2 and HA segments do not reassort across the Victoria-Yamagata boundary}
-By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in diversity, which indicate that an inter-lineage reassortment event has taken place.
+By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between lineage diversity, which indicate that an inter-lineage reassortment event has taken place.
 Because we are taking the mean of pairwise comparisons, this method gives a qualitative measure of panmixis between Victoria and Yamagata lineages in two trees.
@@
 	\includegraphics[width=0.65\textwidth]	{figures/InfB_LineageRatiosOverTime.png}
-	\caption{\textbf{Ratio of lineages in the dataset.}}
+	\caption{\textbf{Ratio of lineages in the dataset.}
+The ratio of Victoria (red) to Yamagata (blue) lineage sequences in each segment over time.
+Yamagata lineage PA, NP, NA and MP segments and Victoria lineage NS segment eventually become fixed in the influenza B population.
+PB1, PB2 and HA segments maintain separate Victoria and Yamagata lineages.}
 	\label{lineageRatiosOverTime}
@@
 Inter-lineage reassortment events are indicated by lines entering a different genome.
-The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the parent node and mean TMRCA of reassortant clades).
+The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the reassortant node and its parent node).
 Lineage extinction dates are not shown accurately.}
@@
 We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
-An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
+An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
 
 In all cases, PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
-In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each having two sequences), exhibit poor sampling and short circulation times (up to two years in the case of B/Nanchang/630/1994-like viruses).
-In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past (up to 4 years for B/Bankok/163/1990-like, although B/Waikato/6/2005-like only circulated for 5 months).
+In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each represented by two isolates), exhibit poor sampling and short circulation times.
+In the case of B/Nanchang/630/1994-like viruses this is up to two years.
+In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past: up to 4 years for B/Bankok/163/1990-like, although the more recently isolated B/Waikato/6/2005-like viruses circulated for only 5 months.
 
@@
 If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to see small-scale similarities between phylogenetic trees of those segments.
-In our case we expect SPR distances (which are proportional to the numbers of reassortment that have taken place between trees) to reflect the raw reassortment rate.
+In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the raw reassortment rate.
 
@@
 
-In Figure \ref{SPRdistances} the 95\% HPD intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, data not shown) between all segments.
+In Figure \ref{SPRdistances} the 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, see Supplementary information) between all segments.
 We note that our SPR distance analysis might simply lack power.
@@
 In addition, we use approximate SPR distances, rather than exact SPR distances.
-However, we do estimate exact SPR distances for a limited number of segment pairs (PB1-PB2 and PB1-HA) and find that after normalization exact and approximate SPR distances are not significantly different.
+However, we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (see Supplementary information).
 
@@
 Figure \ref{deltaTMRCA} shows normalized mean $\Delta$TMRCA values for all pairs of trees.
-Most segment pairs show very low values for this statistic, indicating differences in TMRCAs.
-PB1, PB2 and HA trees, on the other hand, exhibit TMRCAs between themselves which are, on average, much more similar to the replicates of those trees.
-This suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, \textit{i.e.} there is isolation by distance.
-In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees.
+Most segment pairs show very low values for this statistic, indicating considerable differences between TMRCAs of the trees being compared.
+A large number of these comparisons have TMRCA differences up to 10 times larger ($\Delta$TMRCA$\approx$0.1) than TMRCA differences that occur between replicate BEAST analyses of one alignment over the course of the MCMC chain.
+PB1, PB2 and HA trees, on the other hand, exhibit normalized $\Delta$TMRCA values that are much higher.
+This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occassionally very similar to uncertainty in tip-to-tip TMRCAs within either of these segments.
+Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, \textit{i.e.} there is isolation by distance.
+In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.
 Previously, associations between MP and NA segments have been suspected in influenza A viruses (Samantha Lycett, personal communication).
@@
 These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
-In addition, there are sites on PB1, PB2 and HA proteins which also show high (albeit smaller) LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
+In addition, there are sites on PB1, PB2 and HA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
 
@@
 In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were much more similar at the nucleotide and amino acid levels.
-The most recent lineage of influenza B viruses with mixed-lineage PB1-PB2-HA complexes (B/Waikato/6/2005-like) only circulated for 5 months before going extinct.
+The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.
 We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands \textit{i.e.} small numbers of genes that resist being homogenized through gene flow.
@@
 Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, \textit{e.g.} PB1-PB2-PA and HA-NA \cite{mccullers2004}.
-Though it would be easy to explain co-adaptation between these segments by referring to their functional roles (\textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities), our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.
+Though it would be easy to explain co-adaptation between these segments by referring to their functional roles \textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities, our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.
 
-It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations (which did not become fixed in the population) and our estimates of the number of reassortments do not differ significantly between all segments.
-Figure \ref{SPRdistances} and recent evidence suggests that, at least in influenza A, reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
+It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations which did not become fixed in the population and our estimates of the number of reassortments do not differ significantly between all segments.
+Our finding that influenza B virus segments do not significantly differ in reassortment rate (Figure \ref{SPRdistances}) are in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
 In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we offer several alternatives.
@@
 Additional evidence for a relationship between PB1 and HA segments in influenza A viruses is given by previous influenza pandemics, which were caused by avian-human influenza A virus reassortants.
-It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments (and in the case of A/H2N2, NA as well) derived from avian influenza A viruses \cite{kawaoka1989}.
+It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments derived from avian influenza A viruses \cite{kawaoka1989}.
 However, there have also been reassortant influenza viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \cite{gregory2002} and H1N1/09 in 2009 \cite{smith2009}.
@@
 A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct in 2002 (Figures \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot})), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
+We find this scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
+Our main dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bankok/163/1990-like genome constellations.
+The larger dataset with 1433 PB1, PB2 and HA sequences has an additional example of this: isolates with B/Waikato/6/2005-like PB1-PB2-HA constellations.
+We note that in both cases the genome constellations of these reassortants were not fixed in the influenza B population and the more recent B/Waikato/6/2005-like reassortants persisted for a much shorter period of time (5 months versus 4 years), suggestive of declining PB1+2/HA reassortant fitness over time.
 
@@
 If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
-%%AR - I am not sure I understand the following. I guess this is simply a null model that what we observe is simply chance (HA/P1&2 ressortants failed to fix). Why does this need a name?
+
 It is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
+In this case PB1, PB2 and HA segments of Vic and Yam lineage accumulate substitutions that improve their ability to co-operate with segments of the same lineage and worse at interacting segments of a different lineage.
 This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \cite{lee2008}.
-
-%%AR - this is the partner hypothesis for the one on line 384. Put this up there. Also what would stop one or other of the PB1+PB2 complexes from fixing? 
-Finally it is possible that PB1 and PB2 segments are co-adapted whereas the HA segment is simply hitchhiking along stochastically. 
-We find it unlikely given that B/Waikato/6/2005-like strains, the most recent isolates with a mixed-lineage PB1-PB2-HA complex, circulated for a limited amount of time within a limited geographic area, suggesting that this particular lineage was outcompeted by influenza B strains with `pure' PB1-PB2-HA complexes.
 
@@
 
-%%AR - rather than predictions, I think this would be better described as a possible experiment with different outcomes supporting one hypothesis or the other.
-\subsection*{Predictions}
+\subsection*{Suggested experiments}
 Using previously developed plasmid systems \cite{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
@@
 Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
+We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (see Supplementary information).
+The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
+However, our analyses do not indicate any obvious or dramatic differences between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two.
 
@@
 We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins that fall within the regions that form contacts within the influenza B polymerase heterotrimer, suggesting more subtle roles for sites we have identified.
-
-%%AR I would be very wary of this. Don't go out on a limb unnecessarily. Perhaps if there is experimental evidence to support it.
-%\subsection*{Implications for influenza B virus control and prevention}
-%It is our hope that understanding the mechanism(s) underlying the links between Vic and Yam lineage PB1-PB2-HA complexes might suggest a more varied approach to both influenza B surveillance and control.
-%For one, our findings suggest that viruses bearing mixed-lineage PB1-PB2-HA complexes, if unveiled by surveillance, are unlikely to contribute to future influenza B evolution (as seen with B/Waikato/6/2005-like viruses).
-%Similarly, in the absence of sequence data, it should be possible to predict the lineage of PB1, PB2 and HA segments of influenza B viruses by sequencing only one of those segments.
-%
-%It remains to be seen whether equivalent or similar links exist between segments of influenza A viruses, which present a much greater threat to human health worldwide than influenza B viruses.
-%By discovering the underlying mechanism by which PB1, PB2 and HA are co-adapted might reveal potential targets for new classes of antiviral drugs.
 
@@
 
+\section*{Acknowledgements}
+
 \bibliographystyle{pnas}

### 2014-02-03 fluB.tex
Copy and paste some acknowledgements...

@@
 \section*{Acknowledgements}
+GD was supported by a NERC studentship..... 
+TB was supported by a Newton International Fellowship from the Royal Society. 
+The research leading to these results has received funding from the European Research Council under the European Community's Seventh Framework Programme (FP7/2007-2013) under Grant Agreement no.
+278433-PREDEMICS and ERC Grant agreement no.
+260864.
+AR and SL acknowledge the support of the Wellcome Trust (grant no. 092807).
 

### 2014-02-03 fluB.tex
Template for direct referencing of supporting information from main text.

@@
 
-\section*{Conclusions}
+\section*{Conclusion}
 In this study we have investigated the patterns of reassortment amongst two influenza B virus segment lineages and found preserved diversity and consistent co-reassortment of 3 segments: PB1, PB2 and HA.
@@
 Given sufficient time we expect the two PB1-PB2-HA complexes to recruit the rest of the genomic segments into co-adapted and co-reassorting segment complexes eventually resulting in sympatric speciation of influenza B viruses.
+
+\section*{Supporting information}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+
+\begin{figure}[h]
+\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and PB2 segments based on $\chi^{2}_{df}$ values.}}
+\label{ChiPB1PB2}
+\end{figure}
+
+\begin{figure}[h]
+\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and HA segments based on $\chi^{2}_{df}$ values.}}
+\label{ChiPB1HA}
+\end{figure}
+
+% TB: FILL IN WITH REST OF SUPP FIGURES
 

### 2014-02-03 fluB.tex
Revise Results.

@@
 We argue that this difference between segments is due to selection against reassortant viruses with mixed lineage PB1, PB2 and HA segments.
-Given sufficient time we expect influenza B viruses to undergo sympatric speciation, if more segments get recruited by the already reassortment-isolated PB1-PB2-HA gene complex.
+Given sufficient time and continued recruitment to the reassortment-isolated PB1-PB2-HA gene complex, we expect influenza B viruses to eventually undergo sympatric speciation.
 
@@
 \section*{Results}
-The differentiation into Vic and Yam lineages can be seen in all segments \cite{chen2008} \tbc{also reference supplemental tree figure}.
+
+\subsection*{Phylogenetic analyses reveal maintenance of diversity and co-assortment of PB1, PB2 and HA gene segments}
+
+\tbc{Make this paragraph a broad description of the phylogenies across the segments...
+basically the current paragraph with a few tweaks.} The differentiation into Vic and Yam lineages can be seen in all segments \cite{chen2008} \tbc{also reference tree figure}.
 Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
@@
 
-Looking at the most recent common ancestor of all lineages existing at yearly time slices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity with sequences sampled more recently tracing their ancestry to increasingly more recent common ancestors.
+Looking at patterns of ancestry across contemporaneous viruses for each segment, it is clear that the PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, while the PB1, PB2 and HA segments have maintained diversity up to the present (Figure \ref{tmrcaOT}).
+The PB1, PB2 and HA segments maintain a common ancestor in $\sim$1983 up to the present and thus accumulate genetic diversity since the split of those segments into Vic and Yam lineages, while other segments often lose diversity with ancestors to present day viruses appearing between $\sim$1991 and $\sim$1999.
 
@@
 	\centering		
-	\includegraphics[width=0.65\textwidth]{figures/InfB_tmrcaOT_lines.png}
+	\includegraphics[width=0.75\textwidth]{figures/InfB_tmrcaOT_lines.png}
 	\caption{\textbf{Oldest TMRCA of all surviving lineages over time.}
@@
 PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
-All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs). \tbc{Describe that each point estimate is a slice through the tree.}}
 	\label{tmrcaOT}
@@
 
-Whereas PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, we see no evidence of broad-scale lineage turnover for PB1, PB2 and HA segments.
-Figure \ref{tmrcaOT} shows that lineages of PB1, PB2 and HA segments are consistently derived from an old TMRCA, indicative of continued accumulation of diversity since the split of those segments into Vic and Yam lineages.
+By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between lineage diversity, which indicate that an inter-lineage reassortment event has taken place (Figure \ref{betweenDiversity}).
+Because we are taking the mean of pairwise comparisons, this method gives a qualitative measure of panmixis between Victoria and Yamagata lineages in two trees.
+We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
+Increasing mean TMRCA of Vic and Yam lineage pairs for all 5 segments (Figure \ref{betweenDiversity}) indicate that every segment has reassorted with respect to the Victoria and Yamagata lineages of PB1, PB2 and HA segments.
+However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
+This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments (\textit{i.e.} PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees).
+In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1 and PB2 segments on one hand and HA on the other.
+These reductions in diversity represent small clades with PB1 and PB2 segments from one lineage and the HA segment from another, which go extinct by 1997.
+In addition, the assignment of Vic or Yam lineages of PB1, PB2 and HA segments to branches of other segment trees is very similar and often identical.
+This results in PB1, PB2 and HA lineage labels switching between Vic and Yam simultaneously in all trees after 1997, suggesting co-reassortment of Vic and Yam lineages of PB1, PB2 and HA segments.
 
-\subsection*{PB1, PB2 and HA segments do not reassort across the Victoria-Yamagata boundary}
-By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between lineage diversity, which indicate that an inter-lineage reassortment event has taken place.
-Because we are taking the mean of pairwise comparisons, this method gives a qualitative measure of panmixis between Victoria and Yamagata lineages in two trees.
-
-We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
-Increasing mean between label TMRCAs for all segments in Figure \ref{betweenDiversity} indicate that every segment has reassorted with respect to Victoria and Yamagata lineages of PB1, PB2 and HA segments.
-
-\begin{figure}
+\begin{figure}[h]
 	\centering		
@@
 \end{figure}
+
+By plotting the ratios of sequences assigned to either Vic or Yam lineage in each segment (Figure \ref{lineageRatiosOverTime}) it is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
+These losses of diversity correspond to fixation of the Vic lineage in NS and fixation of the Yam lineage in PA, NP, NA and MP.
+Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen (Figure \ref{lineageRatiosOverTime}), where the two lineages have been isolated at a ratio close to 50\% over long periods of time.
+On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically suggesting that one lineage may predominate within a given season \tbc{CITE}.
 
@@
 
-However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
-This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments (\textit{i.e.} PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees).
-In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1 and PB2 segments on one hand and HA on the other.
-These reductions in diversity represent small clades with PB1 and PB2 segments from one lineage and the HA segment from another, which go extinct by 1997.
-
-In addition, the assignment of Vic or Yam lineages of PB1, PB2 and HA segments to branches of other segment trees is very similar and often identical.
-This results in PB1, PB2 and HA lineage labels switching between Vic and Yam simultaneously in all trees after 1997, suggesting co-reassortment of Vic and Yam lineages of PB1, PB2 and HA segments.
-
-By plotting the ratios of sequences assigned to either Vic or Yam lineage in each segment (Figure \ref{lineageRatiosOverTime}) it is evident that losses of diversity seen in Figures \ref{tmrcaOT} and \ref{betweenDiversity} are related to the repeated fixation of either the Vic or the Yam lineage in PA, NP, NA, MP and NS segments.
-These losses of diversity correspond to extinctions of Vic lineage in PA, NP, NA and MP segments and Yam lineage in NS segment.
-Similarly, the lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA segments is reflected in Figure \ref{lineageRatiosOverTime} where the two lineages have been isolated at a ratio close to 0.5 over long periods of time.
-On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically suggesting frequency-dependent domination of one or the other lineage within a given season.
-
 %% AR - perhaps this section should go in the discussion as it a summary of the reconstructions?
-\subsection*{Schematic plot of influenza B reassortment history}
+%\subsection*{Schematic plot of influenza B reassortment history}
 
@@
 We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV).
+\tbc{Repeat finding that the railroad plot shows introgression and fixation of Vic lineages into a Yam background and vice versa, and that PB1, PB2 and HA go together in reassortment splits.}
+Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
+For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to B/Iowa/03/2002-like isolates.
+A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has been derived from Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
 
@@
 The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the reassortant node and its parent node).
-Lineage extinction dates are not shown accurately.}
+Lineage extinction dates are not shown accurately.
+\tbc{Giving this a bit more vertical spread (increasing aspect ratio) would likely clarify things a bit.}}
 	\label{railroadPlot}
 \end{figure}
-
-Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
-For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to B/Iowa/03/2002-like isolates.
-A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has been derived from Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
 
@@
 \subsection*{PB1, PB2 and HA trees experience reassortment `distance' effects}
+\tbc{Briefly define SPR, or just say that SPR measures tree similarity.}
 Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of trees after normalization (see Methods).
@@
 \subsection*{Linkage disequilibrium analysis confirms phylogenetic findings}
+\tbc{Briefly define LD and say that it doesn't depend on Vic and Yam lineage labels.}
 Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.

### 2014-02-03 fluB.tex
Revise discussion.

@@
 We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete states using an asymmetric transition matrix \cite{lemey2009} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates.
-Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing state transitions, \textit{i.e.} Yam to Vic or Vic to Yam (see Figure \ref{methodFig}).
+Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing state transitions, i.e.\ Yam to Vic or Vic to Yam (see Figure \ref{methodFig}).
 In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
@@
 Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
-Some segment lineages associated with mixed-lineage genomes have become fixed in the influenza B virus population \textit{i.e.} became the `trunk' of the phylogenetic tree of a segment.
+Some segment lineages associated with mixed-lineage genomes have become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment.
 
@@
 However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
-This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments (\textit{i.e.} PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees).
+This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments, i.e.\ PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees.
 In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1 and PB2 segments on one hand and HA on the other.
@@
 This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occassionally very similar to uncertainty in tip-to-tip TMRCAs within either of these segments.
-Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, \textit{i.e.} there is isolation by distance.
+Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, i.e.\ there is isolation by distance.
 In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.
@@
 	\centering	
-	\includegraphics[width=0.65\textwidth]	{figures/InfB_aaMeanLD.png}
+	\includegraphics[width=0.65\textwidth]{figures/InfB_aaMeanLD.png}
 	\caption{\textbf{LD comparison between influenza B proteins.}
@@
 
-\subsection*{Evidence of incipient sympatric speciation}
-In this paper we have shown that PB1, PB2 and HA segments of influenza B viruses are the only ones that have maintained both a Vic and a Yam lineage (Figure \ref{lineageRatiosOverTime}).
-Evidence suggests that this is a result of prolonged lack of reassortment between Vic and Yam lineages of these segments (Figure \ref{betweenDiversity}) which have evolved co-assorting amino acid sites detectable as high linkage disequilibrium (Figure \ref{segmentLD}).
+\subsection*{Evidence of a co-adapted PB1-PB2-HA gene complex}
+In this paper we show that the PB1, PB2 and HA segments of influenza B viruses are the only ones that have continuously maintained separate Vic and Yam lineages, while other segments have had undergone fixation by either Vic or Yam (Figures \ref{tmrcaOT}--\ref{railroadPlot}).
+Evidence suggests that this is a result of prolonged lack of reassortment between Vic and Yam lineages in PB1, PB2 and HA (Figure \ref{betweenDiversity}) which possess co-assorting sequences detectable as high linkage disequilibrium (Figure \ref{segmentLD}).
 We propose that this pattern of non-reassortment is due to the action of selection and not simply biased reassortment.
@@
 
-Reassortment `distances' represent the temporal separation of reassorting lineages and suggests that reassortments between PB1, PB2 and HA segments tend to combine lineages that have relatively recently been part of the same genome (\textit{i.e.} that have shared a recent common virion).
+Reassortment `distances' represent the temporal separation of reassorting lineages and suggests that reassortments between PB1, PB2 and HA segments tend to combine lineages that have relatively recently been part of the same genome, i.e.\ that have shared a recent common virion.
 In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were much more similar at the nucleotide and amino acid levels.
 The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.
-We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands \textit{i.e.} small numbers of genes that resist being homogenized through gene flow.
-Thus, if both PB1-PB2-HA complex lineages continue circulating in humans we expect more segments to be recruited to the two complexes, resulting in eventual sympatric speciation of influenza B viruses.
+We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \tbc{CITE}.
+\tb{Thus, if both PB1-PB2-HA complex lineages continue circulating in humans we expect more segments to be recruited to the two complexes, resulting in eventual sympatric speciation of influenza B viruses.}{}
+\tbc{Suggest moving this idea to `Future' section below.}
 
@@
 Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \cite{bergeron2010,fulvini2011}.
-Early influenza A vaccines used to be derived by infecting chicken eggs with an egg-adapted strain and a seasonal human isolate in the presence of antisera raised against the HA and NA proteins of the egg-adapted strain, thus selecting for reassortants which were egg-adapted (\textit{i.e.} had internal segments derived from the egg-adapted strain) but possessed the HA and NA segments of the seasonal strain.
+Early influenza A vaccines used to be derived by infecting chicken eggs with an egg-adapted strain and a seasonal human isolate in the presence of antisera raised against the HA and NA proteins of the egg-adapted strain, thus selecting for reassortants which were egg-adapted, i.e.\ had internal segments derived from the egg-adapted strain, but possessed the HA and NA segments of the seasonal strain.
 However, in addition to producing reassortants with HA-NA derived from the seasonal strain, as intended, the second most frequent class of reassortants produced using this methods were those with PB1, HA and NA derived from seasonal strains \cite{bergeron2010,fulvini2011}.
@@
 Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
-A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct in 2002 (Figures \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot})), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
+A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot})).
 We find this scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
@@
 If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
+\tbc{However, avian PB1 and PB2 are not more diverse than other internal segments, yes?}
 
@@
 This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \cite{lee2008}.
+\tbc{I would think this would still require a functional linkage between PB1, PB2 and HA.  This is the mechanism by which a functional linkage results in a co-adapted gene complex.}
 
@@
 
-Sympatric speciation in other systems usually requires strong barriers to introgression, \textit{e.g.} infertility of F1 hybrids can lead to the evolution of prezygotic reproductive isolation otherwise known as reinforcement or Wallace effect.
+Sympatric speciation in other systems usually requires strong barriers to introgression, \textit{e.g.} infertility of F1 hybrids can lead to the evolution of prezygotic reproductive isolation otherwise known as reinforcement or the Wallace effect.
 To what extent this would apply to influenza B viruses remains unknown.
@@
 If influenza B viruses are undergoing sympatric speciation, does co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or is co-infection so rare that speciation occurs via mutation-driven co-evolution?
+\tbc{I don't understand this.}
 Given that PB1 and PB2 segments are functionally linked, is it reasonable to expect that the next segment to be recruited to the PB1-PB2-HA segment complexes will be functionally linked to either PB1/PB2 or HA?
@@
 This is the second scenario we might expect to occur, whereby relatively frequent reassortments occur between Vic and Yam lineages and are followed by selective sweeps.
-In this case influenza B viruses would undergo periodic genome homogenization with the exception of PB1, PB2 and HA segments.
+In this case influenza B viruses would undergo periodic genome homogenization with the exception of PB1, PB2 and HA segments, as strong antigenic dissimilarity between Vic HA and Yam HA would cause even strong selective sweeps to be unable to result in extinction of the opposing PB1-PB2-HA gene complex.
 However, this model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.
@@
 Using previously developed plasmid systems \cite{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
-We predict that artificially produced viruses with mixed-lineage PB1-PB2-HA complexes will have reduced fitness when compared to viruses with pure-lineage (\textit{i.e.} entirely Vic or entirely Yam) PB1-PB2-HA complexes.
+We predict that artificially produced viruses with mixed-lineage PB1-PB2-HA complexes will have reduced fitness when compared to viruses with pure-lineage, i.e.\ entirely Vic or entirely Yam, PB1-PB2-HA complexes.
 In addition, we expect the relationship between Vic and Yam lineage PB1, PB2 and HA segments to be dependent on date of segment isolation, as viruses with mixed-lineage PB1-PB2-HA complexes isolated earlier should perform better than viruses with PB1-PB2-HA segments isolated more recently.
@@
 
-\section*{Conclusion}
+\subsection*{Conclusion}
 In this study we have investigated the patterns of reassortment amongst two influenza B virus segment lineages and found preserved diversity and consistent co-reassortment of 3 segments: PB1, PB2 and HA.
@@
 Given sufficient time we expect the two PB1-PB2-HA complexes to recruit the rest of the genomic segments into co-adapted and co-reassorting segment complexes eventually resulting in sympatric speciation of influenza B viruses.
+
+\tbc{This conclusion restates results too much.
+This is more of a beginning to a discussion than an end to it (you might even consider moving it there).
+Normally a conclusion is focused either on future work that this makes possible or public health implications.
+`Suggest experiments' kind of takes care of  the further work angle, and I'm not actually sure about public health implications.
+Or perhaps emphasize novelty of your phylogenetic approach to reassortment and say it could be applied to other systems.}
 

### 2014-02-04 fluB.tex
Addressing points made by Trevor.
Addressing some points and adding a vague beginning of a different
conclusion. Also found a paper co-authored with Darren that's relevant
and also managed to find a paper that uses our ΔTMRCA statistic (but
calls it Δt and uses it in a completely different way).

@@
 PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
-All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs). \tbc{Describe that each point estimate is a slice through the tree.}}
+Each point is the TMRCA of all surviving lineages existing at each time slice through the tree and vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
 	\label{tmrcaOT}
@@
 Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen (Figure \ref{lineageRatiosOverTime}), where the two lineages have been isolated at a ratio close to 50\% over long periods of time.
-On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically suggesting that one lineage may predominate within a given season \tbc{CITE}.
+On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically suggesting that one lineage may predominate within a given season, consistent with surveillance data \cite{reed2012}.
 
@@
 Figure \ref{railroadPlot} focuses only on reassortments that have occured after 1990.
-We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV).
-\tbc{Repeat finding that the railroad plot shows introgression and fixation of Vic lineages into a Yam background and vice versa, and that PB1, PB2 and HA go together in reassortment splits.}
+We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV) in Figure \ref{railroadPlot}.
+Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata lineage derived background.
+All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata lineage derived PB1-B2-HA.
 Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
@@
 \subsection*{PB1, PB2 and HA trees experience reassortment `distance' effects}
-\tbc{Briefly define SPR, or just say that SPR measures tree similarity.}
-Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of trees after normalization (see Methods).
+SPR (subtree prune and regraft) distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between the two trees and is a measure of tree-to-tree similarity.
+Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
 If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to see small-scale similarities between phylogenetic trees of those segments.
@@
 \subsection*{Linkage disequilibrium analysis confirms phylogenetic findings}
-\tbc{Briefly define LD and say that it doesn't depend on Vic and Yam lineage labels.}
+Linkage disequilibrium (LD) is a measure of association between alleles at different polymorphic loci within a population.
+To estimate LD no other information, except for alleles at polymorphic amino acid sites in the dataset, is required.
 Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.
@@
 The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.
-We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \tbc{CITE}.
-\tb{Thus, if both PB1-PB2-HA complex lineages continue circulating in humans we expect more segments to be recruited to the two complexes, resulting in eventual sympatric speciation of influenza B viruses.}{}
-\tbc{Suggest moving this idea to `Future' section below.}
+We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \cite{turner2005}.
 
@@
 
-Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments', which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
+Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
 This is easiest to interpret as frequency-dependent balancing selection acting to preserve antigenic diversity.
 If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
-\tbc{However, avian PB1 and PB2 are not more diverse than other internal segments, yes?}
 
@@
 This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \cite{lee2008}.
-\tbc{I would think this would still require a functional linkage between PB1, PB2 and HA.  This is the mechanism by which a functional linkage results in a co-adapted gene complex.}
+Similarly to the previous scenario, however, an association of some kind between PB1+2 and HA segments would be required to explain the low observation frequency and poor fitness of PB1-PB2/HA reassortants.
 
@@
 If influenza B viruses are undergoing sympatric speciation, does co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or is co-infection so rare that speciation occurs via mutation-driven co-evolution?
-\tbc{I don't understand this.}
-Given that PB1 and PB2 segments are functionally linked, is it reasonable to expect that the next segment to be recruited to the PB1-PB2-HA segment complexes will be functionally linked to either PB1/PB2 or HA?
+We think that co-dependence between PB1 and PB2 segments can be explained by the fact that they are functionally linked: together with the PA protein they form part of the influenza B virus RNA-dependent RNA polymerase heterotrimer.
+If more segments get recruited to the PB1-PB2-HA segment complex, is it reasonable to assume that the next segment to be recruited will also code for a protein functionally linked to either the two polymerase proteins (PB1 and PB2) or haemagglutinin (HA)?
 
@@
 \subsection*{Conclusion}
+%The $\Delta$TMRCA method, which is an extension of patristic (or phylogenetic) distance methods, can be applied to cases where reticulate evolution is suspected and phylogenetic trees can readily be made.
+%Patristic distance methods are far from new, although their use in conjunction with temporally calibrated phylogenies has been limited.
+For example, one study focused on patristic (or phylogenetic) distances to predict viral titer in \textit{Drosophila} species infected by viruses of other \textit{Drosophila} species \cite{longdon2011}.
+Another study has used the equivalent statistic to our $\Delta$TMRCA (called $\Delta$\textit{t}) to assess the temporal incongruence in phylogenies of 171 amphibian species induced by using different calibrations \cite{ruane2011}.
+%By using relatively straightforward measures of diversity, such as TMRCA of all surviving lineages over time and mean pairwise TMRCA between labels...
+
 In this study we have investigated the patterns of reassortment amongst two influenza B virus segment lineages and found preserved diversity and consistent co-reassortment of 3 segments: PB1, PB2 and HA.

### 2014-02-05 fluB.tex
Found semi-serious labeling errors.
Got mixed up with PB1/PB2 labels in the subtype maps (occurred quite
early on).
Looks like the right trees were run but PB1 and PB2 labels were
switched. It only affects the visibility of PB1-PB2 reassortants on
trees when using discrete traits though, so nothing has to be redone.

Also including the genomic trees grid figure for supplementary info.

@@
 
-We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
+We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
 An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.

### 2014-02-06 fluB.tex
Add language to distinguish lineages that associate from lineages that derive. Per #7.

@@
 	\caption{\textbf{Ratio of lineages in the dataset.}
-The ratio of Victoria (red) to Yamagata (blue) lineage sequences in each segment over time.
+The ratio of sequences that derive from the original Victoria clade (red) to sequences that drive from the Yamagata clade (blue) lineage sequences in each segment over time.
 Yamagata lineage PA, NP, NA and MP segments and Victoria lineage NS segment eventually become fixed in the influenza B population.
@@
 	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
-Victoria and Yamagata lineage genomes are represented by 8 (red or blue, respectively) parallel lines.
+Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored red and lineages that derive from the original Yamagata clade colored blue.
 Inter-lineage reassortment events are indicated by lines entering a different genome.

### 2014-02-07 fluB.tex
Lineage should now only refer to Vic/Yam.
Addressing #7.
Also adding alternative Vic/Yam colour scheme (orange/gray) to the
current one (orange/green). If everyone is happy with the new one I'll
update all 3 figures with the new one.

@@
 	\includegraphics[width=0.75\textwidth]{figures/InfB_tmrcaOT_lines.png}
-	\caption{\textbf{Oldest TMRCA of all surviving lineages over time.}
-PA, NP, NA, MP and NS segments of influenza B viruses show periodic losses of diversity, indicating lineage turn over.
+	\caption{\textbf{Oldest TMRCA of all surviving branches over time.}
+PA, NP, NA, MP and NS segments of influenza B viruses show periodic losses of diversity, indicating lineage turnover.
 PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
@@
 Lineage extinction dates are not shown accurately.
-\tbc{Giving this a bit more vertical spread (increasing aspect ratio) would likely clarify things a bit.}}
 	\label{railroadPlot}
@@
 
-Reassortment `distances' represent the temporal separation of reassorting lineages and suggests that reassortments between PB1, PB2 and HA segments tend to combine lineages that have relatively recently been part of the same genome, i.e.\ that have shared a recent common virion.
+Reassortment `distances' represent the temporal separation of reassorting branches and suggests that reassortments between PB1, PB2 and HA segments tend to combine branches that have relatively recently been part of the same genome, i.e.\ that have shared a recent common virion.
 In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were much more similar at the nucleotide and amino acid levels.
@@
 The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
-However, our analyses do not indicate any obvious or dramatic differences between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two.
+However, our analyses do not indicate any obvious or dramatic differences between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two (see Supplementary information).
 
@@
 Another study has used the equivalent statistic to our $\Delta$TMRCA (called $\Delta$\textit{t}) to assess the temporal incongruence in phylogenies of 171 amphibian species induced by using different calibrations \cite{ruane2011}.
-%By using relatively straightforward measures of diversity, such as TMRCA of all surviving lineages over time and mean pairwise TMRCA between labels...
+%By using relatively straightforward measures of diversity, such as TMRCA of all surviving branches over time and mean pairwise TMRCA between labels...
 

### 2014-02-07 fluB.tex
Moving  supp to end of manuscript.
Also beginning to write some supplementary text.

@@
 The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the reassortant node and its parent node).
-Lineage extinction dates are not shown accurately.
+Lineage extinction dates are not shown accurately.}
 	\label{railroadPlot}
@@
 
-We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
+We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bangkok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
 An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
@@
 In the case of B/Nanchang/630/1994-like viruses this is up to two years.
-In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past: up to 4 years for B/Bankok/163/1990-like, although the more recently isolated B/Waikato/6/2005-like viruses circulated for only 5 months.
+In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bangkok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past: up to 4 years for B/Bangkok/163/1990-like, although the more recently isolated B/Waikato/6/2005-like viruses circulated for only 5 months.
 
@@
 We find this scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
-Our main dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bankok/163/1990-like genome constellations.
+Our main dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bangkok/163/1990-like genome constellations.
 The larger dataset with 1433 PB1, PB2 and HA sequences has an additional example of this: isolates with B/Waikato/6/2005-like PB1-PB2-HA constellations.
@@
 
-\begin{figure}[h]
+\section*{MCC trees}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.95\textwidth]  {supp_figures/InfB_genomeGrid.pdf}
+\caption{\textbf{MCC trees of 8 segments.}}
+\label{}
+\end{figure}
+
+\section*{Linkage disequilibrium data}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB1_PB2.png}
 \caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and PB2 segments based on $\chi^{2}_{df}$ values.}}
@@
 
-\begin{figure}[h]
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB1_HA.png}
 \caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and HA segments based on $\chi^{2}_{df}$ values.}}
 \label{ChiPB1HA}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB2_HA.png}
+\caption{\textbf{Hierarchical clustering of amino acid sites on PB2 and HA segments based on $\chi^{2}_{df}$ values.}}
+\label{ChiPB2HA}
+\end{figure}
+
+We have identified amino acid sites which exhibit a high degree of linkage disequilibrium between PB1, PB2 and HA segments.
+Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three proteins at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.
+
+None of the sites we have identified on PB1 and PB2 proteins fall within the previously described regions that form the contacts between the two polymerase subunits \cite{sugiyama2009}.
+
+Sites 211, 213 and 214 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
+Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211, 213 and 214 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
+Previous research \cite{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
+Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 positions 211 and 214 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at positions 211 and 214 and threonine (T) at position 213.
+It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
+Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \cite{fodor2004}.
+Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \cite{guilligay2008}.
+Again, we cannot determine whether this finding is significant or not.
+
+\section*{Tree-to-tree comparisons}
+
+Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates} are showing approximate SPR distances prior to the normalization procedure (see Methods).
+Figure \ref{SPRdistancesTrees} shows that MP and NS trees show greater approximate SPR distances when compared to any other segment tree, which is consistent with the fact that only M1 and NS1 coding sequences, which are shorter than any other coding sequences, were used to infer the trees.
+Figure \ref{SPRdistancesReplicates} confirms this by showing greater variability in tree topology (as greater approximate SPR distance) when MP and NS trees are compared against their replicates.
+
+For all other segments approximate SPR distances between all pairs of trees prior to normalization suggests that most segment pairs have similar posterior distributions of approximate SPR distances.
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_trees.png}
+\caption{\textbf{Approximate SPR distances between all pairs of trees of segments.}}
+\label{SPRdistancesTrees}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_replicates.png}
+\caption{\textbf{Approximate SPR distances between replicate trees of each segment.}}
+\label{SPRdistancesReplicates}
+\end{figure}
+
+Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates} show $\Delta$TMRCA statistics for inter-segment and intra-segment, respectively, comparisons prior to normalization.
+Figure \ref{deltaTMRCAtrees} shows that similarity in TMRCAs (as low $\Delta$TMRCA) between PB1, PB2 and HA segments is due to considerably smaller $\Delta$TMRCA statistic for the comparison between these segments.
+It is also noteworthy that NA and MP comparisons have similarly low values for this statistic.
+Figure \ref{deltaTMRCAreplicates} shows that TMRCAs are quite robust over the course of MCMC and node times do not, even in the most extreme cases, deviate from their replicates by more than 1 year.
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
+\caption{\textbf{Mean $\Delta$TMRCA between all pairs of trees of segments.}}
+\label{deltaTMRCAtrees}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
+\caption{\textbf{Mean $\Delta$TMRCA between replicate trees of each segment.}}
+\label{deltaTMRCAreplicates}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_normRErate.png}
+\caption{\textbf{Normalized reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_trees.png}
+\caption{\textbf{Reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_replicates.png}
+\caption{\textbf{Reassortment rate (approximate number of SPR moves per total time in the tree) between replicate trees of each segment.}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist.png}
+\caption{\textbf{Difference}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist2.png}
+\caption{\textbf{Ratio}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist.png}
+\caption{\textbf{Difference}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist2.png}
+\caption{\textbf{Ratio}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist.png}
+\caption{\textbf{Difference}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
+\caption{\textbf{Ratio}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
+\caption{\textbf{Correlation}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
+\caption{\textbf{Correlation}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB2-HA_corr.png}
+\caption{\textbf{Correlation}}
+\label{}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_aaLD_MDS.png}
+\caption{\textbf{2-dimensional MDS embedding of segment based on mean amino acid linkage disequilibrium ($\chi\_{df}^{2}$).}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_aaLD_MatrixMDScorr.png}
+\caption{\textbf{Correlation between mean amino acid linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_ntLD_MDS.png}
+\caption{\textbf{2-dimensional MDS embedding of segment based on mean nucleotide linkage disequilibrium ($\chi\_{df}^{2}$)}}
+\label{}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_ntLD_MatrixMDScorr.png}
+\caption{\textbf{Correlation between mean nucleotide linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
+\label{}
 \end{figure}

### 2014-02-08 fluB.tex
Adding to conclusion and supplementary info.
Adding a more diverse conclusion, some text to supplementary information
and extra text to some figures.

@@
 Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis), and are thought to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir.
-Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2013}.
+Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2014}.
 
@@
 After assessing convergence of all 3 MCMC chains by visual inspection using Tracer \cite{tracer}, we combined samples across chains to give a total of 27,000 samples from the posterior distribution of trees.
-\tbc{The MCC tree for each segment should be shown in supplemental information.  I would imagine one figure with eight panels.}
+Maximum clade credibility trees of all 8 influenza B segments is shown in Figure \ref{genomeGrid}.
 
@@
 \subsection*{Phylogenetic analyses reveal maintenance of diversity and co-assortment of PB1, PB2 and HA gene segments}
-
-\tbc{Make this paragraph a broad description of the phylogenies across the segments...
-basically the current paragraph with a few tweaks.} The differentiation into Vic and Yam lineages can be seen in all segments \cite{chen2008} \tbc{also reference tree figure}.
+The differentiation into Vic and Yam lineages can be seen in all segments \cite{chen2008} (see Figure \ref{genomeGrid}).
 Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
 Some segment lineages associated with mixed-lineage genomes have become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment.
+Overall, we see many old bifurcations within the trees of all segments.
+Many of these correspond to reassortment events involving the PB1-PB2-HA segment complexes of Vic and Yam lineages, highlighting the role reassortment plays in the long-term evolution of influenza B viruses.
 
-Looking at patterns of ancestry across contemporaneous viruses for each segment, it is clear that the PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, while the PB1, PB2 and HA segments have maintained diversity up to the present (Figure \ref{tmrcaOT}).
+By quantifying patterns of ancestry across contemporaneous viruses for each segment, it is clear that the PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, while the PB1, PB2 and HA segments have maintained diversity up to the present (Figure \ref{tmrcaOT}).
 The PB1, PB2 and HA segments maintain a common ancestor in $\sim$1983 up to the present and thus accumulate genetic diversity since the split of those segments into Vic and Yam lineages, while other segments often lose diversity with ancestors to present day viruses appearing between $\sim$1991 and $\sim$1999.
@@
 \end{figure}
-
-%% AR - perhaps this section should go in the discussion as it a summary of the reconstructions?
-%\subsection*{Schematic plot of influenza B reassortment history}
 
@@
 	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
+Following the normalization procedure approximate SPR distances are similar across all pairwise comparisons.
+We interpret this as lack of evidence for small-scale topological similarities between trees of all segments, which we expect to arise if any two segments were being co-packaged.
 All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
@@
 	\caption{\textbf{Normalized mean $\Delta$TMRCA statistics between pairs of segments.}
+PB1, PB2 and HA trees exhibit reciprocally highly similar TMRCAs, unlike most other pairwise comparisons.
 All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
@@
 	\caption{\textbf{LD comparison between influenza B proteins.}
-Pairwise comparisons of linkage disequilibrium between amino acid sites on influenza B proteins.}
+Pairwise comparisons of linkage disequilibrium between amino acid sites on influenza B proteins.
+PB1, PB2 and HA proteins exhibit high mean amino acid LD between themselves.
+This is evidence of a considerable number of co-assorting alleles within these proteins.}
 	\label{segmentLD}
@@
 \subsection*{Conclusion}
-%The $\Delta$TMRCA method, which is an extension of patristic (or phylogenetic) distance methods, can be applied to cases where reticulate evolution is suspected and phylogenetic trees can readily be made.
-%Patristic distance methods are far from new, although their use in conjunction with temporally calibrated phylogenies has been limited.
-For example, one study focused on patristic (or phylogenetic) distances to predict viral titer in \textit{Drosophila} species infected by viruses of other \textit{Drosophila} species \cite{longdon2011}.
-Another study has used the equivalent statistic to our $\Delta$TMRCA (called $\Delta$\textit{t}) to assess the temporal incongruence in phylogenies of 171 amphibian species induced by using different calibrations \cite{ruane2011}.
-%By using relatively straightforward measures of diversity, such as TMRCA of all surviving branches over time and mean pairwise TMRCA between labels...
+A version of our lineage turnover metric has previously been used to investigate the tempo and strength of selection in influenza A viruses during seasonal circulation \cite{bedford2011}.
+Similarly our $\Delta$TMRCA statistic, an extension of patristic distance methods, has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \cite{longdon2011} and as \textit{$\Delta$t} to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \cite{ruane2011}.
 
-In this study we have investigated the patterns of reassortment amongst two influenza B virus segment lineages and found preserved diversity and consistent co-reassortment of 3 segments: PB1, PB2 and HA.
-Our findings reveal that Vic and Yam lineages of PB1, PB2 and HA segments co-associate amongst themselves with segments of their own lineage, thus resulting in an entirely Vic-lineage-derived PB1-PB2-HA segment complex and an entirely Yam-lineage-derived PB1-PB2-HA segment complex that co-circulate within the influenza B virus population, occasionally exchanging non-PB1-PB2-HA segments.
-Though we detect strains with mixed-lineage PB1-PB2-HA complexes, we find that they have performed poorly and have circulated for short amounts of evolutionary time when compared to viruses with `pure' lineage PB1-PB2-HA complexes.
+In this paper we have used this statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
+We believe that patristic distance methods, though far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies, including identifying clades or taxa that are products of reticulate evolution.
+To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
 
-We argue that this is evidence for a selectively maintained relationship between PB1-PB2-HA complex of Vic and Yam lineages and not due to biased reassortment.
+In this paper we apply a combination of relatively straightforward population genetics and phylogenetic methods to full genome sequences in order to describe and quantify reassortment patterns in influenza B viruses circulating in humans from 1980 to the present day.
+Our main finding is that in influenza B viruses only PB1, PB2 and HA segments maintain both Victoria and Yamagata lineages and that PB1-PB2-HA complexes of both lineages co-associate preferentially. 
+We argue that this is due to selection against viruses with mixed-lineage PB1-PB2-HA complexes.
 Given sufficient time we expect the two PB1-PB2-HA complexes to recruit the rest of the genomic segments into co-adapted and co-reassorting segment complexes eventually resulting in sympatric speciation of influenza B viruses.
 
-\tbc{This conclusion restates results too much.
-This is more of a beginning to a discussion than an end to it (you might even consider moving it there).
-Normally a conclusion is focused either on future work that this makes possible or public health implications.
-`Suggest experiments' kind of takes care of  the further work angle, and I'm not actually sure about public health implications.
-Or perhaps emphasize novelty of your phylogenetic approach to reassortment and say it could be applied to other systems.}
+\newpage
 
@@
 
+We summarised posterior sets of trees using TreeAnnotator (part of the BEAST package \cite{drummond2012}) by discarding the first 1000 trees of each analysis as burn-in and keeping median node heights.
+Figure \ref{genomeGrid} shows the maximum clade credibility (MCC) phylogenies of each segment.
+Each tree in Figure \ref{genomeGrid} corresponds to the summarised output from single analyses comprised of 9000 trees sampled from the posterior distribution of trees.
+
 \begin{figure}
@@
 \includegraphics[width=0.95\textwidth]  {supp_figures/InfB_genomeGrid.pdf}
-\caption{\textbf{MCC trees of 8 segments.}}
-\label{}
+\caption{\textbf{MCC trees of 8 segments.}
+Maximum clade credibility trees of all 8 genome segments of influenza B viruses isolated since 1980.
+Trees are coloured based on inferred PB1-PB2-HA lineage.
+Vertical bars indicate the original Victoria and Yamagata lineages within each segment.}
+\label{genomeGrid}
 \end{figure}
@@
 Figure \ref{SPRdistancesReplicates} confirms this by showing greater variability in tree topology (as greater approximate SPR distance) when MP and NS trees are compared against their replicates.
-
 For all other segments approximate SPR distances between all pairs of trees prior to normalization suggests that most segment pairs have similar posterior distributions of approximate SPR distances.
@@
 
+We also tested the accuracy of approximate SPR distances using exact SPR distances calculated from PB1, PB2 and HA comparisons.
+Figures \ref{SPR_PB1-PB2_difference}--\ref{SPR_PB2-HA_ratio} show that after normalization the approximate and exact SPR distance estimates are not significantly different.
+There is also good correlation between the two SPR distances (Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation}).
+
 \begin{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist.png}
-\caption{\textbf{Difference}}
-\label{}
+\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances after normalization.}}
+\label{SPR_PB1-PB2_difference}
 \end{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist2.png}
-\caption{\textbf{Ratio}}
-\label{}
+\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances after normalization.}}
+\label{SPR_PB1-PB2_ratio}
 \end{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist.png}
-\caption{\textbf{Difference}}
-\label{}
+\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances after normalization.}}
+\label{SPR_PB1-HA_difference}
 \end{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist2.png}
-\caption{\textbf{Ratio}}
-\label{}
+\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances after normalization.}}
+\label{SPR_PB1-HA_ratio}
 \end{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist.png}
-\caption{\textbf{Difference}}
-\label{}
+\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances after normalization.}}
+\label{SPR_PB2-HA_difference}
 \end{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
-\caption{\textbf{Ratio}}
-\label{}
+\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after normalization.}}
+\label{SPR_PB2-HA_ratio}
 \end{figure}
@@
 \includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
-\caption{\textbf{Correlation}}
-\label{}
+\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances after normalization.}}
+\label{SPR_PB1-PB2_correlation}
 \end{figure}
@@
 \includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
-\caption{\textbf{Correlation}}
-\label{}
+\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances after normalization.}}
+\label{SPR_PB1-HA_correlation}
 \end{figure}
@@
 \includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB2-HA_corr.png}
-\caption{\textbf{Correlation}}
-\label{}
+\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances after normalization.}}
+\label{SPR_PB2-HA_correlation}
 \end{figure}
@@
 
+Figures \ref{MDSaaLD} and \ref{MDSntLD} visualize the relationships between segments based on mean amino acid and nucleotide site linkage disequilibrium estimates, respectively.
+PB1, PB2 and HA segments appear to be by far the most similar segments, as indicated by their proximity in the MDS embedding.
+Figures \ref{MDSaaCorr} and \ref{MDSntCorr} compare the observed mean LD values between segments. 
+Based on mean amino acid site LD (Figure \ref{MDSaaCorr}) only PB1, PB2 and HA segments show considerable clustering, all other segments being equidistant from each other in the MDS embedding.
+MDS distance versus mean nucleotide site LD scatter plot, on the other hand, shows a continuum of relationships between segments, which is probably caused by more variability at the nucleotide level than the amino acid level.
+This difference in variability would build up linkage disequilibrium between nucleotide sites and would reduce it via convergent substitutions over relatively short periods of time.
+
 \begin{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_aaLD_MDS.png}
-\caption{\textbf{2-dimensional MDS embedding of segment based on mean amino acid linkage disequilibrium ($\chi\_{df}^{2}$).}}
-\label{}
+\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean amino acid linkage disequilibrium ($\chi\_{df}^{2}$).}}
+\label{MDSaaLD}
 \end{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_aaLD_MatrixMDScorr.png}
-\caption{\textbf{Correlation between mean amino acid linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
-\label{}
+\caption{\textbf{Correlation between mean amino acid site linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
+\label{MDSaaCorr}
 \end{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_ntLD_MDS.png}
-\caption{\textbf{2-dimensional MDS embedding of segment based on mean nucleotide linkage disequilibrium ($\chi\_{df}^{2}$)}}
-\label{}
+\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean nucleotide linkage disequilibrium ($\chi\_{df}^{2}$).}}
+\label{MDSntLD}
 \end{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_ntLD_MatrixMDScorr.png}
-\caption{\textbf{Correlation between mean nucleotide linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
-\label{}
+\caption{\textbf{Correlation between mean nucleotide site linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
+\label{MDSntCorr}
 \end{figure}
-
-% TB: FILL IN WITH REST OF SUPP FIGURES
 

### 2014-02-09 fluB.tex
Updating most figures, adding new supp ones.
Updating all figures with (hopefully) their latest versions. Changes are
either the M1-MP/NS1-NS issue or colour scheme updates. I didn't update
some, since I'm re-running some trees due to labeling issues.
Also adding a lot of supplementary figures relating to exact-approximate
SPR distances. Basically showing that approximate SPR distances are
reasonable overestimates of exact SPR distances.

@@
 
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_trees.png}
-\caption{\textbf{Reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
-\label{}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_replicates.png}
-\caption{\textbf{Reassortment rate (approximate number of SPR moves per total time in the tree) between replicate trees of each segment.}}
-\label{}
-\end{figure}
-
 We also tested the accuracy of approximate SPR distances using exact SPR distances calculated from PB1, PB2 and HA comparisons.
@@
 \centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist.png}
+\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances after normalization.}}
+\label{NormSPR_PB1-PB2_difference}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist2.png}
+\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances after normalization.}}
+\label{NormSPR_PB1-PB2_ratio}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist.png}
+\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances after normalization.}}
+\label{NormSPR_PB1-HA_difference}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist2.png}
+\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances after normalization.}}
+\label{NormSPR_PB1-HA_ratio}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_hist.png}
+\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances after normalization.}}
+\label{NormSPR_PB2-HA_difference}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
+\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after normalization.}}
+\label{NormSPR_PB2-HA_ratio}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_corr.png}
+\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances after normalization.}}
+\label{NormSPR_PB1-PB2_correlation}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_corr.png}
+\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances after normalization.}}
+\label{NormSPR_PB1-HA_correlation}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_corr.png}
+\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances after normalization.}}
+\label{NormSPR_PB2-HA_correlation}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering  
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances after normalization.}}
+\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances.}}
 \label{SPR_PB1-PB2_difference}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances after normalization.}}
+\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances.}}
 \label{SPR_PB1-PB2_ratio}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances after normalization.}}
+\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances.}}
 \label{SPR_PB1-HA_difference}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances after normalization.}}
+\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances.}}
 \label{SPR_PB1-HA_ratio}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances after normalization.}}
+\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances.}}
 \label{SPR_PB2-HA_difference}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after normalization.}}
+\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after.}}
 \label{SPR_PB2-HA_ratio}
@@
 \centering  
-\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances after normalization.}}
+\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
+\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances after.}}
 \label{SPR_PB1-PB2_correlation}
@@
 \centering  
-\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances after normalization.}}
+\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
+\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances after.}}
 \label{SPR_PB1-HA_correlation}
@@
 \centering  
-\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB2-HA_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances after normalization.}}
+\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB2-HA_corr.png}
+\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances.}}
 \label{SPR_PB2-HA_correlation}

### 2014-02-10 fluB.tex
Minor changes to text.

@@
 Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments, a process known as reassortment. 
-Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of circulating human strains and with non-human influenza A strains.
-Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis), and are thought to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir.
+Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains.
+Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis), and are widely considered to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir.
 Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2014}.
@@
 In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
-This has occured through repeated reassortments and subsequent fixation of reassortant viruses within the influenza B population.
+This has occured through repeated reassortments and subsequent fixation of reassortant genome constellations within the influenza B population.
 
@@
 	Label switches thus indicate reassortment events that combine lineages falling on different sides of the Vic/Yam bifurcation.
-	C) A reassortment graph depiction shows that tip number 6 is determined to be a reassortant.
+	C) A reassortment graph depiction shows that tip number 6 is determined to be a reassortant based on B).
 	}
@@
 The longest protein coding region of each segment was extracted and used for all further analyses.
-%% GD - the following bit is meant to explain why we use M1 and NS1 alignments to make trees, but refer to them as MP and NS respectively in the text and figures, but I also realized that those labels have been used very inconsistently in the figures.
-I'll correct this when I re-make the figures to be slightly better resolution and also fix some minor typeface issues I've noticed.
-"M1" and "NS1" will only be used in the LD figure, since that is relevant if measuring amino acid LD, whereas whenever trees of M1 or NS1 are used we can be reasonably sure that the history of the MP and NS segment is accurately captured by M1 and NS1 regions.
 We thus assume that homologous recombination has not taken place and that the evolutionary history of the whole segment can be inferred from the longest coding sequence in the segment.
@@
 The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \cite{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
-
-Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y.
+Each strain was thus assigned 8 lineages depending on the combination of lineages from which their genomes were derived, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y.
 
@@
 Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
-Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \cite{obrien2009}.
+Here, we modeled the substitution process using the HKY model of nucleotide substitution \cite{hky1985}, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \cite{obrien2009}.
 We used a flexible Bayesian skyride demographic model \cite{minin2008}.
@@
 We also calculated the total amount of evolutionary time spent by each segment with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments.
-We do this by summing the branch lengths in each tree under 3 different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
+We do this by summing the branch lengths in each tree under 3 (2 in the case of PB1, PB2 and HA trees) different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
 This gives a measure of how successful, over long periods of time, each particular PB1-PB2-HA constellation has been.
@@
 \end{equation}
-where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $g(A_{ij},B_{ij})$ is the difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree A and the \textit{i}th posterior sample of tree B.
-Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
+where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n |g(A_{ij}, B_{ij})|$ and $|g(A_{ij},B_{ij})|$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree A and the \textit{i}th posterior sample of tree B.
+Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree A (which we refer to as A'), to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
+We had 3 replicate analyses of each segment and in order to calculate $f(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as A and analyses numbered 2, 3 and 1 as A', in that order.
 
-These parameters are estimated for each pairing of trees at each MCMC state and provides us with 95\% highest posterior density intervals for those parameters, thus giving us the ability to test specific hypotheses regarding similarities between the trees of different segments.
+Calculating $\Delta(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this parameter allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
 Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
@@
 We calculated approximate SPR distances \cite{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
-Approximate SPR distances were normalized using the procedure described above.
+Approximate SPR distances were normalized using the procedure described above, where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from trees A, B and indepent analyses thereof (A' and B').
 
@@
 For quantifying LD we used the $\chi^{2}_{df}$ statistic \cite{zhao2005}, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
-LD was estimated only at loci where each nucleotide allele was represented by at least two isolates.
+LD was estimated only at loci where each nucleotide or amino acid allele was represented by at least two isolates.
 We ignored gaps in the alignment and did not consider them as polymorphisms.
@@
 Overall, we see many old bifurcations within the trees of all segments.
-Many of these correspond to reassortment events involving the PB1-PB2-HA segment complexes of Vic and Yam lineages, highlighting the role reassortment plays in the long-term evolution of influenza B viruses.
+Many of these correspond to reassortment events involving the PB1, PB2 and HA segments of Vic and Yam lineages, highlighting the role reassortment plays in the long-term evolution of influenza B viruses.
 
@@
 We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
-Increasing mean TMRCA of Vic and Yam lineage pairs for all 5 segments (Figure \ref{betweenDiversity}) indicate that every segment has reassorted with respect to the Victoria and Yamagata lineages of PB1, PB2 and HA segments.
+Losses of diversity (represented by more recent mean pairwise TMRCAs between Vic and Yam labels) in Figure \ref{betweenDiversity} in all 8 segment trees indicate that every segment has reassorted with respect to the Victoria and Yamagata lineages of PB1, PB2 and HA segments.
 However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
-This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments, i.e.\ PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees.
-In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1 and PB2 segments on one hand and HA on the other.
-These reductions in diversity represent small clades with PB1 and PB2 segments from one lineage and the HA segment from another, which go extinct by 1997.
-In addition, the assignment of Vic or Yam lineages of PB1, PB2 and HA segments to branches of other segment trees is very similar and often identical.
+This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments.
+After 1997 in PB1, PB2 and HA trees PB1, PB2 and HA traits only `meet' at the root.
+In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1, PB2 and HA segments.
+These reductions in diversity represent small clades with reassortant PB1-PB2-HA constellations, which go extinct by 1997.
+We also observe that the assignment of Vic or Yam lineages of PB1, PB2 and HA segments to branches of other segment trees is very similar and often identical.
 This results in PB1, PB2 and HA lineage labels switching between Vic and Yam simultaneously in all trees after 1997, suggesting co-reassortment of Vic and Yam lineages of PB1, PB2 and HA segments.
@@
 	\caption{\textbf{Ratio of lineages in the dataset.}
-The ratio of sequences that derive from the original Victoria clade (red) to sequences that drive from the Yamagata clade (blue) lineage sequences in each segment over time.
+The ratio of sequences that derive from the original Victoria clade (orange) to sequences that drive from the Yamagata clade (grey) lineage sequences in each segment over time.
 Yamagata lineage PA, NP, NA and MP segments and Victoria lineage NS segment eventually become fixed in the influenza B population.
@@
 We reconstructed reassortment events that were detected by using lineage labels.
-Figure \ref{railroadPlot} focuses only on reassortments that have occured after 1990.
+Figure \ref{railroadPlot} focuses only on inter-lineage reassortments that have occured after 1990.
 We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV) in Figure \ref{railroadPlot}.
 Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata lineage derived background.
-All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata lineage derived PB1-B2-HA.
+All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata lineage derived PB1-PB2-HA segments.
 Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
-For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to B/Iowa/03/2002-like isolates.
-A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has been derived from Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
+For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to the rise of viruses with B/Iowa/03/2002-like genome constellations.
+A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has originally been derived from the Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
 
@@
 	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
-Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored red and lineages that derive from the original Yamagata clade colored blue.
+Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored orange and lineages that derive from the original Yamagata clade colored grey.
 Inter-lineage reassortment events are indicated by lines entering a different genome.
@@
 \subsection*{Recent PB1-PB2-HA constellations are entirely derived from Victoria or Yamagata lineage segments}
-
-We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary since 1997 (Figure \ref{betweenDiversity}), form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
 Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
-Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).
+Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) since 1997 all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).
 
@@
 We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bangkok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
-An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
+An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating the larger dataset with only PB1, PB2 and HA sequences.
 
@@
 If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to see small-scale similarities between phylogenetic trees of those segments.
-In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the raw reassortment rate.
+In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the raw (\textit{i.e} both within and between lineages) reassortment rate.
 
@@
 In Figure \ref{SPRdistances} the 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, see Supplementary information) between all segments.
-We note that our SPR distance analysis might simply lack power.
-SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
-In addition, we use approximate SPR distances, rather than exact SPR distances.
-However, we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (see Supplementary information).
+We note, however, that our SPR distance analysis might simply lack power.
+Analyses between independent analyses of the same segments yield distances that are comparable to distances between different segments (see Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
+In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
+Thus we caution against over-interpreting Figure \ref{SPRdistances}.
+Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (see supplementary Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_correlation}).
 
@@
 PB1, PB2 and HA trees, on the other hand, exhibit normalized $\Delta$TMRCA values that are much higher.
-This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occassionally very similar to uncertainty in tip-to-tip TMRCAs within either of these segments.
+This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occassionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
 Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, i.e.\ there is isolation by distance.
@@
 In Figure \ref{segmentLD} their signature is greater mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
-In some cases this value exceeds the estimated LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA).
+In some cases this value exceeds the estimated mean LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA have higher mean LD than mean LD when PB1 is compared against itself).
 
@@
 \subsection*{Evidence of a co-adapted PB1-PB2-HA gene complex}
-In this paper we show that the PB1, PB2 and HA segments of influenza B viruses are the only ones that have continuously maintained separate Vic and Yam lineages, while other segments have had undergone fixation by either Vic or Yam (Figures \ref{tmrcaOT}--\ref{railroadPlot}).
+In this paper we show that the PB1, PB2 and HA segments of influenza B viruses are the only ones that have continuously maintained separate Vic and Yam lineages, while other segments have fixed either Vic or Yam lineages (Figures \ref{tmrcaOT}--\ref{railroadPlot}).
 Evidence suggests that this is a result of prolonged lack of reassortment between Vic and Yam lineages in PB1, PB2 and HA (Figure \ref{betweenDiversity}) which possess co-assorting sequences detectable as high linkage disequilibrium (Figure \ref{segmentLD}).
@@
 
-It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations which did not become fixed in the population and our estimates of the number of reassortments do not differ significantly between all segments.
-Our finding that influenza B virus segments do not significantly differ in reassortment rate (Figure \ref{SPRdistances}) are in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
+It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations which did not become fixed in the population and our estimates of the number of reassortments do not differ significantly between all segments (Figure \ref{SPRdistances}).
+The latter is in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
 In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we offer several alternatives.
@@
 It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments derived from avian influenza A viruses \cite{kawaoka1989}.
-However, there have also been reassortant influenza viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \cite{gregory2002} and H1N1/09 in 2009 \cite{smith2009}.
+However, there have also been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \cite{gregory2002} and H1N1/09 in 2009 \cite{smith2009}.
 
@@
 This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \cite{lee2008}.
-Similarly to the previous scenario, however, an association of some kind between PB1+2 and HA segments would be required to explain the low observation frequency and poor fitness of PB1-PB2/HA reassortants.
+Similarly to the previous scenario, however, an association of some kind between PB1+2 and HA segments would be required to explain the low observation frequency and poor fitness of PB1+2/HA reassortants.
 
@@
 This is the second scenario we might expect to occur, whereby relatively frequent reassortments occur between Vic and Yam lineages and are followed by selective sweeps.
-In this case influenza B viruses would undergo periodic genome homogenization with the exception of PB1, PB2 and HA segments, as strong antigenic dissimilarity between Vic HA and Yam HA would cause even strong selective sweeps to be unable to result in extinction of the opposing PB1-PB2-HA gene complex.
-However, this model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.
+In this case influenza B viruses would undergo periodic genome homogenization events with the exception of PB1, PB2 and HA segments.
+Because Vic HA and Yam HA are antigenically dissimilar we think that balancing selection would prevent even strong selective sweeps from driving the opposing PB1-PB2-HA gene complex to extinction.
+This model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.
 
@@
 We find it unlikely that either Victoria or Yamagata lineage PB1-PB2-HA complexes will go extinct stochastically in the near future, as they have co-circulated for prolonged periods at a ratio close to 0.5, suggesting the action of balancing selection (Figure \ref{lineageRatiosOverTime}).
-Depletion of susceptible individuals through influenza pandemics or vaccination seem unlikely as well.
-Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines do not produce lifelong immunity.
+Extinction through depletion of susceptible individuals, such as following influenza pandemics or mass vaccination seem unlikely as well.
+Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines do not produce lifelong immunity and are usually applied to specific subsets of the population.
 
 \subsection*{Suggested experiments}
+The association between Victoria and Yamagata lineage PB1-PB2-HA complexes should be relatively straightforward to test in the lab.
 Using previously developed plasmid systems \cite{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
@@
 The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
-However, our analyses do not indicate any obvious or dramatic differences between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two (see Supplementary information).
+However, our analyses do not indicate any obvious or dramatic differences between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two (see supplementary Figure \ref{robustCounting}).
 
@@
 
-In this paper we have used this statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
+For our purposes we have used this statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
 We believe that patristic distance methods, though far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies, including identifying clades or taxa that are products of reticulate evolution.
@@
 
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_robustCounting.png}
+\caption{\textbf{Synonymous, non-synonymous and nucleotide substitution rates in segments under different PB1-PB2-HA complexes.}}
+\label{robustCounting}
+\end{figure}
+
 \section*{Acknowledgements}

### 2014-02-10 fluB.tex
Some more minor changes to text.
Nothing major, just eyballing the manuscript in its entirety and
adjusting minor oversights and expanding some bits.

@@
 
-We express the similarity of two trees by comparing the means of the distribution of TMRCA differences between pairs of tips in the two trees.
+We express the similarity of two trees by comparing the means of the distribution of TMRCA differences (which we refer to as $\Delta$TMRCA) between pairs of tips in the two trees.
 In comparing TMRCAs of trees A and B we calculate
@@
 This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \cite{lee2008}.
-Similarly to the previous scenario, however, an association of some kind between PB1+2 and HA segments would be required to explain the low observation frequency and poor fitness of PB1+2/HA reassortants.
+It is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \cite{nerome1998,shaw2002}, offering a potential explanation for why the budding Victoria lineage segments were not homogenized via reassortment with Yamagata lineage segment in the early years of the split between the two lineages.
+The restriction to eastern Asia would presumably also give sufficient time for PB1, PB2 and HA segments to co-evolve together.
+Similarly to the previous scenario, however, mutation-driven co-evolution would require an association of some kind between PB1+2 and HA segments to explain the low observed frequency and poor fitness of PB1+2/HA reassortants.
 
@@
 Ideally, the co-adaptation would be easier to understand by referring to the structures of PB1 and PB2 proteins, as the link between these would be intuitive.
-We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins that fall within the regions that form contacts within the influenza B polymerase heterotrimer, suggesting more subtle roles for sites we have identified.
+We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins that fall within the regions that form contacts within the influenza B polymerase heterotrimer \cite{sugiyama2009}, suggesting more subtle roles for sites we have identified.
 
@@
 In this paper we apply a combination of relatively straightforward population genetics and phylogenetic methods to full genome sequences in order to describe and quantify reassortment patterns in influenza B viruses circulating in humans from 1980 to the present day.
-Our main finding is that in influenza B viruses only PB1, PB2 and HA segments maintain both Victoria and Yamagata lineages and that PB1-PB2-HA complexes of both lineages co-associate preferentially. 
+Our main finding is that in influenza B viruses only PB1, PB2 and HA segments maintain both Victoria and Yamagata lineages which associate with segments of their own lineage, yielding two co-circulating PB1-PB2-HA complexes: one entirely derived from Victoria and one from Yamagata lineage segments.
 We argue that this is due to selection against viruses with mixed-lineage PB1-PB2-HA complexes.
@@
 
+Figures \ref{SPR_PB1-PB2_difference}--\ref{SPR_PB2-HA_ratio} focus on differences between exact and approximate SPR distances and show that prior to normalization approximate SPR distances are overestimates of exact SPR distances.
+Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation} show that there is a similar degree of correspondence between approximate and exact SPR distances to when approximate and exact SPR distances are normalized (Figures \ref{NormSPR_PB1-PB2_correlation}--\ref{NormSPR_PB2-HA_correlation}) using the procedure described in Methods.
+
 \begin{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after.}}
+\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances.}}
 \label{SPR_PB2-HA_ratio}
@@
 \includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances after.}}
+\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances.}}
 \label{SPR_PB1-PB2_correlation}
@@
 \includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances after.}}
+\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances.}}
 \label{SPR_PB1-HA_correlation}
@@
 \clearpage
-
+\section*{Multidimensional scaling}
 Figures \ref{MDSaaLD} and \ref{MDSntLD} visualize the relationships between segments based on mean amino acid and nucleotide site linkage disequilibrium estimates, respectively.
@@
 
+\section*{Differences between Vic and Yam PB1-PB2-HA}
+Through the combination of inferred PB1-PB2-HA background and realized synonymous and non-synonymous substitution counts on branches \cite{obrien2009} it was possible to investigate whether Victoria and Yamagata lineage backgrounds differ in substitution rates.
+Figure \ref{robustCounting} shows the inferred synonymous, non-synonymous (both as substitutions per year) and overall nucleotide substitution rate (as substitutions per site per year) under different lineages of PB1-PB2-HA.
+
 \begin{figure}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_robustCounting.png}
-\caption{\textbf{Synonymous, non-synonymous and nucleotide substitution rates in segments under different PB1-PB2-HA complexes.}}
+\caption{\textbf{Synonymous, non-synonymous and nucleotide substitution rates in segments under different PB1-PB2-HA complexes.}
+Vertical bars indicating uncertainty are 95\% HPDs, black bars indicate 95\% HPDs that do not overlap.}
 \label{robustCounting}

### 2014-02-12 fluB.tex
Minor text changes.
Adding references to specific figures in supplementary info and extra
text under some figures.

@@
 \section*{Methods}
-We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in Supplementary information).
+We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information).
 The longest protein coding region of each segment was extracted and used for all further analyses.
@@
 We calculated approximate SPR distances \cite{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
-Approximate SPR distances were normalized using the procedure described above, where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from trees A, B and indepent analyses thereof (A' and B').
+Approximate SPR distances were normalized using the procedure described above, where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from trees A, B and independent analyses thereof (A' and B').
 
@@
 
-In Figure \ref{SPRdistances} the 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, see Supplementary information) between all segments.
+In Figure \ref{SPRdistances} the 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in both trees, see Figure \ref{NormSPR_RErate} in supplementary information) between all segments.
 We note, however, that our SPR distance analysis might simply lack power.
@@
 
-Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD (see Supplementary information).
+Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD (see Figures \ref{ChiGenome} and \ref{ChiCore} in supplementary information).
 These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
@@
 Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
-We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (see Supplementary information).
+We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (see Figures \ref{NormSPR_PB2-HA_difference}, \ref{NormSPR_PB2-HA_ratio}, \ref{SPR_PB2-HA_difference}, \ref{SPR_PB2-HA_ratio} and \ref{SPR_PB2-HA_correlation} in supplementary information).
 The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
@@
 \centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB1_PB2.png}
-\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and PB2 segments based on $\chi^{2}_{df}$ values.}}
-\label{ChiPB1PB2}
+\includegraphics[width=0.95\textwidth]  {supp_figures/aaChi_Genome.png}
+\caption{\textbf{Heatmap of genome-wide linkage disequilibrium ($\chi^{2}_{df}$) between polymorphic amino acid sites.}
+Patterns of LD across the genome suggest a network of reciprocally linked amino acid sites on PB1, PB2, HA and to some extent NA proteins.
+Proximity of sites on heatmaps might not correspond to proximity of sites within proteins.}
+\label{ChiGenome}
 \end{figure}
@@
 \centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB1_HA.png}
-\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and HA segments based on $\chi^{2}_{df}$ values.}}
-\label{ChiPB1HA}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB2_HA.png}
-\caption{\textbf{Hierarchical clustering of amino acid sites on PB2 and HA segments based on $\chi^{2}_{df}$ values.}}
-\label{ChiPB2HA}
+\includegraphics[width=0.85\textwidth]  {supp_figures/aaChi_PB1_PB2_HA.png}
+\caption{\textbf{Heatmap of linkage disequilibrium ($\chi^{2}_{df}$) between amino acid sites on PB1, PB2 and HA proteins.}
+Numbers next to each row and column correspond to amino acid site number within a given protein starting from methionine.}
+\label{ChiCore}
 \end{figure}
@@
 \caption{\textbf{Normalized reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
-\label{}
+\label{NormSPR_RErate}
 \end{figure}
@@
 There is also good correlation between the two SPR distances (Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation}). 
+Due to excessively long computation time of exact SPR distances very few PB2--HA comparisons were made.
 

### 2014-02-17 fluB.tex
Moving `InfB_LineageRatios.png` to supp.
Also using it in supp info.
Moving and altering text a bit.
Moving a few paragraphs on deltaTMRCA from Conclusion to Methods,
elaborating Conclusion to compensate.
Also some minor text changes.

@@
 Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments, a process known as reassortment. 
-Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains.
+Influenza A viruses are widely considered to be a major threat to human health worldwide due to their ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains.
 Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis), and are widely considered to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir.
@@
 These two lineages now possess antigenically distinct HA surface glycoproteins \cite{kanegae1990,rota1990,nerome1998,nakagawa2002,ansaldi2003} allowing them to co-circulate in the human population.
-Phylogenetic analysis of evolutionary rate, selective pressures and reassortment history of major clades of influenza B has shown extensive and often complicated patterns of reassortment between all segments of influenza B viruses between and within the Vic and Yam lineages \cite{chen2008}.
+Phylogenetic analysis of evolutionary rate, selective pressures and reassortment history of influenza B has shown extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within the Vic and Yam lineages \cite{chen2008}.
 
@@
 
-We show that despite extensive reassortment, three of the eight segments, PB1, PB2 and HA, still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
+We show that despite extensive reassortment, three of the eight segments -- two segments coding for components of the influenza B virus polymerase, PB1 and PB1, and the surface glycoprotein HA -- still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
 In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
@@
 We used a flexible Bayesian skyride demographic model \cite{minin2008}.
-We accounted for incomplete sampling dates for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
+We accounted for incomplete sampling dates for 94 sequences (of which 93 had only year and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
 
@@
 We inferred the diversity of each segment at a single point in time by estimating the date of the most recent common ancestor of all branches at yearly intervals, which places an upper bound on the maximum amount of diversity existing at each time period.
+A version of this lineage turnover metric has previously been used to investigate the tempo and strength of selection in influenza A viruses during seasonal circulation \cite{bedford2011}.
 In addition, we calculated mean pairwise time of most recent common ancestor (TMRCA) between branches labelled as Vic and Yam for PB1, PB2 and HA.
-This gave us a measure of how much a particular segment reassorts with respect to PB1, PB2 and HA segments.
+This gave us a measure of how much a particular segment reassorts with respect to Vic and Yam lineages of PB1, PB2 and HA segments.
 If Vic and Yam lineages of PB1, PB2 and HA segments were to be considered as being separate populations this measure would be equivalent to `between population' diversity.
@@
 Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
+Our $\Delta$TMRCA statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \cite{longdon2011} and as \textit{$\Delta$t} to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \cite{ruane2011}.
 
@@
 We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
-For quantifying LD we used the $\chi^{2}_{df}$ statistic \cite{zhao2005}, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
+To quantify LD we used the $\chi^{2}_{df}$ statistic \cite{zhao2005}, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
 LD was estimated only at loci where each nucleotide or amino acid allele was represented by at least two isolates.
@@
 We also calculated mean LD for all pairs of segments to quantify the overall association between them.
-Linkage disequilibrium was not estimated for the larger PB1, PB2 and HA dataset, as it would be impossible to put these measurements into a genome-wide perspective of linkage disequilibrium using just the data at hand.
+Linkage disequilibrium was not estimated for the larger PB1, PB2 and HA dataset, because it lacked sequences from the rest of the influenza B genome and thus would have failed to put LD between PB1, PB2 and HA proteins into a genome-wide perspective.
 
@@
 By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between lineage diversity, which indicate that an inter-lineage reassortment event has taken place (Figure \ref{betweenDiversity}).
-Because we are taking the mean of pairwise comparisons, this method gives a qualitative measure of panmixis between Victoria and Yamagata lineages in two trees.
+Because we are taking the mean of pairwise comparisons, this method gives a quantitative measure of diversity of panmixis between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree.
 We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
@@
 	\caption{\textbf{Ratio of lineages in the dataset.}
-The ratio of sequences that derive from the original Victoria clade (orange) to sequences that drive from the Yamagata clade (grey) lineage sequences in each segment over time.
+The ratio of sequences that derive from the original Victoria clade (orange) to sequences that derive from the Yamagata clade (grey) in each segment over time.
 Yamagata lineage PA, NP, NA and MP segments and Victoria lineage NS segment eventually become fixed in the influenza B population.
@@
 We note, however, that our SPR distance analysis might simply lack power.
-Analyses between independent analyses of the same segments yield distances that are comparable to distances between different segments (see Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
+Comparisons between independent analyses of the same segments yield distances that are comparable to distances between different segments (see Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
 In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
@@
 Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.
-In Figure \ref{segmentLD} their signature is greater mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
+In Figure \ref{segmentLD} the signature of Vic and Yam lineage-specific substitutions is greater mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
 In some cases this value exceeds the estimated mean LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA have higher mean LD than mean LD when PB1 is compared against itself).
@@
 Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
-This is easiest to interpret as frequency-dependent balancing selection acting to preserve antigenic diversity.
+This is easiest to interpret as frequency-dependent balancing selection acting to preserve antigenic diversity \cite{worobey2014}.
 If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
@@
 Similarly to the previous scenario, however, mutation-driven co-evolution would require an association of some kind between PB1+2 and HA segments to explain the low observed frequency and poor fitness of PB1+2/HA reassortants.
+In addition, we see no reason why these three particular segments, and not the rest of the genome, would become co-adapted to each other in geographic isolation.
 
@@
 
-If influenza B viruses are undergoing sympatric speciation, does co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or is co-infection so rare that speciation occurs via mutation-driven co-evolution?
+If influenza B viruses are undergoing sympatric speciation, it is imperative to determine whether co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or whether co-infection so rare that speciation occurs via mutation-driven co-evolution.
 We think that co-dependence between PB1 and PB2 segments can be explained by the fact that they are functionally linked: together with the PA protein they form part of the influenza B virus RNA-dependent RNA polymerase heterotrimer.
-If more segments get recruited to the PB1-PB2-HA segment complex, is it reasonable to assume that the next segment to be recruited will also code for a protein functionally linked to either the two polymerase proteins (PB1 and PB2) or haemagglutinin (HA)?
 
-Are the reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background entering a Victoria PB1-PB2-HA background via reassortment indicative of rare events followed by selective sweeps or stochastic fixation?
+It would be interesting to find out, from a speciation point of view, whether the next segments to be recruited into the PB1-PB2-HA segment complex will be functionally linked to the two polymerase proteins (PB1 and PB2) or haemagglutinin (HA).
+We already observe little mixing between the NA and MP segments and the PB1-PB2-HA complex in Figure \ref{betweenDiversity}, which is suggestive of recruitment to the complex.
+The NA protein has somewhat higher LD with PB1, PB2 and HA proteins in Figure \ref{segmentLD} and appears to have developed amino acid polymorphisms linked to lineage-specific sites on PB1, PB2 and HA proteins (Figure \ref{ChiGenome}).
+We cannot, however, dimiss the possibility that differentiation of the NA segment under the two PB1-PB2-HA complexes is caused by the rapid rate of nucleotide substitutions in the NA segment.
+Due to lack of polymorphic amino acid sites in the M1 protein it is difficult to determine whether the MP segment is linked to the NA segment or the PB1-PB2-HA segment complex to any appreciable degree.
+We do note that NA and MP segments exhibit a curious pattern of TMRCA similarities in Figure \ref{deltaTMRCA}.
+Together with Figure \ref{betweenDiversity} it suggests either that NA and MP segments are being recruited to PB1-PB2-HA segment complex together or that NA and MP form a second co-adapted and co-assorting segment complex.
+Recruitment to PB1-PB2-HA and formation of a second co-adapted segment complex are not mutually exclusive scenarios.
 
-It's not unfeasible for selective sweeps following reassortments to breakdown developing co-adaptation of segments, especially if they are not functionally linked and have themselves been reassorted into a new PB1-PB2-HA background recently.
+It also remains unknown whether reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background reassorting into a Victoria PB1-PB2-HA background indicative of rare events followed by selective sweeps or stochastic fixation.
+It's not unfeasible for selective sweeps following reassortments to break down developing co-adaptation of segments, especially if they are not functionally linked and have themselves been reassorted into a new PB1-PB2-HA background recently.
 This is the second scenario we might expect to occur, whereby relatively frequent reassortments occur between Vic and Yam lineages and are followed by selective sweeps.
@@
 Extinction through depletion of susceptible individuals, such as following influenza pandemics or mass vaccination seem unlikely as well.
-Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines do not produce lifelong immunity and are usually applied to specific subsets of the population.
+Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines, which are usually applied to specific subsets of the population, do not produce lifelong immunity.
 
@@
 The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
-However, our analyses do not indicate any obvious or dramatic differences between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two (see supplementary Figure \ref{robustCounting}).
+However, our analyses do not indicate any obvious differences in synonymous, non-synonymous or nucleotide substitution rates between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two (see supplementary Figure \ref{robustCounting}).
 
@@
 \subsection*{Conclusion}
-A version of our lineage turnover metric has previously been used to investigate the tempo and strength of selection in influenza A viruses during seasonal circulation \cite{bedford2011}.
-Similarly our $\Delta$TMRCA statistic, an extension of patristic distance methods, has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \cite{longdon2011} and as \textit{$\Delta$t} to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \cite{ruane2011}.
+We have used the $\Delta$TMRCA statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
+We believe that patristic distance methods, though far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies.
+One of many useful applications of the $\Delta$TMRCA method would be identifying clades or taxa that are products of reticulate evolution.
+Within-analysis $\Delta$TMRCA could be used as a cutoff to detect outlier taxa with greater than expected $\Delta$TMRCA values.
+To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
+In addition, by treating the relative position of each isolate within a phylogeny of one segment as a label and modelling it on phylogenies of other segments we have also developed a metric similar to `between population' diversity used in calculating $F_{ST}$, which is capable of quantifying reticulate evolution-induced loss of diversity between partitions of taxa in two or more phylogenies.
 
-For our purposes we have used this statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
-We believe that patristic distance methods, though far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies, including identifying clades or taxa that are products of reticulate evolution.
-To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
-
-In this paper we apply a combination of relatively straightforward population genetics and phylogenetic methods to full genome sequences in order to describe and quantify reassortment patterns in influenza B viruses circulating in humans from 1980 to the present day.
+In this paper we apply a novel combination of population genetics and phylogenetic methods to full genome sequences in order to describe and quantify reassortment patterns in influenza B viruses circulating in humans from 1980 to the present day.
 Our main finding is that in influenza B viruses only PB1, PB2 and HA segments maintain both Victoria and Yamagata lineages which associate with segments of their own lineage, yielding two co-circulating PB1-PB2-HA complexes: one entirely derived from Victoria and one from Yamagata lineage segments.
 We argue that this is due to selection against viruses with mixed-lineage PB1-PB2-HA complexes.
-Given sufficient time we expect the two PB1-PB2-HA complexes to recruit the rest of the genomic segments into co-adapted and co-reassorting segment complexes eventually resulting in sympatric speciation of influenza B viruses.
+We also suspect that NA-MP segments might be linked in a similar way and potentially become the next segments to be recruited by the Vic and Yam PB1-PB2-HA complexes.
+Given sufficient time we expect the two PB1-PB2-HA complexes to eventually recruit the rest of the genomic segments into co-adapted and co-reassorting segment complexes resulting in sympatric speciation of human influenza B viruses.
 
@@
 
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_lineageRatios.png}
+\caption{\textbf{Total counts and ratios of Victoria and Yamagata lineage sequences in each segment.}}
+\label{lineageRatios}
+\end{figure}
+
 \section*{Acknowledgements}

### 2014-02-19 correspondence/coverLetter.tex
Adding cover letter.
A few points on initial submission:

1. eLife asks for a single paragraph cover letter, current one has 4
(because it's mimicking Trevor's flux cover letter). Having said that,
no one at eLife seemed to mind Trevor's cover letter. Should we cut it
down to the first paragraph and let the abstract do the rest or keep it
as is?

2. eLife love love loves early-career scientists and I might be
worrying prematurely here, but I can't see any easy way of specifying
that I'm a PhD student (which is about as early as it gets) other than
putting my title in as PhD student Gytis Dudas

\documentclass[stdletter,letterpaper,addrfromright,orderfromdateto,dateleft,11pt,noaddrto,sigleft]{newlfm}
\topmarginskip{-0.4in}
\bottommarginskip{-1.5in}
\leftmarginsize{1in}
\rightmarginsize{1.25in}
\sigskipbefore{0.2in}
\sigskipafter{0in}
\noLines
\nolines
\noHeadline
\noheadline
\signature{Gytis Dudas}

\namefrom{}
\addrfrom{Inst.\ of Evolutionary Biology \\ University of Edinburgh \\ Ashworth Laboratories \\ King's Buildings \\ Edinburgh, UK}
\emailfrom{g.dudas@sms.ed.ac.uk}

\greetto{Dear Editor,}
\closeline{Sincerely,}

% comments
\usepackage{color}
\usepackage{ulem}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}}
\def\tbc#1{\textcolor{purple}{[#1]}}

\begin{document}

\begin{newlfm}
Please find our manuscript (attached) entitled ``Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex''. 
We would be grateful if you considered it for publication in \textit{eLife}.
It is our opinion that \textit{eLife} is a fitting journal for this work, as the manuscript provides a comprehensive view of the reassortment history of influenza B viruses since the 1980s and identifies 3 segments exhibiting dynamics analogous to what is considered a prelude to sympatric speciation in multicellular taxa.

Reassortment itself is of great importance in influenza virus evolution and epidemiology, as it is the only currently known process through which pandemic influenza viruses arise.
Influenza B viruses have been noted for their propensity for reassortment between two co-circulating subtypes, called Victoria and Yamagata, in humans, but little has been done to formalize the observed patterns of reassortment.
We apply tools traditionally used in population genetics to quantify gene flow between labeled phylogenies, detect co-assorting alleles on different segments and exploit the properties of time scaled phylogenies to calculate the temporal discordance between lineages being combined by reassortment.

Our main finding is that out of the 8 genomic segments of influenza B viruses, 3 segments exhibit limited reassortment with respect to each other.
Although each segment initially had a Victoria and a Yamagata lineage, we detect substantial homogenization through reassortment in the influenza B genome since 1990s, with one or the other lineage being fixed in 5 segments in the influenza B population.
The 3 segments maintaining both Victoria and Yamagata lineages are segments coding for components of the influenza B polymerase: PB1 and PB2 and the segment coding for the virus' main surface glycoprotein HA.

We believe that this finding is of great interest to evolutionary biologists, as it bears all the signs of sympatric speciation, namely co-circulation and the presence of differentiated regions of the genome resisting gene flow.
Additionally, current understanding of the molecular biology of influenza viruses does not offer any explanations for why PB1, PB2 and HA segments should be associated, implying a gap in our knowledge of these viruses.
We also think that the methods we have developed will be of benefit to future studies on reticulate evolution.

\end{newlfm}

\end{document}

### 2014-02-19 fluB.tex
A few edits/comments

@@
 Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages. 
-Frequent reassortment between the segments of these two lineages has been noted in the past, but the effects of these reassortments have not been investigated in much detail.
+Frequent exchange of genomic segments of these two lineages has been noted in the past, but the effects %% AR - effects or 
+of these reassortments have not been investigated in detail.
 We investigate inter-lineage reassortments by comparing phylogenetic trees across genomic segments.
@@
 \section*{Introduction}
-Seasonal influenza causes between 250,000 and 500,000 deaths annually and is comprised of three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \cite{flufactsheet}.
+Seasonal influenza causes between 250,000 and 500,000 deaths annually and comprises lineages from three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \cite{flufactsheet}.
 However, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
@@
 Influenza A viruses are widely considered to be a major threat to human health worldwide due to their ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains.
-Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis), and are widely considered to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir.
+Although influenza B viruses have been observed to infect seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis, they are thought to primarily infect humans and are thus are unlikely to exhibit pandemics due to the absence of an animal reservoir from which to acquire antigenic novelty.
 Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2014}.
+%%AR - some earlier references here
 
@@
 In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
-This has occured through repeated reassortments and subsequent fixation of reassortant genome constellations within the influenza B population.
+This has occurred through repeated reassortments and subsequent fixation of reassortant genome constellations within the influenza B population.
 

### 2014-02-19 fluB.tex
Edits

@@
 Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages. 
-Frequent exchange of genomic segments of these two lineages has been noted in the past, but the effects %% AR - effects or 
-of these reassortments have not been investigated in detail.
+Frequent exchange of genomic segments of these two lineages has been noted in the past, but the effects of these reassortments have not been investigated in detail.
 We investigate inter-lineage reassortments by comparing phylogenetic trees across genomic segments.

### 2014-02-19 correspondence/coverLetter.tex
Revisions.

@@
 \begin{newlfm}
-Please find our manuscript (attached) entitled ``Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex''. 
+Please find, attached, our manuscript entitled ``Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex''.
 We would be grateful if you considered it for publication in \textit{eLife}.
@@
 Reassortment itself is of great importance in influenza virus evolution and epidemiology, as it is the only currently known process through which pandemic influenza viruses arise.
-Influenza B viruses have been noted for their propensity for reassortment between two co-circulating subtypes, called Victoria and Yamagata, in humans, but little has been done to formalize the observed patterns of reassortment.
+Influenza B viruses have been noted for their propensity for reassortment between two co-circulating lineages, called Victoria and Yamagata, in humans, but little has been done to formalize the observed patterns of reassortment.
 We apply tools traditionally used in population genetics to quantify gene flow between labeled phylogenies, detect co-assorting alleles on different segments and exploit the properties of time scaled phylogenies to calculate the temporal discordance between lineages being combined by reassortment.
@@
 Our main finding is that out of the 8 genomic segments of influenza B viruses, 3 segments exhibit limited reassortment with respect to each other.
-Although each segment initially had a Victoria and a Yamagata lineage, we detect substantial homogenization through reassortment in the influenza B genome since 1990s, with one or the other lineage being fixed in 5 segments in the influenza B population.
+Although all segments initially had a Victoria and a Yamagata lineage, we detect substantial homogenization through reassortment in the influenza B genome since 1990s, with one or the other lineage being fixed in 5 segments in the influenza B population.
 The 3 segments maintaining both Victoria and Yamagata lineages are segments coding for components of the influenza B polymerase: PB1 and PB2 and the segment coding for the virus' main surface glycoprotein HA.

### 2014-02-19 fluB.tex
Revisions.

@@
 	B) We then transfer lineage labels from one tree to the same tips in another tree.
-	Label switches thus indicate reassortment events that combine lineages falling on different sides of the Vic/Yam bifurcation.
+	Transitions between labels along this second tree thus indicate reassortment events that combine lineages falling on different sides of the Vic/Yam bifurcation in the first tree.
 	C) A reassortment graph depiction shows that tip number 6 is determined to be a reassortant based on B).
@@
 The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \cite{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
-Each strain was thus assigned 8 lineages depending on the combination of lineages from which their genomes were derived, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y.
+Each strain was thus assigned 8 lineages depending on the combination of lineages from which their genomes were derived, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as (V,V,V,V,V,V,V,Y).
 
@@
 We accounted for incomplete sampling dates for 94 sequences (of which 93 had only year and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
-
 We ran 3 independent MCMC chains, each with 200 million states, sampled every 20,000 steps and discarded the first 10\% of the MCMC states as burn-in.
 After assessing convergence of all 3 MCMC chains by visual inspection using Tracer \cite{tracer}, we combined samples across chains to give a total of 27,000 samples from the posterior distribution of trees.
-Maximum clade credibility trees of all 8 influenza B segments is shown in Figure \ref{genomeGrid}.
 
@@
 We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete states using an asymmetric transition matrix \cite{lemey2009} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates.
-Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing state transitions, i.e.\ Yam to Vic or Vic to Yam (see Figure \ref{methodFig}).
+Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing state transitions, i.e.\ Yam to Vic or Vic to Yam (Figure \ref{methodFig}).
 In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
@@
 \subsection*{Tree to tree similarities}
-We subsampled our combined posterior distribution of trees to give a total of 2700 trees.
-Pairwise comparisons between each pair of segments were performed by comparing individual pairs of sampled trees.
 
-We express the similarity of two trees by comparing the means of the distribution of TMRCA differences (which we refer to as $\Delta$TMRCA) between pairs of tips in the two trees.
-In comparing TMRCAs of trees A and B we calculate
+\newcommand{\dtmrca}{d_\mathrm{TMRCA}} % TB: This could be replaced with \Delta_\mathrm{TMRCA} or \Delta as you prefer
+\newcommand{\dspr}{d_\mathrm{SPR}}
+
+We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ by comparing TMRCA differences between pairs of tips in the two trees.
+In comparing of trees of segments A and B we calculate
 \begin{equation}
-\Delta(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
+\dtmrca(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
 \end{equation}
-where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n |g(A_{ij}, B_{ij})|$ and $|g(A_{ij},B_{ij})|$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree A and the \textit{i}th posterior sample of tree B.
-Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree A (which we refer to as A'), to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
-We had 3 replicate analyses of each segment and in order to calculate $f(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as A and analyses numbered 2, 3 and 1 as A', in that order.
+where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $g(A_{ij},B_{ij})$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree $A$ and the \textit{i}th posterior sample of tree $B$.
+Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
+We had 3 replicate analyses of each segment and in order to calculate $f(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
+We subsampled our combined posterior distribution of trees to give a total of 2700 trees on which to analyze $\dtmrca$.
 
-Calculating $\Delta(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this parameter allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
+Calculating $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
 Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
-Our $\Delta$TMRCA statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \cite{longdon2011} and as \textit{$\Delta$t} to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \cite{ruane2011}.
+Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \cite{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \cite{ruane2011}.
 
@@
 We calculated approximate SPR distances \cite{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
-Approximate SPR distances were normalized using the procedure described above, where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from trees A, B and independent analyses thereof (A' and B').
+Approximate SPR distances $\dspr$ were normalized using the procedure described above, where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from segments $A$, $B$ and independent analyses thereof ($A'$ and $B'$).
 
@@
 We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
-To quantify LD we used the $\chi^{2}_{df}$ statistic \cite{zhao2005}, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
+To quantify LD we used the $\chi^{2}$ statistic, as it is equal to the widely used $r^2$ LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus \cite{zhao2005}.
 LD was estimated only at loci where each nucleotide or amino acid allele was represented by at least two isolates.
@@
 \subsection*{Phylogenetic analyses reveal maintenance of diversity and co-assortment of PB1, PB2 and HA gene segments}
-The differentiation into Vic and Yam lineages can be seen in all segments \cite{chen2008} (see Figure \ref{genomeGrid}).
+The differentiation into Vic and Yam lineages can be seen in all segments (Figure \ref{genomeGrid}).
 Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
-Some segment lineages associated with mixed-lineage genomes have become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment.
-Overall, we see many old bifurcations within the trees of all segments.
-Many of these correspond to reassortment events involving the PB1, PB2 and HA segments of Vic and Yam lineages, highlighting the role reassortment plays in the long-term evolution of influenza B viruses.
+On some segments, either the Victoria or Yamagata lineage has become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment, seen as modern viruses deriving completely from either Victoria or Yamagata (yellow vs purple bars in Figure \ref{genomeGrid}).
+This pattern is apparent in the PA, NP, NA, MP and NS segments.
+However, the PB1, PB2 and HA segments possess modern viruses deriving from both Victoria and Yamagata.
+Consistent with fixation of Victoria or Yamagata lineages, the PA, NP, NA, MP and NS segments periodically lose diversity, while maintenance of parallel Victoria and Yamagata lineages results in continually increasing diversity in segments PB1, PB2 and HA (Figure \ref{tmrcaOT}).
+The PB1, PB2 and HA segments maintain a common ancestor in $\sim$1983 in present-day viruses and thus accumulate genetic diversity since the split of those segments into Vic and Yam lineages, while other segments often lose diversity with ancestors to present-day viruses appearing between $\sim$1991 and $\sim$1999.
 
-By quantifying patterns of ancestry across contemporaneous viruses for each segment, it is clear that the PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, while the PB1, PB2 and HA segments have maintained diversity up to the present (Figure \ref{tmrcaOT}).
-The PB1, PB2 and HA segments maintain a common ancestor in $\sim$1983 up to the present and thus accumulate genetic diversity since the split of those segments into Vic and Yam lineages, while other segments often lose diversity with ancestors to present day viruses appearing between $\sim$1991 and $\sim$1999.
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.95\textwidth]{supp_figures/InfB_genomeGrid.pdf}
+\caption{\textbf{Maximum clade credibility (MCC) trees of all 8 genome segments of influenza B viruses isolated since 1980.}
+Trees are coloured based on inferred PB1-PB2-HA lineage.
+Vertical bars indicate the original Victoria and Yamagata lineages within each segment.
+Each tree corresponds to the summarised output from single analyses comprised of 9000 trees sampled from the posterior distribution of trees.
+}
+\label{genomeGrid}
+\end{figure}
 
@@
 We note, however, that our SPR distance analysis might simply lack power.
-Comparisons between independent analyses of the same segments yield distances that are comparable to distances between different segments (see Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
+Comparisons between independent analyses of the same segments yield distances that are comparable to distances between different segments (Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
 In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
 Thus we caution against over-interpreting Figure \ref{SPRdistances}.
-Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (see supplementary Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_correlation}).
+Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_correlation}).
 
@@
 
-Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD (see Figures \ref{ChiGenome} and \ref{ChiCore} in supplementary information).
+Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
 These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
@@
 Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
-We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (see Figures \ref{NormSPR_PB2-HA_difference}, \ref{NormSPR_PB2-HA_ratio}, \ref{SPR_PB2-HA_difference}, \ref{SPR_PB2-HA_ratio} and \ref{SPR_PB2-HA_correlation} in supplementary information).
+We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (Figures \ref{NormSPR_PB2-HA_difference}, \ref{NormSPR_PB2-HA_ratio}, \ref{SPR_PB2-HA_difference}, \ref{SPR_PB2-HA_ratio} and \ref{SPR_PB2-HA_correlation}).
 The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
-However, our analyses do not indicate any obvious differences in synonymous, non-synonymous or nucleotide substitution rates between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two (see supplementary Figure \ref{robustCounting}).
+However, our analyses do not indicate any obvious differences in synonymous, non-synonymous or nucleotide substitution rates between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two (Figure \ref{robustCounting}).
 
@@
 \renewcommand{\thetable}{S\arabic{table}}
-
-\section*{MCC trees}
-
-We summarised posterior sets of trees using TreeAnnotator (part of the BEAST package \cite{drummond2012}) by discarding the first 1000 trees of each analysis as burn-in and keeping median node heights.
-Figure \ref{genomeGrid} shows the maximum clade credibility (MCC) phylogenies of each segment.
-Each tree in Figure \ref{genomeGrid} corresponds to the summarised output from single analyses comprised of 9000 trees sampled from the posterior distribution of trees.
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.95\textwidth]  {supp_figures/InfB_genomeGrid.pdf}
-\caption{\textbf{MCC trees of 8 segments.}
-Maximum clade credibility trees of all 8 genome segments of influenza B viruses isolated since 1980.
-Trees are coloured based on inferred PB1-PB2-HA lineage.
-Vertical bars indicate the original Victoria and Yamagata lineages within each segment.}
-\label{genomeGrid}
-\end{figure}
 

### 2014-02-20 fluB.tex
Minor adjustments to text.
Cleaning up references to our LD statistics, adding some new references
and making minor alterations for better readability.
Also adding one of the LD papers that Trevor can't access.

@@
 Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages. 
-Frequent exchange of genomic segments of these two lineages has been noted in the past, but the effects of these reassortments have not been investigated in detail.
+Frequent exchange of genomic segments of these two lineages has been noted in the past, but the observed patterns of reassortment have not been formalized in detail.
 We investigate inter-lineage reassortments by comparing phylogenetic trees across genomic segments.
@@
 Although influenza B viruses have been observed to infect seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis, they are thought to primarily infect humans and are thus are unlikely to exhibit pandemics due to the absence of an animal reservoir from which to acquire antigenic novelty.
-Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2014}.
-%%AR - some earlier references here
+Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{burnet1955,hay2001,bedford2014}.
 
@@
 
-\newcommand{\dtmrca}{d_\mathrm{TMRCA}} % TB: This could be replaced with \Delta_\mathrm{TMRCA} or \Delta as you prefer
+\newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}} % TB: This could be replaced with \Delta_\mathrm{TMRCA} or \Delta as you prefer
 \newcommand{\dspr}{d_\mathrm{SPR}}
@@
 
+\newcommand{\chiSq}{\chi^{2}_{df}}
+
 \subsection*{Linkage disequilibrium across the influenza B genome}
 We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
-To quantify LD we used the $\chi^{2}$ statistic, as it is equal to the widely used $r^2$ LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus \cite{zhao2005}.
+To quantify LD we used the $\chiSq$ statistic \cite{hedrick1986}, as it is equal to the widely used $r^2$ LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus \cite{zhao2005}.
 LD was estimated only at loci where each nucleotide or amino acid allele was represented by at least two isolates.
@@
 Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
-On some segments, either the Victoria or Yamagata lineage has become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment, seen as modern viruses deriving completely from either Victoria or Yamagata (yellow vs purple bars in Figure \ref{genomeGrid}).
+On some segments, either the Victoria or Yamagata lineage has become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment, seen as modern viruses deriving completely from either Victoria or Yamagata lineage (yellow vs purple bars in Figure \ref{genomeGrid}).
 This pattern is apparent in the PA, NP, NA, MP and NS segments.
-However, the PB1, PB2 and HA segments possess modern viruses deriving from both Victoria and Yamagata.
+However, the PB1, PB2 and HA segments of modern viruses are derived from both Victoria and Yamagata lineages.
 Consistent with fixation of Victoria or Yamagata lineages, the PA, NP, NA, MP and NS segments periodically lose diversity, while maintenance of parallel Victoria and Yamagata lineages results in continually increasing diversity in segments PB1, PB2 and HA (Figure \ref{tmrcaOT}).
-The PB1, PB2 and HA segments maintain a common ancestor in $\sim$1983 in present-day viruses and thus accumulate genetic diversity since the split of those segments into Vic and Yam lineages, while other segments often lose diversity with ancestors to present-day viruses appearing between $\sim$1991 and $\sim$1999.
+The PB1, PB2 and HA segments from present-day viruses maintain a common ancestor in $\sim$1983 and thus accumulate genetic diversity since the split of those segments into Vic and Yam lineages, while other segments often lose diversity with ancestors to present-day viruses appearing between $\sim$1991 and $\sim$1999.
 
@@
 Figure \ref{railroadPlot} focuses only on inter-lineage reassortments that have occured after 1990.
-We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV) in Figure \ref{railroadPlot}.
+We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS with prime (') indicating independently acquired segments) circulating 1992 -- 2011: B/Alaska/12/1996-like (Y-Y-Y-Y-Y-Y-Y-V), B/Nanchang/2/1997-like (V-V-Y-V-Y-V-Y-V), B/Iowa/03/2002-like (V-V-Y'-V-Y-Y-Y'-V'), B/California/NHRC0001/2006-like (V-V-Y-V-Y'-Y-Y'-V') and B/Brisbane/33/2008-like (V-V-Y-V-Y'-Y-Y-V) in Figure \ref{railroadPlot}.
 Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata lineage derived background.
@@
 Although we do not find evidence of differences in numbers of raw reassortments between all pairs of trees, we find support for a reassortment `distance' effect.
-Figure \ref{deltaTMRCA} shows normalized mean $\Delta$TMRCA values for all pairs of trees.
+Figure \ref{deltaTMRCA} shows normalized mean $\dtmrca$ values for all pairs of trees.
 Most segment pairs show very low values for this statistic, indicating considerable differences between TMRCAs of the trees being compared.
-A large number of these comparisons have TMRCA differences up to 10 times larger ($\Delta$TMRCA$\approx$0.1) than TMRCA differences that occur between replicate BEAST analyses of one alignment over the course of the MCMC chain.
-PB1, PB2 and HA trees, on the other hand, exhibit normalized $\Delta$TMRCA values that are much higher.
+A large number of these comparisons have TMRCA differences up to 10 times larger ($\dtmrca$$\approx$0.1) than TMRCA differences that occur between replicate BEAST analyses of one alignment over the course of the MCMC chain.
+PB1, PB2 and HA trees, on the other hand, exhibit normalized $\dtmrca$ values that are much higher.
 This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occassionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
@@
 	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedMuDeltaTMRCA.png}
-	\caption{\textbf{Normalized mean $\Delta$TMRCA statistics between pairs of segments.}
+	\caption{\textbf{Normalized mean $\dtmrca$ statistics between pairs of segments.}
 PB1, PB2 and HA trees exhibit reciprocally highly similar TMRCAs, unlike most other pairwise comparisons.
@@
 Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.
-In Figure \ref{segmentLD} the signature of Vic and Yam lineage-specific substitutions is greater mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
+In Figure \ref{segmentLD} the signature of Vic and Yam lineage-specific substitutions is greater mean $\chiSq$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
 In some cases this value exceeds the estimated mean LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA have higher mean LD than mean LD when PB1 is compared against itself).
@@
 \subsection*{Conclusion}
-We have used the $\Delta$TMRCA statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
+We have used the $\dtmrca$ statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
 We believe that patristic distance methods, though far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies.
-One of many useful applications of the $\Delta$TMRCA method would be identifying clades or taxa that are products of reticulate evolution.
-Within-analysis $\Delta$TMRCA could be used as a cutoff to detect outlier taxa with greater than expected $\Delta$TMRCA values.
+One of many useful applications of the $\dtmrca$ method would be identifying clades or taxa that are products of reticulate evolution.
+Within-analysis $\dtmrca$ could be used as a cutoff to detect outlier taxa with greater than expected $\dtmrca$ values.
 To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
@@
 \includegraphics[width=0.95\textwidth]  {supp_figures/aaChi_Genome.png}
-\caption{\textbf{Heatmap of genome-wide linkage disequilibrium ($\chi^{2}_{df}$) between polymorphic amino acid sites.}
+\caption{\textbf{Heatmap of genome-wide linkage disequilibrium ($\chiSq$) between polymorphic amino acid sites.}
 Patterns of LD across the genome suggest a network of reciprocally linked amino acid sites on PB1, PB2, HA and to some extent NA proteins.
@@
 \includegraphics[width=0.85\textwidth]  {supp_figures/aaChi_PB1_PB2_HA.png}
-\caption{\textbf{Heatmap of linkage disequilibrium ($\chi^{2}_{df}$) between amino acid sites on PB1, PB2 and HA proteins.}
+\caption{\textbf{Heatmap of linkage disequilibrium ($\chiSq$) between amino acid sites on PB1, PB2 and HA proteins.}
 Numbers next to each row and column correspond to amino acid site number within a given protein starting from methionine.}
@@
 
-Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates} show $\Delta$TMRCA statistics for inter-segment and intra-segment, respectively, comparisons prior to normalization.
-Figure \ref{deltaTMRCAtrees} shows that similarity in TMRCAs (as low $\Delta$TMRCA) between PB1, PB2 and HA segments is due to considerably smaller $\Delta$TMRCA statistic for the comparison between these segments.
+Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates} show $\dtmrca$ statistics for inter-segment and intra-segment, respectively, comparisons prior to normalization.
+Figure \ref{deltaTMRCAtrees} shows that similarity in TMRCAs (as low $\dtmrca$) between PB1, PB2 and HA segments is due to considerably smaller $\dtmrca$ statistic for the comparison between these segments.
 It is also noteworthy that NA and MP comparisons have similarly low values for this statistic.
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
-\caption{\textbf{Mean $\Delta$TMRCA between all pairs of trees of segments.}}
+\caption{\textbf{Mean $\dtmrca$ between all pairs of trees of segments.}}
 \label{deltaTMRCAtrees}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
-\caption{\textbf{Mean $\Delta$TMRCA between replicate trees of each segment.}}
+\caption{\textbf{Mean $\dtmrca$ between replicate trees of each segment.}}
 \label{deltaTMRCAreplicates}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_aaLD_MDS.png}
-\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean amino acid linkage disequilibrium ($\chi\_{df}^{2}$).}}
+\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean amino acid linkage disequilibrium ($\chiSq$).}}
 \label{MDSaaLD}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_ntLD_MDS.png}
-\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean nucleotide linkage disequilibrium ($\chi\_{df}^{2}$).}}
+\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean nucleotide linkage disequilibrium ($\chiSq$).}}
 \label{MDSntLD}

### 2014-02-20 correspondence/coverLetter.tex
Adding a more favourable title.
Made the multi-paragraph covering letter into a single paragraph by removing line breaks (plus a few edits)

@@
 \noheadline
-\signature{Gytis Dudas}
+\signature{Gytis Dudas, Ph.D. candidate}
 
@@
 We would be grateful if you considered it for publication in \textit{eLife}.
-It is our opinion that \textit{eLife} is a fitting journal for this work, as the manuscript provides a comprehensive view of the reassortment history of influenza B viruses since the 1980s and identifies 3 segments exhibiting dynamics analogous to what is considered a prelude to sympatric speciation in multicellular taxa.
-
-Reassortment itself is of great importance in influenza virus evolution and epidemiology, as it is the only currently known process through which pandemic influenza viruses arise.
+The manuscript provides a comprehensive view of the reassortment history of influenza B viruses since the 1980s and identifies 3 segments exhibiting dynamics analogous to what is considered a prelude to sympatric speciation in multicellular taxa.
+Reassortment is of great importance in influenza virus evolution and epidemiology, as it is the only currently known process through which pandemic influenza viruses arise.
 Influenza B viruses have been noted for their propensity for reassortment between two co-circulating lineages, called Victoria and Yamagata, in humans, but little has been done to formalize the observed patterns of reassortment.
 We apply tools traditionally used in population genetics to quantify gene flow between labeled phylogenies, detect co-assorting alleles on different segments and exploit the properties of time scaled phylogenies to calculate the temporal discordance between lineages being combined by reassortment.
-
-Our main finding is that out of the 8 genomic segments of influenza B viruses, 3 segments exhibit limited reassortment with respect to each other.
+Our primary finding is that out of the 8 genomic segments of influenza B viruses, 3 segments exhibit very limited reassortment with respect to each other.
 Although all segments initially had a Victoria and a Yamagata lineage, we detect substantial homogenization through reassortment in the influenza B genome since 1990s, with one or the other lineage being fixed in 5 segments in the influenza B population.
 The 3 segments maintaining both Victoria and Yamagata lineages are segments coding for components of the influenza B polymerase: PB1 and PB2 and the segment coding for the virus' main surface glycoprotein HA.
-
 We believe that this finding is of great interest to evolutionary biologists, as it bears all the signs of sympatric speciation, namely co-circulation and the presence of differentiated regions of the genome resisting gene flow.

### 2014-02-21 fluB.tex
Move Methods to end, remove most supp figs.
Moved Methods to end of manuscript, commented out most supplemental
figures, the rest were put after Methods.
Also minor wording changes here and there, including one extra equation
for LD.

@@
 
+% symbols
+\newcommand{\chiSq}{\chi^{2}_{df}}
+\newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}}
+\newcommand{\dspr}{d_\mathrm{SPR}}
+
 %%% TITLE %%%
@@
 Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
-Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages. 
+Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata. 
 Frequent exchange of genomic segments of these two lineages has been noted in the past, but the observed patterns of reassortment have not been formalized in detail.
@@
 
-\section*{Methods}
-We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information).
-The longest protein coding region of each segment was extracted and used for all further analyses.
-We thus assume that homologous recombination has not taken place and that the evolutionary history of the whole segment can be inferred from the longest coding sequence in the segment.
-To date there has been little evidence of homologous recombination in influenza viruses \cite{chare2003,boni2008,han2010}.
-The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \cite{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
-Each strain was thus assigned 8 lineages depending on the combination of lineages from which their genomes were derived, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as (V,V,V,V,V,V,V,Y).
-
-We also downloaded all available sequences of influenza B isolates sampled 1984--2013 from GISAID for which PB1, PB2 and HA segments were sequenced.
-This comprised a dataset of 1433 isolates in total which became available only after the primary analyses were performed using the smaller dataset and had too many sequences to analyze using the methods described later.
-Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences, as before.
-Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.
-
-Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
-Here, we modeled the substitution process using the HKY model of nucleotide substitution \cite{hky1985}, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \cite{obrien2009}.
-We used a flexible Bayesian skyride demographic model \cite{minin2008}.
-We accounted for incomplete sampling dates for 94 sequences (of which 93 had only year and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
-We ran 3 independent MCMC chains, each with 200 million states, sampled every 20,000 steps and discarded the first 10\% of the MCMC states as burn-in.
-After assessing convergence of all 3 MCMC chains by visual inspection using Tracer \cite{tracer}, we combined samples across chains to give a total of 27,000 samples from the posterior distribution of trees.
-
-Every sequence was assigned 7 discrete traits in BEAUti corresponding to the lineages of all other segments with which a strain was isolated \textit{e.g.} PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values.
-We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete states using an asymmetric transition matrix \cite{lemey2009} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates.
-Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing state transitions, i.e.\ Yam to Vic or Vic to Yam (Figure \ref{methodFig}).
-In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
-
-\subsection*{Measures of diversity}
-We inferred the diversity of each segment at a single point in time by estimating the date of the most recent common ancestor of all branches at yearly intervals, which places an upper bound on the maximum amount of diversity existing at each time period.
-A version of this lineage turnover metric has previously been used to investigate the tempo and strength of selection in influenza A viruses during seasonal circulation \cite{bedford2011}.
-In addition, we calculated mean pairwise time of most recent common ancestor (TMRCA) between branches labelled as Vic and Yam for PB1, PB2 and HA.
-This gave us a measure of how much a particular segment reassorts with respect to Vic and Yam lineages of PB1, PB2 and HA segments.
-If Vic and Yam lineages of PB1, PB2 and HA segments were to be considered as being separate populations this measure would be equivalent to `between population' diversity.
-
-We also calculated the total amount of evolutionary time spent by each segment with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments.
-We do this by summing the branch lengths in each tree under 3 (2 in the case of PB1, PB2 and HA trees) different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
-This gives a measure of how successful, over long periods of time, each particular PB1-PB2-HA constellation has been.
-
-\subsection*{Tree to tree similarities}
-
-\newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}} % TB: This could be replaced with \Delta_\mathrm{TMRCA} or \Delta as you prefer
-\newcommand{\dspr}{d_\mathrm{SPR}}
-
-We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ by comparing TMRCA differences between pairs of tips in the two trees.
-In comparing of trees of segments A and B we calculate
-\begin{equation}
-\dtmrca(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
-\end{equation}
-where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $g(A_{ij},B_{ij})$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree $A$ and the \textit{i}th posterior sample of tree $B$.
-Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
-We had 3 replicate analyses of each segment and in order to calculate $f(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
-We subsampled our combined posterior distribution of trees to give a total of 2700 trees on which to analyze $\dtmrca$.
-
-Calculating $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
-Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
-Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \cite{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \cite{ruane2011}.
-
-Subtree prune and regraft (SPR) distances between phylogenetic trees are an approximate measure of the numbers of reassortment or recombination events.
-Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
-We calculated approximate SPR distances \cite{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
-Approximate SPR distances $\dspr$ were normalized using the procedure described above, where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from segments $A$, $B$ and independent analyses thereof ($A'$ and $B'$).
-
-\newcommand{\chiSq}{\chi^{2}_{df}}
-
-\subsection*{Linkage disequilibrium across the influenza B genome}
-We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
-To quantify LD we used the $\chiSq$ statistic \cite{hedrick1986}, as it is equal to the widely used $r^2$ LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus \cite{zhao2005}.
-LD was estimated only at loci where each nucleotide or amino acid allele was represented by at least two isolates.
-We ignored gaps in the alignment and did not consider them as polymorphisms.
-We also calculated mean LD for all pairs of segments to quantify the overall association between them.
-Linkage disequilibrium was not estimated for the larger PB1, PB2 and HA dataset, because it lacked sequences from the rest of the influenza B genome and thus would have failed to put LD between PB1, PB2 and HA proteins into a genome-wide perspective.
-
 \section*{Results}
@@
 However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
-This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments.
-After 1997 in PB1, PB2 and HA trees PB1, PB2 and HA traits only `meet' at the root.
-In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1, PB2 and HA segments.
+This suggests that after 1997 no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments and their lineage labels only `meet' at the root.
+We do see reduced diversity between Vic and Yam labels of PB1, PB2 and HA segments in a time period close to the initial split of Vic and Yam lineages (1986--1996).
 These reductions in diversity represent small clades with reassortant PB1-PB2-HA constellations, which go extinct by 1997.
@@
 	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
-Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored orange and lineages that derive from the original Yamagata clade colored grey.
+Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored yellow/brown and lineages that derive from the original Yamagata clade colored lilac/purple.
 Inter-lineage reassortment events are indicated by lines entering a different genome.
@@
 	\includegraphics[width=0.65\textwidth]{figures/InfB_stateTime.png}
-	\caption{\textbf{Amount of evolutionary time spent under different PB1-PB2-HA constellations.}
+	\caption{\textbf{Amount of evolutionary time each segment has spent under different PB1-PB2-HA constellations.}
 All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.
@@
 Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
-If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to see small-scale similarities between phylogenetic trees of those segments.
+If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
 In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the raw (\textit{i.e} both within and between lineages) reassortment rate.
@@
 Following the normalization procedure approximate SPR distances are similar across all pairwise comparisons.
-We interpret this as lack of evidence for small-scale topological similarities between trees of all segments, which we expect to arise if any two segments were being co-packaged.
+We interpret this as lack of evidence for small-scale topological similarities between trees of all segments, which we expect to arise if any two segments were being co-packaged and co-reassorted.
 All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
@@
 Thus we caution against over-interpreting Figure \ref{SPRdistances}.
-Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_correlation}).
+Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_ratio}).
 
@@
 This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occassionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
-Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, i.e.\ there is isolation by distance.
+Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, i.e.\ we observe isolation by distance.
 In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.
@@
 \subsection*{Evidence of a co-adapted PB1-PB2-HA gene complex}
-In this paper we show that the PB1, PB2 and HA segments of influenza B viruses are the only ones that have continuously maintained separate Vic and Yam lineages, while other segments have fixed either Vic or Yam lineages (Figures \ref{tmrcaOT}--\ref{railroadPlot}).
+In this paper we show that the PB1, PB2 and HA segments of influenza B viruses are the only ones that have continuously maintained separate Vic and Yam lineages, while other segments have fixed either Vic or Yam lineages (Figures \ref{genomeGrid}, \ref{lineageRatiosOverTime} and \ref{railroadPlot}).
 Evidence suggests that this is a result of prolonged lack of reassortment between Vic and Yam lineages in PB1, PB2 and HA (Figure \ref{betweenDiversity}) which possess co-assorting sequences detectable as high linkage disequilibrium (Figure \ref{segmentLD}).
 We propose that this pattern of non-reassortment is due to the action of selection and not simply biased reassortment.
-Figures \ref{SPRdistances} and \ref{deltaTMRCA} show that despite all pairs of segments having the same reassortment frequency, suggesting there is no bias in segment packaging, PB1, PB2 and HA segments exhibit much lower reassortment `distances'.
+We do not find evidence of bias in reassortment or segment packaging bias between PB1, PB2 and HA segments (Figure \ref{SPRdistances}), but evidence points to these segments having considerably lower reassortment `distances' (Figure \ref{deltaTMRCA}).
 
 Reassortment `distances' represent the temporal separation of reassorting branches and suggests that reassortments between PB1, PB2 and HA segments tend to combine branches that have relatively recently been part of the same genome, i.e.\ that have shared a recent common virion.
-In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were much more similar at the nucleotide and amino acid levels.
+In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were presumably much more similar at the nucleotide and amino acid levels.
 The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.
@@
 Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \cite{bergeron2010,fulvini2011}.
-Early influenza A vaccines used to be derived by infecting chicken eggs with an egg-adapted strain and a seasonal human isolate in the presence of antisera raised against the HA and NA proteins of the egg-adapted strain, thus selecting for reassortants which were egg-adapted, i.e.\ had internal segments derived from the egg-adapted strain, but possessed the HA and NA segments of the seasonal strain.
-However, in addition to producing reassortants with HA-NA derived from the seasonal strain, as intended, the second most frequent class of reassortants produced using this methods were those with PB1, HA and NA derived from seasonal strains \cite{bergeron2010,fulvini2011}.
-
+A previously used technique for producing vaccine seed strains involved selecting for HA-NA reassortants, which often resulted in PB1-HA-NA reassortants, the second most frequent class of reassortants \cite{bergeron2010,fulvini2011}.
 Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \cite{cobbin2013}.
@@
 Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
-A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot})).
+A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{genomeGrid}, \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot})).
 We find this scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
@@
 Because Vic HA and Yam HA are antigenically dissimilar we think that balancing selection would prevent even strong selective sweeps from driving the opposing PB1-PB2-HA gene complex to extinction.
-This model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.
+This `\textit{status quo}' model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.
 
@@
 Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
-We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (Figures \ref{NormSPR_PB2-HA_difference}, \ref{NormSPR_PB2-HA_ratio}, \ref{SPR_PB2-HA_difference}, \ref{SPR_PB2-HA_ratio} and \ref{SPR_PB2-HA_correlation}).
+We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (Figures \ref{NormSPR_PB2-HA_difference} and \ref{NormSPR_PB2-HA_ratio}).
 The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
@@
 Ideally, the co-adaptation would be easier to understand by referring to the structures of PB1 and PB2 proteins, as the link between these would be intuitive.
-We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins that fall within the regions that form contacts within the influenza B polymerase heterotrimer \cite{sugiyama2009}, suggesting more subtle roles for sites we have identified.
+We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins (Figure \ref{ChiCore}) that fall within the regions that form contacts within the influenza B polymerase heterotrimer \cite{sugiyama2009}, suggesting more subtle roles for sites we have identified.
 
@@
 
-\section*{Supporting information}
+\section*{Methods}
+We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information).
+The longest protein coding region of each segment was extracted and used for all further analyses.
+We thus assume that homologous recombination has not taken place and that the evolutionary history of the whole segment can be inferred from the longest coding sequence in the segment.
+To date there has been little evidence of homologous recombination in influenza viruses \cite{chare2003,boni2008,han2010}.
+The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \cite{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
+Each strain was thus assigned 8 lineages depending on the combination of lineages from which their genomes were derived, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as (V,V,V,V,V,V,V,Y).
+
+We also downloaded all available sequences of influenza B isolates sampled 1984--2013 from GISAID for which PB1, PB2 and HA segments were sequenced.
+This comprised a dataset of 1433 isolates in total which became available only after the primary analyses were performed using the smaller dataset and had too many sequences to analyze using the methods described later.
+Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences, as before.
+Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.
+
+Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
+Here, we modeled the substitution process using the HKY model of nucleotide substitution \cite{hky1985}, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \cite{obrien2009}.
+We used a flexible Bayesian skyride demographic model \cite{minin2008}.
+We accounted for incomplete sampling dates for 94 sequences (of which 93 had only year and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
+We ran 3 independent MCMC chains, each with 200 million states, sampled every 20,000 steps and discarded the first 10\% of the MCMC states as burn-in.
+After assessing convergence of all 3 MCMC chains by visual inspection using Tracer \cite{tracer}, we combined samples across chains to give a total of 27,000 samples from the posterior distribution of trees.
+
+Every sequence was assigned 7 discrete traits in BEAUti corresponding to the lineages of all other segments with which a strain was isolated \textit{e.g.} PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values.
+We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete states using an asymmetric transition matrix \cite{lemey2009} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates.
+Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing state transitions, i.e.\ Yam to Vic or Vic to Yam (Figure \ref{methodFig}).
+In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
+
+\subsection*{Measures of diversity}
+We inferred the diversity of each segment at a single point in time by estimating the date of the most recent common ancestor of all branches at yearly intervals, which places an upper bound on the maximum amount of diversity existing at each time point.
+A version of this lineage turnover metric has previously been used to investigate the tempo and strength of selection in influenza A viruses during seasonal circulation \cite{bedford2011}.
+In addition, we calculated mean pairwise time of most recent common ancestor (TMRCA) between branches labelled as Vic and Yam for PB1, PB2 and HA.
+This gave us a measure of how much a particular segment reassorts with respect to Vic and Yam lineages of PB1, PB2 and HA segments.
+If Vic and Yam lineages of PB1, PB2 and HA segments were to be considered as being separate populations this measure would be equivalent to `between population' diversity.
+
+We also calculated the total amount of evolutionary time spent by each segment with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments.
+We do this by summing the branch lengths in each tree under 3 (2 in the case of PB1, PB2 and HA trees) different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
+This gives a measure of how successful, over long periods of time, each particular PB1-PB2-HA constellation has been.
+
+\subsection*{Tree to tree similarities}
+
+We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ by comparing TMRCA differences between pairs of tips in the two trees.
+In comparing of trees of segments A and B we calculate
+\begin{equation}
+\dtmrca(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
+\end{equation}
+where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $g(A_{ij},B_{ij})$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree $A$ and the \textit{i}th posterior sample of tree $B$.
+Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
+We had 3 replicate analyses of each segment and in order to calculate $f(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
+We subsampled our combined posterior distribution of trees to give a total of 2700 trees on which to analyze $\dtmrca$.
+
+Calculating $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
+Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
+Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \cite{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \cite{ruane2011}.
+
+Subtree prune and regraft (SPR) distances between phylogenetic trees are an approximate measure of the numbers of reassortment or recombination events.
+Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
+We calculated approximate SPR distances \cite{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
+Approximate SPR distances $\dspr$ were normalized using the procedure described above, where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from segments $A$, $B$ and independent analyses thereof ($A'$ and $B'$).
+
+\subsection*{Linkage disequilibrium across the influenza B genome}
+We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
+To quantify LD we adapt the $\chiSq$ statistic from \cite{hedrick1986}:
+
+\begin{equation}
+\chiSq=\frac{\chi^{2}}{N\,(k-1)\,(m-1)},
+\end{equation}
+
+where $\chi^{2}$ is calculated from a classical contingency table, $N$ is the number of haplotypes and $(k-1)(m-1)$ are the degrees of freedom.
+This statistic is equal to the widely used $r^2$ LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus \cite{zhao2005}.
+LD was estimated only at loci where each nucleotide or amino acid allele was present in at least two isolates.
+We ignored gaps in the alignment and did not consider them as polymorphisms.
+We also calculated mean LD for all pairs of segments to quantify the overall association between them.
+Linkage disequilibrium was not estimated for the larger PB1, PB2 and HA dataset, because it lacked sequences from the rest of the influenza B genome and thus would have failed to put LD between PB1, PB2 and HA proteins into a genome-wide perspective.
+
 \setcounter{figure}{0}
@@
 
-\section*{Linkage disequilibrium data}
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_normRErate.png}
+\caption{\textbf{Normalized reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
+\label{NormSPR_RErate}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_trees.png}
+\caption{\textbf{Approximate SPR distances between all pairs of trees of segments.}
+There is a visible trend where comparisons between shorter segments have larger SPR distances, consistent with decreasing tree topology stability over the course of MCMC for shorter segments.}
+\label{SPRdistancesTrees}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_replicates.png}
+\caption{\textbf{Approximate SPR distances between replicate trees of each segment.}
+Approximate SPR distances between replicates of MP and NS trees are much higher ($\approx$260) than any other segments, suggesting greater variability in tree topology over the course of MCMC.
+SPR distances between replicates of most other segments are $\approx$200.
+}
+\label{SPRdistancesReplicates}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist.png}
+\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances after normalization.}
+95\% HPD interval (blue) overlaps zero, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
+\label{NormSPR_PB1-PB2_difference}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist2.png}
+\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances after normalization.}
+95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
+\label{NormSPR_PB1-PB2_ratio}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist.png}
+\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances after normalization.}
+95\% HPD interval (blue) overlaps zero, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
+\label{NormSPR_PB1-HA_difference}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist2.png}
+\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances after normalization.}
+95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
+\label{NormSPR_PB1-HA_ratio}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_hist.png}
+\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances after normalization.}
+95\% HPD interval (blue) overlaps zero, suggesting no evidence of differences between approximate and exact SPR distances following normalization.
+Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
+\label{NormSPR_PB2-HA_difference}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_hist2.png}
+\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after normalization.}
+95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.
+Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
+\label{NormSPR_PB2-HA_ratio}
+\end{figure}
 
@@
 \caption{\textbf{Heatmap of linkage disequilibrium ($\chiSq$) between amino acid sites on PB1, PB2 and HA proteins.}
-Numbers next to each row and column correspond to amino acid site number within a given protein starting from methionine.}
-\label{ChiCore}
-\end{figure}
-
-We have identified amino acid sites which exhibit a high degree of linkage disequilibrium between PB1, PB2 and HA segments.
-Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three proteins at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.
-
-None of the sites we have identified on PB1 and PB2 proteins fall within the previously described regions that form the contacts between the two polymerase subunits \cite{sugiyama2009}.
-
+Numbers next to each row and column correspond to amino acid site number within a given protein starting from methionine.
 Sites 211, 213 and 214 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
@@
 Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \cite{fodor2004}.
-Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \cite{guilligay2008}.
-Again, we cannot determine whether this finding is significant or not.
-
-\section*{Tree-to-tree comparisons}
-
-Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates} are showing approximate SPR distances prior to the normalization procedure (see Methods).
-Figure \ref{SPRdistancesTrees} shows that MP and NS trees show greater approximate SPR distances when compared to any other segment tree, which is consistent with the fact that only M1 and NS1 coding sequences, which are shorter than any other coding sequences, were used to infer the trees.
-Figure \ref{SPRdistancesReplicates} confirms this by showing greater variability in tree topology (as greater approximate SPR distance) when MP and NS trees are compared against their replicates.
-For all other segments approximate SPR distances between all pairs of trees prior to normalization suggests that most segment pairs have similar posterior distributions of approximate SPR distances.
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_trees.png}
-\caption{\textbf{Approximate SPR distances between all pairs of trees of segments.}}
-\label{SPRdistancesTrees}
+Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \cite{guilligay2008}.}
+\label{ChiCore}
 \end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_replicates.png}
-\caption{\textbf{Approximate SPR distances between replicate trees of each segment.}}
-\label{SPRdistancesReplicates}
-\end{figure}
-
-Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates} show $\dtmrca$ statistics for inter-segment and intra-segment, respectively, comparisons prior to normalization.
-Figure \ref{deltaTMRCAtrees} shows that similarity in TMRCAs (as low $\dtmrca$) between PB1, PB2 and HA segments is due to considerably smaller $\dtmrca$ statistic for the comparison between these segments.
-It is also noteworthy that NA and MP comparisons have similarly low values for this statistic.
-Figure \ref{deltaTMRCAreplicates} shows that TMRCAs are quite robust over the course of MCMC and node times do not, even in the most extreme cases, deviate from their replicates by more than 1 year.
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
-\caption{\textbf{Mean $\dtmrca$ between all pairs of trees of segments.}}
-\label{deltaTMRCAtrees}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
-\caption{\textbf{Mean $\dtmrca$ between replicate trees of each segment.}}
-\label{deltaTMRCAreplicates}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_normRErate.png}
-\caption{\textbf{Normalized reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
-\label{NormSPR_RErate}
-\end{figure}
-
-We also tested the accuracy of approximate SPR distances using exact SPR distances calculated from PB1, PB2 and HA comparisons.
-Figures \ref{SPR_PB1-PB2_difference}--\ref{SPR_PB2-HA_ratio} show that after normalization the approximate and exact SPR distance estimates are not significantly different.
-There is also good correlation between the two SPR distances (Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation}). 
-Due to excessively long computation time of exact SPR distances very few PB2--HA comparisons were made.
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances after normalization.}}
-\label{NormSPR_PB1-PB2_difference}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances after normalization.}}
-\label{NormSPR_PB1-PB2_ratio}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances after normalization.}}
-\label{NormSPR_PB1-HA_difference}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances after normalization.}}
-\label{NormSPR_PB1-HA_ratio}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances after normalization.}}
-\label{NormSPR_PB2-HA_difference}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after normalization.}}
-\label{NormSPR_PB2-HA_ratio}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances after normalization.}}
-\label{NormSPR_PB1-PB2_correlation}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances after normalization.}}
-\label{NormSPR_PB1-HA_correlation}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances after normalization.}}
-\label{NormSPR_PB2-HA_correlation}
-\end{figure}
-
-\clearpage
-
-Figures \ref{SPR_PB1-PB2_difference}--\ref{SPR_PB2-HA_ratio} focus on differences between exact and approximate SPR distances and show that prior to normalization approximate SPR distances are overestimates of exact SPR distances.
-Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation} show that there is a similar degree of correspondence between approximate and exact SPR distances to when approximate and exact SPR distances are normalized (Figures \ref{NormSPR_PB1-PB2_correlation}--\ref{NormSPR_PB2-HA_correlation}) using the procedure described in Methods.
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances.}}
-\label{SPR_PB1-PB2_difference}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances.}}
-\label{SPR_PB1-PB2_ratio}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances.}}
-\label{SPR_PB1-HA_difference}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances.}}
-\label{SPR_PB1-HA_ratio}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances.}}
-\label{SPR_PB2-HA_difference}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances.}}
-\label{SPR_PB2-HA_ratio}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances.}}
-\label{SPR_PB1-PB2_correlation}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances.}}
-\label{SPR_PB1-HA_correlation}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB2-HA_corr.png}
-\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances.}}
-\label{SPR_PB2-HA_correlation}
-\end{figure}
-
-\clearpage
-\section*{Multidimensional scaling}
-Figures \ref{MDSaaLD} and \ref{MDSntLD} visualize the relationships between segments based on mean amino acid and nucleotide site linkage disequilibrium estimates, respectively.
-PB1, PB2 and HA segments appear to be by far the most similar segments, as indicated by their proximity in the MDS embedding.
-Figures \ref{MDSaaCorr} and \ref{MDSntCorr} compare the observed mean LD values between segments. 
-Based on mean amino acid site LD (Figure \ref{MDSaaCorr}) only PB1, PB2 and HA segments show considerable clustering, all other segments being equidistant from each other in the MDS embedding.
-MDS distance versus mean nucleotide site LD scatter plot, on the other hand, shows a continuum of relationships between segments, which is probably caused by more variability at the nucleotide level than the amino acid level.
-This difference in variability would build up linkage disequilibrium between nucleotide sites and would reduce it via convergent substitutions over relatively short periods of time.
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_aaLD_MDS.png}
-\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean amino acid linkage disequilibrium ($\chiSq$).}}
-\label{MDSaaLD}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_aaLD_MatrixMDScorr.png}
-\caption{\textbf{Correlation between mean amino acid site linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
-\label{MDSaaCorr}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_ntLD_MDS.png}
-\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean nucleotide linkage disequilibrium ($\chiSq$).}}
-\label{MDSntLD}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_ntLD_MatrixMDScorr.png}
-\caption{\textbf{Correlation between mean nucleotide site linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
-\label{MDSntCorr}
-\end{figure}
-
-\section*{Differences between Vic and Yam PB1-PB2-HA}
-Through the combination of inferred PB1-PB2-HA background and realized synonymous and non-synonymous substitution counts on branches \cite{obrien2009} it was possible to investigate whether Victoria and Yamagata lineage backgrounds differ in substitution rates.
-Figure \ref{robustCounting} shows the inferred synonymous, non-synonymous (both as substitutions per year) and overall nucleotide substitution rate (as substitutions per site per year) under different lineages of PB1-PB2-HA.
 
@@
 \caption{\textbf{Synonymous, non-synonymous and nucleotide substitution rates in segments under different PB1-PB2-HA complexes.}
+Evolutionary rate dissimilarities under Vic and Yam PB1-PB2-HA complexes are not systematic and appear negligible.
+Synonymous and non-synonymous rates were calculated by dividing the sum of all substitutions of a given class by the total amount of evolutionary time under each PB1-PB2-HA constellation.
+Nucleotide rates were calculated by multiplying the inferred nucleotide substitution rate on each branch by the branch length, then dividing this by the total amount of evolutionary time under each PB1-PB2-HA constellation.
 Vertical bars indicating uncertainty are 95\% HPDs, black bars indicate 95\% HPDs that do not overlap.}
@@
 
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_lineageRatios.png}
-\caption{\textbf{Total counts and ratios of Victoria and Yamagata lineage sequences in each segment.}}
-\label{lineageRatios}
-\end{figure}
+%Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates} are showing approximate SPR distances prior to the normalization procedure (see Methods).
+%Figure \ref{SPRdistancesTrees} shows that MP and NS trees show greater approximate SPR distances when compared to any other segment tree, which is consistent with the fact that only M1 and NS1 coding sequences, which are shorter than any %other coding sequences, were used to infer the trees.
+%Figure \ref{SPRdistancesReplicates} confirms this by showing greater variability in tree topology (as greater approximate SPR distance) when MP and NS trees are compared against their replicates.
+%For all other segments approximate SPR distances between all pairs of trees prior to normalization suggests that most segment pairs have similar posterior distributions of approximate SPR distances.
+
+%Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates} show $\dtmrca$ statistics for inter-segment and intra-segment, respectively, comparisons prior to normalization.
+%Figure \ref{deltaTMRCAtrees} shows that similarity in TMRCAs (as low $\dtmrca$) between PB1, PB2 and HA segments is due to considerably smaller $\dtmrca$ statistic for the comparison between these segments.
+%It is also noteworthy that NA and MP comparisons have similarly low values for this statistic.
+%Figure \ref{deltaTMRCAreplicates} shows that TMRCAs are quite robust over the course of MCMC and node times do not, even in the most extreme cases, deviate from their replicates by more than 1 year.
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
+%\caption{\textbf{Mean $\dtmrca$ between all pairs of trees of segments.}}
+%\label{deltaTMRCAtrees}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
+%\caption{\textbf{Mean $\dtmrca$ between replicate trees of each segment.}}
+%\label{deltaTMRCAreplicates}
+%\end{figure}
+
+%We also tested the accuracy of approximate SPR distances using exact SPR distances calculated from PB1, PB2 and HA comparisons.
+%Figures \ref{SPR_PB1-PB2_difference}--\ref{SPR_PB2-HA_ratio} show that after normalization the approximate and exact SPR distance estimates are not significantly different.
+%There is also good correlation between the two SPR distances (Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation}). 
+%Due to excessively long computation time of exact SPR distances very few PB2--HA comparisons were made.
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances after normalization.}}
+%\label{NormSPR_PB1-PB2_correlation}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances after normalization.}}
+%\label{NormSPR_PB1-HA_correlation}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances after normalization.}
+%Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
+%\label{NormSPR_PB2-HA_correlation}
+%\end{figure}
+
+\clearpage
+
+%Figures \ref{SPR_PB1-PB2_difference}--\ref{SPR_PB2-HA_ratio} focus on differences between exact and approximate SPR distances and show that prior to normalization approximate SPR distances are overestimates of exact SPR distances.
+%Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation} show that there is a similar degree of correspondence between approximate and exact SPR distances to when approximate and exact SPR distances are normalized (Figures %\ref{NormSPR_PB1-PB2_correlation}--\ref{NormSPR_PB2-HA_correlation}) using the procedure described in Methods.
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist.png}
+%\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances.}}
+%\label{SPR_PB1-PB2_difference}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances.}}
+%\label{SPR_PB1-PB2_ratio}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist.png}
+%\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances.}}
+%\label{SPR_PB1-HA_difference}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances.}}
+%\label{SPR_PB1-HA_ratio}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist.png}
+%\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances.}}
+%\label{SPR_PB2-HA_difference}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances.}}
+%\label{SPR_PB2-HA_ratio}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances.}}
+%\label{SPR_PB1-PB2_correlation}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances.}}
+%\label{SPR_PB1-HA_correlation}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB2-HA_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances.}}
+%\label{SPR_PB2-HA_correlation}
+%\end{figure}
+
+%\clearpage
+%\section*{Multidimensional scaling}
+%Figures \ref{MDSaaLD} and \ref{MDSntLD} visualize the relationships between segments based on mean amino acid and nucleotide site linkage disequilibrium estimates, respectively.
+%PB1, PB2 and HA segments appear to be by far the most similar segments, as indicated by their proximity in the MDS embedding.
+%Figures \ref{MDSaaCorr} and \ref{MDSntCorr} compare the observed mean LD values between segments. 
+%Based on mean amino acid site LD (Figure \ref{MDSaaCorr}) only PB1, PB2 and HA segments show considerable clustering, all other segments being equidistant from each other in the MDS embedding.
+%MDS distance versus mean nucleotide site LD scatter plot, on the other hand, shows a continuum of relationships between segments, which is probably caused by more variability at the nucleotide level than the amino acid level.
+%This difference in variability would build up linkage disequilibrium between nucleotide sites and would reduce it via convergent substitutions over relatively short periods of time.
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_aaLD_MDS.png}
+%\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean amino acid linkage disequilibrium ($\chiSq$).}}
+%\label{MDSaaLD}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_aaLD_MatrixMDScorr.png}
+%\caption{\textbf{Correlation between mean amino acid site linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
+%\label{MDSaaCorr}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_ntLD_MDS.png}
+%\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean nucleotide linkage disequilibrium ($\chiSq$).}}
+%\label{MDSntLD}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_ntLD_MatrixMDScorr.png}
+%\caption{\textbf{Correlation between mean nucleotide site linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
+%\label{MDSntCorr}
+%\end{figure}
+
+%\section*{Differences between Vic and Yam PB1-PB2-HA}
+%Through the combination of inferred PB1-PB2-HA background and realized synonymous and non-synonymous substitution counts on branches \cite{obrien2009} it was possible to investigate whether Victoria and Yamagata lineage backgrounds differ in substitution rates.
+%Figure \ref{robustCounting} shows the inferred synonymous, non-synonymous (both as substitutions per year) and overall nucleotide substitution rate (as substitutions per site per year) under different lineages of PB1-PB2-HA.
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_lineageRatios.png}
+%\caption{\textbf{Total counts and ratios of Victoria and Yamagata lineage sequences in each segment.}}
+%\label{lineageRatios}
+%\end{figure}
 

### 2014-02-24 fluB.tex
Add my funding.
Also altering GISAID's reference.
Uncomment some future "figure supplements".
I figure if the aSPR figure gets supplements, so should the deltaTMRCA
figure (both show normalised values), especially since the dTMRCA
supplements show something very dramatic and explains what's going on a
bit better.

@@
 
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
-%\caption{\textbf{Mean $\dtmrca$ between all pairs of trees of segments.}}
-%\label{deltaTMRCAtrees}
-%\end{figure}
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
+\caption{\textbf{Mean $\dtmrca$ between all pairs of trees of segments.}}
+\label{deltaTMRCAtrees}
+\end{figure}
 
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
-%\caption{\textbf{Mean $\dtmrca$ between replicate trees of each segment.}}
-%\label{deltaTMRCAreplicates}
-%\end{figure}
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
+\caption{\textbf{Mean $\dtmrca$ between replicate trees of each segment.}}
+\label{deltaTMRCAreplicates}
+\end{figure}
 
@@
 \section*{Acknowledgements}
-GD was supported by a NERC studentship..... 
+GD was supported by a NERC studentship D76739X.
 TB was supported by a Newton International Fellowship from the Royal Society. 

### 2014-02-27 fluB.tex
Fix a few typos.
Changing a few subheaders per Andrew's suggestion.
Inserted some minor text formatting to make sure
"B/California/NHRC0001/2006" doesn't go out of the margins.
eLife wants to know dates of personal communications and it turns out I
did that on Christmas day.

Small question - for source data we can either include raw
post-processed MCMC output (in which case betweenDiversity data are the
largest and occupy over 250MB) or summary values (mean, median, HPDs)?

By the way, all of this is hopefully going up on arXiv and eLife on
Monday (runs that required re-doing should be finish

@@
 
-\subsection*{Phylogenetic analyses reveal maintenance of diversity and co-assortment of PB1, PB2 and HA gene segments}
+\subsection*{Maintenance of diversity and co-assortment of PB1, PB2 and HA gene segments}
 The differentiation into Vic and Yam lineages can be seen in all segments (Figure \ref{genomeGrid}).
@@
 PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
-Each point is the TMRCA of all surviving lineages existing at each time slice through the tree and vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+Each point is the mean time of most recent common ancestor (TMRCA) of all surviving lineages existing at each time slice through the tree and vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
 	\label{tmrcaOT}
@@
 By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between lineage diversity, which indicate that an inter-lineage reassortment event has taken place (Figure \ref{betweenDiversity}).
-Because we are taking the mean of pairwise comparisons, this method gives a quantitative measure of diversity of panmixis between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree.
+Because we are taking the mean of pairwise comparisons, this method gives a quantitative measure of panmixis between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree.
 We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
@@
 	\caption{\textbf{Ratio of lineages in the dataset.}
-The ratio of sequences that derive from the original Victoria clade (orange) to sequences that derive from the Yamagata clade (grey) in each segment over time.
+The ratio of sequences that derive from the original Victoria clade (yellow) to sequences that derive from the Yamagata clade (purple) in each segment over time.
 Yamagata lineage PA, NP, NA and MP segments and Victoria lineage NS segment eventually become fixed in the influenza B population.
@@
 Figure \ref{railroadPlot} focuses only on inter-lineage reassortments that have occured after 1990.
-We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS with prime (') indicating independently acquired segments) circulating 1992 -- 2011: B/Alaska/12/1996-like (Y-Y-Y-Y-Y-Y-Y-V), B/Nanchang/2/1997-like (V-V-Y-V-Y-V-Y-V), B/Iowa/03/2002-like (V-V-Y'-V-Y-Y-Y'-V'), B/California/NHRC0001/2006-like (V-V-Y-V-Y'-Y-Y'-V') and B/Brisbane/33/2008-like (V-V-Y-V-Y'-Y-Y-V) in Figure \ref{railroadPlot}.
-Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata lineage derived background.
+We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS with prime (') indicating independently acquired segments) circulating 1992 -- 2011 (Figure \ref{railroadPlot}):
+\begin{itemize}
+  \item B/Alaska/12/1996-like (Y-Y-Y-Y-Y-Y-Y-V)
+  \item B/Nanchang/2/1997-like (V-V-Y-V-Y-V-Y-V)
+  \item B/Iowa/03/2002-like (V-V-Y'-V-Y-Y-Y'-V')
+  \item B/California/NHRC0001/2006-like (V-V-Y-V-Y'-Y-Y'-V')
+  \item B/Brisbane/33/2008-like (V-V-Y-V-Y'-Y-Y-V)
+\end{itemize}
+Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, \\B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata lineage derived background.
 All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata lineage derived PB1-PB2-HA segments.
@@
 
-\subsection*{Recent PB1-PB2-HA constellations are entirely derived from Victoria or Yamagata lineage segments}
+\subsection*{Lineage composition of circulating PB1-PB2-HA constellations}
 Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
@@
 In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.
-Previously, associations between MP and NA segments have been suspected in influenza A viruses (Samantha Lycett, personal communication).
+Previously, associations between MP and NA segments have been suspected in influenza A viruses (Samantha Lycett, personal communication, 24 Dec 2013).
 
@@
 
-\subsection*{Linkage disequilibrium analysis confirms phylogenetic findings}
+\subsection*{Lineage-specific substitutions in PB1, PB2 and HA segments}
 Linkage disequilibrium (LD) is a measure of association between alleles at different polymorphic loci within a population.
@@
 
-If influenza B viruses are undergoing sympatric speciation, it is imperative to determine whether co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or whether co-infection so rare that speciation occurs via mutation-driven co-evolution.
+If influenza B viruses are undergoing sympatric speciation, it is imperative to determine whether co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or whether co-infection is so rare that speciation occurs via mutation-driven co-evolution.
 We think that co-dependence between PB1 and PB2 segments can be explained by the fact that they are functionally linked: together with the PA protein they form part of the influenza B virus RNA-dependent RNA polymerase heterotrimer.
@@
 
-It also remains unknown whether reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background reassorting into a Victoria PB1-PB2-HA background indicative of rare events followed by selective sweeps or stochastic fixation.
+It also remains unknown whether reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background reassorting into a Victoria PB1-PB2-HA background are indicative of rare events followed by selective sweeps or stochastic fixation.
 It's not unfeasible for selective sweeps following reassortments to break down developing co-adaptation of segments, especially if they are not functionally linked and have themselves been reassorted into a new PB1-PB2-HA background recently.
@@
 We accounted for incomplete sampling dates for 94 sequences (of which 93 had only year and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
+A relaxed molecular clock was used, where branch lengths are drawn from a lognormal distribution \cite{drummond2006}.
 We ran 3 independent MCMC chains, each with 200 million states, sampled every 20,000 steps and discarded the first 10\% of the MCMC states as burn-in.
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_normRErate.png}
-\caption{\textbf{Normalized reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
+\caption{\textbf{Normalized reassortment rate}
+Reassortment rate is calculated as approximate number of SPR moves per sum of total time in both trees.}
 \label{NormSPR_RErate}
@@
 \caption{\textbf{Heatmap of genome-wide linkage disequilibrium ($\chiSq$) between polymorphic amino acid sites.}
-Patterns of LD across the genome suggest a network of reciprocally linked amino acid sites on PB1, PB2, HA and to some extent NA proteins.
+Patterns of LD across the genome suggest a network of reciprocally linked amino acid sites on PB1, PB2, HA and, to some extent NA, proteins.
 Proximity of sites on heatmaps might not correspond to proximity of sites within proteins.}
@@
 Numbers next to each row and column correspond to amino acid site number within a given protein starting from methionine.
+Amino acid sites exhibiting reciprocally high LD between PB1, PB2 and HA proteins are: 176, 211, 213, 214, 508 (PB1), 115, 301, 382, 383, 397, 468, 483, 639 (PB2) and 86, 91, 96, 141, 152, 163, 164, 183, 190, 218, 270, 277, 282 (HA).
 Sites 211, 213 and 214 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).

### 2014-02-28 fluB.tex
Change subsection headers.
New subsection headers for merged paragraphs.

@@
 
-\subsection*{Maintenance of diversity and co-assortment of PB1, PB2 and HA gene segments}
+\subsection*{Analysis of reassortment patterns across Victoria and Yamagata lineages}
 The differentiation into Vic and Yam lineages can be seen in all segments (Figure \ref{genomeGrid}).
@@
 
-\subsection*{Lineage composition of circulating PB1-PB2-HA constellations}
 Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
@@
 
-\subsection*{PB1, PB2 and HA trees experience reassortment `distance' effects}
+\subsection*{Analysis of within-lineage reassortment patterns}
 SPR (subtree prune and regraft) distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between the two trees and is a measure of tree-to-tree similarity.
@@
 
-\subsection*{Lineage-specific substitutions in PB1, PB2 and HA segments}
 Linkage disequilibrium (LD) is a measure of association between alleles at different polymorphic loci within a population.

### 2014-02-28 fluB.tex
Revisions and comments.

@@
 
-By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between lineage diversity, which indicate that an inter-lineage reassortment event has taken place (Figure \ref{betweenDiversity}).
-Because we are taking the mean of pairwise comparisons, this method gives a quantitative measure of panmixis between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree.
+By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between-lineage diversity, which indicate that an inter-lineage reassortment event has taken place (Figure \ref{betweenDiversity}).
+Because we are taking the mean of pairwise comparisons, this method gives a quantitative measure of panmixis between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree, but instead gives the TMRCA of pairs of viruses.
+\tbc{I'm confused, I would think panmixis would need a comparison of within-lineage diversity compared to between-lineage diversity.
+The $F_{ST}$ statistics we talked about previously.
+I don't think you need to resurrect the $F_{ST}$ plot, but instead just change the language here.}
 We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
-Losses of diversity (represented by more recent mean pairwise TMRCAs between Vic and Yam labels) in Figure \ref{betweenDiversity} in all 8 segment trees indicate that every segment has reassorted with respect to the Victoria and Yamagata lineages of PB1, PB2 and HA segments.
+Losses of diversity (represented by more recent mean pairwise TMRCAs between Vic and Yam labels) in Figure \ref{betweenDiversity} indicate that every segment has reassorted with respect to the Victoria and Yamagata lineages of PB1, PB2 and HA segments.
 However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
@@
 
-By plotting the ratios of sequences assigned to either Vic or Yam lineage in each segment (Figure \ref{lineageRatiosOverTime}) it is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
+By plotting the ratios of sequences deriving from either the Vic or Yam lineage in each segment (Figure \ref{lineageRatiosOverTime}) it is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
 These losses of diversity correspond to fixation of the Vic lineage in NS and fixation of the Yam lineage in PA, NP, NA and MP.
-Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen (Figure \ref{lineageRatiosOverTime}), where the two lineages have been isolated at a ratio close to 50\% over long periods of time.
-On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically suggesting that one lineage may predominate within a given season, consistent with surveillance data \cite{reed2012}.
+Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen, where the two lineages have been isolated at a ratio close to 50\% over long periods of time (Figure \ref{lineageRatiosOverTime}).
+On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically consistent with one lineage predominating within a given season, in agreement with surveillance data \cite{reed2012}.
 
@@
 We reconstructed reassortment events that were detected by using lineage labels.
-Figure \ref{railroadPlot} focuses only on inter-lineage reassortments that have occured after 1990.
-We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS with prime (') indicating independently acquired segments) circulating 1992 -- 2011 (Figure \ref{railroadPlot}):
+Figure \ref{railroadPlot} focuses only on inter-lineage reassortments that have occurred after 1990.
+We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS with prime (') indicating independently acquired segments) circulating between 1992 and 2011 (Figure \ref{railroadPlot}):
 \begin{itemize}
@@
 \end{itemize}
-Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, \\B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata lineage derived background.
-All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata lineage derived PB1-PB2-HA segments.
+Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, \\B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata derived background.
+All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata derived PB1-PB2-HA segments.
 Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
@@
 A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has originally been derived from the Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
+\tbc{Concluding sentence that summarizes this rather complicated paragraph.}
 
@@
 
+\tbc{Opening sentence: something like mixed-linage PB1-PB2-HA constellations emerge, but are rare.}
 Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
 Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) since 1997 all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).
+We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bangkok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
+An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating the larger dataset with only PB1, PB2 and HA sequences.
+In all cases, PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
+In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each represented by two isolates), exhibit poor sampling and short circulation times.
+In the case of B/Nanchang/630/1994-like viruses this is up to two years.
+In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bangkok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past: up to 4 years for B/Bangkok/163/1990-like, although the more recently isolated B/Waikato/6/2005-like viruses circulated for only 5 months.
 
@@
 
-We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bangkok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
-An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating the larger dataset with only PB1, PB2 and HA sequences.
-
-In all cases, PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
-In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each represented by two isolates), exhibit poor sampling and short circulation times.
-In the case of B/Nanchang/630/1994-like viruses this is up to two years.
-In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bangkok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past: up to 4 years for B/Bangkok/163/1990-like, although the more recently isolated B/Waikato/6/2005-like viruses circulated for only 5 months.
-
 \subsection*{Analysis of within-lineage reassortment patterns}
-SPR (subtree prune and regraft) distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between the two trees and is a measure of tree-to-tree similarity.
+SPR (subtree prune and regraft) provides a measure of tree-to-tree distance, in which distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between them \tbc{CITE}.
 Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
-If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
-In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the raw (\textit{i.e} both within and between lineages) reassortment rate.
+If there are biases in the way segments reassort, so that some segments tend to co-assort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
+In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the overall (\textit{i.e} both within and between lineages) reassortment rate.
 
@@
 
-In Figure \ref{SPRdistances} the 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in both trees, see Figure \ref{NormSPR_RErate} in supplementary information) between all segments.
-We note, however, that our SPR distance analysis might simply lack power.
+The 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments between segments (Figure \ref{SPRdistances}).
+Reassortment rate given as number of SPR moves per total time in both trees shows similar results (Figure \ref{NormSPR_RErate}).
+We note, however, that because of phylogenetic uncertainty our estimate of SPR distance might simply lack power.
 Comparisons between independent analyses of the same segments yield distances that are comparable to distances between different segments (Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
+Still, we find that independent replicates from the same segment (Figure \ref{SPRdistancesReplicates}) show lower SPR distances that comparisons between segments (Figure \ref{SPRdistancesTrees}), suggesting that phylogenetic noise is not completely overwhelming reassortment signal.
 In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
@@
 Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_ratio}).
+\tbc{I don't think you need both differences and ratios in Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_ratio}.  Just pick one.}
+\tbc{The difference between \textit{normalized}, \textit{exact} and \textit{approximate} SPR distances needs to be clarified with having to send the reader to Methods.}
 
-Although we do not find evidence of differences in numbers of raw reassortments between all pairs of trees, we find support for a reassortment `distance' effect.
+Although we do not find evidence of differences in numbers of overall reassortments between segments, we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
 Figure \ref{deltaTMRCA} shows normalized mean $\dtmrca$ values for all pairs of trees.
@@
 PB1, PB2 and HA trees, on the other hand, exhibit normalized $\dtmrca$ values that are much higher.
-This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occassionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
-Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, i.e.\ we observe isolation by distance.
+This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
+\tbc{The preceding needs to be clarified / expanded.
+It's overly technical and terse at the moment.
+Give the basic rationale of why you're looking at differences in TMRCA (perhaps cite Andrew's 2008 paper).
+Motivate this with why $\dtmrca$ may differ from SPR distance, i.e.
+it looks at deeper phylogenetic events.}
+Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort among themselves unless both reassorting segments have similar TMRCAs, i.e.\ we observe isolation by distance.
 In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.
 Previously, associations between MP and NA segments have been suspected in influenza A viruses (Samantha Lycett, personal communication, 24 Dec 2013).
+\tbc{I'm wary of including `personal communications' in manuscripts for the same reason as I'm wary of `data not shown'.
+Could this sentence be dropped (keeping the previous about MP and NA)?
+Sam's already an author on this paper.}
 
@@
 
-Linkage disequilibrium (LD) is a measure of association between alleles at different polymorphic loci within a population.
-To estimate LD no other information, except for alleles at polymorphic amino acid sites in the dataset, is required.
-Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.
-In Figure \ref{segmentLD} the signature of Vic and Yam lineage-specific substitutions is greater mean $\chiSq$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
-In some cases this value exceeds the estimated mean LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA have higher mean LD than mean LD when PB1 is compared against itself).
+Linkage disequilibrium (LD) is a measure of non-random association between pairs of alleles at different polymorphic loci within a population.
+To estimate LD no other information, except for allele frequencies at polymorphic amino acid sites, is required.
+Although not an external validation of previous phylogenetic results, we observe significantly greater LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}).
+This suggests that the Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions that increase LD.
+\tbc{LD would also be increased if within Victoria there was strong association of alleles between segments.  It's not just Vic/Yam, yes?}
+Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
+\tbc{Do these sites show high LD because they're the only variable ones and hence the only sites that can show high LD?}
+These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
+In addition, there are sites on PB1, PB2 and HA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
 
@@
 
-Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
-These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
-In addition, there are sites on PB1, PB2 and HA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
-
 \section*{Discussion}
@@
 Evidence suggests that this is a result of prolonged lack of reassortment between Vic and Yam lineages in PB1, PB2 and HA (Figure \ref{betweenDiversity}) which possess co-assorting sequences detectable as high linkage disequilibrium (Figure \ref{segmentLD}).
-We propose that this pattern of non-reassortment is due to the action of selection and not simply biased reassortment.
-We do not find evidence of bias in reassortment or segment packaging bias between PB1, PB2 and HA segments (Figure \ref{SPRdistances}), but evidence points to these segments having considerably lower reassortment `distances' (Figure \ref{deltaTMRCA}).
+We propose that this pattern of coassortment is due to the action of selection and not simply biased reassortment.
+We do not find evidence of bias in reassortment or segment packaging bias between PB1, PB2 and HA segments (Figure \ref{SPRdistances}), but evidence points to ressortments between these segments being disfavored over longer spans of evolutionary time (Figure \ref{deltaTMRCA}).
 
 Reassortment `distances' represent the temporal separation of reassorting branches and suggests that reassortments between PB1, PB2 and HA segments tend to combine branches that have relatively recently been part of the same genome, i.e.\ that have shared a recent common virion.
-In addition, most strains with mixed-lineage PB1-PB2-HA complexes occured in the early years of the Vic--Yam split, when the two lineages were presumably much more similar at the nucleotide and amino acid levels.
+\tbc{Reassortment `distances' is too vague, I'd suggest talking more directly about TMRCAs.}
+In addition, most strains with mixed-lineage PB1-PB2-HA complexes occurred in the early years of the Vic--Yam split, when the two lineages were presumably much more similar at the nucleotide and amino acid levels.
 The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.
-We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \cite{turner2005}.
 
@@
 Indeed, trees of PB1 and PB2 segments exhibit high similarity in tip--tip TMRCAs (Figure \ref{deltaTMRCA}), suggesting highly similar TMRCAs.
-In addition, PB1--PB2 reassortants are the rarest and least persistent amongst mixed-lineage PB1-PB2-HA strains and have not been isolated since 1996.
+In addition, PB1--PB2 reassortants are the rarest and least persistent among mixed-lineage PB1-PB2-HA strains and have not been isolated since 1996.
 
@@
 
+\tbc{It sounds like this is the main hypothesis you're proposing, you're just elaborating on how functional linkage results in lack of reassortment.
+There is an unknown functional linkage between PB1, PB2 and HA and that the Vic and Yam associations have drifted away from each other.
+The use of `also' is confusing.}
 It is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
@@
 In addition, we see no reason why these three particular segments, and not the rest of the genome, would become co-adapted to each other in geographic isolation.
-
-\subsection*{The future of influenza B viruses}
-We see three potential paths of evolution for influenza B viruses.
-More segments could be recruited into the two currently circulating co-adapted segment complexes (PB1, PB2 and HA segments being the genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
-
-Sympatric speciation in other systems usually requires strong barriers to introgression, \textit{e.g.} infertility of F1 hybrids can lead to the evolution of prezygotic reproductive isolation otherwise known as reinforcement or the Wallace effect.
-To what extent this would apply to influenza B viruses remains unknown.
-
-If influenza B viruses are undergoing sympatric speciation, it is imperative to determine whether co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or whether co-infection is so rare that speciation occurs via mutation-driven co-evolution.
-We think that co-dependence between PB1 and PB2 segments can be explained by the fact that they are functionally linked: together with the PA protein they form part of the influenza B virus RNA-dependent RNA polymerase heterotrimer.
-
-It would be interesting to find out, from a speciation point of view, whether the next segments to be recruited into the PB1-PB2-HA segment complex will be functionally linked to the two polymerase proteins (PB1 and PB2) or haemagglutinin (HA).
-We already observe little mixing between the NA and MP segments and the PB1-PB2-HA complex in Figure \ref{betweenDiversity}, which is suggestive of recruitment to the complex.
-The NA protein has somewhat higher LD with PB1, PB2 and HA proteins in Figure \ref{segmentLD} and appears to have developed amino acid polymorphisms linked to lineage-specific sites on PB1, PB2 and HA proteins (Figure \ref{ChiGenome}).
-We cannot, however, dimiss the possibility that differentiation of the NA segment under the two PB1-PB2-HA complexes is caused by the rapid rate of nucleotide substitutions in the NA segment.
-Due to lack of polymorphic amino acid sites in the M1 protein it is difficult to determine whether the MP segment is linked to the NA segment or the PB1-PB2-HA segment complex to any appreciable degree.
-We do note that NA and MP segments exhibit a curious pattern of TMRCA similarities in Figure \ref{deltaTMRCA}.
-Together with Figure \ref{betweenDiversity} it suggests either that NA and MP segments are being recruited to PB1-PB2-HA segment complex together or that NA and MP form a second co-adapted and co-assorting segment complex.
-Recruitment to PB1-PB2-HA and formation of a second co-adapted segment complex are not mutually exclusive scenarios.
-
-It also remains unknown whether reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background reassorting into a Victoria PB1-PB2-HA background are indicative of rare events followed by selective sweeps or stochastic fixation.
-It's not unfeasible for selective sweeps following reassortments to break down developing co-adaptation of segments, especially if they are not functionally linked and have themselves been reassorted into a new PB1-PB2-HA background recently.
-This is the second scenario we might expect to occur, whereby relatively frequent reassortments occur between Vic and Yam lineages and are followed by selective sweeps.
-In this case influenza B viruses would undergo periodic genome homogenization events with the exception of PB1, PB2 and HA segments.
-Because Vic HA and Yam HA are antigenically dissimilar we think that balancing selection would prevent even strong selective sweeps from driving the opposing PB1-PB2-HA gene complex to extinction.
-This `\textit{status quo}' model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.
-
-Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether dynamics similar to Victoria and Yamagata lineages have not occurred in influenza B virus genomes before and left no trace through extinction.
-We find it unlikely that either Victoria or Yamagata lineage PB1-PB2-HA complexes will go extinct stochastically in the near future, as they have co-circulated for prolonged periods at a ratio close to 0.5, suggesting the action of balancing selection (Figure \ref{lineageRatiosOverTime}).
-Extinction through depletion of susceptible individuals, such as following influenza pandemics or mass vaccination seem unlikely as well.
-Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines, which are usually applied to specific subsets of the population, do not produce lifelong immunity.
 
@@
 
+\subsection*{The future of influenza B viruses}
+We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \cite{turner2005}.
+In this context, we see three potential paths of evolution for influenza B viruses.
+More segments could be recruited into the two currently circulating co-adapted segment complexes (PB1, PB2 and HA segments being the genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
+\tbc{Briefly outline all three possibilities here.}
+
+Sympatric speciation in other systems usually requires strong barriers to introgression, \textit{e.g.} infertility of F1 hybrids can lead to the evolution of prezygotic reproductive isolation otherwise known as reinforcement or the Wallace effect.
+To what extent this would apply to influenza B viruses remains unknown.
+If influenza B viruses are undergoing sympatric speciation, it is imperative to determine whether co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or whether co-infection is so rare that speciation occurs via mutation-driven co-evolution.
+We think that co-dependence between PB1 and PB2 segments can be explained by the fact that they are functionally linked: together with the PA protein they form part of the influenza B virus RNA-dependent RNA polymerase heterotrimer.
+
+It would be interesting to find out, from a speciation point of view, whether the next segments to be recruited into the PB1-PB2-HA segment complex will be functionally linked to the two polymerase proteins (PB1 and PB2) or haemagglutinin (HA).
+We already observe little mixing between the NA and MP segments and the PB1-PB2-HA complex (Figure \ref{betweenDiversity}), which is suggestive of recruitment to the complex.
+The NA protein has somewhat higher LD with PB1, PB2 and HA proteins in Figure \ref{segmentLD} and appears to have developed amino acid polymorphisms linked to lineage-specific sites on PB1, PB2 and HA proteins (Figure \ref{ChiGenome}).
+We cannot, however, dismiss the possibility that differentiation of the NA segment under the two PB1-PB2-HA complexes is caused by the rapid rate of nucleotide substitutions in the NA segment.
+Due to lack of polymorphic amino acid sites in the M1 protein it is difficult to determine whether the MP segment is linked to the NA segment or the PB1-PB2-HA segment complex to any appreciable degree.
+We do note that NA and MP segments exhibit a curious pattern of TMRCA similarities in Figure \ref{deltaTMRCA}.
+Together with Figure \ref{betweenDiversity} it suggests either that NA and MP segments are being recruited to PB1-PB2-HA segment complex together or that NA and MP form a second co-adapted and co-assorting segment complex.
+Recruitment to PB1-PB2-HA and formation of a second co-adapted segment complex are not mutually exclusive scenarios.
+\tbc{This whole paragraph seems overly speculative, extrapolating from very little signal (and signal that is not consistent between analyses --- phylogeny, LD and $\dtmrca$ give conflicting stories, as you note.
+I'd favor cutting the entire paragraph.}
+
+It also remains unknown whether reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background reassorting into a Victoria PB1-PB2-HA background are indicative of rare events followed by selective sweeps or stochastic fixation.
+It's not unfeasible for selective sweeps following reassortments to break down developing co-adaptation of segments, especially if they are not functionally linked and have themselves been reassorted into a new PB1-PB2-HA background recently.
+This is the second scenario we might expect to occur, whereby relatively frequent reassortments occur between Vic and Yam lineages and are followed by selective sweeps.
+In this case influenza B viruses would undergo periodic genome homogenization events with the exception of PB1, PB2 and HA segments.
+Because Vic HA and Yam HA are antigenically dissimilar we think that balancing selection would prevent even strong selective sweeps from driving the opposing PB1-PB2-HA gene complex to extinction.
+This `\textit{status quo}' model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.
+
+Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether dynamics similar to Victoria and Yamagata lineages have not occurred in influenza B virus genomes before and left no trace through extinction.
+We find it unlikely that either Victoria or Yamagata lineage PB1-PB2-HA complexes will go extinct stochastically in the near future, as they have co-circulated for prolonged periods at a ratio close to 0.5, suggesting the action of balancing selection (Figure \ref{lineageRatiosOverTime}).
+Extinction through depletion of susceptible individuals, such as following influenza pandemics or mass vaccination seem unlikely as well.
+Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines, which are usually applied to specific subsets of the population, do not produce lifelong immunity.
+
+\tbc{What's the third scenario?}
+
 \subsection*{Conclusion}
 We have used the $\dtmrca$ statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
-We believe that patristic distance methods, though far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies.
+We believe that patristic distance methods such as this, though themselves far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies.
 One of many useful applications of the $\dtmrca$ method would be identifying clades or taxa that are products of reticulate evolution.
-Within-analysis $\dtmrca$ could be used as a cutoff to detect outlier taxa with greater than expected $\dtmrca$ values.
+Within-analysis \tbc{within what?} $\dtmrca$ could be used as a cutoff to detect outlier taxa with greater than expected $\dtmrca$ values.
 To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
@@
 Given sufficient time we expect the two PB1-PB2-HA complexes to eventually recruit the rest of the genomic segments into co-adapted and co-reassorting segment complexes resulting in sympatric speciation of human influenza B viruses.
+\tbc{This was just one scenario, yes?}
 
@@
 
+\tbc{I'm more used to Supp figures appearing after Acknowledgements and References.}
+
 \section*{Acknowledgements}

### 2014-03-03 fluB.tex
Added another affiliation.
Add Trevor's suggestions.

@@
 \author[1,3]{Samantha Lycett}
-\author[1,4]{Andrew Rambaut}
+\author[1,4,5]{Andrew Rambaut}
 
@@
 \affil[4]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
+\affil[5]{Centre for Immunology, Infection and Evolution at the University of Edinburgh, Edinburgh, UK}
 
@@
 By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between-lineage diversity, which indicate that an inter-lineage reassortment event has taken place (Figure \ref{betweenDiversity}).
-Because we are taking the mean of pairwise comparisons, this method gives a quantitative measure of panmixis between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree, but instead gives the TMRCA of pairs of viruses.
-\tbc{I'm confused, I would think panmixis would need a comparison of within-lineage diversity compared to between-lineage diversity.
-The $F_{ST}$ statistics we talked about previously.
-I don't think you need to resurrect the $F_{ST}$ plot, but instead just change the language here.}
+Because we are taking the mean of pairwise comparisons, this method gives a quantitative measure of reassortment-induced loss of diversity between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree, but instead gives the TMRCA of pairs of viruses.
 We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
@@
 A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has originally been derived from the Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
-\tbc{Concluding sentence that summarizes this rather complicated paragraph.}
+Overall, we see that successful inter-lineage reassortment events do not break up Vic and Yam PB1-PB2-HA complexes, but any other segments reassorted into one or the other PB1-PB2-HA background evolve with a considerable degree of autonomy.
 
@@
 
-\tbc{Opening sentence: something like mixed-linage PB1-PB2-HA constellations emerge, but are rare.}
+The vast majority of influenza B isolates possess either Vic or Yam lineage derived PB1-PB2-HA complexes, although on rare occasions mixed-lineage PB1-PB2-HA constellations emerge.
 Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
@@
 \subsection*{Analysis of within-lineage reassortment patterns}
-SPR (subtree prune and regraft) provides a measure of tree-to-tree distance, in which distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between them \tbc{CITE}.
+SPR (subtree prune and regraft) provides a measure of tree-to-tree distance, in which distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between them \cite{svinti2013}, as long as each branch experiences, at most, one reassortment.
 Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
@@
 
+We normalized our SPR distance comparisons to account for differences in tree topology stability over the course of the MCMC chain and also use an approximation of SPR distance \cite{whidden2009,whidden2010,whidden2013} (see Methods).
 The 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments between segments (Figure \ref{SPRdistances}).
@@
 Thus we caution against over-interpreting Figure \ref{SPRdistances}.
-Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_ratio}).
-\tbc{I don't think you need both differences and ratios in Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_ratio}.  Just pick one.}
-\tbc{The difference between \textit{normalized}, \textit{exact} and \textit{approximate} SPR distances needs to be clarified with having to send the reader to Methods.}
+Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_difference}).
 
 Although we do not find evidence of differences in numbers of overall reassortments between segments, we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
+Unlike SPR distances, which are counts of the numbers of reassortment events, $\dtmrca$ takes time into account, and is most sensitive when only one of the two trees being compared loses diversity via reassortment.
+Similarly to SPR distances, we normalized our $\dtmrca$ comparisons to account for node date stability over the course of the MCMC chain (see Methods).
 Figure \ref{deltaTMRCA} shows normalized mean $\dtmrca$ values for all pairs of trees.
-Most segment pairs show very low values for this statistic, indicating considerable differences between TMRCAs of the trees being compared.
-A large number of these comparisons have TMRCA differences up to 10 times larger ($\dtmrca$$\approx$0.1) than TMRCA differences that occur between replicate BEAST analyses of one alignment over the course of the MCMC chain.
+Most segment pairs show very low values for this statistic with $\dtmrca\approx$0.1, indicating that $\dtmrca$ measurements between independent analyses of the same segment are up to 10 times smaller than $\dtmrca$ values between analyses of different segments.
 PB1, PB2 and HA trees, on the other hand, exhibit normalized $\dtmrca$ values that are much higher.
-This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
-\tbc{The preceding needs to be clarified / expanded.
-It's overly technical and terse at the moment.
-Give the basic rationale of why you're looking at differences in TMRCA (perhaps cite Andrew's 2008 paper).
-Motivate this with why $\dtmrca$ may differ from SPR distance, i.e.
-it looks at deeper phylogenetic events.}
+This shows that TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
+For example, the distribution of comparison between PB1 and PB2 trees has a mean $\dtmrca$$\approx0.4$, indicating that the sum of $\dtmrca$ values from PB1-PB1' and PB2-PB2' comparisons (prime (') indicates an independent MCMC chain) are 40\% of twice the $\dtmrca$ value from the PB1-PB2 comparison, but this value can range anywhere between $\approx$0.15 and $\approx$0.75.
 Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort among themselves unless both reassorting segments have similar TMRCAs, i.e.\ we observe isolation by distance.
 In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.
-Previously, associations between MP and NA segments have been suspected in influenza A viruses (Samantha Lycett, personal communication, 24 Dec 2013).
-\tbc{I'm wary of including `personal communications' in manuscripts for the same reason as I'm wary of `data not shown'.
-Could this sentence be dropped (keeping the previous about MP and NA)?
-Sam's already an author on this paper.}
 
@@
 Although not an external validation of previous phylogenetic results, we observe significantly greater LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}).
-This suggests that the Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions that increase LD.
-\tbc{LD would also be increased if within Victoria there was strong association of alleles between segments.  It's not just Vic/Yam, yes?}
+This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
+We conclude that Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions.
 Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
-\tbc{Do these sites show high LD because they're the only variable ones and hence the only sites that can show high LD?}
+Figure \ref{ChiGenome} also shows that the number of variable amino acid sites in PB1 and PB2 proteins is comparable to most other proteins investigated (see Methods for cut-offs used to exclude sites from LD analysis).
 These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
@@
 
-Reassortment `distances' represent the temporal separation of reassorting branches and suggests that reassortments between PB1, PB2 and HA segments tend to combine branches that have relatively recently been part of the same genome, i.e.\ that have shared a recent common virion.
-\tbc{Reassortment `distances' is too vague, I'd suggest talking more directly about TMRCAs.}
+When comparing the mean tip-tip TMRCA deviations we find evidence of isolation by distance - when reassortments occur between PB1, PB2 and HA segments those events tend to involve branches that have similar TMRCAs.
+Other segment pairs, with the exception of MP-NA, seem to show little signal of this and are able to reassort with branches that have been evolving under temporally distant genomic backgrounds.
 In addition, most strains with mixed-lineage PB1-PB2-HA complexes occurred in the early years of the Vic--Yam split, when the two lineages were presumably much more similar at the nucleotide and amino acid levels.
@@
 
-\tbc{It sounds like this is the main hypothesis you're proposing, you're just elaborating on how functional linkage results in lack of reassortment.
-There is an unknown functional linkage between PB1, PB2 and HA and that the Vic and Yam associations have drifted away from each other.
-The use of `also' is confusing.}
-It is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
+It is possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
 In this case PB1, PB2 and HA segments of Vic and Yam lineage accumulate substitutions that improve their ability to co-operate with segments of the same lineage and worse at interacting segments of a different lineage.
@@
 Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
-We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (Figures \ref{NormSPR_PB2-HA_difference} and \ref{NormSPR_PB2-HA_ratio}).
+We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (Figures \ref{NormSPR_PB2-HA_difference}).
 The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
@@
 More segments could be recruited into the two currently circulating co-adapted segment complexes (PB1, PB2 and HA segments being the genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
-\tbc{Briefly outline all three possibilities here.}
+This is the speciation scenario.
+The two alternatives to this are the `\textit{status quo}' model, where the influenza B genome continues to be homogenized via gene flow with the exception of PB1, PB2 and HA segments and the extinction scenario, whereby one of the two PB1-PB2-HA complexes goes extinct, marking the return of single-strain dynamics in the influenza B virus population.
 
@@
 We think that co-dependence between PB1 and PB2 segments can be explained by the fact that they are functionally linked: together with the PA protein they form part of the influenza B virus RNA-dependent RNA polymerase heterotrimer.
-
-It would be interesting to find out, from a speciation point of view, whether the next segments to be recruited into the PB1-PB2-HA segment complex will be functionally linked to the two polymerase proteins (PB1 and PB2) or haemagglutinin (HA).
-We already observe little mixing between the NA and MP segments and the PB1-PB2-HA complex (Figure \ref{betweenDiversity}), which is suggestive of recruitment to the complex.
-The NA protein has somewhat higher LD with PB1, PB2 and HA proteins in Figure \ref{segmentLD} and appears to have developed amino acid polymorphisms linked to lineage-specific sites on PB1, PB2 and HA proteins (Figure \ref{ChiGenome}).
-We cannot, however, dismiss the possibility that differentiation of the NA segment under the two PB1-PB2-HA complexes is caused by the rapid rate of nucleotide substitutions in the NA segment.
-Due to lack of polymorphic amino acid sites in the M1 protein it is difficult to determine whether the MP segment is linked to the NA segment or the PB1-PB2-HA segment complex to any appreciable degree.
-We do note that NA and MP segments exhibit a curious pattern of TMRCA similarities in Figure \ref{deltaTMRCA}.
-Together with Figure \ref{betweenDiversity} it suggests either that NA and MP segments are being recruited to PB1-PB2-HA segment complex together or that NA and MP form a second co-adapted and co-assorting segment complex.
-Recruitment to PB1-PB2-HA and formation of a second co-adapted segment complex are not mutually exclusive scenarios.
-\tbc{This whole paragraph seems overly speculative, extrapolating from very little signal (and signal that is not consistent between analyses --- phylogeny, LD and $\dtmrca$ give conflicting stories, as you note.
-I'd favor cutting the entire paragraph.}
 
@@
 
-\tbc{What's the third scenario?}
-
 \subsection*{Conclusion}
@@
 One of many useful applications of the $\dtmrca$ method would be identifying clades or taxa that are products of reticulate evolution.
-Within-analysis \tbc{within what?} $\dtmrca$ could be used as a cutoff to detect outlier taxa with greater than expected $\dtmrca$ values.
+$\dtmrca$ measures between independent analyses of the same alignment (`within-alignment') could be used as a cutoff to detect outlier taxa with greater than expected `between-alignment' $\dtmrca$ values.
 To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
@@
 We also suspect that NA-MP segments might be linked in a similar way and potentially become the next segments to be recruited by the Vic and Yam PB1-PB2-HA complexes.
-Given sufficient time we expect the two PB1-PB2-HA complexes to eventually recruit the rest of the genomic segments into co-adapted and co-reassorting segment complexes resulting in sympatric speciation of human influenza B viruses.
-\tbc{This was just one scenario, yes?}
+Given sufficient time it should become clear whether PB1-PB2-HA complexes of human influenza B viruses are continuing to resist gene flow whilst the rest of the influenza B virus genome is repeatedly homogenized or whether the two PB1-PB2-HA complexes are recruiting the rest of the genomic segments into co-adapted and co-reassorting segment complexes on their way to sympatric speciation.
 
@@
 
+\section*{Acknowledgements}
+GD was supported by a NERC studentship D76739X.
+TB was supported by a Newton International Fellowship from the Royal Society. 
+The research leading to these results has received funding from the European Research Council under the European Community's Seventh Framework Programme (FP7/2007-2013) under Grant Agreement no.
+278433-PREDEMICS and ERC Grant agreement no.
+260864.
+AR and SL acknowledge the support of the Wellcome Trust (grant no. 092807).
+
+\bibliographystyle{pnas}
+\bibliography{fluB}
+
+\clearpage
+
 \setcounter{figure}{0}
@@
 
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances after normalization.}
-95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
-\label{NormSPR_PB1-PB2_ratio}
-\end{figure}
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances after normalization.}
+%95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
+%\label{NormSPR_PB1-PB2_ratio}
+%\end{figure}
 
@@
 
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances after normalization.}
-95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
-\label{NormSPR_PB1-HA_ratio}
-\end{figure}
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances after normalization.}
+%95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
+%\label{NormSPR_PB1-HA_ratio}
+%\end{figure}
 
@@
 
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_hist2.png}
-\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after normalization.}
-95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.
-Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
-\label{NormSPR_PB2-HA_ratio}
-\end{figure}
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after normalization.}
+%95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.
+%Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
+%\label{NormSPR_PB2-HA_ratio}
+%\end{figure}
 
@@
 %\end{figure}
-
-\clearpage
 
@@
 
-\tbc{I'm more used to Supp figures appearing after Acknowledgements and References.}
-
-\section*{Acknowledgements}
-GD was supported by a NERC studentship D76739X.
-TB was supported by a Newton International Fellowship from the Royal Society. 
-The research leading to these results has received funding from the European Research Council under the European Community's Seventh Framework Programme (FP7/2007-2013) under Grant Agreement no.
-278433-PREDEMICS and ERC Grant agreement no.
-260864.
-AR and SL acknowledge the support of the Wellcome Trust (grant no. 092807).
-
-\bibliographystyle{pnas}
-\bibliography{fluB}
 \end{document}

### 2014-03-03 fluB.tex
Add a blurb under dTMRCA figure supplements.

@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
-\caption{\textbf{Mean $\dtmrca$ between all pairs of trees of segments.}}
+\caption{\textbf{Mean $\dtmrca$ between all pairs of trees of segments.}
+Mean $\dtmrca$ between trees of segments reveal that tip pairs in PB1, PB2 and HA trees have very similar TMRCAs.
+The upper tail of the 95\% HPD (HPDs are represented as vertical lines) interval of mean $\dtmrca$ values for PB1-PB2-HA and MP-NA trees do not exceed 3 years.}
 \label{deltaTMRCAtrees}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
-\caption{\textbf{Mean $\dtmrca$ between replicate trees of each segment.}}
+\caption{\textbf{Mean $\dtmrca$ between replicate trees of each segment.}
+Mean $\dtmrca$ values between independent analyses of each segment show that mean $\dtmrca$ values rarely exceed 1 year.}
 \label{deltaTMRCAreplicates}

### 2014-03-14 scripts/Hrothvitnir_eLife.py
Add scripts used to get data.
Hrothvitnir ("fame-wolf", another name for Fenrir) outputs diversity
statistics from posterior sets of trees.
Hydra (named so because it deals with 4 trees at a time) outputs
deltaTMRCA and SPR statistics from posterior sets of trees.
LD_calculator_ChiSqdf does what it says on the tin.

Will add the IPython notebook soon too.

@@ -1,0 +1,737 @@
+#####################################################################################
+##
+##  This script analyzes posterior sets of trees produced by BEAST.
+##  It is meant to accompany the manuscript "Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex"
+##  and has only been tested with trees produced specifically for it.
+##  You must have python and numpy installed on your machine to use this script.
+##
+##  Usage:
+##  open command line, go to folder where the script is located and type in:
+##  "python Hrothvitnir_eLife.py -i <<path to .trees.txt file>> -m <<mode - X, diversityOt, FstOt or stateTime>> 1> <<path to output file>>" (ignore <<, >> and ")
+##  
+##  The following figures in the manuscript were made using these modes:
+##  Figure 3 - diversityOT
+##  Figure 4 - FstOt
+##  Figures 7 and S9 - stateTime
+##
+##
+##  Gytis Dudas
+##  Institute of Evolutionary Biology
+##  University of Edinburgh
+##  EH9 3JT
+##  Edinburgh, UK
+##
+#####################################################################################
+
+import re
+from datetime import datetime as dt
+import time
+import calendar
+import numpy as np
+import random
+import glob
+import os
+import sys
+
+def unique(o, idfun=repr):
+    """
+    Returns a list of unique values in a list.
+    """
+    seen = {}
+    return [seen.setdefault(idfun(e),e) for e in o if idfun(e) not in seen]
+
+def column(data,col):
+    """
+    Returns a list that is a column in a list of lists.
+    """
+    return [row [col] for row in data]
+
+def index(data,item):
+    """
+    Returns a list of indices where an item can be found within a given list.
+    """
+    return [i for i,x in enumerate(data) if x == item]
+
+def makeSliceList(slices,start,end,bins):
+    """
+    Return a list of dates spaced out regularly within a year.
+    """
+    for i in np.arange(start,end,1/float(bins)):
+        slices.append([i])
+
+def mostRecent(some_list):
+    """
+    Identify the most recent tip in the tree and its date.
+    """
+    full=[]
+    month=[]
+    year=[]
+    ## Summarize tip names in regular expressions
+    ## Enclose the bit of the sequence name containing the date in brackets
+    ## Date should be in format yyyy-mm-dd
+    dateRegex='[A-Z]\/[A-Za-z\/\-\_\.0-9]+\/[0-9]+\_([0-9\-]+)'
+    for i in some_list:
+        C=re.search('%s'%(dateRegex),i[1])
[…]

### 2014-03-14 scripts/Hydra_eLife.py

@@ -1,0 +1,1220 @@
+#####################################################################################
+##
+##  This script analyzes posterior sets of trees produced by BEAST.
+##  It is meant to accompany the manuscript "Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex"
+##  and has only been tested with trees produced specifically for it.
+##  You must have python, numpy and RSPR (available at http://kiwi.cs.dal.ca/Software/RSPR) installed on your machine to use this script.
+##
+##  You also require at least 4 tree files to run this script.
+##  It compares trees made from alignments A and B (trees A and B), then normalizes this comparison by independent analyses of alignments A and B (trees A' and B').
+##  
+##
+##  Usage:
+##  go to OPTIONS at the bottom of the script and alter as required
+##  if running in batchProcessing mode point to tree files (filenames and path variables)
+##  otherwise the script should prompt you to select the files manually
+##
+##  open command line, go to folder where the script is located and type in:
+##  "python Hydra_eLife.py" (ignore ")
+##  
+##  The following figures in the manuscript were made using this script:
+##  Figures 8, 9, S1-S6, S10-S11
+##
+##
+##  Gytis Dudas
+##  Institute of Evolutionary Biology
+##  University of Edinburgh
+##  EH9 3JT
+##  Edinburgh, UK
+##
+#####################################################################################
+
+import re
+from datetime import datetime as dt
+import time
+import calendar
+import numpy as np
+import random
+import glob
+import os
+import sys
+import subprocess
+
+def unique(o, idfun=repr):
+    """
+    Returns a list of unique values in a list.
+    """
+    seen = {}
+    return [seen.setdefault(idfun(e),e) for e in o if idfun(e) not in seen]
+
+def column(data,col):
+    """
+    Returns a list that is a column in a list of lists.
+    """
+    return [row [col] for row in data]
+
+def index(data,item):
+    """
+    Returns a list of indices where an item can be found within a given list.
+    """
+    return [i for i,x in enumerate(data) if x == item]
+
+def makeSliceList(slices,start,end,bins):
+    """
+    Return a list of dates spaced out regularly within a year.
+    """
+    for i in np.arange(start,end,1/float(bins)):
+        slices.append([i])
+
+def mostRecent(some_list,objectContainer):
+    """
+    Identify the most recent tip in the tree and its date.
+    """
+    full=[]
[…]

### 2014-03-14 scripts/LD_calculator_ChiSqdf_eLife.py

@@ -1,0 +1,380 @@
+#####################################################################################
+##
+##  This script calculates linkage disequilibrium (LD) between sites in different alignments (nucleotide or amino acid).
+##  It is meant to accompany the manuscript "Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex"
+##  and has only been tested with alignments produced specifically for it.
+##  You must have python, numpy and biopython installed on your machine to use this script.
+##
+##  You also require either protein or nucleotide alignments (in nexus format) with the same number of taxa that have the same names.
+##  The sequences used for this project are listed in the acknowledgment table accompanying the manuscript:
+##  https://github.com/evogytis/fluB/tree/master/acknowledgement%20tables
+##
+##  and can be downloaded from GISAID available at:
+##  http://platform.gisaid.org
+##  
+##  open command line, go to folder where the script is located and type in:
+##  "python LD_calculatorChiSqdf_eLife.py" (ignore ")
+##
+##  To switch between nucleotide and amino acid modes comment 
+##
+##  The following figures in the manuscript were made using this script:
+##  Figures 10, S7-S8
+##
+##
+##  Alternatively, instead of running the script the output is available here:
+##  https://github.com/evogytis/fluB/tree/master/data
+##  
+##
+##  Gytis Dudas
+##  Institute of Evolutionary Biology
+##  University of Edinburgh
+##  EH9 3JT
+##  Edinburgh, UK
+##
+#####################################################################################
+
+
+import numpy as np
+from Bio.Seq import Seq
+from Bio.Alphabet import IUPAC
+import re
+from datetime import datetime as dt
+import time
+import calendar
+from Bio import SeqIO
+from Bio import AlignIO
+
+def unique(o, idfun=repr):
+    """
+    Returns a list of unique values in a list.
+    """
+    seen = {}
+    return [seen.setdefault(idfun(e),e) for e in o if idfun(e) not in seen]
+
+def column(data,col):
+    """
+    Returns a list that is a column in a list of lists.
+    """
+    return [row [col] for row in data]
+
+def index(data,item):
+    """
[…]

### 2014-05-05 fluB.tex
Remove reference to NA-MP.
Move genomeGrid from supp figures to figures.
Change text accordingly.
Implement some of the reviewers' comments.
Some clarifications should fix #13, #14, #15, #16, #18 and #19.

Added an alternative `stateTime' figure which excludes PB1, PB2 and HA
trees to avoid confusion.

Changed `LineageRatiosOverTime' figure to include genome counts at the
top to clarify that it shows something extremely basic.

Also altered LD calculator a bit, it didn't work with time-slicing
before and now does. I was trying to animate LD over time, didn't look
as impressive as expected.

@@
 \centering
-\includegraphics[width=0.95\textwidth]{supp_figures/InfB_genomeGrid.pdf}
+\includegraphics[width=0.95\textwidth]{figures/InfB_genomeGrid.pdf}
 \caption{\textbf{Maximum clade credibility (MCC) trees of all 8 genome segments of influenza B viruses isolated since 1980.}
@@
 
-By plotting the ratios of sequences deriving from either the Vic or Yam lineage in each segment (Figure \ref{lineageRatiosOverTime}) it is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
+We show the ratio of Vic and Yam lineages in our sequence data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on lineage assignment performed earlier.
+It is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
 These losses of diversity correspond to fixation of the Vic lineage in NS and fixation of the Yam lineage in PA, NP, NA and MP.
@@
 The ratio of sequences that derive from the original Victoria clade (yellow) to sequences that derive from the Yamagata clade (purple) in each segment over time.
+Numbers at the top of the figure show the total number of genomes analyzed within each influenza season.
+Numbers at the bottom are influenza seasons from the 1984/1985 (84/85) season to the 2011/2012 season.
 Yamagata lineage PA, NP, NA and MP segments and Victoria lineage NS segment eventually become fixed in the influenza B population.
@@
 \end{itemize}
-Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, \\B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata derived background.
+B/Alaska/12/1996-like, B/Nanchang/2/1997-like and B/Iowa/03/2002-like constellations were observed by Chen \& Holmes \cite{chen2008}, but sequences from B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like constellations were not available at the time.
+Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata derived background.
 All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata derived PB1-PB2-HA segments.
@@
 Evidence suggests that this is a result of prolonged lack of reassortment between Vic and Yam lineages in PB1, PB2 and HA (Figure \ref{betweenDiversity}) which possess co-assorting sequences detectable as high linkage disequilibrium (Figure \ref{segmentLD}).
+The vast majority of the sampled evolutionary history of each segment of influenza B viruses since the split of Vic and Yam lineages has been spent in association with either completely Victoria or completely Yamagata lineage derived PB1-PB2-HA complexes (Figure \ref{stateTime}), suggesting that having `pure' lineage PB1-PB2-HA complexes is important for whole-genome fitness.
 We propose that this pattern of coassortment is due to the action of selection and not simply biased reassortment.
-We do not find evidence of bias in reassortment or segment packaging bias between PB1, PB2 and HA segments (Figure \ref{SPRdistances}), but evidence points to ressortments between these segments being disfavored over longer spans of evolutionary time (Figure \ref{deltaTMRCA}).
+One could image that due to some artefact of segment packaging PB1, PB2 and HA segments preferentially co-assort together and thus have a reduced reassortment rate when compared to the rest of the genomic segments.
+We find no evidence of this -- PB1, PB2 and HA segments reassort to the same extent as other segments (Figure \ref{SPRdistances}) and instead only reassortments combining lineages across greater temporal distances are disfavoured (Figure \ref{deltaTMRCA}), which does not seem to drive the overall reassortment rate down.
+If PB1, PB2 and HA trees were embedded within a `species tree' we would see many reassortments within both major branches of this tree -- the Vic and Yam lineages -- but not between them.
 
@@
 The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.
+
+We should clarify that when we say that Vic and Yam PB1-PB2-HA complexes of the influenza B virus are co-adapted and have co-evolved we mean that these 3 segments do not appear to function well when derived from both lineages and that the evolution to this state has been intimate, in the sense that the two lineages have not reassorted across great temporal distances, unlike the rest of the genome.
 
@@
 We argue that this is due to selection against viruses with mixed-lineage PB1-PB2-HA complexes.
-We also suspect that NA-MP segments might be linked in a similar way and potentially become the next segments to be recruited by the Vic and Yam PB1-PB2-HA complexes.
 Given sufficient time it should become clear whether PB1-PB2-HA complexes of human influenza B viruses are continuing to resist gene flow whilst the rest of the influenza B virus genome is repeatedly homogenized or whether the two PB1-PB2-HA complexes are recruiting the rest of the genomic segments into co-adapted and co-reassorting segment complexes on their way to sympatric speciation.
@@
 Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
+This method operates under the assumption that the segment trees capture the overall `species tree' (or `virus tree') of influenza B viruses quite well.
+It is not an unreasonable assumption, given the seasonal bottlenecks of these viruses.
+The $\dtmrca$ statistic essentially captures a facet of this influenza B `species tree' by quantifying the temporal distance between an admixture event and the original lineage that was replaced by the incoming lineage.
 Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \cite{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \cite{ruane2011}.

### 2014-05-05 scripts/LD_calculator_ChiSqdf_eLife.py

@@ -132,8 +132,8 @@ def overlap(a,b):
 
 ## Makes a list of dates if partitioning of LD is desired
 # Default value makes a single bin for all sequences isolated between 1900 and 2014
-step=114
-timeline=np.arange(1900,2014,step)
+step=3
+timeline=np.arange(1984,2014,step)
 
 ## Segments to analyze                                                                       
 segments=['PB1','PB2','PA','HA','NP','NA','M1','NS1']
@@ -144,9 +144,11 @@ segments=['PB1','PB2','PA','HA','NP','NA','M1','NS1']
 mixedLineage=[]
 
 ## Select mode - nt for nucleotide alignments, aa for amino acid alignments
-mode='nt'
-#mode='aa'
+#mode='nt'
+mode='aa'
 
+
+#segments=['NA','M1','NS1']
 
 ## Iterate over all pairs of segments
 for mem1 in range(len(segments)):
@@ -169,12 +171,12 @@ for mem1 in range(len(segments)):
             ## Define alignment name
             seg1='InfB_%s translation alignment.nex'%(segment1)
             seg2='InfB_%s translation alignment.nex'%(segment2)
-            f = open('/Users/admin/Documents/Viral sequences/InfB reassortment/Alignments/Translated/%s_%s_rDChicomparisonNoSelfCommonTEST.csv'%(segment1,segment2),'w')
+            f = open('/Users/admin/Documents/Viral sequences/InfB reassortment/Alignments/Translated/%s_%s_rDChicomparisonNoSelfTimed3.csv'%(segment1,segment2),'w')
         elif mode=='nt':
             path='/Users/admin/Documents/Viral sequences/InfB reassortment/Alignments/Without ancient seqs and pruned/'
             seg1='InfB_%s.nex'%(segment1)
             seg2='InfB_%s.nex'%(segment2)
-            f = open('/Users/admin/Documents/Viral sequences/InfB reassortment/Alignments/Without ancient seqs and pruned/%s_%s_rDChiNtcomparisonNoSelf.csv'%(segment1,segment2),'w')
+            f = open('/Users/admin/Documents/Viral sequences/InfB reassortment/Alignments/Without ancient seqs and pruned/%s_%s_rDChiNtcomparisonNoSelfRemake.csv'%(segment1,segment2),'w')
         
         handle1 = open(path+seg1, "rU")
         handle2 = open(path+seg2, "rU")
@@ -185,30 +187,34 @@ for mem1 in range(len(segments)):
[…]

### 2014-05-07 scripts/LD_calculator.py
Streamline LD calculator.
Now calculates ChiSq.df *AND* D' statistics of linkage disequilibrium.
D', which essentially calculates how frequently things recombine, gives
incredibly high LD when one haplotype is not observed.

@@ -1,0 +1,346 @@
+#####################################################################################
+##
+##  This script calculates linkage disequilibrium (LD) between sites in different alignments (nucleotide or amino acid).
+##  It calculates both the Chi-Squared df and D' statistics of linkage disequilibrium.
+##  It is meant to accompany the manuscript "Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex"
+##  and has only been tested with alignments produced specifically for it.
+##  You must have python, numpy and biopython installed on your machine to use this script.
+##
+##  You also require either protein or nucleotide alignments (in nexus format) with the same number of taxa that have the same names.
+##  The sequences used for this project are listed in the acknowledgment table accompanying the manuscript:
+##  https://github.com/evogytis/fluB/tree/master/acknowledgement%20tables
+##
+##  and can be downloaded from GISAID available at:
+##  http://platform.gisaid.org
+##  
+##  open command line, go to folder where the script is located and type in:
+##  "python LD_calculator.py" (ignore ")
+##
+##  To switch between nucleotide and amino acid modes comment out mode='aa' or mode='nt', respectively.
+##
+##  The following figures in the manuscript were made using this script:
+##  Figures 10, S7-S8
+##
+##
+##  Alternatively, instead of running the script the output is available here:
+##  https://github.com/evogytis/fluB/tree/master/data
+##  
+##
+##  Gytis Dudas
+##  Institute of Evolutionary Biology
+##  University of Edinburgh
+##  EH9 3JT
+##  Edinburgh, UK
+##
+#####################################################################################
+
+
+import numpy as np
+from Bio.Seq import Seq
+from Bio.Alphabet import IUPAC
+import re
+from datetime import datetime as dt
+import time
+import calendar
+from Bio import SeqIO
+from Bio import AlignIO
+import itertools
+
+def unique(o, idfun=repr):
+    """
+    Returns a list of unique values in a list.
+    """
+    seen = {}
+    return [seen.setdefault(idfun(e),e) for e in o if idfun(e) not in seen]
+
+def column(data,col):
+    """
[…]

### 2014-05-07 scripts/LD_calculator_ChiSqdf_eLife.py

@@ -1,357 +1,0 @@
-#####################################################################################
-##
-##  This script calculates linkage disequilibrium (LD) between sites in different alignments (nucleotide or amino acid).
-##  It is meant to accompany the manuscript "Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex"
-##  and has only been tested with alignments produced specifically for it.
-##  You must have python, numpy and biopython installed on your machine to use this script.
-##
-##  You also require either protein or nucleotide alignments (in nexus format) with the same number of taxa that have the same names.
-##  The sequences used for this project are listed in the acknowledgment table accompanying the manuscript:
-##  https://github.com/evogytis/fluB/tree/master/acknowledgement%20tables
-##
-##  and can be downloaded from GISAID available at:
-##  http://platform.gisaid.org
-##  
-##  open command line, go to folder where the script is located and type in:
-##  "python LD_calculatorChiSqdf_eLife.py" (ignore ")
-##
-##  To switch between nucleotide and amino acid modes comment 
-##
-##  The following figures in the manuscript were made using this script:
-##  Figures 10, S7-S8
-##
-##
-##  Alternatively, instead of running the script the output is available here:
-##  https://github.com/evogytis/fluB/tree/master/data
-##  
-##
-##  Gytis Dudas
-##  Institute of Evolutionary Biology
-##  University of Edinburgh
-##  EH9 3JT
-##  Edinburgh, UK
-##
-#####################################################################################
-
-
-import numpy as np
-from Bio.Seq import Seq
-from Bio.Alphabet import IUPAC
-import re
-from datetime import datetime as dt
-import time
-import calendar
-from Bio import SeqIO
-from Bio import AlignIO
-
-def unique(o, idfun=repr):
-    """
-    Returns a list of unique values in a list.
-    """
-    seen = {}
-    return [seen.setdefault(idfun(e),e) for e in o if idfun(e) not in seen]
-
-def column(data,col):
-    """
-    Returns a list that is a column in a list of lists.
-    """
-    return [row [col] for row in data]
-
-def index(data,item):
-    """
[…]

### 2014-05-08 fluB.tex
Change citation style and text/scripts related to LD.
Changed citation style to MBE-like. Added more information about D' and
why it's horrible for flu sequences.
Also improved efficiency of LD calculator. LD results will be derived
from ~1600 complete influenza B genomes and will stick with statistics
related to r-squared. Might implement minor allele frequency cutoff
though.

@@
 % cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-%\renewcommand\citeleft{(}
-%\renewcommand\citeright{)}
-%\renewcommand\citeform[1]{\textsl{#1}}
+%\usepackage{cite}
+\usepackage{natbib}
+%\renewcommand\citepleft{(}
+%\renewcommand\citepright{)}
+%\renewcommand\citepform[1]{\textsl{#1}}
+
+\usepackage{amsmath}
 
@@
 \section*{Introduction}
-Seasonal influenza causes between 250,000 and 500,000 deaths annually and comprises lineages from three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \cite{flufactsheet}.
-However, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
-Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.
+Seasonal influenza causes between 250,000 and 500,000 deaths annually and comprises lineages from three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \citep{flufactsheet}.
+However, influenza B viruses are increasingly being recognized as important human pathogens \citep{paul-glezen2013}.
+Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \citep{ECDC1213}.
 
@@
 Influenza A viruses are widely considered to be a major threat to human health worldwide due to their ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains.
-Although influenza B viruses have been observed to infect seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis, they are thought to primarily infect humans and are thus are unlikely to exhibit pandemics due to the absence of an animal reservoir from which to acquire antigenic novelty.
-Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{burnet1955,hay2001,bedford2014}.
+Although influenza B viruses have been observed to infect seals \citep{osterhaus2000,bodewes2013} through a reverse zoonosis, they are thought to primarily infect humans and are thus are unlikely to exhibit pandemics due to the absence of an animal reservoir from which to acquire antigenic novelty.
+Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \citep{burnet1955,hay2001,bedford2014}.
 
-Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/2/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}.
-These two lineages now possess antigenically distinct HA surface glycoproteins \cite{kanegae1990,rota1990,nerome1998,nakagawa2002,ansaldi2003} allowing them to co-circulate in the human population.
-Phylogenetic analysis of evolutionary rate, selective pressures and reassortment history of influenza B has shown extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within the Vic and Yam lineages \cite{chen2008}.
+Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/2/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \citep{rota1990}.
+These two lineages now possess antigenically distinct HA surface glycoproteins \citep{kanegae1990,rota1990,nerome1998,nakagawa2002,ansaldi2003} allowing them to co-circulate in the human population.
+Phylogenetic analysis of evolutionary rate, selective pressures and reassortment history of influenza B has shown extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within the Vic and Yam lineages \citep{chen2008}.
 
@@
 Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen, where the two lineages have been isolated at a ratio close to 50\% over long periods of time (Figure \ref{lineageRatiosOverTime}).
-On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically consistent with one lineage predominating within a given season, in agreement with surveillance data \cite{reed2012}.
+On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically consistent with one lineage predominating within a given season, in agreement with surveillance data \citep{reed2012}.
 
@@
 \end{itemize}
-B/Alaska/12/1996-like, B/Nanchang/2/1997-like and B/Iowa/03/2002-like constellations were observed by Chen \& Holmes \cite{chen2008}, but sequences from B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like constellations were not available at the time.
+B/Alaska/12/1996-like, B/Nanchang/2/1997-like and B/Iowa/03/2002-like constellations were observed by Chen \& Holmes \citep{chen2008}, but sequences from \\B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like constellations were not available at the time.
 Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata derived background.
@@
 \subsection*{Analysis of within-lineage reassortment patterns}
-SPR (subtree prune and regraft) provides a measure of tree-to-tree distance, in which distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between them \cite{svinti2013}, as long as each branch experiences, at most, one reassortment.
+SPR (subtree prune and regraft) provides a measure of tree-to-tree distance, in which distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between them \citep{svinti2013}, as long as each branch experiences, at most, one reassortment.
 Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
@@
 
-We normalized our SPR distance comparisons to account for differences in tree topology stability over the course of the MCMC chain and also use an approximation of SPR distance \cite{whidden2009,whidden2010,whidden2013} (see Methods).
+We normalized our SPR distance comparisons to account for differences in tree topology stability over the course of the MCMC chain and also use an approximation of SPR distance \citep{whidden2009,whidden2010,whidden2013} (see Methods).
 The 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments between segments (Figure \ref{SPRdistances}).
@@
 For example, the distribution of comparison between PB1 and PB2 trees has a mean $\dtmrca$$\approx0.4$, indicating that the sum of $\dtmrca$ values from PB1-PB1' and PB2-PB2' comparisons (prime (') indicates an independent MCMC chain) are 40\% of twice the $\dtmrca$ value from the PB1-PB2 comparison, but this value can range anywhere between $\approx$0.15 and $\approx$0.75.
-Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort among themselves unless both reassorting segments have similar TMRCAs, i.e.\ we observe isolation by distance.
+Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort among themselves unless both reassorting segments have similar TMRCAs, i.e.\ we observe isolation by temporal distance.
 In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.
@@
 Linkage disequilibrium (LD) is a measure of non-random association between pairs of alleles at different polymorphic loci within a population.
-To estimate LD no other information, except for allele frequencies at polymorphic amino acid sites, is required.
+To estimate LD no other information, except for haplotype and allele frequencies at polymorphic amino acid sites, is required.
 Although not an external validation of previous phylogenetic results, we observe significantly greater LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}).
@@
 
-When comparing the mean tip-tip TMRCA deviations we find evidence of isolation by distance - when reassortments occur between PB1, PB2 and HA segments those events tend to involve branches that have similar TMRCAs.
+When comparing the mean tip-tip TMRCA deviations we find evidence of isolation by temporal distance - when reassortments occur between PB1, PB2 and HA segments those events tend to involve branches that have similar TMRCAs.
 Other segment pairs, with the exception of MP-NA, seem to show little signal of this and are able to reassort with branches that have been evolving under temporally distant genomic backgrounds.
@@
 \subsection*{Causes of co-dependence between PB1-PB2-HA}
-Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, \textit{e.g.} PB1-PB2-PA and HA-NA \cite{mccullers2004}.
+Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, \textit{e.g.} PB1-PB2-PA and HA-NA \citep{mccullers2004}.
 Though it would be easy to explain co-adaptation between these segments by referring to their functional roles \textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities, our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.
@@
 It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations which did not become fixed in the population and our estimates of the number of reassortments do not differ significantly between all segments (Figure \ref{SPRdistances}).
-The latter is in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
+The latter is in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \citep{marshall2013}.
 In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we offer several alternatives.
@@
 Explaining the co-dependence of PB1+2 and HA segments is more difficult.
-Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \cite{bergeron2010,fulvini2011}.
-A previously used technique for producing vaccine seed strains involved selecting for HA-NA reassortants, which often resulted in PB1-HA-NA reassortants, the second most frequent class of reassortants \cite{bergeron2010,fulvini2011}.
-Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \cite{cobbin2013}.
+Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \citep{bergeron2010,fulvini2011}.
+A previously used technique for producing vaccine seed strains involved selecting for HA-NA reassortants, which often resulted in PB1-HA-NA reassortants, the second most frequent class of reassortants \citep{bergeron2010,fulvini2011}.
+Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \citep{cobbin2013}.
 Additional evidence for a relationship between PB1 and HA segments in influenza A viruses is given by previous influenza pandemics, which were caused by avian-human influenza A virus reassortants.
-It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments derived from avian influenza A viruses \cite{kawaoka1989}.
-However, there have also been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \cite{gregory2002} and H1N1/09 in 2009 \cite{smith2009}.
+It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments derived from avian influenza A viruses \citep{kawaoka1989}.
+However, there have also been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \citep{gregory2002} and H1N1/09 in 2009 \citep{smith2009}.
 
@@
 
-Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
-This is easiest to interpret as frequency-dependent balancing selection acting to preserve antigenic diversity \cite{worobey2014}.
+Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \citep{chen2006,obenauer2006}.
+This is easiest to interpret as frequency-dependent balancing selection acting to preserve antigenic diversity \citep{worobey2014}.
 If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
@@
 In this case PB1, PB2 and HA segments of Vic and Yam lineage accumulate substitutions that improve their ability to co-operate with segments of the same lineage and worse at interacting segments of a different lineage.
-This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \cite{lee2008}.
-It is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \cite{nerome1998,shaw2002}, offering a potential explanation for why the budding Victoria lineage segments were not homogenized via reassortment with Yamagata lineage segment in the early years of the split between the two lineages.
+This process, termed mutation-driven co-evolution \citep{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \citep{lee2008}.
+It is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering a potential explanation for why the budding Victoria lineage segments were not homogenized via reassortment with Yamagata lineage segment in the early years of the split between the two lineages.
 The restriction to eastern Asia would presumably also give sufficient time for PB1, PB2 and HA segments to co-evolve together.
@@
 The association between Victoria and Yamagata lineage PB1-PB2-HA complexes should be relatively straightforward to test in the lab.
-Using previously developed plasmid systems \cite{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
+Using previously developed plasmid systems \citep{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
 We predict that artificially produced viruses with mixed-lineage PB1-PB2-HA complexes will have reduced fitness when compared to viruses with pure-lineage, i.e.\ entirely Vic or entirely Yam, PB1-PB2-HA complexes.
@@
 Ideally, the co-adaptation would be easier to understand by referring to the structures of PB1 and PB2 proteins, as the link between these would be intuitive.
-We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins (Figure \ref{ChiCore}) that fall within the regions that form contacts within the influenza B polymerase heterotrimer \cite{sugiyama2009}, suggesting more subtle roles for sites we have identified.
+We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins (Figure \ref{ChiCore}) that fall within the regions that form contacts within the influenza B polymerase heterotrimer \citep{sugiyama2009}, suggesting more subtle roles for sites we have identified.
 
 \subsection*{The future of influenza B viruses}
-We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \cite{turner2005}.
+We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \citep{turner2005}.
 In this context, we see three potential paths of evolution for influenza B viruses.
@@
 \section*{Methods}
-We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information).
+We compiled a dataset of 452 complete influenza B genomes from GISAID \citep{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information).
 The longest protein coding region of each segment was extracted and used for all further analyses.
 We thus assume that homologous recombination has not taken place and that the evolutionary history of the whole segment can be inferred from the longest coding sequence in the segment.
-To date there has been little evidence of homologous recombination in influenza viruses \cite{chare2003,boni2008,han2010}.
-The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \cite{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
+To date there has been little evidence of homologous recombination in influenza viruses \citep{chare2003,boni2008,han2010}.
+The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \citep{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \citep{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
 Each strain was thus assigned 8 lineages depending on the combination of lineages from which their genomes were derived, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as (V,V,V,V,V,V,V,Y).
 
-We also downloaded all available sequences of influenza B isolates sampled 1984--2013 from GISAID for which PB1, PB2 and HA segments were sequenced.
-This comprised a dataset of 1433 isolates in total which became available only after the primary analyses were performed using the smaller dataset and had too many sequences to analyze using the methods described later.
-Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences, as before.
-Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.
+We also collated all complete influenza B virus genomes available on Genbank as of May 5, 2014.
+After removing isolates that had considerable portions of any sequence missing, were isolated prior to 1980 or were suspected of having a contaminant sequence in any segment, we were left with (probably about 1603 sequences) sequences.
+This dataset only became available after all primary analyses were performed, are mainly from Australia, New Zealand and the United States and are too numerous to analyze in BEAST \citep{drummond2012}.
+PhyML \citep{guindon2003} was used to produce phylogenies of each segment and the lineage of each isolate was determined based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences, as described above.
+By associating strains with lineage identity of each of their segments, we reconstructed the most parsimonious inter-lineage reassortment history.
 
-Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
-Here, we modeled the substitution process using the HKY model of nucleotide substitution \cite{hky1985}, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \cite{obrien2009}.
-We used a flexible Bayesian skyride demographic model \cite{minin2008}.
+Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \citep{drummond2012}.
+Here, we modeled the substitution process using the HKY model of nucleotide substitution \citep{hky1985}, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \citep{obrien2009}.
+We used a flexible Bayesian skyride demographic model \citep{minin2008}.
 We accounted for incomplete sampling dates for 94 sequences (of which 93 had only year and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
-A relaxed molecular clock was used, where branch lengths are drawn from a lognormal distribution \cite{drummond2006}.
+A relaxed molecular clock was used, where branch lengths are drawn from a lognormal distribution \citep{drummond2006}.
 We ran 3 independent MCMC chains, each with 200 million states, sampled every 20,000 steps and discarded the first 10\% of the MCMC states as burn-in.
-After assessing convergence of all 3 MCMC chains by visual inspection using Tracer \cite{tracer}, we combined samples across chains to give a total of 27,000 samples from the posterior distribution of trees.
+After assessing convergence of all 3 MCMC chains by visual inspection using Tracer \citep{tracer}, we combined samples across chains to give a total of 27,000 samples from the posterior distribution of trees.
 
 Every sequence was assigned 7 discrete traits in BEAUti corresponding to the lineages of all other segments with which a strain was isolated \textit{e.g.} PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values.
-We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete states using an asymmetric transition matrix \cite{lemey2009} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates.
+We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete states using an asymmetric transition matrix \citep{lemey2009} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates.
 Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing state transitions, i.e.\ Yam to Vic or Vic to Yam (Figure \ref{methodFig}).
@@
 We inferred the diversity of each segment at a single point in time by estimating the date of the most recent common ancestor of all branches at yearly intervals, which places an upper bound on the maximum amount of diversity existing at each time point.
-A version of this lineage turnover metric has previously been used to investigate the tempo and strength of selection in influenza A viruses during seasonal circulation \cite{bedford2011}.
+A version of this lineage turnover metric has previously been used to investigate the tempo and strength of selection in influenza A viruses during seasonal circulation \citep{bedford2011}.
 In addition, we calculated mean pairwise time of most recent common ancestor (TMRCA) between branches labelled as Vic and Yam for PB1, PB2 and HA.
@@
 Calculating $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
-Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
+Our approach exploits the branch scaling used by BEAST \citep{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
 This method operates under the assumption that the segment trees capture the overall `species tree' (or `virus tree') of influenza B viruses quite well.
@@
 The $\dtmrca$ statistic essentially captures a facet of this influenza B `species tree' by quantifying the temporal distance between an admixture event and the original lineage that was replaced by the incoming lineage.
-Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \cite{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \cite{ruane2011}.
+Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \citep{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \citep{ruane2011}.
 
@@
 Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
-We calculated approximate SPR distances \cite{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
+We calculated approximate SPR distances \citep{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
 Approximate SPR distances $\dspr$ were normalized using the procedure described above, where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from segments $A$, $B$ and independent analyses thereof ($A'$ and $B'$).
@@
 \subsection*{Linkage disequilibrium across the influenza B genome}
-We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
-To quantify LD we adapt the $\chiSq$ statistic from \cite{hedrick1986}:
+We used the expanded Genbank dataset with (probably about 1600 sequences) sequences to estimate linkage disequilibrium (LD) between amino acid loci across the longest proteins encoded by each segment of the influenza B virus genome.
+To quantify LD we adapt the $\chiSq$ statistic from \citep{hedrick1986}:
 
@@
 where $\chi^{2}$ is calculated from a classical contingency table, $N$ is the number of haplotypes and $(k-1)(m-1)$ are the degrees of freedom.
-This statistic is equal to the widely used $r^2$ LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus \cite{zhao2005}.
+This statistic is equal to the widely used $r^2$ LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus \citep{zhao2005}.
 LD was estimated only at loci where each nucleotide or amino acid allele was present in at least two isolates.
@@
 We also calculated mean LD for all pairs of segments to quantify the overall association between them.
-Linkage disequilibrium was not estimated for the larger PB1, PB2 and HA dataset, because it lacked sequences from the rest of the influenza B genome and thus would have failed to put LD between PB1, PB2 and HA proteins into a genome-wide perspective.
+
+We also calculated another LD statistic, D' \citep{lewontin1964}:
+
+\begin{equation}
+D'_{ij}=\frac{D_{ij}}{D_{ij}^{max}},
+\end{equation}
+
+where
+
+\begin{equation}
+D_{ij}=p(A_i B_j) - p(A_i) p(B_j)
+\end{equation}
+
+and
+
+\begin{equation}
+\begin{split}
+D_{ij}^{max}=min[ p(A_i) p(B_j) , (1-p(A_i))(1-p(B_j)) ] \mbox{ when } D_{ij} < 0\\
+D_{ij}^{max}=min[ (1-p(A_i)) p(B_j) , p(A_i)(1-p(B_j)) ] \mbox{ when } D_{ij} \geq 0,
+\end{split}
+\end{equation}
+
+where $p(A_i)$ is the frequency of allele $A_i$ at locus A, $p(B_j)$ is the frequency of allele $B_j$ at locus B and $p(A_i B_j)$ is the frequency of haplotype $A_i B_j$.
+D' is inflated when some haplotypes are not observed \textit{e.g.} when the minor allele frequency is low.
+We find that D' is almost uniformly high across the influenza B virus genome and close to 1.0 for almost any pair of polymorphic loci.
+This is because most amino acid alleles in the dataset exist transiently, meaning that they don't get a chance to reassort and we only observe them within the backgrounds of more persistent alleles, which D' quantifies as complete LD.
+We think that metrics related to $r^2$, like $\chiSq$, perform much better on temporal data such as ours in finding associations between alleles and are easier to interpret.
 
 \section*{Acknowledgements}
-GD was supported by a NERC studentship D76739X.
+GD was supported by a Natural Environment Research Council studentship D76739X.
 TB was supported by a Newton International Fellowship from the Royal Society. 
@@
 
-\bibliographystyle{pnas}
+\bibliographystyle{mbe}
 \bibliography{fluB}
@@
 Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211, 213 and 214 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
-Previous research \cite{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
+Previous research \citep{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
 Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 positions 211 and 214 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at positions 211 and 214 and threonine (T) at position 213.
 It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
-Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \cite{fodor2004}.
-Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \cite{guilligay2008}.}
+Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \citep{fodor2004}.
+Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \citep{guilligay2008}.}
 \label{ChiCore}
@@
 %\section*{Differences between Vic and Yam PB1-PB2-HA}
-%Through the combination of inferred PB1-PB2-HA background and realized synonymous and non-synonymous substitution counts on branches \cite{obrien2009} it was possible to investigate whether Victoria and Yamagata lineage backgrounds differ in substitution rates.
+%Through the combination of inferred PB1-PB2-HA background and realized synonymous and non-synonymous substitution counts on branches \citep{obrien2009} it was possible to investigate whether Victoria and Yamagata lineage backgrounds differ in substitution rates.
 %Figure \ref{robustCounting} shows the inferred synonymous, non-synonymous (both as substitutions per year) and overall nucleotide substitution rate (as substitutions per site per year) under different lineages of PB1-PB2-HA.

### 2014-05-08 scripts/LD_calculator.py

@@ -34,7 +34,7 @@
 ##
 #####################################################################################
 
-
+import sys
 import numpy as np
 from Bio.Seq import Seq
 from Bio.Alphabet import IUPAC
@@ -145,46 +145,53 @@ segments=['PB1','PB2','PA','HA','NP','NA','M1','NS1']
 #mixedLineage=['B/Alaska/03/1992_1992-06-13', 'B/Alaska/03/1992_1992-06-13', 'B/Alaska/03/1992_1992-06-13', 'B/Cordoba/2979/1991_1991-06-13', 'B/Cordoba/2979/1991_1991-06-13', 'B/Cordoba/2979/1991_1991-06-13', 'B/Bangkok/163/1990_1990', 'B/Bangkok/163/1990_1990', 'B/Bangkok/163/1990_1990', 'B/Norway/1/84_1984', 'B/Norway/1/84_1984', 'B/Norway/1/84_1984', 'B/Houston/1/91_1991', 'B/Houston/1/91_1991', 'B/Houston/1/91_1991', 'B/Memphis/5/93_1993', 'B/Memphis/5/93_1993', 'B/Memphis/5/93_1993', 'B/Bangkok/153/1990_1990', 'B/Bangkok/153/1990_1990', 'B/Bangkok/153/1990_1990', 'B/Johannesburg/06/1994_1994', 'B/Johannesburg/06/1994_1994', 'B/Johannesburg/06/1994_1994', 'B/Lisbon/02/1994_1994-10-26', 'B/Lisbon/02/1994_1994-10-26', 'B/Lisbon/02/1994_1994-10-26', 'B/Connecticut/02/1995_1995-01-05', 'B/Connecticut/02/1995_1995-01-05', 'B/Connecticut/02/1995_1995-01-05', 'B/Texas/14/1991_1991-01-11', 'B/Texas/14/1991_1991-01-11', 'B/Texas/14/1991_1991-01-11', 'B/Connecticut/07/1993_1993-01-26', 'B/Connecticut/07/1993_1993-01-26', 'B/Connecticut/07/1993_1993-01-26', 'B/Ann_Arbor/1994_1994', 'B/Ann_Arbor/1994_1994', 'B/Ann_Arbor/1994_1994', 'B/Oita/15/92_1992-10-16', 'B/Oita/15/92_1992-10-16', 'B/Oita/15/92_1992-10-16', 'B/New_York/24/1993_1993-12-08', 'B/New_York/24/
[…]

### 2014-05-09 scripts/LD_calculator.py
Optimize LD calculator further, fix bug.

@@ -152,7 +152,7 @@ mode='aa'
 ## Iterate over all pairs of segments
 for mem1 in range(0,len(segments)):
     #for mem2 in range(mem1,len(segments)):
-    for mem2 in range(0,len(segments)):
+    for mem2 in range(mem1,len(segments)):
         print '\n\nComparing %s and %s'%(segments[mem1],segments[mem2])
 
         ## dictA/B are dicts that map each strain in each time block to that strain's sequence
@@ -253,10 +253,16 @@ for mem1 in range(0,len(segments)):
                     polymorphicLociB.append(j)
 
 
-            ## construct polymorphic site pairs                                       
-            for loc1 in polymorphicLociA:
-                for loc2 in polymorphicLociB:
-                    polymorphicPairs.append((loc1,loc2))
+
+            ## construct polymorphic site pairs, optimize self-comparisons
+            if mem1==mem2:
+                for loc1 in range(len(polymorphicLociA)-1):
+                    for loc2 in range(loc1+1,len(polymorphicLociB)):
+                        polymorphicPairs.append((polymorphicLociA[loc1],polymorphicLociB[loc2]))
+            else:                                                     
+                for loc1 in polymorphicLociA:
+                    for loc2 in polymorphicLociB:
+                        polymorphicPairs.append((loc1,loc2))
 
             print '\nNumber of polymorphic pairs of loci between %s and %s: %s'%(segments[mem1],segments[mem2],len(polymorphicPairs))
 

### 2014-05-10 scripts/LD_calculator.py
Fix nucleotide LD bug.

@@ -235,7 +235,7 @@ for mem1 in range(0,len(segments)):
                 ## ignore invariant sites at locus 1
                 if mode=='aa' and len(removeItem(unique(column(alignmentA[tp],i)),['-','?']))==1:
                     pass
-                elif mode=='nt' and len(removeItem2(unique(column(alignmentA[tp],i)),['A','C','T','G']))==1:
+                elif mode=='nt' and len(removeItem(unique(column(alignmentA[tp],i)),['R','Y','S','W','K','M','B','D','H','H','V','N','-']))==1:
                     pass
                 else:
                     ## log polymorphic loci in alignmentA
@@ -246,7 +246,7 @@ for mem1 in range(0,len(segments)):
                 ## ignore invariant sites at locus 2
                 if mode=='aa' and len(removeItem(unique(column(alignmentB[tp],j)),['-','?']))==1 or (seg1==seg2 and i==j):
                     pass
-                elif mode=='nt' and len(removeItem2(unique(column(alignmentB[tp],j)),['A','C','T','G']))==1 or (seg1==seg2 and i==j):
+                elif mode=='nt' and len(removeItem(unique(column(alignmentB[tp],j)),['R','Y','S','W','K','M','B','D','H','H','V','N','-']))==1 or (seg1==seg2 and i==j):
                     pass
                 else:
                     ## log polymorphic loci in alignmentB

### 2014-05-12 scripts/LD_calculator.py
Fix LD calculations for nucleotides.
Fixes #26.
Earlier changes to text should also fix #23.

@@ -146,8 +146,15 @@ segments=['PB1','PB2','PA','HA','NP','NA','M1','NS1']
 mixedLineage=[]
 
 ## Select mode - nt for nucleotide alignments, aa for amino acid alignments, used later to filter out invalid residues in the alignment
-#mode='nt'
-mode='aa'
+mode='nt'
+#mode='aa'
+
+
+if mode=='nt':
+    print 'Running in nucleotide LD mode'
+elif mode=='aa':
+    print 'Running in amino acid LD mode'
+
 
 ## Iterate over all pairs of segments
 for mem1 in range(0,len(segments)):
@@ -168,22 +175,24 @@ for mem1 in range(0,len(segments)):
         segment1=segments[mem1]
         segment2=segments[mem2]
 
+        ## input and output paths
+        path='/Users/admin/Documents/Viral sequences/InfB reassortment/Analyses/huge LD/'
+        outpath='/Users/admin/Dropbox/'
+
         if mode=='aa':
-            ## Point to folder containing alignments
-            path='/Users/admin/Documents/Viral sequences/InfB reassortment/Analyses/huge LD/'
-            outpath='/Users/admin/Dropbox/'
-            #outpath=path
             ## Define alignment name
             seg1='%s translation alignment.nex'%(segment1)
             seg2='%s translation alignment.nex'%(segment2)
-            f = open(outpath+'%s_%s_ChiSqdf-Dprime.HUGE.csv'%(segment1,segment2),'w')
+            f = open(outpath+'%s_%s_ChiSqdf-Dprime.aa.HUGE.csv'%(segment1,segment2),'w')
             header='time point,time step size,N strains,N haplotypes usable,site 1,site 2,number of polymoprhisms at site 1,number of polymoprhisms at site 2,minor allele freq site 1,minor allele freq site 2,D,|D\'|,Chi,ChiSqDf'
             print>>f,header
         elif mode=='nt':
-            path='/Users/admin/Documents/Viral sequences/InfB reassortment/Alignments/Without ancient seqs and pruned/'
-            seg1='InfB_%s.nex'%(segment1)
-            seg2='InfB_%s.nex'%(segment2)
-            f = open('/Users/admin/Documents/Viral sequences/InfB reassortment/Alignments/Without ancient seqs and pruned/%s_%s_rDChiNtcomparisonNoSelfRemake.csv'%(segment1,segment2),'w')
+            ## Define alignment name
[…]

### 2014-05-14 fluB.tex
Minor text changes, mostly clarifying.
Kind of fixes #25 in saying that NA (+MP) might be the next segments to
be recruited.

@@
 By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between-lineage diversity, which indicate that an inter-lineage reassortment event has taken place (Figure \ref{betweenDiversity}).
-Because we are taking the mean of pairwise comparisons, this method gives a quantitative measure of reassortment-induced loss of diversity between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree, but instead gives the TMRCA of pairs of viruses.
+Because we are taking the mean of pairwise comparisons, this method gives a quantitative measure of reassortment-induced loss of diversity between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree, but can be interpreted as mean coalescence date between Vic and Yam lineages of PB1, PB2 and HA segments in all other trees.
 We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
@@
 Figure \ref{railroadPlot} focuses only on inter-lineage reassortments that have occurred after 1990.
-We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS with prime (') indicating independently acquired segments) circulating between 1992 and 2011 (Figure \ref{railroadPlot}):
+We identify 5 major (in terms of persistence) reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS with prime (') indicating independently acquired segments) circulating between 1992 and 2011 (Figure \ref{railroadPlot}):
 \begin{itemize}
@@
 This is the speciation scenario.
-The two alternatives to this are the `\textit{status quo}' model, where the influenza B genome continues to be homogenized via gene flow with the exception of PB1, PB2 and HA segments and the extinction scenario, whereby one of the two PB1-PB2-HA complexes goes extinct, marking the return of single-strain dynamics in the influenza B virus population.
+Due to the rarity of inter-lineage reassortment events it is unclear whether this process is already under way with NA and MP, based on Figures \ref{genomeGrid}, \ref{betweenDiversity} and \ref{deltaTMRCA}, being the next segments to be recruited to PB1-PB2-HA complexes.
+The two alternatives to the speciation hypothesis are the `\textit{status quo}' model, where the influenza B genome continues to be homogenized via gene flow with the exception of PB1, PB2 and HA segments and the extinction scenario, whereby one of the two PB1-PB2-HA complexes goes extinct, marking the return of single-strain dynamics in the influenza B virus population.
 

### 2014-05-14 fluB.tex
Add link to GH repo for data availability.
Fix #20 for now.

@@
 This is because most amino acid alleles in the dataset exist transiently, meaning that they don't get a chance to reassort and we only observe them within the backgrounds of more persistent alleles, which D' quantifies as complete LD.
-We think that metrics related to $r^2$, like $\chiSq$, perform much better on temporal data such as ours in finding associations between alleles and are easier to interpret.
+We think that metrics related to $r^2$, like $\chiSq$, perform much better on temporal data such as ours in finding persistent associations between alleles and are easier to interpret.
+
+\subsection*{Data availability}
+Python scripts used to process trees and sequences, as well as their output are publicly available at:
+\\https://github.com/evogytis/fluB/tree/master/data.
 

### 2014-05-15 fluB.tex
Fix #12
Not sure what else we can do to address this point. We're saying these
3 segments will probably never reassort apart and the reviewers are
saying they still might do.
Any discussion over what "co-adapted" or  "co-evolved" mean were
already addressed by e5285d1438947bd8ecf13871f4ac67acc9909d32.

@@
 A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has originally been derived from the Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
-Overall, we see that successful inter-lineage reassortment events do not break up Vic and Yam PB1-PB2-HA complexes, but any other segments reassorted into one or the other PB1-PB2-HA background evolve with a considerable degree of autonomy.
+Overall, we see that successful inter-lineage reassortment events do not break up Vic and Yam PB1-PB2-HA complexes. 
+This is a very unlikely outcome -- if `fit' reassortant lineages were generated at random the probability of not picking any 3 segments across 5 reassortment events is $p = \binom{8}{3} \times (\frac{2^{5} \times 2 - 2}{2^{8} - 2})^{5} = 0.0485$..
+Other segments, however, appear to evolve with a considerable degree of autonomy.
 

### 2014-05-15 fluB.tex
Add extra acknowledgments.
Darren now owes me a beer :D
Clarify null probability calculation per #12.

--- 2014-05-14 fluB.tex
@@
 A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has originally been derived from the Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
-Overall, we see that successful inter-lineage reassortment events do not break up Vic and Yam PB1-PB2-HA complexes, but any other segments reassorted into one or the other PB1-PB2-HA background evolve with a considerable degree of autonomy.
+We observe that in 5 successful inter-lineage reassortment events, none break up the PB1-PB2-HA complex.
+This is an unlikely outcome -- the probability of not breaking up PB1-PB2-HA across 5 reassortment events is $p = (\frac{2^{5} \times 2 - 2}{2^{8} - 2})^{5} = 0.0009$, where reassortment events are considered to sample from the Vic and Yam lineages at random for each of the 8 segments.
+If we correct for multiple testing with the assumption that co-assortment of any 3 segments is of interest we find that the probability of not breaking up an arbitrary 3 segments across 5 reassortment events is $p = \binom{8}{3} \times (\frac{2^{5} \times 2 - 2}{2^{8} - 2})^{5} = 0.0485$.
 
@@
 \section*{Acknowledgements}
+We would like to thank Darren Obbard and Paul Wikramaratna for helpful discussions.
 GD was supported by a Natural Environment Research Council studentship D76739X.

### 2014-05-16 fluB.tex
Moving SPR stuff into supplementary information.
Also added a new methods figure that uses the species tree analogy to
show how what deltaTMRCA calculates.

@@
 \end{itemize}
-B/Alaska/12/1996-like, B/Nanchang/2/1997-like and B/Iowa/03/2002-like constellations were observed by Chen \& Holmes \citep{chen2008}, but sequences from \\B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like constellations were not available at the time.
+B/Alaska/12/1996-like, B/Nanchang/2/1997-like and B/Iowa/03/2002-like constellations were observed by \citet{chen2008}, but sequences from \\B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like constellations were not available at the time.
+In their study \citet{chen2008} also recovered the co-assortment pattern of PB1, PB2 and HA lineages, but did not remark upon it.
 Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata derived background.
@@
 
-\subsection*{Analysis of within-lineage reassortment patterns}
-SPR (subtree prune and regraft) provides a measure of tree-to-tree distance, in which distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between them \citep{svinti2013}, as long as each branch experiences, at most, one reassortment.
-Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
-If there are biases in the way segments reassort, so that some segments tend to co-assort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
-In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the overall (\textit{i.e} both within and between lineages) reassortment rate.
-
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedApproxSPR.png}
-	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
-Following the normalization procedure approximate SPR distances are similar across all pairwise comparisons.
-We interpret this as lack of evidence for small-scale topological similarities between trees of all segments, which we expect to arise if any two segments were being co-packaged and co-reassorted.
-All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
-	\label{SPRdistances}
-\end{figure}
-
-We normalized our SPR distance comparisons to account for differences in tree topology stability over the course of the MCMC chain and also use an approximation of SPR distance \citep{whidden2009,whidden2010,whidden2013} (see Methods).
-The 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments between segments (Figure \ref{SPRdistances}).
-Reassortment rate given as number of SPR moves per total time in both trees shows similar results (Figure \ref{NormSPR_RErate}).
-We note, however, that because of phylogenetic uncertainty our estimate of SPR distance might simply lack power.
-Comparisons between independent analyses of the same segments yield distances that are comparable to distances between different segments (Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
-Still, we find that independent replicates from the same segment (Figure \ref{SPRdistancesReplicates}) show lower SPR distances that comparisons between segments (Figure \ref{SPRdistancesTrees}), suggesting that phylogenetic noise is not completely overwhelming reassortment signal.
-In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
-Thus we caution against over-interpreting Figure \ref{SPRdistances}.
-Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_difference}).
-
-Although we do not find evidence of differences in numbers of overall reassortments between segments, we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
-Unlike SPR distances, which are counts of the numbers of reassortment events, $\dtmrca$ takes time into account, and is most sensitive when only one of the two trees being compared loses diversity via reassortment.
-Similarly to SPR distances, we normalized our $\dtmrca$ comparisons to account for node date stability over the course of the MCMC chain (see Methods).
+\subsection*{Analysis of reassortment properties}
+We attempted to quantify the temporal discordance between lineages reassorting into new genomic constellations.
+If one were able to recover an influenza `species tree', including admixture/reassortment events, it would be possible to estimate the reassortment or recombination `distance', which is the time between a split in the species tree in the past and a reassortment event (see Figure \ref{speciesTree}).
+Although we do not find evidence of differences in numbers of overall reassortments between segments (see Supplementary information), we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
+$\dtmrca$ is most sensitive when only one of the two trees being compared loses diversity via reassortment.
+We normalize our $\dtmrca$ comparisons to account for alignment length induced uncertainty in tree topology stability over the course of the MCMC chain (see Methods).
 Figure \ref{deltaTMRCA} shows normalized mean $\dtmrca$ values for all pairs of trees.
@@
 We propose that this pattern of coassortment is due to the action of selection and not simply biased reassortment.
-One could image that due to some artefact of segment packaging PB1, PB2 and HA segments preferentially co-assort together and thus have a reduced reassortment rate when compared to the rest of the genomic segments.
-We find no evidence of this -- PB1, PB2 and HA segments reassort to the same extent as other segments (Figure \ref{SPRdistances}) and instead only reassortments combining lineages across greater temporal distances are disfavoured (Figure \ref{deltaTMRCA}), which does not seem to drive the overall reassortment rate down.
-If PB1, PB2 and HA trees were embedded within a `species tree' we would see many reassortments within both major branches of this tree -- the Vic and Yam lineages -- but not between them.
 
@@
 Though it would be easy to explain co-adaptation between these segments by referring to their functional roles \textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities, our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.
-
-It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations which did not become fixed in the population and our estimates of the number of reassortments do not differ significantly between all segments (Figure \ref{SPRdistances}).
-The latter is in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \citep{marshall2013}.
-In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we offer several alternatives.
 
@@
 Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
-We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (Figures \ref{NormSPR_PB2-HA_difference}).
+We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (Figures \ref{NormSPR_PB1-PB2_difference} -- \ref{NormSPR_PB2-HA_difference}).
 The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
@@
 \subsection*{Tree to tree similarities}
-
 We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ by comparing TMRCA differences between pairs of tips in the two trees.
@@
 
+\begin{figure}[h]
+	\centering	
+	\includegraphics[width=0.85\textwidth]{supp_figures/InfB_methodSpeciesTree.pdf}
+	\caption{\textbf{Calculating $\dtmrca$ from a species tree perspective.}
+Consider an organism that has diverged into 3 taxa (A, B, C) with a genome comprised of 3 segments (blue, green and red).
+Due to reassortments taxa A and B share a slightly more recent TMRCA in the red segment, likewise for taxa B and C in the green segment.
+By comparing differences in TMRCAs between taxa A-B, A-C and B-C in blue, red and green segments we would find that the red segment has a lower `reassortment distance' (X) than the green segment (Y).
+}
+	\label{speciesTree}
+\end{figure}
+
 Calculating $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
@@
 This method operates under the assumption that the segment trees capture the overall `species tree' (or `virus tree') of influenza B viruses quite well.
-It is not an unreasonable assumption, given the seasonal bottlenecks of these viruses.
-The $\dtmrca$ statistic essentially captures a facet of this influenza B `species tree' by quantifying the temporal distance between an admixture event and the original lineage that was replaced by the incoming lineage.
+It is not an unreasonable assumption, given the seasonal bottlenecks influenza viruses experience.
+This makes it almost certain that influenza viruses circulating at any given time point are derived from a single genome that existed in the recent past.
+The $\dtmrca$ statistic essentially captures a facet of this influenza B `species tree' by quantifying the temporal distance between an admixture event and the original lineage that was replaced by the incoming lineage (see Figure \ref{speciesTree}).
 Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \citep{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \citep{ruane2011}.
-
-Subtree prune and regraft (SPR) distances between phylogenetic trees are an approximate measure of the numbers of reassortment or recombination events.
-Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
-We calculated approximate SPR distances \citep{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
-Approximate SPR distances $\dspr$ were normalized using the procedure described above, where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from segments $A$, $B$ and independent analyses thereof ($A'$ and $B'$).
 
@@
 
+%%% TITLE %%%
+\title{\vspace{1.0cm} \huge \bf Supplemental information:\\ \LARGE Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex}
+
+\author[1]{Gytis Dudas}
+\author[2]{Trevor Bedford}
+\author[1,3]{Samantha Lycett}
+\author[1,4,5]{Andrew Rambaut}
+
+\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
+\affil[2]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
+\affil[3]{Institute of Biodiversity Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK}
+\affil[4]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
+\affil[5]{Centre for Immunology, Infection and Evolution at the University of Edinburgh, Edinburgh, UK}
+
+\date{\today}
+
+\maketitle
+
 \setcounter{figure}{0}
@@
 \renewcommand{\thetable}{S\arabic{table}}
+
+\subsection*{Analysis of within-lineage reassortment patterns}
+Subtree prune and regraft (SPR) distances between phylogenetic trees are an approximate measure of the numbers of reassortment or recombination events \citep{svinti2013}.
+Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
+We calculated approximate SPR distances \citep{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
+Approximate SPR distances were normalized using the procedure used to normalize $\dspr$ described in Methods:
+\begin{equation}
+\dspr(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
+\end{equation}
+where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from segments $A$, $B$ and independent analyses thereof ($A'$ and $B'$).
+Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization.
+If there are biases in the way segments reassort, so that some segments tend to co-assort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
+In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the overall (\textit{i.e} both within and between lineages) reassortment rate.
+
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedApproxSPR.png}
+	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
+Following the normalization procedure approximate SPR distances are similar across all pairwise comparisons.
+We interpret this as lack of evidence for small-scale topological similarities between trees of all segments, which we expect to arise if any two segments were being co-packaged and co-reassorted.
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+	\label{SPRdistances}
+\end{figure}
+
+The 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments between segments (Figure \ref{SPRdistances}).
+Reassortment rate given as number of SPR moves per total time in both trees shows similar results (Figure \ref{NormSPR_RErate}).
+This is in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \citep{marshall2013}.
+We note, however, that because of phylogenetic uncertainty our estimate of SPR distance might simply lack power.
+Comparisons between independent analyses of the same segments yield distances that are comparable to distances between different segments (Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
+Still, we find that independent replicates from the same segment (Figure \ref{SPRdistancesReplicates}) show lower SPR distances that comparisons between segments (Figure \ref{SPRdistancesTrees}), suggesting that phylogenetic noise is not completely overwhelming reassortment signal.
+In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
+Thus we caution against over-interpreting Figure \ref{SPRdistances}.
+Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_difference}).
 

### 2014-05-16 scripts/Hydra_eLife.py
Check out what more conservative SPR distances do.
Basically reports SPR distance only for clades with 2 or more taxa.
Just in case singletons were the cause of most noise previously.

@@ -1062,11 +1062,17 @@ def parseTreeFile(treefile1,treefile2,repfile1,repfile2,modes,additionalInfo):
                     oot=subprocess.check_output([mainCommand],shell=True)
                     ootCerberus=re.search('approx drSPR\=([0-9]+)',oot.split('\n')[-3])
 
+                    AB_clusters2=[x for x in oot.split('\n')[-4].split(' ')[2:] if x.count(',')>0]
+
                     normOot1=subprocess.check_output([normalize1],shell=True)
                     norm1Cerberus=re.search('approx drSPR\=([0-9]+)',normOot1.split('\n')[-3])
 
+                    AA_clusters2=[x for x in normOot1.split('\n')[-4].split(' ')[2:] if x.count(',')>0]
+
                     normOot2=subprocess.check_output([normalize2],shell=True)
                     norm2Cerberus=re.search('approx drSPR\=([0-9]+)',normOot2.split('\n')[-3])
+
+                    BB_clusters2=[x for x in normOot2.split('\n')[-4].split(' ')[2:] if x.count(',')>0]
 
                 elif approximation==0:
                     mainCommand='echo \"%s\n%s\" | %srspr'%(string1,string2,rsprPath)
@@ -1094,12 +1100,18 @@ def parseTreeFile(treefile1,treefile2,repfile1,repfile2,modes,additionalInfo):
                     if ootCerberus is None:
                         print 'Output not caught by regex:'
                         print oot.split('\n')
-                    mm=float(ootCerberus.group(1))
-                    n1=float(norm1Cerberus.group(1))
-                    n2=float(norm2Cerberus.group(1))
+##                    mm=float(ootCerberus.group(1))
+##                    n1=float(norm1Cerberus.group(1))
+##                    n2=float(norm2Cerberus.group(1))
+##                    normA=float(n1+n2)/float(mm*2)
+##                    normB=float((n1*n2)**0.5)/float(mm)
+
+                    mm=float(len(AB_clusters2))
+                    n1=float(len(AA_clusters2))
+                    n2=float(len(BB_clusters2))
                     normA=float(n1+n2)/float(mm*2)
[…]

### 2014-05-19 fluB.tex
Add extra information from the 1600 dataset.
Mention B/Malaysia/1829782/2007-like virus, clarify that LD data is now
from the 1600 dataset (and use it in places where available), revise
some LD findings, add more to the description of LineageRatiosOverTime
figure.

@@
 \usepackage{graphicx}
-\DeclareGraphicsExtensions{.pdf,.png,.jpg}
+\usepackage{grffile}
+%\DeclareGraphicsExtensions{.pdf,.png,.jpg}
 
@@
 These reductions in diversity represent small clades with reassortant PB1-PB2-HA constellations, which go extinct by 1997.
-We also observe that the assignment of Vic or Yam lineages of PB1, PB2 and HA segments to branches of other segment trees is very similar and often identical.
-This results in PB1, PB2 and HA lineage labels switching between Vic and Yam simultaneously in all trees after 1997, suggesting co-reassortment of Vic and Yam lineages of PB1, PB2 and HA segments.
+We also observe that the assignment of Vic or Yam lineages of PB1, PB2 and HA segments to branches of other segment trees is very similar and often identical after 1997.
+This suggests that PB1, PB2 and HA lineage labels are switching between Vic and Yam simultaneously in all trees after 1997.
 
@@
 The ratio of sequences that derive from the original Victoria clade (yellow) to sequences that derive from the Yamagata clade (purple) in each segment over time.
-Numbers at the top of the figure show the total number of genomes analyzed within each influenza season.
-Numbers at the bottom are influenza seasons from the 1984/1985 (84/85) season to the 2011/2012 season.
+Black lines indicate where this ratio lies in the larger 1600 genomes dataset.
+Numbers at the top of the figure show the total number of genomes analyzed within each influenza season in the main dataset comprised of 452 genomes, while the numbers in brackets correspond to numbers of sequences in the larger 1600 genomes dataset.
+Numbers at the bottom are influenza seasons from the 1987/1988 (87/88) season to the 2011/2012 season.
 Yamagata lineage PA, NP, NA and MP segments and Victoria lineage NS segment eventually become fixed in the influenza B population.
@@
 We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bangkok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
-An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating the larger dataset with only PB1, PB2 and HA sequences.
-In all cases, PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
-In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each represented by two isolates), exhibit poor sampling and short circulation times.
-In the case of B/Nanchang/630/1994-like viruses this is up to two years.
-In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bangkok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past: up to 4 years for B/Bangkok/163/1990-like, although the more recently isolated B/Waikato/6/2005-like viruses circulated for only 5 months.
+We detected two new reassortant lineages when investigating the larger (1600 genomes) dataset -- B/Waikato/6/2005-like viruses with PB1-PB2-HA constellation YYV (17 sequences isolated 9 May -- 12 October in 2005) and B/Malaysia/1829782/2007 with PB1-PB2-HA constellation YVY (1 sequence isolated 2 August 2007).
 
@@
 	\centering		
-	\includegraphics[width=0.65\textwidth]{figures/InfB_stateTime.png}
+	\includegraphics[width=0.65\textwidth]{figures/InfB_stateTimeAlt.png}
 	\caption{\textbf{Amount of evolutionary time each segment has spent under different PB1-PB2-HA constellations.}
@@
 Although we do not find evidence of differences in numbers of overall reassortments between segments (see Supplementary information), we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
-$\dtmrca$ is most sensitive when only one of the two trees being compared loses diversity via reassortment.
+$\dtmrca$ is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
 We normalize our $\dtmrca$ comparisons to account for alignment length induced uncertainty in tree topology stability over the course of the MCMC chain (see Methods).
@@
 For example, the distribution of comparison between PB1 and PB2 trees has a mean $\dtmrca$$\approx0.4$, indicating that the sum of $\dtmrca$ values from PB1-PB1' and PB2-PB2' comparisons (prime (') indicates an independent MCMC chain) are 40\% of twice the $\dtmrca$ value from the PB1-PB2 comparison, but this value can range anywhere between $\approx$0.15 and $\approx$0.75.
-Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort among themselves unless both reassorting segments have similar TMRCAs, i.e.\ we observe isolation by temporal distance.
-In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.
 
@@
 We conclude that Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions.
-Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
+Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 4 sites on PB1, 4 on PB2 and 4 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
 Figure \ref{ChiGenome} also shows that the number of variable amino acid sites in PB1 and PB2 proteins is comparable to most other proteins investigated (see Methods for cut-offs used to exclude sites from LD analysis).
@@
 	\centering	
-	\includegraphics[width=0.65\textwidth]{figures/InfB_aaMeanLD.png}
+	\includegraphics[width=0.65\textwidth]{supp_figures/InfB_LDcircle.1600.aa.ChiSqdf.minorCutoff1percent.png}
 	\caption{\textbf{LD comparison between influenza B proteins.}
 Pairwise comparisons of linkage disequilibrium between amino acid sites on influenza B proteins.
-PB1, PB2 and HA proteins exhibit high mean amino acid LD between themselves.
+Many polymorphic amino acid sites on PB1, PB2 and HA proteins exhibit high LD between themselves, followed by the NA protein.
 This is evidence of a considerable number of co-assorting alleles within these proteins.}
@@
 In addition, most strains with mixed-lineage PB1-PB2-HA complexes occurred in the early years of the Vic--Yam split, when the two lineages were presumably much more similar at the nucleotide and amino acid levels.
-The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.
+The most recent influenza B viruses with mixed-lineage PB1-PB2-HA complexes are B/Waikato/6/2005-like and B/Malaysia/1829782/2007-like viruses which circulated for only 5 months or are known from a single isolate, respectively.
 
@@
 Our main dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bangkok/163/1990-like genome constellations.
-The larger dataset with 1433 PB1, PB2 and HA sequences has an additional example of this: isolates with B/Waikato/6/2005-like PB1-PB2-HA constellations.
-We note that in both cases the genome constellations of these reassortants were not fixed in the influenza B population and the more recent B/Waikato/6/2005-like reassortants persisted for a much shorter period of time (5 months versus 4 years), suggestive of declining PB1+2/HA reassortant fitness over time.
+The larger dataset with 1600 complete genomes has two additional examples of this: isolates with B/Waikato/6/2005-like and B/Malaysia/1829782/2007-like PB1-PB2-HA constellations.
+We note that in all cases the genome constellations of these reassortants were not fixed in the influenza B population and the more recent B/Waikato/6/2005-like reassortants persisted for a much shorter period of time (5 months versus 4 years), suggestive of declining PB1+2/HA reassortant fitness over time.
 
@@
 
-Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
+Given the short circulation times of recent PB1-PB2-HA reassortants B/Waikato/6/2005-like and B/Malaysia/1829782/2007-like viruses we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
 We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (Figures \ref{NormSPR_PB1-PB2_difference} -- \ref{NormSPR_PB2-HA_difference}).
@@
 \subsection*{Linkage disequilibrium across the influenza B genome}
-We used the expanded Genbank dataset with (probably about 1600 sequences) sequences to estimate linkage disequilibrium (LD) between amino acid loci across the longest proteins encoded by each segment of the influenza B virus genome.
+We used the expanded Genbank dataset with approximately 1600 complete genome sequences to estimate linkage disequilibrium (LD) between amino acid loci across the longest proteins encoded by each segment of the influenza B virus genome.
 To quantify LD we adapt the $\chiSq$ statistic from \citep{hedrick1986}:
@@
 \centering  
-\includegraphics[width=0.95\textwidth]  {supp_figures/aaChi_Genome.png}
+\includegraphics[width=0.95\textwidth]  {supp_figures/InfB_LDheatmap.1600.aa.ChiSqdf.minorCutoff1percent.png}
 \caption{\textbf{Heatmap of genome-wide linkage disequilibrium ($\chiSq$) between polymorphic amino acid sites.}
@@
 Numbers next to each row and column correspond to amino acid site number within a given protein starting from methionine.
-Amino acid sites exhibiting reciprocally high LD between PB1, PB2 and HA proteins are: 176, 211, 213, 214, 508 (PB1), 115, 301, 382, 383, 397, 468, 483, 639 (PB2) and 86, 91, 96, 141, 152, 163, 164, 183, 190, 218, 270, 277, 282 (HA).
-Sites 211, 213 and 214 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
-Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211, 213 and 214 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
+Amino acid sites exhibiting reciprocally high LD between PB1, PB2 and HA proteins are: 176, 211, 213, 508 (PB1), 115, 399, 470, 642 (PB2) and 96, 163, 520 and 570 (HA).
+Sites 211 and 213 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
+Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211 and 213 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
 Previous research \citep{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
-Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 positions 211 and 214 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at positions 211 and 214 and threonine (T) at position 213.
+Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 position 211 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at position 211 and threonine (T) at position 213.
 It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
 Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \citep{fodor2004}.
-Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \citep{guilligay2008}.}
+Similarly, site 399 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \citep{guilligay2008}.}
 \label{ChiCore}

### 2014-05-21 fluB.tex
Hopefully final draft for MBE submission.
Some minor text changes, correcting font sizes in one of the figures.

@@
 % Remove brackets from numbering in list of References
-\renewcommand\refname{\large References}
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
+%\renewcommand\refname{\large References}
+%\makeatletter
+%\renewcommand{\@biblabel}[1]{\quad#1.}
+%\makeatother
 
@@
 Seasonal influenza causes between 250,000 and 500,000 deaths annually and comprises lineages from three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \citep{flufactsheet}.
-However, influenza B viruses are increasingly being recognized as important human pathogens \citep{paul-glezen2013}.
-Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \citep{ECDC1213}.
+However, influenza B viruses are increasingly being recognized as important human pathogens \citep{paul-glezen2013} and occasionally come to predominate an influenza season.
+For example in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \citep{ECDC1213}.
 
@@
 Influenza A viruses are widely considered to be a major threat to human health worldwide due to their ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains.
-Although influenza B viruses have been observed to infect seals \citep{osterhaus2000,bodewes2013} through a reverse zoonosis, they are thought to primarily infect humans and are thus are unlikely to exhibit pandemics due to the absence of an animal reservoir from which to acquire antigenic novelty.
+Although influenza B viruses have been observed to infect seals \citep{osterhaus2000,bodewes2013} through a reverse zoonosis, they are thought to primarily infect humans and are thus unlikely to exhibit pandemics due to the absence of an animal reservoir from which to acquire antigenic novelty.
 Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \citep{burnet1955,hay2001,bedford2014}.
@@
 Here, we extend previous methods to reveal an evolutionarily intriguing pattern of reassortment in influenza B.
-In our approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual viruses in the tree of the other segments.
+In our approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual isolates in the tree of the other segments.
 By modelling the transition between labels on a phylogenetic tree, reassortment events which result in the replacement of one segment's lineage by another show up as label changes along a branch (Figure \ref{methodFig}).
@@
 By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between-lineage diversity, which indicate that an inter-lineage reassortment event has taken place (Figure \ref{betweenDiversity}).
-Because we are taking the mean of pairwise comparisons, this method gives a quantitative measure of reassortment-induced loss of diversity between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree, but can be interpreted as mean coalescence date between Vic and Yam lineages of PB1, PB2 and HA segments in all other trees.
+This method gives a quantitative measure of reassortment-induced loss of diversity between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree, but can be interpreted as mean coalescence date between Vic and Yam lineages of PB1, PB2 and HA segments in all other trees.
 We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
 Losses of diversity (represented by more recent mean pairwise TMRCAs between Vic and Yam labels) in Figure \ref{betweenDiversity} indicate that every segment has reassorted with respect to the Victoria and Yamagata lineages of PB1, PB2 and HA segments.
-However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
+However, we also see that the labels for these 3 segments show reciprocal preservation of diversity after 1997.
 This suggests that after 1997 no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments and their lineage labels only `meet' at the root.
@@
 These reductions in diversity represent small clades with reassortant PB1-PB2-HA constellations, which go extinct by 1997.
-We also observe that the assignment of Vic or Yam lineages of PB1, PB2 and HA segments to branches of other segment trees is very similar and often identical after 1997.
-This suggests that PB1, PB2 and HA lineage labels are switching between Vic and Yam simultaneously in all trees after 1997.
+We also observe that the assignment of these 3 segment labels to branches of other segment trees is very similar and often identical after 1997.
+This suggests that PB1, PB2 and HA lineage labels switch simultaneously in all trees after 1997.
 
@@
 
-We show the ratio of Vic and Yam lineages in our sequence data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on lineage assignment performed earlier.
+We show the ratio of Vic and Yam lineages in our sequence data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on lineage assignment performed earlier (see Methods).
 It is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
@@
 Black lines indicate where this ratio lies in the larger 1600 genomes dataset.
-Numbers at the top of the figure show the total number of genomes analyzed within each influenza season in the main dataset comprised of 452 genomes, while the numbers in brackets correspond to numbers of sequences in the larger 1600 genomes dataset.
+Numbers at the top of the figure show the total number of genomes analyzed within each influenza season in the main dataset comprised of 452 genomes from which the ratio was calculated, while the numbers in brackets correspond to numbers of sequences in the larger 1600 genomes dataset.
 Numbers at the bottom are influenza seasons from the 1987/1988 (87/88) season to the 2011/2012 season.
@@
 In their study \citet{chen2008} also recovered the co-assortment pattern of PB1, PB2 and HA lineages, but did not remark upon it.
-Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata derived background.
+Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like,\\ B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata derived background.
 All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata derived PB1-PB2-HA segments.
@@
 A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has originally been derived from the Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
+
 We observe that in 5 successful inter-lineage reassortment events, none break up the PB1-PB2-HA complex.
@@
 
-The vast majority of influenza B isolates possess either Vic or Yam lineage derived PB1-PB2-HA complexes, although on rare occasions mixed-lineage PB1-PB2-HA constellations emerge.
+Although the vast majority of influenza B isolates possess either Vic or Yam lineage derived PB1-PB2-HA complexes, on rare occasions mixed-lineage PB1-PB2-HA constellations emerge.
 Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
@@
 All segments have spent significantly more of their history with entirely Vic or entirely Yam-derived PB1-PB2-HA complexes.
-All horizontal lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
 	\label{stateTime}
@@
 Although we do not find evidence of differences in numbers of overall reassortments between segments (see Supplementary information), we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
-$\dtmrca$ is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
+This difference in TMRCAs, or $\dtmrca$, is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
 We normalize our $\dtmrca$ comparisons to account for alignment length induced uncertainty in tree topology stability over the course of the MCMC chain (see Methods).
@@
 PB1, PB2 and HA trees, on the other hand, exhibit normalized $\dtmrca$ values that are much higher.
-This shows that TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
+This shows that TMRCA differences between trees of PB1, PB2 and HA segments are, though noisy, occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
 For example, the distribution of comparison between PB1 and PB2 trees has a mean $\dtmrca$$\approx0.4$, indicating that the sum of $\dtmrca$ values from PB1-PB1' and PB2-PB2' comparisons (prime (') indicates an independent MCMC chain) are 40\% of twice the $\dtmrca$ value from the PB1-PB2 comparison, but this value can range anywhere between $\approx$0.15 and $\approx$0.75.
@@
 To estimate LD no other information, except for haplotype and allele frequencies at polymorphic amino acid sites, is required.
-Although not an external validation of previous phylogenetic results, we observe significantly greater LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}).
+Although not an external validation of previous phylogenetic results, we observe greater LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}) in a large (approximately 1600 genomes) dataset.
 This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
@@
 Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 4 sites on PB1, 4 on PB2 and 4 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
-Figure \ref{ChiGenome} also shows that the number of variable amino acid sites in PB1 and PB2 proteins is comparable to most other proteins investigated (see Methods for cut-offs used to exclude sites from LD analysis).
 These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
-In addition, there are sites on PB1, PB2 and HA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
+In addition, there are sites on PB1, PB2, HA and NA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
 
@@
 The vast majority of the sampled evolutionary history of each segment of influenza B viruses since the split of Vic and Yam lineages has been spent in association with either completely Victoria or completely Yamagata lineage derived PB1-PB2-HA complexes (Figure \ref{stateTime}), suggesting that having `pure' lineage PB1-PB2-HA complexes is important for whole-genome fitness.
-We propose that this pattern of coassortment is due to the action of selection and not simply biased reassortment.
+We propose that this pattern of coassortment is due to the action of selection and not simply biased or rare reassortment.
 
@@
 
-We should clarify that when we say that Vic and Yam PB1-PB2-HA complexes of the influenza B virus are co-adapted and have co-evolved we mean that these 3 segments do not appear to function well when derived from both lineages and that the evolution to this state has been intimate, in the sense that the two lineages have not reassorted across great temporal distances, unlike the rest of the genome.
+We should clarify that when we say that Vic and Yam PB1-PB2-HA complexes of the influenza B virus are co-adapted and have co-evolved we mean that these 3 segments do not appear to function well when derived from both lineages and that the evolution to this state has been intimate, in the sense that there has not been a successful reassortant combining these two lineages across great temporal distances, unlike the rest of the genome.
 
@@
 Indeed, trees of PB1 and PB2 segments exhibit high similarity in tip--tip TMRCAs (Figure \ref{deltaTMRCA}), suggesting highly similar TMRCAs.
-In addition, PB1--PB2 reassortants are the rarest and least persistent among mixed-lineage PB1-PB2-HA strains and have not been isolated since 1996.
+In addition, PB1--PB2 reassortants are the rarest and least persistent among mixed-lineage PB1-PB2-HA strains and have not been isolated in great numbers.
 
@@
 Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \citep{bergeron2010,fulvini2011}.
-A previously used technique for producing vaccine seed strains involved selecting for HA-NA reassortants, which often resulted in PB1-HA-NA reassortants, the second most frequent class of reassortants \citep{bergeron2010,fulvini2011}.
+A previously used technique for producing vaccine seed strains involved selecting for HA-NA reassortants, which often yielded PB1-HA-NA reassortants as a side-effect \citep{bergeron2010,fulvini2011}.
 Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \citep{cobbin2013}.
@@
 Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
-A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{genomeGrid}, \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot})).
-We find this scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
-Our main dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bangkok/163/1990-like genome constellations.
-The larger dataset with 1600 complete genomes has two additional examples of this: isolates with B/Waikato/6/2005-like and B/Malaysia/1829782/2007-like PB1-PB2-HA constellations.
-We note that in all cases the genome constellations of these reassortants were not fixed in the influenza B population and the more recent B/Waikato/6/2005-like reassortants persisted for a much shorter period of time (5 months versus 4 years), suggestive of declining PB1+2/HA reassortant fitness over time.
-
+A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{genomeGrid}, \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot}).
 Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \citep{chen2006,obenauer2006}.
@@
 If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
+
+We find this hitchhiking scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
+Our main dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bangkok/163/1990-like genome constellations.
+The larger dataset with 1600 complete genomes has one additional example of this: isolates with a B/Waikato/6/2005-like PB1-PB2-HA constellation.
+We note that in all cases the genome constellations of these reassortants were not fixed in the influenza B population and the more recent B/Waikato/6/2005-like reassortants persisted for a much shorter period of time (5 months versus 4 years), suggestive of declining PB1+2/HA reassortant fitness over time.
 
@@
 We argue that this is due to selection against viruses with mixed-lineage PB1-PB2-HA complexes.
-Given sufficient time it should become clear whether PB1-PB2-HA complexes of human influenza B viruses are continuing to resist gene flow whilst the rest of the influenza B virus genome is repeatedly homogenized or whether the two PB1-PB2-HA complexes are recruiting the rest of the genomic segments into co-adapted and co-reassorting segment complexes on their way to sympatric speciation.
+Given sufficient time it should become clear whether PB1-PB2-HA complexes of human influenza B viruses are simply resisting gene flow whilst the rest of the influenza B virus genome is repeatedly homogenized or whether the two PB1-PB2-HA complexes are recruiting the rest of the genomic segments into co-adapted and co-reassorting segment complexes on their way to sympatric speciation.
 
@@
 \section*{Methods}
-We compiled a dataset of 452 complete influenza B genomes from GISAID \citep{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information).
+We compiled a dataset of 452 complete influenza B genomes from GISAID \citep{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in Supplementary information).
 The longest protein coding region of each segment was extracted and used for all further analyses.
@@
 We also collated all complete influenza B virus genomes available on Genbank as of May 5, 2014.
-After removing isolates that had considerable portions of any sequence missing, were isolated prior to 1980 or were suspected of having a contaminant sequence in any segment, we were left with (probably about 1603 sequences) sequences.
+After removing isolates that had considerable portions of any sequence missing, were isolated prior to 1980 or were suspected of having a contaminant sequence in any segment, we were left with 1603 sequences.
 This dataset only became available after all primary analyses were performed, are mainly from Australia, New Zealand and the United States and are too numerous to analyze in BEAST \citep{drummond2012}.
@@
 
-Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \citep{drummond2012}.
+Temporally-calibrated phylogenies were recovered for each segment using Markov chain Monte Carlo (MCMC) methods in the BEAST software package \citep{drummond2012}.
 Here, we modeled the substitution process using the HKY model of nucleotide substitution \citep{hky1985}, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \citep{obrien2009}.
@@
 
-Every sequence was assigned 7 discrete traits in BEAUti corresponding to the lineages of all other segments with which a strain was isolated \textit{e.g.} PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values.
+Every sequence was assigned 7 discrete traits in BEAUti corresponding to the lineages of all other segments with which a strain was isolated \textit{e.g.} PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as values for each trait.
 We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete states using an asymmetric transition matrix \citep{lemey2009} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates.
@@
 A version of this lineage turnover metric has previously been used to investigate the tempo and strength of selection in influenza A viruses during seasonal circulation \citep{bedford2011}.
-In addition, we calculated mean pairwise time of most recent common ancestor (TMRCA) between branches labelled as Vic and Yam for PB1, PB2 and HA.
+In addition, we calculated mean pairwise time of most recent common ancestor (TMRCA) between branches labelled as Vic and Yam for PB1, PB2 and HA traits.
 This gave us a measure of how much a particular segment reassorts with respect to Vic and Yam lineages of PB1, PB2 and HA segments.
@@
 We also calculated the total amount of evolutionary time spent by each segment with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments.
-We do this by summing the branch lengths in each tree under 3 (2 in the case of PB1, PB2 and HA trees) different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
+We do this by summing the branch lengths in each tree under 3 different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
 This gives a measure of how successful, over long periods of time, each particular PB1-PB2-HA constellation has been.
@@
 We ignored gaps in the alignment and did not consider them as polymorphisms.
-We also calculated mean LD for all pairs of segments to quantify the overall association between them.
-
+In all cases we used a minor allele frequency cutoff of 1\%.
 We also calculated another LD statistic, D' \citep{lewontin1964}:
@@
 
-\bibliographystyle{mbe}
-\bibliography{fluB}
-
 \clearpage
@@
 We calculated approximate SPR distances \citep{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
-Approximate SPR distances were normalized using the procedure used to normalize $\dspr$ described in Methods:
+Approximate SPR distances were normalized using the procedure used to normalize $\dspr$ (see Methods):
 \begin{equation}
@@
 \centering  
-\includegraphics[width=0.85\textwidth]  {supp_figures/aaChi_PB1_PB2_HA.png}
+\includegraphics[width=0.85\textwidth]  {supp_figures/InfB_LDheatmap.PB1_PB2_HA.1600.aa.ChiSqdf.minorCutoff1percent.png}
 \caption{\textbf{Heatmap of linkage disequilibrium ($\chiSq$) between amino acid sites on PB1, PB2 and HA proteins.}
@@
 
+\bibliographystyle{mbe}
+\bibliography{fluB}
+
 \end{document}

### 2014-05-25 fluB.tex
Address datasets per #27.

@@
 By modelling the transition between labels on a phylogenetic tree, reassortment events which result in the replacement of one segment's lineage by another show up as label changes along a branch (Figure \ref{methodFig}).
-We use this method to reconstruct major reassortment events and quantify reassortment dynamics over time in influenza B viruses.
+We use this method to reconstruct major reassortment events and quantify reassortment dynamics over time in a dataset of 452 influenza B genomes, and conduct secondary analyses in a dataset of 1603 influenza B genomes.
 

### 2014-05-25 fluB.tex
Wording and notation fixes.

@@
 Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
-On some segments, either the Victoria or Yamagata lineage has become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment, seen as modern viruses deriving completely from either Victoria or Yamagata lineage (yellow vs purple bars in Figure \ref{genomeGrid}).
+On some segments, either the Victoria or Yamagata lineage fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment, seen as modern viruses deriving completely from either the Victoria or Yamagata lineage (yellow vs purple bars in Figure \ref{genomeGrid}).
 This pattern is apparent in the PA, NP, NA, MP and NS segments.
@@
 Figure \ref{deltaTMRCA} shows normalized mean $\dtmrca$ values for all pairs of trees.
-Most segment pairs show very low values for this statistic with $\dtmrca\approx$0.1, indicating that $\dtmrca$ measurements between independent analyses of the same segment are up to 10 times smaller than $\dtmrca$ values between analyses of different segments.
+Most segment pairs show very low values for this statistic with $\dtmrca \approx 0.1$, indicating that $\dtmrca$ measurements between replicate posterior samples from the same segment are up to 10 times smaller than $\dtmrca$ values between posterior samples from different segments.
 PB1, PB2 and HA trees, on the other hand, exhibit normalized $\dtmrca$ values that are much higher.
 This shows that TMRCA differences between trees of PB1, PB2 and HA segments are, though noisy, occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
-For example, the distribution of comparison between PB1 and PB2 trees has a mean $\dtmrca$$\approx0.4$, indicating that the sum of $\dtmrca$ values from PB1-PB1' and PB2-PB2' comparisons (prime (') indicates an independent MCMC chain) are 40\% of twice the $\dtmrca$ value from the PB1-PB2 comparison, but this value can range anywhere between $\approx$0.15 and $\approx$0.75.
 
@@
 \subsection*{Tree to tree similarities}
-We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ by comparing TMRCA differences between pairs of tips in the two trees.
-In comparing of trees of segments A and B we calculate
+We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ by averaging distances over posterior samples $(1, \dots, L)$, so that
 \begin{equation}
-\dtmrca(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
+	\dtmrca(A, B) = \frac{1}{L}\sum_{i=1}^L \delta(A_i, B_i).
 \end{equation}
-where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $g(A_{ij},B_{ij})$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree $A$ and the \textit{i}th posterior sample of tree $B$.
-Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
+We calculate $\delta$ for an individual posterior sample following
+\begin{equation}
+\delta(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
+\end{equation}
+where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $n$ is the total number of pairwise comparisons available between sets of tips.
+Thus, $g(A_{ij},B_{ij})$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree $A$ and the \textit{i}th posterior sample of tree $B$.
+Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), to control for variability in tree topology stability over the course of the MCMC chain.
 We had 3 replicate analyses of each segment and in order to calculate $f(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
@@
 In all cases we used a minor allele frequency cutoff of 1\%.
-We also calculated another LD statistic, D' \citep{lewontin1964}:
+We also calculated another LD statistic, $D'$ \citep{lewontin1964}:
 
@@
 where $p(A_i)$ is the frequency of allele $A_i$ at locus A, $p(B_j)$ is the frequency of allele $B_j$ at locus B and $p(A_i B_j)$ is the frequency of haplotype $A_i B_j$.
-D' is inflated when some haplotypes are not observed \textit{e.g.} when the minor allele frequency is low.
-We find that D' is almost uniformly high across the influenza B virus genome and close to 1.0 for almost any pair of polymorphic loci.
-This is because most amino acid alleles in the dataset exist transiently, meaning that they don't get a chance to reassort and we only observe them within the backgrounds of more persistent alleles, which D' quantifies as complete LD.
+$D'$ is inflated when some haplotypes are not observed \textit{e.g.} when the minor allele frequency is low.
+We find that $D'$ is almost uniformly high across the influenza B virus genome and close to 1.0 for almost any pair of polymorphic loci.
+This is because most amino acid alleles in the dataset exist transiently, meaning that they don't get a chance to reassort and we only observe them within the backgrounds of more persistent alleles, which $D'$ quantifies as complete LD.
 We think that metrics related to $r^2$, like $\chiSq$, perform much better on temporal data such as ours in finding persistent associations between alleles and are easier to interpret.

### 2014-05-25 fluB.tex
More small notation fixes.

@@
 In all cases we used a minor allele frequency cutoff of 1\%.
-We also calculated another LD statistic, $D'$ \citep{lewontin1964}:
-
-\begin{equation}
-D'_{ij}=\frac{D_{ij}}{D_{ij}^{max}},
-\end{equation}
-
-where
-
-\begin{equation}
-D_{ij}=p(A_i B_j) - p(A_i) p(B_j)
-\end{equation}
-
-and
-
+We also calculated another LD statistic, $D'$ \citep{lewontin1964} as $D'_{ij}=D_{ij}/D_{ij}^{max}$,
+where $D_{ij}=p(A_i B_j) - p(A_i) p(B_j)$ and
 \begin{equation}
@@
 \end{equation}
-
 where $p(A_i)$ is the frequency of allele $A_i$ at locus A, $p(B_j)$ is the frequency of allele $B_j$ at locus B and $p(A_i B_j)$ is the frequency of haplotype $A_i B_j$.

### 2014-05-27 fluB.tex
Fix deltaTMRCA notation per @evogytis.

@@
 \subsection*{Tree to tree similarities}
-We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ by averaging distances over posterior samples $(1, \dots, L)$, so that
+We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ for a particular posterior sample $i$, following
 \begin{equation}
-	\dtmrca(A, B) = \frac{1}{L}\sum_{i=1}^L \delta(A_i, B_i).
-\end{equation}
-We calculate $\delta$ for an individual posterior sample following
-\begin{equation}
-\delta(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
+\dtmrca(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
 \end{equation}

### 2014-05-29 fluB.tex
Merge by hacking

--- 2014-05-25 fluB.tex
@@
 \section*{Methods}
-We compiled a dataset of 452 complete influenza B genomes from GISAID \citep{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in Supplementary information).
+We compiled a primary dataset of 452 complete influenza B genomes from GISAID \citep{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in Supplementary information).
 The longest protein coding region of each segment was extracted and used for all further analyses.
@@
 
-We also collated all complete influenza B virus genomes available on Genbank as of May 5, 2014.
+We also collated a secondary dataset from all complete influenza B virus genomes available on Genbank as of May 5, 2014.
 After removing isolates that had considerable portions of any sequence missing, were isolated prior to 1980 or were suspected of having a contaminant sequence in any segment, we were left with 1603 sequences.
@@
 PhyML \citep{guindon2003} was used to produce phylogenies of each segment and the lineage of each isolate was determined based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences, as described above.
-By associating strains with lineage identity of each of their segments, we reconstructed the most parsimonious inter-lineage reassortment history.
+By associating strains with lineage identity of each of their segments, we reconstructed the most parsimonious inter-lineage reassortment history for the secondary dataset.
+The secondary dataset was used to check how representative the primary dataset was, to estimate LD and to broadly confirm our results.
+All analyses pertain to the primary dataset unless stated otherwise.
 
-Temporally-calibrated phylogenies were recovered for each segment using Markov chain Monte Carlo (MCMC) methods in the BEAST software package \citep{drummond2012}.
+Temporally-calibrated phylogenies were recovered for each segment in the primary dataset using Markov chain Monte Carlo (MCMC) methods in the BEAST software package \citep{drummond2012}.
 Here, we modeled the substitution process using the HKY model of nucleotide substitution \citep{hky1985}, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \citep{obrien2009}.
@@
 \subsection*{Measures of diversity}
-We inferred the diversity of each segment at a single point in time by estimating the date of the most recent common ancestor of all branches at yearly intervals, which places an upper bound on the maximum amount of diversity existing at each time point.
+We inferred the diversity of each segment from their phylogenetic tree by estimating the date of the most recent common ancestor of all branches at yearly time points, which places an upper bound on the maximum amount of diversity existing at each time point.
 A version of this lineage turnover metric has previously been used to investigate the tempo and strength of selection in influenza A viruses during seasonal circulation \citep{bedford2011}.
@@
 
-We also calculated the total amount of evolutionary time spent by each segment with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments.
+We also calculated the total amount of sampled evolutionary time spent by each segment with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments.
 We do this by summing the branch lengths in each tree under 3 different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
@@
 \subsection*{Linkage disequilibrium across the influenza B genome}
-We used the expanded Genbank dataset with approximately 1600 complete genome sequences to estimate linkage disequilibrium (LD) between amino acid loci across the longest proteins encoded by each segment of the influenza B virus genome.
+We used the secondary Genbank dataset with approximately 1600 complete genome sequences to estimate linkage disequilibrium (LD) between amino acid loci across the longest proteins encoded by each segment of the influenza B virus genome.
 To quantify LD we adapt the $\chiSq$ statistic from \citep{hedrick1986}:

### 2014-05-29 scripts/LD_calculator.py

@@ -134,8 +134,9 @@ def overlap(a,b):
 
 ## Makes a list of dates if partitioning of LD is desired
 # Default value makes a single bin for all sequences isolated between 1900 and 2014
-step=40
-timeline=np.arange(1980,2020,step)
+step=0.25
+windowSize=3
+timeline=np.arange(1984,2014-windowSize,step)
 
 ## Segments to analyze                                                                       
 segments=['PB1','PB2','PA','HA','NP','NA','M1','NS1']
@@ -146,8 +147,8 @@ segments=['PB1','PB2','PA','HA','NP','NA','M1','NS1']
 mixedLineage=[]
 
 ## Select mode - nt for nucleotide alignments, aa for amino acid alignments, used later to filter out invalid residues in the alignment
-mode='nt'
-#mode='aa'
+#mode='nt'
+mode='aa'
 
 
 if mode=='nt':
@@ -176,21 +177,21 @@ for mem1 in range(0,len(segments)):
         segment2=segments[mem2]
 
         ## input and output paths
-        path='/Users/admin/Documents/Viral sequences/InfB reassortment/Analyses/huge LD/'
+        path='/Users/admin/Documents/Viral_sequences/InfB reassortment/Analyses/huge LD/'
         outpath='/Users/admin/Dropbox/'
 
         if mode=='aa':
             ## Define alignment name
             seg1='%s translation alignment.nex'%(segment1)
             seg2='%s translation alignment.nex'%(segment2)
-            f = open(outpath+'%s_%s_ChiSqdf-Dprime.aa.HUGE.csv'%(segment1,segment2),'w')
+            f = open(outpath+'%s_%s_ChiSqdf-Dprime.aa.HUGE_ts3.csv'%(segment1,segment2),'w')
             header='time point,time step size,N strains,N haplotypes usable,site 1,site 2,number of polymoprhisms at site 1,number of polymoprhisms at site 2,minor allele freq site 1,minor allele freq site 2,D,|D\'|,Chi,ChiSqDf'
             print>>f,header
         elif mode=='nt':
             ## Define alignment name
             seg1='%s alignment.nex'%(segment1)
             seg2='%s alignment.nex'%(segment2)
-            f = open(outpath+'%s_%s_ChiSqdf-Dprime.nt.HUGE.csv'%(segment1,segment2),'w')
[…]

### 2014-05-29 fluB.tex
Add more primary/secondary distinctions.
Fix #27
Also move acknowledgment tables to data, add genbank accessions.
Earlier commits more or less fixed #11.

@@
 
-We show the ratio of Vic and Yam lineages in our sequence data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on lineage assignment performed earlier (see Methods).
+We show the ratio of Vic and Yam lineages in our primary and secondary sequence data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on lineage assignment performed earlier (see Methods).
 It is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
@@
 	\caption{\textbf{Ratio of lineages in the dataset.}
-The ratio of sequences that derive from the original Victoria clade (yellow) to sequences that derive from the Yamagata clade (purple) in each segment over time.
-Black lines indicate where this ratio lies in the larger 1600 genomes dataset.
-Numbers at the top of the figure show the total number of genomes analyzed within each influenza season in the main dataset comprised of 452 genomes from which the ratio was calculated, while the numbers in brackets correspond to numbers of sequences in the larger 1600 genomes dataset.
+The ratio of sequences that derive from the original Victoria clade (yellow) to sequences that derive from the Yamagata clade (purple) in each segment from the primary dataset over time.
+Black lines indicate where this ratio lies in the larger secondary dataset.
+Numbers at the top of the figure show the total number of genomes analyzed within each influenza season in the primary dataset comprised of 452 genomes from which the ratio was calculated, while the numbers in brackets correspond to numbers of sequences in the larger secondary genomes dataset.
 Numbers at the bottom are influenza seasons from the 1987/1988 (87/88) season to the 2011/2012 season.
@@
 Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) since 1997 all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).
-We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bangkok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
-We detected two new reassortant lineages when investigating the larger (1600 genomes) dataset -- B/Waikato/6/2005-like viruses with PB1-PB2-HA constellation YYV (17 sequences isolated 9 May -- 12 October in 2005) and B/Malaysia/1829782/2007 with PB1-PB2-HA constellation YVY (1 sequence isolated 2 August 2007).
+We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the primary dataset with the following PB1-PB2-HA constellations: VVY (B/Bangkok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
+We detected two new reassortant lineages when investigating the larger secondary dataset -- B/Waikato/6/2005-like viruses with PB1-PB2-HA constellation YYV (17 sequences isolated 9 May -- 12 October in 2005) and B/Malaysia/1829782/2007 with PB1-PB2-HA constellation YVY (1 sequence isolated 2 August 2007).
 
@@
 Linkage disequilibrium (LD) is a measure of non-random association between pairs of alleles at different polymorphic loci within a population.
-To estimate LD no other information, except for haplotype and allele frequencies at polymorphic amino acid sites, is required.
-Although not an external validation of previous phylogenetic results, we observe greater LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}) in a large (approximately 1600 genomes) dataset.
-This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
+To estimate LD no other information, except for haplotype and allele frequencies at polymorphic amino acid or nucleotide sites, is required.
+Although not an external validation of previous phylogenetic results, we observe greater amino acid LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}) in a large secondary dataset (see Methods).
+This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting non-synonymous alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
 We conclude that Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions.
@@
 	\caption{\textbf{LD comparison between influenza B proteins.}
-Pairwise comparisons of linkage disequilibrium between amino acid sites on influenza B proteins.
+Pairwise comparisons of linkage disequilibrium between amino acid sites on influenza B proteins in the secondary dataset.
 Many polymorphic amino acid sites on PB1, PB2 and HA proteins exhibit high LD between themselves, followed by the NA protein.
@@
 We find this hitchhiking scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
-Our main dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bangkok/163/1990-like genome constellations.
-The larger dataset with 1600 complete genomes has one additional example of this: isolates with a B/Waikato/6/2005-like PB1-PB2-HA constellation.
+Our primary dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bangkok/163/1990-like genome constellations.
+The larger secondary dataset with 1603 complete genomes has one additional example of this: isolates with a B/Waikato/6/2005-like PB1-PB2-HA constellation.
 We note that in all cases the genome constellations of these reassortants were not fixed in the influenza B population and the more recent B/Waikato/6/2005-like reassortants persisted for a much shorter period of time (5 months versus 4 years), suggestive of declining PB1+2/HA reassortant fitness over time.
@@
 \subsection*{Linkage disequilibrium across the influenza B genome}
-We used the secondary Genbank dataset with approximately 1600 complete genome sequences to estimate linkage disequilibrium (LD) between amino acid loci across the longest proteins encoded by each segment of the influenza B virus genome.
+We used the secondary Genbank dataset with 1603 complete genome sequences to estimate linkage disequilibrium (LD) between amino acid loci across the longest proteins encoded by each segment of the influenza B virus genome.
 To quantify LD we adapt the $\chiSq$ statistic from \citep{hedrick1986}:
@@
 We find that $D'$ is almost uniformly high across the influenza B virus genome and close to 1.0 for almost any pair of polymorphic loci.
-This is because most amino acid alleles in the dataset exist transiently, meaning that they don't get a chance to reassort and we only observe them within the backgrounds of more persistent alleles, which $D'$ quantifies as complete LD.
+This is because most amino acid alleles in the population exist transiently, meaning that they don't get a chance to reassort and we only observe them within the backgrounds of more persistent alleles, which $D'$ quantifies as complete LD.
 We think that metrics related to $r^2$, like $\chiSq$, perform much better on temporal data such as ours in finding persistent associations between alleles and are easier to interpret.

### 2014-06-02 fluB.tex
Split text into main and supp, change names.
Readdress the cover letter, add supp info as a separate file for
submission to MBE, change script names to something more generic.

@@
 
-\clearpage
-
-%%% TITLE %%%
-\title{\vspace{1.0cm} \huge \bf Supplemental information:\\ \LARGE Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex}
-
-\author[1]{Gytis Dudas}
-\author[2]{Trevor Bedford}
-\author[1,3]{Samantha Lycett}
-\author[1,4,5]{Andrew Rambaut}
-
-\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
-\affil[2]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
-\affil[3]{Institute of Biodiversity Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK}
-\affil[4]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
-\affil[5]{Centre for Immunology, Infection and Evolution at the University of Edinburgh, Edinburgh, UK}
-
-\date{\today}
-
-\maketitle
-
-\setcounter{figure}{0}
-\setcounter{table}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
-
-\subsection*{Analysis of within-lineage reassortment patterns}
-Subtree prune and regraft (SPR) distances between phylogenetic trees are an approximate measure of the numbers of reassortment or recombination events \citep{svinti2013}.
-Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
-We calculated approximate SPR distances \citep{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
-Approximate SPR distances were normalized using the procedure used to normalize $\dspr$ (see Methods):
-\begin{equation}
-\dspr(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
-\end{equation}
-where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from segments $A$, $B$ and independent analyses thereof ($A'$ and $B'$).
-Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization.
-If there are biases in the way segments reassort, so that some segments tend to co-assort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
-In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the overall (\textit{i.e} both within and between lineages) reassortment rate.
-
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedApproxSPR.png}
-	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
-Following the normalization procedure approximate SPR distances are similar across all pairwise comparisons.
-We interpret this as lack of evidence for small-scale topological similarities between trees of all segments, which we expect to arise if any two segments were being co-packaged and co-reassorted.
-All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
-	\label{SPRdistances}
-\end{figure}
-
-The 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments between segments (Figure \ref{SPRdistances}).
-Reassortment rate given as number of SPR moves per total time in both trees shows similar results (Figure \ref{NormSPR_RErate}).
-This is in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \citep{marshall2013}.
-We note, however, that because of phylogenetic uncertainty our estimate of SPR distance might simply lack power.
-Comparisons between independent analyses of the same segments yield distances that are comparable to distances between different segments (Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
-Still, we find that independent replicates from the same segment (Figure \ref{SPRdistancesReplicates}) show lower SPR distances that comparisons between segments (Figure \ref{SPRdistancesTrees}), suggesting that phylogenetic noise is not completely overwhelming reassortment signal.
-In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
-Thus we caution against over-interpreting Figure \ref{SPRdistances}.
-Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_difference}).
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_normRErate.png}
-\caption{\textbf{Normalized reassortment rate}
-Reassortment rate is calculated as approximate number of SPR moves per sum of total time in both trees.}
-\label{NormSPR_RErate}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_trees.png}
-\caption{\textbf{Approximate SPR distances between all pairs of trees of segments.}
-There is a visible trend where comparisons between shorter segments have larger SPR distances, consistent with decreasing tree topology stability over the course of MCMC for shorter segments.}
-\label{SPRdistancesTrees}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_replicates.png}
-\caption{\textbf{Approximate SPR distances between replicate trees of each segment.}
-Approximate SPR distances between replicates of MP and NS trees are much higher ($\approx$260) than any other segments, suggesting greater variability in tree topology over the course of MCMC.
-SPR distances between replicates of most other segments are $\approx$200.
-}
-\label{SPRdistancesReplicates}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances after normalization.}
-95\% HPD interval (blue) overlaps zero, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
-\label{NormSPR_PB1-PB2_difference}
-\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist2.png}
-%\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances after normalization.}
-%95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
-%\label{NormSPR_PB1-PB2_ratio}
-%\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances after normalization.}
-95\% HPD interval (blue) overlaps zero, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
-\label{NormSPR_PB1-HA_difference}
-\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist2.png}
-%\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances after normalization.}
-%95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
-%\label{NormSPR_PB1-HA_ratio}
-%\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_hist.png}
-\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances after normalization.}
-95\% HPD interval (blue) overlaps zero, suggesting no evidence of differences between approximate and exact SPR distances following normalization.
-Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
-\label{NormSPR_PB2-HA_difference}
-\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_hist2.png}
-%\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after normalization.}
-%95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.
-%Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
-%\label{NormSPR_PB2-HA_ratio}
-%\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.95\textwidth]  {supp_figures/InfB_LDheatmap.1600.aa.ChiSqdf.minorCutoff1percent.png}
-\caption{\textbf{Heatmap of genome-wide linkage disequilibrium ($\chiSq$) between polymorphic amino acid sites.}
-Patterns of LD across the genome suggest a network of reciprocally linked amino acid sites on PB1, PB2, HA and, to some extent NA, proteins.
-Proximity of sites on heatmaps might not correspond to proximity of sites within proteins.}
-\label{ChiGenome}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.85\textwidth]  {supp_figures/InfB_LDheatmap.PB1_PB2_HA.1600.aa.ChiSqdf.minorCutoff1percent.png}
-\caption{\textbf{Heatmap of linkage disequilibrium ($\chiSq$) between amino acid sites on PB1, PB2 and HA proteins.}
-Numbers next to each row and column correspond to amino acid site number within a given protein starting from methionine.
-Amino acid sites exhibiting reciprocally high LD between PB1, PB2 and HA proteins are: 176, 211, 213, 508 (PB1), 115, 399, 470, 642 (PB2) and 96, 163, 520 and 570 (HA).
-Sites 211 and 213 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
-Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211 and 213 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
-Previous research \citep{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
-Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 position 211 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at position 211 and threonine (T) at position 213.
-It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
-Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \citep{fodor2004}.
-Similarly, site 399 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \citep{guilligay2008}.}
-\label{ChiCore}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_robustCounting.png}
-\caption{\textbf{Synonymous, non-synonymous and nucleotide substitution rates in segments under different PB1-PB2-HA complexes.}
-Evolutionary rate dissimilarities under Vic and Yam PB1-PB2-HA complexes are not systematic and appear negligible.
-Synonymous and non-synonymous rates were calculated by dividing the sum of all substitutions of a given class by the total amount of evolutionary time under each PB1-PB2-HA constellation.
-Nucleotide rates were calculated by multiplying the inferred nucleotide substitution rate on each branch by the branch length, then dividing this by the total amount of evolutionary time under each PB1-PB2-HA constellation.
-Vertical bars indicating uncertainty are 95\% HPDs, black bars indicate 95\% HPDs that do not overlap.}
-\label{robustCounting}
-\end{figure}
-
-%Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates} are showing approximate SPR distances prior to the normalization procedure (see Methods).
-%Figure \ref{SPRdistancesTrees} shows that MP and NS trees show greater approximate SPR distances when compared to any other segment tree, which is consistent with the fact that only M1 and NS1 coding sequences, which are shorter than any %other coding sequences, were used to infer the trees.
-%Figure \ref{SPRdistancesReplicates} confirms this by showing greater variability in tree topology (as greater approximate SPR distance) when MP and NS trees are compared against their replicates.
-%For all other segments approximate SPR distances between all pairs of trees prior to normalization suggests that most segment pairs have similar posterior distributions of approximate SPR distances.
-
-%Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates} show $\dtmrca$ statistics for inter-segment and intra-segment, respectively, comparisons prior to normalization.
-%Figure \ref{deltaTMRCAtrees} shows that similarity in TMRCAs (as low $\dtmrca$) between PB1, PB2 and HA segments is due to considerably smaller $\dtmrca$ statistic for the comparison between these segments.
-%It is also noteworthy that NA and MP comparisons have similarly low values for this statistic.
-%Figure \ref{deltaTMRCAreplicates} shows that TMRCAs are quite robust over the course of MCMC and node times do not, even in the most extreme cases, deviate from their replicates by more than 1 year.
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
-\caption{\textbf{Mean $\dtmrca$ between all pairs of trees of segments.}
-Mean $\dtmrca$ between trees of segments reveal that tip pairs in PB1, PB2 and HA trees have very similar TMRCAs.
-The upper tail of the 95\% HPD (HPDs are represented as vertical lines) interval of mean $\dtmrca$ values for PB1-PB2-HA and MP-NA trees do not exceed 3 years.}
-\label{deltaTMRCAtrees}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
-\caption{\textbf{Mean $\dtmrca$ between replicate trees of each segment.}
-Mean $\dtmrca$ values between independent analyses of each segment show that mean $\dtmrca$ values rarely exceed 1 year.}
-\label{deltaTMRCAreplicates}
-\end{figure}
-
-%We also tested the accuracy of approximate SPR distances using exact SPR distances calculated from PB1, PB2 and HA comparisons.
-%Figures \ref{SPR_PB1-PB2_difference}--\ref{SPR_PB2-HA_ratio} show that after normalization the approximate and exact SPR distance estimates are not significantly different.
-%There is also good correlation between the two SPR distances (Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation}). 
-%Due to excessively long computation time of exact SPR distances very few PB2--HA comparisons were made.
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_corr.png}
-%\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances after normalization.}}
-%\label{NormSPR_PB1-PB2_correlation}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_corr.png}
-%\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances after normalization.}}
-%\label{NormSPR_PB1-HA_correlation}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_corr.png}
-%\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances after normalization.}
-%Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
-%\label{NormSPR_PB2-HA_correlation}
-%\end{figure}
-
-%Figures \ref{SPR_PB1-PB2_difference}--\ref{SPR_PB2-HA_ratio} focus on differences between exact and approximate SPR distances and show that prior to normalization approximate SPR distances are overestimates of exact SPR distances.
-%Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation} show that there is a similar degree of correspondence between approximate and exact SPR distances to when approximate and exact SPR distances are normalized (Figures %\ref{NormSPR_PB1-PB2_correlation}--\ref{NormSPR_PB2-HA_correlation}) using the procedure described in Methods.
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist.png}
-%\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances.}}
-%\label{SPR_PB1-PB2_difference}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist2.png}
-%\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances.}}
-%\label{SPR_PB1-PB2_ratio}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist.png}
-%\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances.}}
-%\label{SPR_PB1-HA_difference}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist2.png}
-%\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances.}}
-%\label{SPR_PB1-HA_ratio}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist.png}
-%\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances.}}
-%\label{SPR_PB2-HA_difference}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
-%\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances.}}
-%\label{SPR_PB2-HA_ratio}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
-%\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances.}}
-%\label{SPR_PB1-PB2_correlation}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
-%\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances.}}
-%\label{SPR_PB1-HA_correlation}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB2-HA_corr.png}
-%\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances.}}
-%\label{SPR_PB2-HA_correlation}
-%\end{figure}
-
-%\clearpage
-%\section*{Multidimensional scaling}
-%Figures \ref{MDSaaLD} and \ref{MDSntLD} visualize the relationships between segments based on mean amino acid and nucleotide site linkage disequilibrium estimates, respectively.
-%PB1, PB2 and HA segments appear to be by far the most similar segments, as indicated by their proximity in the MDS embedding.
-%Figures \ref{MDSaaCorr} and \ref{MDSntCorr} compare the observed mean LD values between segments. 
-%Based on mean amino acid site LD (Figure \ref{MDSaaCorr}) only PB1, PB2 and HA segments show considerable clustering, all other segments being equidistant from each other in the MDS embedding.
-%MDS distance versus mean nucleotide site LD scatter plot, on the other hand, shows a continuum of relationships between segments, which is probably caused by more variability at the nucleotide level than the amino acid level.
-%This difference in variability would build up linkage disequilibrium between nucleotide sites and would reduce it via convergent substitutions over relatively short periods of time.
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_aaLD_MDS.png}
-%\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean amino acid linkage disequilibrium ($\chiSq$).}}
-%\label{MDSaaLD}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_aaLD_MatrixMDScorr.png}
-%\caption{\textbf{Correlation between mean amino acid site linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
-%\label{MDSaaCorr}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_ntLD_MDS.png}
-%\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean nucleotide linkage disequilibrium ($\chiSq$).}}
-%\label{MDSntLD}
-%\end{figure}
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_ntLD_MatrixMDScorr.png}
-%\caption{\textbf{Correlation between mean nucleotide site linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
-%\label{MDSntCorr}
-%\end{figure}
-
-%\section*{Differences between Vic and Yam PB1-PB2-HA}
-%Through the combination of inferred PB1-PB2-HA background and realized synonymous and non-synonymous substitution counts on branches \citep{obrien2009} it was possible to investigate whether Victoria and Yamagata lineage backgrounds differ in substitution rates.
-%Figure \ref{robustCounting} shows the inferred synonymous, non-synonymous (both as substitutions per year) and overall nucleotide substitution rate (as substitutions per site per year) under different lineages of PB1-PB2-HA.
-
-%\begin{figure}
-%\centering  
-%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_lineageRatios.png}
-%\caption{\textbf{Total counts and ratios of Victoria and Yamagata lineage sequences in each segment.}}
-%\label{lineageRatios}
-%\end{figure}
-
 \bibliographystyle{mbe}

### 2014-06-02 fluB_supp.tex
Found semi-serious labeling errors.
Got mixed up with PB1/PB2 labels in the subtype maps (occurred quite
early on).
Looks like the right trees were run but PB1 and PB2 labels were
switched. It only affects the visibility of PB1-PB2 reassortants on
trees when using discrete traits though, so nothing has to be redone.

Also including the genomic trees grid figure for supplementary info.
Split text into main and supp, change names.
Readdress the cover letter, add supp info as a separate file for
submission to MBE, change script names to something more generic.

--- 2014-01-30 fluB_supp.tex
@@
 % cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-%\renewcommand\citeleft{(}
-%\renewcommand\citeright{)}
-%\renewcommand\citeform[1]{\textsl{#1}}
+%\usepackage{cite}
+\usepackage{natbib}
+%\renewcommand\citepleft{(}
+%\renewcommand\citepright{)}
+%\renewcommand\citepform[1]{\textsl{#1}}
+
+\usepackage{amsmath}
 
 % Remove brackets from numbering in list of References
-\renewcommand\refname{\large References}
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
+%\renewcommand\refname{\large References}
+%\makeatletter
+%\renewcommand{\@biblabel}[1]{\quad#1.}
+%\makeatother
 
@@
 
+% symbols
+\newcommand{\chiSq}{\chi^{2}_{df}}
+\newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}}
+\newcommand{\dspr}{d_\mathrm{SPR}}
+
+\begin{document}
+
 %%% TITLE %%%
@@
 
-\begin{document}
 \maketitle
 
-\section*{Linkage disequilibrium data}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+
+\subsection*{Analysis of within-lineage reassortment patterns}
+Subtree prune and regraft (SPR) distances between phylogenetic trees are an approximate measure of the numbers of reassortment or recombination events \citep{svinti2013}.
+Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
+We calculated approximate SPR distances \citep{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
+Approximate SPR distances were normalized using the procedure used to normalize $\dspr$ (see Methods):
+\begin{equation}
+\dspr(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
+\end{equation}
+where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from segments $A$, $B$ and independent analyses thereof ($A'$ and $B'$).
+Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization.
+If there are biases in the way segments reassort, so that some segments tend to co-assort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
+In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the overall (\textit{i.e} both within and between lineages) reassortment rate.
 
@@
 	\centering		
-\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB1_PB2.png}
-\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and PB2 segments based on $\chi^{2}_{df}$ values.}}
-\label{ChiPB1PB2}
+	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedApproxSPR.png}
+	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
+Following the normalization procedure approximate SPR distances are similar across all pairwise comparisons.
+We interpret this as lack of evidence for small-scale topological similarities between trees of all segments, which we expect to arise if any two segments were being co-packaged and co-reassorted.
+All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
+	\label{SPRdistances}
 \end{figure}
 
-\begin{figure}[h]
+The 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments between segments (Figure \ref{SPRdistances}).
+Reassortment rate given as number of SPR moves per total time in both trees shows similar results (Figure \ref{NormSPR_RErate}).
+This is in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \citep{marshall2013}.
+We note, however, that because of phylogenetic uncertainty our estimate of SPR distance might simply lack power.
+Comparisons between independent analyses of the same segments yield distances that are comparable to distances between different segments (Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
+Still, we find that independent replicates from the same segment (Figure \ref{SPRdistancesReplicates}) show lower SPR distances that comparisons between segments (Figure \ref{SPRdistancesTrees}), suggesting that phylogenetic noise is not completely overwhelming reassortment signal.
+In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
+Thus we caution against over-interpreting Figure \ref{SPRdistances}.
+Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_difference}).
+
+\begin{figure}
 \centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB1_HA.png}
-\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and HA segments based on $\chi^{2}_{df}$ values.}}
-\label{ChiPB1HA}
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_normRErate.png}
+\caption{\textbf{Normalized reassortment rate}
+Reassortment rate is calculated as approximate number of SPR moves per sum of total time in both trees.}
+\label{NormSPR_RErate}
 \end{figure}
 
-\begin{figure}[h]
+\begin{figure}
 \centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB2_HA.png}
-\caption{\textbf{Hierarchical clustering of amino acid sites on PB2 and HA segments based on $\chi^{2}_{df}$ values.}}
-\label{ChiPB2HA}
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_trees.png}
+\caption{\textbf{Approximate SPR distances between all pairs of trees of segments.}
+There is a visible trend where comparisons between shorter segments have larger SPR distances, consistent with decreasing tree topology stability over the course of MCMC for shorter segments.}
+\label{SPRdistancesTrees}
 \end{figure}
 
-We have identified amino acid sites which exhibit a high degree of linkage disequilibrium between PB1, PB2 and HA segments.
-Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three proteins at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.
-
-None of the sites we have identified on PB1 and PB2 proteins fall within the previously described regions that form the contacts between the two polymerase subunits \cite{sugiyama2009}.
-
-Sites 211, 213 and 214 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
-Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211, 213 and 214 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
-Previous research \cite{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
-Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 positions 211 and 214 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at positions 211 and 214 and threonine (T) at position 213.
-It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
-Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \cite{fodor2004}.
-Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \cite{guilligay2008}.
-Again, we cannot determine whether this finding is significant or not.
-
-\begin{figure}[h]
+\begin{figure}
 \centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_trees.png}
-\caption{\textbf{Approximate SPR distances between all pairs of trees of segments.}}
-\label{}
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_replicates.png}
+\caption{\textbf{Approximate SPR distances between replicate trees of each segment.}
+Approximate SPR distances between replicates of MP and NS trees are much higher ($\approx$260) than any other segments, suggesting greater variability in tree topology over the course of MCMC.
+SPR distances between replicates of most other segments are $\approx$200.
+}
+\label{SPRdistancesReplicates}
 \end{figure}
 
-\begin{figure}[h]
+\begin{figure}
 \centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_replicates.png}
-\caption{\textbf{Approximate SPR distances between replicate trees of each segment.}}
-\label{}
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist.png}
+\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances after normalization.}
+95\% HPD interval (blue) overlaps zero, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
+\label{NormSPR_PB1-PB2_difference}
 \end{figure}
 
-\begin{figure}[h]
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances after normalization.}
+%95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
+%\label{NormSPR_PB1-PB2_ratio}
+%\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist.png}
+\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances after normalization.}
+95\% HPD interval (blue) overlaps zero, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
+\label{NormSPR_PB1-HA_difference}
+\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances after normalization.}
+%95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.}
+%\label{NormSPR_PB1-HA_ratio}
+%\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_hist.png}
+\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances after normalization.}
+95\% HPD interval (blue) overlaps zero, suggesting no evidence of differences between approximate and exact SPR distances following normalization.
+Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
+\label{NormSPR_PB2-HA_difference}
+\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances after normalization.}
+%95\% HPD interval (blue) overlaps 1, suggesting no evidence of differences between approximate and exact SPR distances following normalization.
+%Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
+%\label{NormSPR_PB2-HA_ratio}
+%\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.95\textwidth]  {supp_figures/InfB_LDheatmap.1600.aa.ChiSqdf.minorCutoff1percent.png}
+\caption{\textbf{Heatmap of genome-wide linkage disequilibrium ($\chiSq$) between polymorphic amino acid sites.}
+Patterns of LD across the genome suggest a network of reciprocally linked amino acid sites on PB1, PB2, HA and, to some extent NA, proteins.
+Proximity of sites on heatmaps might not correspond to proximity of sites within proteins.}
+\label{ChiGenome}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.85\textwidth]  {supp_figures/InfB_LDheatmap.PB1_PB2_HA.1600.aa.ChiSqdf.minorCutoff1percent.png}
+\caption{\textbf{Heatmap of linkage disequilibrium ($\chiSq$) between amino acid sites on PB1, PB2 and HA proteins.}
+Numbers next to each row and column correspond to amino acid site number within a given protein starting from methionine.
+Amino acid sites exhibiting reciprocally high LD between PB1, PB2 and HA proteins are: 176, 211, 213, 508 (PB1), 115, 399, 470, 642 (PB2) and 96, 163, 520 and 570 (HA).
+Sites 211 and 213 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
+Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211 and 213 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
+Previous research \citep{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
+Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 position 211 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at position 211 and threonine (T) at position 213.
+It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
+Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \citep{fodor2004}.
+Similarly, site 399 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \citep{guilligay2008}.}
+\label{ChiCore}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_robustCounting.png}
+\caption{\textbf{Synonymous, non-synonymous and nucleotide substitution rates in segments under different PB1-PB2-HA complexes.}
+Evolutionary rate dissimilarities under Vic and Yam PB1-PB2-HA complexes are not systematic and appear negligible.
+Synonymous and non-synonymous rates were calculated by dividing the sum of all substitutions of a given class by the total amount of evolutionary time under each PB1-PB2-HA constellation.
+Nucleotide rates were calculated by multiplying the inferred nucleotide substitution rate on each branch by the branch length, then dividing this by the total amount of evolutionary time under each PB1-PB2-HA constellation.
+Vertical bars indicating uncertainty are 95\% HPDs, black bars indicate 95\% HPDs that do not overlap.}
+\label{robustCounting}
+\end{figure}
+
+%Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates} are showing approximate SPR distances prior to the normalization procedure (see Methods).
+%Figure \ref{SPRdistancesTrees} shows that MP and NS trees show greater approximate SPR distances when compared to any other segment tree, which is consistent with the fact that only M1 and NS1 coding sequences, which are shorter than any %other coding sequences, were used to infer the trees.
+%Figure \ref{SPRdistancesReplicates} confirms this by showing greater variability in tree topology (as greater approximate SPR distance) when MP and NS trees are compared against their replicates.
+%For all other segments approximate SPR distances between all pairs of trees prior to normalization suggests that most segment pairs have similar posterior distributions of approximate SPR distances.
+
+%Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates} show $\dtmrca$ statistics for inter-segment and intra-segment, respectively, comparisons prior to normalization.
+%Figure \ref{deltaTMRCAtrees} shows that similarity in TMRCAs (as low $\dtmrca$) between PB1, PB2 and HA segments is due to considerably smaller $\dtmrca$ statistic for the comparison between these segments.
+%It is also noteworthy that NA and MP comparisons have similarly low values for this statistic.
+%Figure \ref{deltaTMRCAreplicates} shows that TMRCAs are quite robust over the course of MCMC and node times do not, even in the most extreme cases, deviate from their replicates by more than 1 year.
+
+\begin{figure}
 \centering  
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
-\caption{\textbf{Mean $\Delta$TMRCA between all pairs of trees of segments.}}
-\label{}
+\caption{\textbf{Mean $\dtmrca$ between all pairs of trees of segments.}
+Mean $\dtmrca$ between trees of segments reveal that tip pairs in PB1, PB2 and HA trees have very similar TMRCAs.
+The upper tail of the 95\% HPD (HPDs are represented as vertical lines) interval of mean $\dtmrca$ values for PB1-PB2-HA and MP-NA trees do not exceed 3 years.}
+\label{deltaTMRCAtrees}
 \end{figure}
 
-\begin{figure}[h]
+\begin{figure}
 \centering  
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
-\caption{\textbf{Mean $\Delta$TMRCA between replicate trees of each segment.}}
-\label{}
+\caption{\textbf{Mean $\dtmrca$ between replicate trees of each segment.}
+Mean $\dtmrca$ values between independent analyses of each segment show that mean $\dtmrca$ values rarely exceed 1 year.}
+\label{deltaTMRCAreplicates}
 \end{figure}
 
-\begin{figure}[h]
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_normRErate.png}
-\caption{\textbf{Normalized reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
-\label{}
-\end{figure}
+%We also tested the accuracy of approximate SPR distances using exact SPR distances calculated from PB1, PB2 and HA comparisons.
+%Figures \ref{SPR_PB1-PB2_difference}--\ref{SPR_PB2-HA_ratio} show that after normalization the approximate and exact SPR distance estimates are not significantly different.
+%There is also good correlation between the two SPR distances (Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation}). 
+%Due to excessively long computation time of exact SPR distances very few PB2--HA comparisons were made.
 
-\begin{figure}[h]
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_trees.png}
-\caption{\textbf{Reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
-\label{}
-\end{figure}
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB1-PB2_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances after normalization.}}
+%\label{NormSPR_PB1-PB2_correlation}
+%\end{figure}
 
-\begin{figure}[h]
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_replicates.png}
-\caption{\textbf{Reassortment rate (approximate number of SPR moves per total time in the tree) between replicate trees of each segment.}}
-\label{}
-\end{figure}
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB1-HA_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances after normalization.}}
+%\label{NormSPR_PB1-HA_correlation}
+%\end{figure}
 
-\begin{figure}[h]
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_locs_aa.png}
-\caption{\textbf{2-dimensional MDS embedding of segment based on mean amino acid linkage disequilibrium ($\chi\_{df}^{2}$).}}
-\label{}
-\end{figure}
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_NormPB2-HA_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances after normalization.}
+%Due to excessively long computation time of exact SPR distances between PB2 and HA trees few comparisons were made.}
+%\label{NormSPR_PB2-HA_correlation}
+%\end{figure}
 
-\begin{figure}[h]
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_corrs_aa.png}
-\caption{\textbf{Correlation between mean amino acid linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
-\label{}
-\end{figure}
+%Figures \ref{SPR_PB1-PB2_difference}--\ref{SPR_PB2-HA_ratio} focus on differences between exact and approximate SPR distances and show that prior to normalization approximate SPR distances are overestimates of exact SPR distances.
+%Figures \ref{SPR_PB1-PB2_correlation}--\ref{SPR_PB2-HA_correlation} show that there is a similar degree of correspondence between approximate and exact SPR distances to when approximate and exact SPR distances are normalized (Figures %\ref{NormSPR_PB1-PB2_correlation}--\ref{NormSPR_PB2-HA_correlation}) using the procedure described in Methods.
 
-\begin{figure}[h]
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_locs_nt.png}
-\caption{\textbf{2-dimensional MDS embedding of segment based on mean nucleotide linkage disequilibrium ($\chi\_{df}^{2}$)}}
-\label{}
-\end{figure}
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist.png}
+%\caption{\textbf{Distribution of differences between exact and approximate PB1-PB2 SPR distances.}}
+%\label{SPR_PB1-PB2_difference}
+%\end{figure}
 
-\begin{figure}[h]
-\centering  
-\includegraphics[width=0.65\textwidth]  {supp_figures/MDS_corrs_nt.png}
-\caption{\textbf{Correlation between mean nucleotide linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
-\label{}
-\end{figure}
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB1-PB2 SPR distances.}}
+%\label{SPR_PB1-PB2_ratio}
+%\end{figure}
 
-\bibliographystyle{pnas}
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist.png}
+%\caption{\textbf{Distribution of differences between exact and approximate PB1-HA SPR distances.}}
+%\label{SPR_PB1-HA_difference}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB1-HA SPR distances.}}
+%\label{SPR_PB1-HA_ratio}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist.png}
+%\caption{\textbf{Distribution of differences between exact and approximate PB2-HA SPR distances.}}
+%\label{SPR_PB2-HA_difference}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
+%\caption{\textbf{Distribution of ratios between exact and approximate PB2-HA SPR distances.}}
+%\label{SPR_PB2-HA_ratio}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB1-PB2 SPR distances.}}
+%\label{SPR_PB1-PB2_correlation}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB1-HA SPR distances.}}
+%\label{SPR_PB1-HA_correlation}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.4\textwidth]  {supp_figures/InfB_supp_PB2-HA_corr.png}
+%\caption{\textbf{Correlation between exact and approximate PB2-HA SPR distances.}}
+%\label{SPR_PB2-HA_correlation}
+%\end{figure}
+
+%\clearpage
+%\section*{Multidimensional scaling}
+%Figures \ref{MDSaaLD} and \ref{MDSntLD} visualize the relationships between segments based on mean amino acid and nucleotide site linkage disequilibrium estimates, respectively.
+%PB1, PB2 and HA segments appear to be by far the most similar segments, as indicated by their proximity in the MDS embedding.
+%Figures \ref{MDSaaCorr} and \ref{MDSntCorr} compare the observed mean LD values between segments. 
+%Based on mean amino acid site LD (Figure \ref{MDSaaCorr}) only PB1, PB2 and HA segments show considerable clustering, all other segments being equidistant from each other in the MDS embedding.
+%MDS distance versus mean nucleotide site LD scatter plot, on the other hand, shows a continuum of relationships between segments, which is probably caused by more variability at the nucleotide level than the amino acid level.
+%This difference in variability would build up linkage disequilibrium between nucleotide sites and would reduce it via convergent substitutions over relatively short periods of time.
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_aaLD_MDS.png}
+%\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean amino acid linkage disequilibrium ($\chiSq$).}}
+%\label{MDSaaLD}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_aaLD_MatrixMDScorr.png}
+%\caption{\textbf{Correlation between mean amino acid site linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
+%\label{MDSaaCorr}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_ntLD_MDS.png}
+%\caption{\textbf{2-dimensional MDS embedding of influenza B virus segments based on mean nucleotide linkage disequilibrium ($\chiSq$).}}
+%\label{MDSntLD}
+%\end{figure}
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_ntLD_MatrixMDScorr.png}
+%\caption{\textbf{Correlation between mean nucleotide site linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
+%\label{MDSntCorr}
+%\end{figure}
+
+%\section*{Differences between Vic and Yam PB1-PB2-HA}
+%Through the combination of inferred PB1-PB2-HA background and realized synonymous and non-synonymous substitution counts on branches \citep{obrien2009} it was possible to investigate whether Victoria and Yamagata lineage backgrounds differ in substitution rates.
+%Figure \ref{robustCounting} shows the inferred synonymous, non-synonymous (both as substitutions per year) and overall nucleotide substitution rate (as substitutions per site per year) under different lineages of PB1-PB2-HA.
+
+%\begin{figure}
+%\centering  
+%\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_lineageRatios.png}
+%\caption{\textbf{Total counts and ratios of Victoria and Yamagata lineage sequences in each segment.}}
+%\label{lineageRatios}
+%\end{figure}
+
+\bibliographystyle{mbe}
 \bibliography{fluB}
+
 \end{document}

### 2014-06-02 scripts/Hrothvitnir.py

@@ -7,12 +7,12 @@
 ##
 ##  Usage:
 ##  open command line, go to folder where the script is located and type in:
-##  "python Hrothvitnir_eLife.py -i <<path to .trees.txt file>> -m <<mode - X, diversityOt, FstOt or stateTime>> 1> <<path to output file>>" (ignore <<, >> and ")
+##  "python Hrothvitnir.py -i <<path to .trees.txt file>> -m <<mode - X, diversityOt, FstOt or stateTime>> 1> <<path to output file>>" (ignore <<, >> and ")
 ##  
 ##  The following figures in the manuscript were made using these modes:
 ##  Figure 3 - diversityOT
 ##  Figure 4 - FstOt
-##  Figures 7 and S9 - stateTime
+##  Figures 7 and S10 - stateTime
 ##
 ##
 ##  Gytis Dudas

### 2014-06-02 scripts/Hydra.py

@@ -15,10 +15,10 @@
 ##  otherwise the script should prompt you to select the files manually
 ##
 ##  open command line, go to folder where the script is located and type in:
-##  "python Hydra_eLife.py" (ignore ")
+##  "python Hydra.py" (ignore ")
 ##  
 ##  The following figures in the manuscript were made using this script:
-##  Figures 8, 9, S1-S6, S10-S11
+##  Figures 8, S1-S7, S11-S12
 ##
 ##
 ##  Gytis Dudas

### 2014-06-02 scripts/LD_calculator.py

@@ -134,9 +134,9 @@ def overlap(a,b):
 
 ## Makes a list of dates if partitioning of LD is desired
 # Default value makes a single bin for all sequences isolated between 1900 and 2014
-step=0.25
-windowSize=3
-timeline=np.arange(1984,2014-windowSize,step)
+step=40
+windowSize=0
+timeline=np.arange(1980,2020-windowSize,step)
 
 ## Segments to analyze                                                                       
 segments=['PB1','PB2','PA','HA','NP','NA','M1','NS1']
@@ -184,16 +184,14 @@ for mem1 in range(0,len(segments)):
             ## Define alignment name
             seg1='%s translation alignment.nex'%(segment1)
             seg2='%s translation alignment.nex'%(segment2)
-            f = open(outpath+'%s_%s_ChiSqdf-Dprime.aa.HUGE_ts3.csv'%(segment1,segment2),'w')
-            header='time point,time step size,N strains,N haplotypes usable,site 1,site 2,number of polymoprhisms at site 1,number of polymoprhisms at site 2,minor allele freq site 1,minor allele freq site 2,D,|D\'|,Chi,ChiSqDf'
-            print>>f,header
         elif mode=='nt':
             ## Define alignment name
             seg1='%s alignment.nex'%(segment1)
             seg2='%s alignment.nex'%(segment2)
-            f = open(outpath+'%s_%s_ChiSqdf-Dprime.nt.HUGE_ts3.csv'%(segment1,segment2),'w')
-            header='time point,time step size,N strains,N haplotypes usable,site 1,site 2,number of polymoprhisms at site 1,number of polymoprhisms at site 2,minor allele freq site 1,minor allele freq site 2,D,|D\'|,Chi,ChiSqDf'
-            print>>f,header
+
+        f = open(outpath+'%s_%s_ChiSqdf-Dprime.%s.HUGE.csv'%(segment1,segment2,mode),'w')
+        header='time point,time step size,N strains,N haplotypes usable,site 1,site 2,number of polymoprhisms at site 1,number of polymoprhisms at site 2,minor allele freq site 1,minor allele freq site 2,D,|D\'|,Chi,ChiSqDf'
+        print>>f,header
 
         ## open both alignments
         handle1 = open(path+seg1, "rU")

### 2014-07-11 fluB.tex
Enable cross-references between main and supp text.
Add 1600 genomes LD, move figures to right places.
Fix #34

@@
 \usepackage{float}
+
+% reference figures across documents
+\usepackage{xr}
+\externaldocument{fluB_supp}
 
@@
 
-We show that despite extensive reassortment, three of the eight segments -- two segments coding for components of the influenza B virus polymerase, PB1 and PB1, and the surface glycoprotein HA -- still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
+We show that despite extensive reassortment, three of the eight segments -- two segments coding for components of the influenza B virus polymerase, PB1 and PB2, and the surface glycoprotein HA -- still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
 In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
@@
  \centering		
-	\includegraphics[width=0.45\textwidth]{figures/TreeFigure2}
+	\includegraphics[width=0.45\textwidth]{figures/TreeFigure2.png}
 	\caption{\textbf{Schematic analysis of reassortment patterns.}
@@
 	\centering	
-	\includegraphics[width=0.65\textwidth]{supp_figures/InfB_LDcircle.1600.aa.ChiSqdf.minorCutoff1percent.png}
+	\includegraphics[width=0.65\textwidth]{figures/InfB_LDcircle.1600.aa.ChiSqdf.minorCutoff1percent.png}
 	\caption{\textbf{LD comparison between influenza B proteins.}
@@
 	\centering	
-	\includegraphics[width=0.85\textwidth]{supp_figures/InfB_methodSpeciesTree.pdf}
+	\includegraphics[width=0.85\textwidth]{figures/InfB_methodSpeciesTree.pdf}
 	\caption{\textbf{Calculating $\dtmrca$ from a species tree perspective.}
@@
 
-\bibliographystyle{mbe}
+\bibliographystyle{mbe.bst}
 \bibliography{fluB}

### 2014-07-11 fluB.tex
Fix #37

@@
 
-Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
+Another possibility is the action of frequency dependent selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
 A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{genomeGrid}, \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot}).
 Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \citep{chen2006,obenauer2006}.
-This is easiest to interpret as frequency-dependent balancing selection acting to preserve antigenic diversity \citep{worobey2014}.
-If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
+This is easiest to interpret as frequency-dependent selection acting to preserve antigenic diversity \citep{worobey2014}.
+If this is the case in influenza B viruses, we expect frequency-dependent selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
 
@@
 In this case influenza B viruses would undergo periodic genome homogenization events with the exception of PB1, PB2 and HA segments.
-Because Vic HA and Yam HA are antigenically dissimilar we think that balancing selection would prevent even strong selective sweeps from driving the opposing PB1-PB2-HA gene complex to extinction.
+Because Vic HA and Yam HA are antigenically dissimilar we think that frequency-dependent selection would prevent even strong selective sweeps from driving the opposing PB1-PB2-HA gene complex to extinction.
 This `\textit{status quo}' model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.
@@
 Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether dynamics similar to Victoria and Yamagata lineages have not occurred in influenza B virus genomes before and left no trace through extinction.
-We find it unlikely that either Victoria or Yamagata lineage PB1-PB2-HA complexes will go extinct stochastically in the near future, as they have co-circulated for prolonged periods at a ratio close to 0.5, suggesting the action of balancing selection (Figure \ref{lineageRatiosOverTime}).
+We find it unlikely that either Victoria or Yamagata lineage PB1-PB2-HA complexes will go extinct stochastically in the near future, as they have co-circulated for prolonged periods at a ratio close to 0.5, suggesting the action of frequency dependent selection (Figure \ref{lineageRatiosOverTime}).
 Extinction through depletion of susceptible individuals, such as following influenza pandemics or mass vaccination seem unlikely as well.

### 2014-07-11 fluB.tex
Fix #35

@@
 
-We should clarify that when we say that Vic and Yam PB1-PB2-HA complexes of the influenza B virus are co-adapted and have co-evolved we mean that these 3 segments do not appear to function well when derived from both lineages and that the evolution to this state has been intimate, in the sense that there has not been a successful reassortant combining these two lineages across great temporal distances, unlike the rest of the genome.
+We should clarify that when we say that Vic and Yam PB1-PB2-HA complexes of the influenza B virus are co-adapted and have co-evolved we mean that these 3 segments do not appear to be fit when derived from both lineages and that the evolution to this state has been intimate, in the sense that there has not been a successful reassortant combining these two lineages across great temporal distances, unlike the rest of the genome.
 

### 2014-07-11 fluB.tex
Address #13
Not exactly a fix, but some progress towards it.

@@
 	\includegraphics[width=0.65\textwidth]	{figures/InfB_LineageRatiosOverTime.png}
-	\caption{\textbf{Ratio of lineages in the dataset.}
-The ratio of sequences that derive from the original Victoria clade (yellow) to sequences that derive from the Yamagata clade (purple) in each segment from the primary dataset over time.
+	\caption{\textbf{Ratio of Vic and Yam sequences in the dataset.}
+The ratio of Victoria (yellow) to Yamagata (purple) sequences in each segment from the primary dataset over time.
 Black lines indicate where this ratio lies in the larger secondary dataset.
-Numbers at the top of the figure show the total number of genomes analyzed within each influenza season in the primary dataset comprised of 452 genomes from which the ratio was calculated, while the numbers in brackets correspond to numbers of sequences in the larger secondary genomes dataset.
+Numbers at the top of the figure show the total number of genomes available for each influenza season in the primary dataset comprised of 452 genomes from which the ratio was calculated, while the numbers in brackets correspond to numbers of sequences in the larger secondary genomes dataset.
 Numbers at the bottom are influenza seasons from the 1987/1988 (87/88) season to the 2011/2012 season.
-Yamagata lineage PA, NP, NA and MP segments and Victoria lineage NS segment eventually become fixed in the influenza B population.
+Yamagata lineage of PA, NP, NA and MP segments and Victoria lineage of the NS segment eventually become fixed in the influenza B population.
 PB1, PB2 and HA segments maintain separate Victoria and Yamagata lineages.}

### 2014-07-14 fluB.tex
Clarify information about Figure 5.
Change references to lineages to references to sequences when talking
about Figure 5. Address #13.

@@
 
-We show the ratio of Vic and Yam lineages in our primary and secondary sequence data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on lineage assignment performed earlier (see Methods).
+We show the ratio of Vic to Yam sequences in our primary and secondary data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on which lineage was assigned to each sequence earlier (see Methods).
 It is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
 These losses of diversity correspond to fixation of the Vic lineage in NS and fixation of the Yam lineage in PA, NP, NA and MP.
-Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen, where the two lineages have been isolated at a ratio close to 50\% over long periods of time (Figure \ref{lineageRatiosOverTime}).
-On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically consistent with one lineage predominating within a given season, in agreement with surveillance data \citep{reed2012}.
+Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen, where the two lineages have been sequenced at a ratio close to 50\% over long periods of time (Figure \ref{lineageRatiosOverTime}).
+On a year-to-year basis, however, the ratios for Vic and Yam sequences PB1, PB2 and HA can fluctuate dramatically consistent with one lineage predominating within a given season, in agreement with surveillance data \citep{reed2012}.
 

### 2014-07-15 scripts/Hrothvitnir.py
Address #29
Add another function to Hrothvitnir.py, heatmap of number of
transitions - 1.

@@ -7,7 +7,7 @@
 ##
 ##  Usage:
 ##  open command line, go to folder where the script is located and type in:
-##  "python Hrothvitnir.py -i <<path to .trees.txt file>> -m <<mode - X, diversityOt, FstOt or stateTime>> 1> <<path to output file>>" (ignore <<, >> and ")
+##  "python Hrothvitnir.py -i <<path to .trees.txt file>> -m <<mode - X, diversityOt, FstOt, N_transitions or stateTime>> 1> <<path to output file>>" (ignore <<, >> and ")
 ##  
 ##  The following figures in the manuscript were made using these modes:
 ##  Figure 3 - diversityOT
@@ -546,6 +546,13 @@ def parseTreeFile(argv):
                     for time in slice_list:
                         sys.stdout.write('\t%s_diversity'%(time[0]))
                 #######################################################
+                if mode=='N_transitions':
+                    headerTraits=[]
+                    for o in trait_list:
+                        headerTraits.append(o[0])
+                    sys.stdout.write('states\t%s'%('\t'.join(headerTraits)))
+                    
+ 
 
             ## After burnin start processing
             if int(cerberus.group(1)) >= burnin:
@@ -698,6 +705,13 @@ def parseTreeFile(argv):
                             ## Only output the oldest TMRCA from the list of TMRCAs
                             sys.stdout.write('\t%s'%(min(temp)))
                 ################################################################################
+                if mode=='N_transitions':
+                    sys.stdout.write('\n%s'%(cerberus.group(1)))
+                    for x in trait_list:
+                        reassortants=[1 for k in object_list[1:] if k.traits[x[0]]!=k.parent.traits[x[0]]]
+                        #sys.stdout.write('\t%s\t%s'%(x[0],sum(reassortants)-1))
+                        sys.stdout.write('\t%s'%(len(reassortants)-1))
+                        
             if 'End;' in line:
                 print '\nEnd of file!'
 
@@ -713,8 +727,8 @@ global burnin
 # stateTime - outputs the non-synonymous (*N), synonymous (*S) and nucleotide (*_nt) substitution rates as well as total amount of time spent under different PB1-PB2-HA constellations.
 # FstOT - outputs mean pairwise diversity between V and Y labeled branches at each time point.
 # diversityOT - outputs time of most recent common ancestor of all lineages existing at each time point.
-
-allModes=['X','stateTime','FstOT','diversityOT']
[…]

### 2014-07-15 fluB.tex
Address #36
Slim down the discussion way down. Remove excessive/irrelevant
speculation.

@@
 
-\subsection*{Evidence of a co-adapted PB1-PB2-HA gene complex}
-In this paper we show that the PB1, PB2 and HA segments of influenza B viruses are the only ones that have continuously maintained separate Vic and Yam lineages, while other segments have fixed either Vic or Yam lineages (Figures \ref{genomeGrid}, \ref{lineageRatiosOverTime} and \ref{railroadPlot}).
-Evidence suggests that this is a result of prolonged lack of reassortment between Vic and Yam lineages in PB1, PB2 and HA (Figure \ref{betweenDiversity}) which possess co-assorting sequences detectable as high linkage disequilibrium (Figure \ref{segmentLD}).
-The vast majority of the sampled evolutionary history of each segment of influenza B viruses since the split of Vic and Yam lineages has been spent in association with either completely Victoria or completely Yamagata lineage derived PB1-PB2-HA complexes (Figure \ref{stateTime}), suggesting that having `pure' lineage PB1-PB2-HA complexes is important for whole-genome fitness.
-We propose that this pattern of coassortment is due to the action of selection and not simply biased or rare reassortment.
+\subsection*{The origins of the linkage remain unknown}
+The origin of the strong genetic linkage between PB1, PB2 and HA segments remains unclear.
+We believe there are two alternative, but similar explanations for the origins of the strong genetic linkage between PB1, PB2 and HA segments: mutation-driven co-evolution \citep{presgraves2010} and Dobzhansky-Muller incompatibility \citep{dobzhansky1937,muller1942}.
+Mutation-driven co-evolution \citep{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \citep{lee2008}, and evolves as a by-product of adaptation.
+If one or the other influenza B lineage has undergone adaptation we might expect these changes to be beneficial in the derived background and incompatible with the ancestral background.
+Dobzhansky-Muller incompatibility operates in a similar way, but the main difference from the scenario described earlier is that the incompatible alleles are neutral or nearly neutral in their native background and become deleterious or lethal when combined with non-native backgrounds.
 
-When comparing the mean tip-tip TMRCA deviations we find evidence of isolation by temporal distance - when reassortments occur between PB1, PB2 and HA segments those events tend to involve branches that have similar TMRCAs.
-Other segment pairs, with the exception of MP-NA, seem to show little signal of this and are able to reassort with branches that have been evolving under temporally distant genomic backgrounds.
-In addition, most strains with mixed-lineage PB1-PB2-HA complexes occurred in the early years of the Vic--Yam split, when the two lineages were presumably much more similar at the nucleotide and amino acid levels.
-The most recent influenza B viruses with mixed-lineage PB1-PB2-HA complexes are B/Waikato/6/2005-like and B/Malaysia/1829782/2007-like viruses which circulated for only 5 months or are known from a single isolate, respectively.
+\subsection*{Potential mechanisms for reassortant incompatibility}
+Unfortunately, the limited amount of genomic data available for the early years of the Vic--Yam split precludes any attempts at answering whether selection or drift have led to the current linkage of PB1, PB2 and HA segments.
+Although the origins of the linkage between these three segments might be difficult to explain, we can speculate on the nature of reassortant incompatibility.
+For example, it is intuitive for why this might be the case for PB1 and PB2: both proteins operate cooperatively in intimate proximity as part of the RNA-dependent RNA polymerase heterotrimer.
+Indeed, we observe that PB1--PB2 reassortants are the rarest and least persistent among mixed-lineage PB1-PB2-HA strains and have not been isolated in great numbers.
+In fact most reassortants breaking the PB1-PB2-HA complex apart have occured in the past, close to the split of Vic and Yam lineages and are very rare.
 
-We should clarify that when we say that Vic and Yam PB1-PB2-HA complexes of the influenza B virus are co-adapted and have co-evolved we mean that these 3 segments do not appear to be fit when derived from both lineages and that the evolution to this state has been intimate, in the sense that there has not been a successful reassortant combining these two lineages across great temporal distances, unlike the rest of the genome.
-
-\subsection*{Causes of co-dependence between PB1-PB2-HA}
-Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, \textit{e.g.} PB1-PB2-PA and HA-NA \citep{mccullers2004}.
-Though it would be easy to explain co-adaptation between these segments by referring to their functional roles \textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities, our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.
-
-It is perhaps easiest to explain co-dependence between PB1 and PB2 segments based on their functions as part of the influenza RNA-dependent RNA polymerase (RdRp) heterotrimer.
-Indeed, trees of PB1 and PB2 segments exhibit high similarity in tip--tip TMRCAs (Figure \ref{deltaTMRCA}), suggesting highly similar TMRCAs.
-In addition, PB1--PB2 reassortants are the rarest and least persistent among mixed-lineage PB1-PB2-HA strains and have not been isolated in great numbers.
-
-Explaining the co-dependence of PB1+2 and HA segments is more difficult.
-Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \citep{bergeron2010,fulvini2011}.
-A previously used technique for producing vaccine seed strains involved selecting for HA-NA reassortants, which often yielded PB1-HA-NA reassortants as a side-effect \citep{bergeron2010,fulvini2011}.
+Although the linkage between the two polymerase segments and HA is more difficult to explain, it is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering ample time for the budding Victoria lineage to develop reassortant incompatibility alleles.
+There is some evidence that the linkage between PB1 and HA might not be a phenomenon restricted to influenza B viruses.
+A technique used previously for producing vaccine seed strains involved selecting for HA-NA reassortants, which often yielded PB1-HA-NA reassortants as a side-effect \citep{bergeron2010,fulvini2011}.
 Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \citep{cobbin2013}.
@@
 
-Another possibility is the action of frequency dependent selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
-A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{genomeGrid}, \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot}).
-Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \citep{chen2006,obenauer2006}.
-This is easiest to interpret as frequency-dependent selection acting to preserve antigenic diversity \citep{worobey2014}.
-If this is the case in influenza B viruses, we expect frequency-dependent selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.
+The association between Victoria and Yamagata lineage PB1-PB2-HA complexes should be relatively straightforward to explore in the lab.
+Using previously developed plasmid systems \citep{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
+There is some evidence, based on the frequency and persistence times of different reassortant classes we have observed, to suggest that there should be a hierarchy of reassortant fitness starting with PB1+PB2+HA reassortants which should be the most fit, followed by PB1+2/HA, then PB1+HA/PB2 and finally PB2+HA/PB1 reassortants with the lowest fitness.
+We believe that this is the most direct approach to unravelling the mechanism responsible for the linkage within the PB1-PB2-HA complex.
 
-We find this hitchhiking scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
-Our primary dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bangkok/163/1990-like genome constellations.
-The larger secondary dataset with 1603 complete genomes has one additional example of this: isolates with a B/Waikato/6/2005-like PB1-PB2-HA constellation.
-We note that in all cases the genome constellations of these reassortants were not fixed in the influenza B population and the more recent B/Waikato/6/2005-like reassortants persisted for a much shorter period of time (5 months versus 4 years), suggestive of declining PB1+2/HA reassortant fitness over time.
-
-It is possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
-In this case PB1, PB2 and HA segments of Vic and Yam lineage accumulate substitutions that improve their ability to co-operate with segments of the same lineage and worse at interacting segments of a different lineage.
-This process, termed mutation-driven co-evolution \citep{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \citep{lee2008}.
-It is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering a potential explanation for why the budding Victoria lineage segments were not homogenized via reassortment with Yamagata lineage segment in the early years of the split between the two lineages.
-The restriction to eastern Asia would presumably also give sufficient time for PB1, PB2 and HA segments to co-evolve together.
-Similarly to the previous scenario, however, mutation-driven co-evolution would require an association of some kind between PB1+2 and HA segments to explain the low observed frequency and poor fitness of PB1+2/HA reassortants.
-In addition, we see no reason why these three particular segments, and not the rest of the genome, would become co-adapted to each other in geographic isolation.
-
-\subsection*{Suggested experiments}
-The association between Victoria and Yamagata lineage PB1-PB2-HA complexes should be relatively straightforward to test in the lab.
-Using previously developed plasmid systems \citep{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
-We predict that artificially produced viruses with mixed-lineage PB1-PB2-HA complexes will have reduced fitness when compared to viruses with pure-lineage, i.e.\ entirely Vic or entirely Yam, PB1-PB2-HA complexes.
-In addition, we expect the relationship between Vic and Yam lineage PB1, PB2 and HA segments to be dependent on date of segment isolation, as viruses with mixed-lineage PB1-PB2-HA complexes isolated earlier should perform better than viruses with PB1-PB2-HA segments isolated more recently.
-
-Given the short circulation times of recent PB1-PB2-HA reassortants B/Waikato/6/2005-like and B/Malaysia/1829782/2007-like viruses we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
-We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (Figures \ref{NormSPR_PB1-PB2_difference} -- \ref{NormSPR_PB2-HA_difference}).
-The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
-However, our analyses do not indicate any obvious differences in synonymous, non-synonymous or nucleotide substitution rates between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two (Figure \ref{robustCounting}).
-
-However, we also see that epistatic effects might interfere with fitness measurements, if for example non-PB1-PB2-HA segments are also temporally mismatched.
-Ideally, the co-adaptation would be easier to understand by referring to the structures of PB1 and PB2 proteins, as the link between these would be intuitive.
-We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins (Figure \ref{ChiCore}) that fall within the regions that form contacts within the influenza B polymerase heterotrimer \citep{sugiyama2009}, suggesting more subtle roles for sites we have identified.
-
-\subsection*{The future of influenza B viruses}
+\subsection*{Will influenza B viruses speciate?}
 We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \citep{turner2005}.
 In this context, we see three potential paths of evolution for influenza B viruses.
-More segments could be recruited into the two currently circulating co-adapted segment complexes (PB1, PB2 and HA segments being the genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
-This is the speciation scenario.
-Due to the rarity of inter-lineage reassortment events it is unclear whether this process is already under way with NA and MP, based on Figures \ref{genomeGrid}, \ref{betweenDiversity} and \ref{deltaTMRCA}, being the next segments to be recruited to PB1-PB2-HA complexes.
+If more segments get recruited to the PB1-PB2-HA complex, the process could continue until none of the segments are able to reassort across the Victoria--Yamagata lineage boundary.
 The two alternatives to the speciation hypothesis are the `\textit{status quo}' model, where the influenza B genome continues to be homogenized via gene flow with the exception of PB1, PB2 and HA segments and the extinction scenario, whereby one of the two PB1-PB2-HA complexes goes extinct, marking the return of single-strain dynamics in the influenza B virus population.
+The eventual fate of influenza B viruses will be mostly determined by the combined effects of reassortment frequency and epistatic interactions between segments over long periods of time.
 
-Sympatric speciation in other systems usually requires strong barriers to introgression, \textit{e.g.} infertility of F1 hybrids can lead to the evolution of prezygotic reproductive isolation otherwise known as reinforcement or the Wallace effect.
-To what extent this would apply to influenza B viruses remains unknown.
-If influenza B viruses are undergoing sympatric speciation, it is imperative to determine whether co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or whether co-infection is so rare that speciation occurs via mutation-driven co-evolution.
-We think that co-dependence between PB1 and PB2 segments can be explained by the fact that they are functionally linked: together with the PA protein they form part of the influenza B virus RNA-dependent RNA polymerase heterotrimer.
-
-It also remains unknown whether reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background reassorting into a Victoria PB1-PB2-HA background are indicative of rare events followed by selective sweeps or stochastic fixation.
-It's not unfeasible for selective sweeps following reassortments to break down developing co-adaptation of segments, especially if they are not functionally linked and have themselves been reassorted into a new PB1-PB2-HA background recently.
-This is the second scenario we might expect to occur, whereby relatively frequent reassortments occur between Vic and Yam lineages and are followed by selective sweeps.
-In this case influenza B viruses would undergo periodic genome homogenization events with the exception of PB1, PB2 and HA segments.
-Because Vic HA and Yam HA are antigenically dissimilar we think that frequency-dependent selection would prevent even strong selective sweeps from driving the opposing PB1-PB2-HA gene complex to extinction.
-This `\textit{status quo}' model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.
-
-Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether dynamics similar to Victoria and Yamagata lineages have not occurred in influenza B virus genomes before and left no trace through extinction.
-We find it unlikely that either Victoria or Yamagata lineage PB1-PB2-HA complexes will go extinct stochastically in the near future, as they have co-circulated for prolonged periods at a ratio close to 0.5, suggesting the action of frequency dependent selection (Figure \ref{lineageRatiosOverTime}).
-Extinction through depletion of susceptible individuals, such as following influenza pandemics or mass vaccination seem unlikely as well.
-Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines, which are usually applied to specific subsets of the population, do not produce lifelong immunity.
-
-\subsection*{Conclusion}
+\section*{Conclusion}
 We have used the $\dtmrca$ statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.

### 2014-07-15 fluB.tex
Address #38
Replace the word "fix" where its use is non-standard. De-sensationalize
flu B. Change a few sentences to incorporate reviewers' suggestions.

@@
 Seasonal influenza causes between 250,000 and 500,000 deaths annually and comprises lineages from three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \citep{flufactsheet}.
-However, influenza B viruses are increasingly being recognized as important human pathogens \citep{paul-glezen2013} and occasionally come to predominate an influenza season.
-For example in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \citep{ECDC1213}.
+Occasionally influenza B viruses become the predominant circulating virus in some locations, for example in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \citep{ECDC1213}.
 
@@
 We show that despite extensive reassortment, three of the eight segments -- two segments coding for components of the influenza B virus polymerase, PB1 and PB2, and the surface glycoprotein HA -- still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
-In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
-This has occurred through repeated reassortments and subsequent fixation of reassortant genome constellations within the influenza B population.
+In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and replaced the previous lineage: all currently circulating influenza B viruses have PA, NP, NA and MP segments derived from Yamagata lineage and NS segments derived from Victoria lineage.
 
@@
 However, the PB1, PB2 and HA segments of modern viruses are derived from both Victoria and Yamagata lineages.
-Consistent with fixation of Victoria or Yamagata lineages, the PA, NP, NA, MP and NS segments periodically lose diversity, while maintenance of parallel Victoria and Yamagata lineages results in continually increasing diversity in segments PB1, PB2 and HA (Figure \ref{tmrcaOT}).
+Periodic loss of diversity in PA, NP, NA, MP and NS segments is consistent with Victoria or Yamagata lineages becoming the `trunk' of the trees of those segments, while maintenance of parallel Victoria and Yamagata lineages results in continually increasing diversity in segments PB1, PB2 and HA (Figure \ref{tmrcaOT}).
 The PB1, PB2 and HA segments from present-day viruses maintain a common ancestor in $\sim$1983 and thus accumulate genetic diversity since the split of those segments into Vic and Yam lineages, while other segments often lose diversity with ancestors to present-day viruses appearing between $\sim$1991 and $\sim$1999.
@@
 This method gives a quantitative measure of reassortment-induced loss of diversity between Victoria and Yamagata lineages in two trees, although care should be taken when interpreting the statistic, as it does not correspond to any real TMRCAs in the tree, but can be interpreted as mean coalescence date between Vic and Yam lineages of PB1, PB2 and HA segments in all other trees.
-We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
+We focus only on PB1, PB2 and HA lineage labels, since all other segments eventually become completely derived from either the Vic or the Yam lineage.
 Losses of diversity (represented by more recent mean pairwise TMRCAs between Vic and Yam labels) in Figure \ref{betweenDiversity} indicate that every segment has reassorted with respect to the Victoria and Yamagata lineages of PB1, PB2 and HA segments.
@@
 We show the ratio of Vic to Yam sequences in our primary and secondary data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on which lineage was assigned to each sequence earlier (see Methods).
-It is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
-These losses of diversity correspond to fixation of the Vic lineage in NS and fixation of the Yam lineage in PA, NP, NA and MP.
+It is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to either the Vic (NS) or the Yam (PA, NP, NA, MP) lineage replacing the other lineage.
 Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen, where the two lineages have been sequenced at a ratio close to 50\% over long periods of time (Figure \ref{lineageRatiosOverTime}).
@@
 	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedMuDeltaTMRCA.png}
-	\caption{\textbf{Normalized mean $\dtmrca$ statistics between pairs of segments.}
+	\caption{\textbf{Normalized mean $\dtmrca$ statistics between for different segment pairs.}
 PB1, PB2 and HA trees exhibit reciprocally highly similar TMRCAs, unlike most other pairwise comparisons.
@@
 Indeed, we observe that PB1--PB2 reassortants are the rarest and least persistent among mixed-lineage PB1-PB2-HA strains and have not been isolated in great numbers.
-In fact most reassortants breaking the PB1-PB2-HA complex apart have occured in the past, close to the split of Vic and Yam lineages and are very rare.
+In fact most reassortants breaking the PB1-PB2-HA complex apart have occured in the past, close to the split of Vic and Yam lineages and have become very rare since.
 
 Although the linkage between the two polymerase segments and HA is more difficult to explain, it is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering ample time for the budding Victoria lineage to develop reassortant incompatibility alleles.
+Given the rarity of complete influenza B genomes from before the Vic--Yam split it is difficult to determine how much of the linkage between PB1, PB2 and HA segments can be attributed to the population structure of influenza B viruses in the past.
 There is some evidence that the linkage between PB1 and HA might not be a phenomenon restricted to influenza B viruses.

### 2014-07-16 fluB_supp.tex
Enable cross-references between main and supp text.
Add 1600 genomes LD, move figures to right places.
Mention supplementary figures.
Address #29 and to some extent #13.

@@
 \usepackage{float}
+
+% reference figures across documents
+\usepackage{xr}
+\externaldocument{fluB}
 
@@
 	\centering		
-	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedApproxSPR.png}
+	\includegraphics[width=0.65\textwidth]{supp_figures/InfB_normalizedApproxSPR.png}
 	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
@@
 \label{SPRdistancesReplicates}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.45\textwidth]  {supp_figures/InfB_seqCountries.452.png}
+\caption{\textbf{Geographic distribution of sequences in the primary dataset.}
+Sequences were assigned to the ``other'' category if there were less than 5 sequences from that country.
+Most of the genomes in the primary dataset were sampled in the USA.}
+\label{geoSeqs4}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.45\textwidth]  {supp_figures/InfB_seqCountries.1600.png}
+\caption{\textbf{Geographic distribution of sequences in the secondary dataset.}
+Sequences were assigned to the ``other'' category if there were less than 5 sequences from that country.
+Most of the genomes in the secondary dataset were sampled in Australia.}
+\label{geoSeqs1600}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.75\textwidth]  {supp_figures/InfB_Ntransitions.png}
+\caption{\textbf{Numbers of trait transitions in each tree.}
+The numbers shown are the mean inferred number of trait transitions (minus one to account for the initial Vic--Yam split) in a given tree and trait combination.
+The numbers in brackets correspond to 95\% highest posterior density intervals.}
+\label{Ntransitions}
 \end{figure}
@@
 
+\clearpage
+
 \bibliographystyle{mbe}

### 2014-07-23 fluB.tex
Clean language per #38
Also add references to similar methods.

@@
 
-Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
+Influenza B viruses make a considerable contribution to morbidity attributed to seasonal influenza. 
 Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata. 
@@
 
-We show that despite extensive reassortment, three of the eight segments -- two segments coding for components of the influenza B virus polymerase, PB1 and PB2, and the surface glycoprotein HA -- still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
+We show that despite extensive reassortment, three of the eight segments -- two segments coding for components of the influenza B virus polymerase, PB1 and PB2, and the surface glycoprotein HA -- still survive as distinct Victoria and Yamagata lineages, which appear to be co-dependent to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
 In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and replaced the previous lineage: all currently circulating influenza B viruses have PA, NP, NA and MP segments derived from Yamagata lineage and NS segments derived from Victoria lineage.
@@
 \subsection*{Analysis of reassortment patterns across Victoria and Yamagata lineages}
-The differentiation into Vic and Yam lineages can be seen in all segments (Figure \ref{genomeGrid}).
-Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
-On some segments, either the Victoria or Yamagata lineage fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment, seen as modern viruses deriving completely from either the Victoria or Yamagata lineage (yellow vs purple bars in Figure \ref{genomeGrid}).
+The differentiation into Vic and Yam lineages can be seen in all segments (Figure \ref{genomeGrid}) and is followed by inter-lineage reassortment events.
+In the phylogenetic trees of some segments either the Victoria or Yamagata lineage has become the `trunk' of the tree (yellow vs purple bars in Figure \ref{genomeGrid}) following reassortment.
 This pattern is apparent in the PA, NP, NA, MP and NS segments.
-However, the PB1, PB2 and HA segments of modern viruses are derived from both Victoria and Yamagata lineages.
+However, the Victoria and Yamagata lineages of PB1, PB2 and HA segments continue to co-circulate to this day.
 Periodic loss of diversity in PA, NP, NA, MP and NS segments is consistent with Victoria or Yamagata lineages becoming the `trunk' of the trees of those segments, while maintenance of parallel Victoria and Yamagata lineages results in continually increasing diversity in segments PB1, PB2 and HA (Figure \ref{tmrcaOT}).
@@
 Vertical bars indicate the original Victoria and Yamagata lineages within each segment.
-Each tree corresponds to the summarised output from single analyses comprised of 9000 trees sampled from the posterior distribution of trees.
-}
+Each tree is the summarised output of a single analysis comprised of 9000 trees sampled from the posterior distribution of trees.}
 \label{genomeGrid}
@@
 	\caption{\textbf{Oldest TMRCA of all surviving branches over time.}
-PA, NP, NA, MP and NS segments of influenza B viruses show periodic losses of diversity, indicating lineage turnover.
+PA, NP, NA, MP and NS segments of influenza B viruses show periodic increases in TMRCAs of all surviving branches (indicative of diversity loss), suggesting lineage turnover.
 PB1, PB2 and HA segments, on the other hand, maintain the diversity dating back to the initial split of Vic and Yam lineages.
@@
 We do see reduced diversity between Vic and Yam labels of PB1, PB2 and HA segments in a time period close to the initial split of Vic and Yam lineages (1986--1996).
-These reductions in diversity represent small clades with reassortant PB1-PB2-HA constellations, which go extinct by 1997.
+These reductions in diversity represent small clades with reassortant PB1-PB2-HA constellations, which go extinct by 1997 (see \ref{genomeGrid}).
 We also observe that the assignment of these 3 segment labels to branches of other segment trees is very similar and often identical after 1997.
@@
 
-We show the ratio of Vic to Yam sequences in our primary and secondary data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on which lineage was assigned to each sequence earlier (see Methods).
-It is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to either the Vic (NS) or the Yam (PA, NP, NA, MP) lineage replacing the other lineage.
+We show the ratio of Vic to Yam sequences in our primary and secondary datasets in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on which lineage each sequence was assigned to (see Methods).
+It is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to either the Vic (NS) or the Yam (PA, NP, NA, MP) lineage replacing the other lineage in the influenza B virus population.
 Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen, where the two lineages have been sequenced at a ratio close to 50\% over long periods of time (Figure \ref{lineageRatiosOverTime}).
@@
 Numbers at the bottom are influenza seasons from the 1987/1988 (87/88) season to the 2011/2012 season.
-Yamagata lineage of PA, NP, NA and MP segments and Victoria lineage of the NS segment eventually become fixed in the influenza B population.
+Yamagata lineage of PA, NP, NA and MP segments and Victoria lineage of the NS segment eventually become fixed (in the population genetics sense of the word) in the influenza B population.
 PB1, PB2 and HA segments maintain separate Victoria and Yamagata lineages.}
@@
 \end{itemize}
-B/Alaska/12/1996-like, B/Nanchang/2/1997-like and B/Iowa/03/2002-like constellations were observed by \citet{chen2008}, but sequences from \\B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like constellations were not available at the time.
+In a previous study B/Alaska/12/1996-like, B/Nanchang/2/1997-like and B/Iowa/03/2002-like constellations were observed \citet{chen2008}, but sequences from \\B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like constellations were not available at the time.
 In their study \citet{chen2008} also recovered the co-assortment pattern of PB1, PB2 and HA lineages, but did not remark upon it.
@@
 Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
-For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to the rise of viruses with B/Iowa/03/2002-like genome constellations.
-A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has originally been derived from the Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
+For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though the other segments it co-assorted with in 1995 -- 1996 (PA and MP) went extinct following the next round of reassortment that led to the rise of B/Iowa/03/2002-like genome constellations.
+A more extreme example is the NS segment, where a Vic sub-lineage was reassorted into an entirely Yam background (B/Alaska/12/1996-like) in 1994--1995, then reassorted back into a mostly Vic background some 5 years later (B/Iowa/03/2002-like) where it has replaced the `original' Vic sub-lineage (see Figure \ref{railroadPlot}).
 
-We observe that in 5 successful inter-lineage reassortment events, none break up the PB1-PB2-HA complex.
+We observe that in all 5 successful inter-lineage reassortment events shown in Figure \ref{railroadPlot}, none break up the PB1-PB2-HA complex.
 This is an unlikely outcome -- the probability of not breaking up PB1-PB2-HA across 5 reassortment events is $p = (\frac{2^{5} \times 2 - 2}{2^{8} - 2})^{5} = 0.0009$, where reassortment events are considered to sample from the Vic and Yam lineages at random for each of the 8 segments.
-If we correct for multiple testing with the assumption that co-assortment of any 3 segments is of interest we find that the probability of not breaking up an arbitrary 3 segments across 5 reassortment events is $p = \binom{8}{3} \times (\frac{2^{5} \times 2 - 2}{2^{8} - 2})^{5} = 0.0485$.
+If we correct for multiple testing with the assumption that co-assortment of any 3 segments is of interest we find that the probability of not breaking up an arbitrary set of 3 segments across 5 reassortment events is $p = \binom{8}{3} \times (\frac{2^{5} \times 2 - 2}{2^{8} - 2})^{5} = 0.0485$.
 
@@
 If one were able to recover an influenza `species tree', including admixture/reassortment events, it would be possible to estimate the reassortment or recombination `distance', which is the time between a split in the species tree in the past and a reassortment event (see Figure \ref{speciesTree}).
-Although we do not find evidence of differences in numbers of overall reassortments between segments (see Supplementary information), we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
+Although we do not find evidence of differences in total number of reassortments between segments (see Supplementary information), we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
 This difference in TMRCAs, or $\dtmrca$, is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
@@
 	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedMuDeltaTMRCA.png}
-	\caption{\textbf{Normalized mean $\dtmrca$ statistics between for different segment pairs.}
+	\caption{\textbf{Normalized mean $\dtmrca$ statistics for different segment pairs.}
 PB1, PB2 and HA trees exhibit reciprocally highly similar TMRCAs, unlike most other pairwise comparisons.
@@
 Mutation-driven co-evolution \citep{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \citep{lee2008}, and evolves as a by-product of adaptation.
-If one or the other influenza B lineage has undergone adaptation we might expect these changes to be beneficial in the derived background and incompatible with the ancestral background.
+If one or the other influenza B lineage has undergone adaptation we might expect these changes to be beneficial in its native background and incompatible with a foreign background.
 Dobzhansky-Muller incompatibility operates in a similar way, but the main difference from the scenario described earlier is that the incompatible alleles are neutral or nearly neutral in their native background and become deleterious or lethal when combined with non-native backgrounds.
@@
 \subsection*{Potential mechanisms for reassortant incompatibility}
-Unfortunately, the limited amount of genomic data available for the early years of the Vic--Yam split precludes any attempts at answering whether selection or drift have led to the current linkage of PB1, PB2 and HA segments.
+Unfortunately, the limited amount of genomic data available for the early years of the Vic--Yam split precludes any attempts of answering whether selection or drift have led to the current linkage of PB1, PB2 and HA segments.
 Although the origins of the linkage between these three segments might be difficult to explain, we can speculate on the nature of reassortant incompatibility.
-For example, it is intuitive for why this might be the case for PB1 and PB2: both proteins operate cooperatively in intimate proximity as part of the RNA-dependent RNA polymerase heterotrimer.
+For example, it is intuitive for why this might be the case for PB1 and PB2: both proteins interact directly as part of the RNA-dependent RNA polymerase heterotrimer.
 Indeed, we observe that PB1--PB2 reassortants are the rarest and least persistent among mixed-lineage PB1-PB2-HA strains and have not been isolated in great numbers.
@@
 
-Although the linkage between the two polymerase segments and HA is more difficult to explain, it is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering ample time for the budding Victoria lineage to develop reassortant incompatibility alleles.
-Given the rarity of complete influenza B genomes from before the Vic--Yam split it is difficult to determine how much of the linkage between PB1, PB2 and HA segments can be attributed to the population structure of influenza B viruses in the past.
-There is some evidence that the linkage between PB1 and HA might not be a phenomenon restricted to influenza B viruses.
-A technique used previously for producing vaccine seed strains involved selecting for HA-NA reassortants, which often yielded PB1-HA-NA reassortants as a side-effect \citep{bergeron2010,fulvini2011}.
-Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \citep{cobbin2013}.
-Additional evidence for a relationship between PB1 and HA segments in influenza A viruses is given by previous influenza pandemics, which were caused by avian-human influenza A virus reassortants.
+Although the linkage between the two polymerase segments and HA is more difficult to explain, it is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering ample time for the budding Victoria lineage to accumulate alleles causing reassortment incompatibility.
+Without more data it is difficult to tell how much of the linkage between PB1, PB2 and HA segments can be attributed to the population structure of influenza B viruses in the past.
+However, there is some evidence that the linkage between PB1 and HA might not be a phenomenon restricted to influenza B viruses.
 It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments derived from avian influenza A viruses \citep{kawaoka1989}.
-However, there have also been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \citep{gregory2002} and H1N1/09 in 2009 \citep{smith2009}.
+In addition, outdated techniques for producing vaccine seed strains through selection for HA-NA reassortants often yielded PB1-HA-NA reassortants as a side-effect \citep{bergeron2010,fulvini2011}.
+Recent experiments have found that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \citep{cobbin2013}.
+Bu we would also like to point out that there have been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \citep{gregory2002} and H1N1/09 in 2009 \citep{smith2009}.
 
-The association between Victoria and Yamagata lineage PB1-PB2-HA complexes should be relatively straightforward to explore in the lab.
-Using previously developed plasmid systems \citep{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
-There is some evidence, based on the frequency and persistence times of different reassortant classes we have observed, to suggest that there should be a hierarchy of reassortant fitness starting with PB1+PB2+HA reassortants which should be the most fit, followed by PB1+2/HA, then PB1+HA/PB2 and finally PB2+HA/PB1 reassortants with the lowest fitness.
+We believe that the association between PB1, PB2 and HA segments should be relatively straightforward to explore in the lab.
+Reverse genetics systems have been developed for influenza B viruses \citep{hoffmann2002}, which would allow the creation of artificial reassortants.
+Based on the frequency and persistence times of different reassortant classes we have observed, we expect a hierarchy of reassortant fitness starting with PB1+PB2+HA reassortants which should be the most fit, followed by PB1+2/HA, then PB1+HA/PB2 and finally PB2+HA/PB1 reassortants with the lowest fitness.
 We believe that this is the most direct approach to unravelling the mechanism responsible for the linkage within the PB1-PB2-HA complex.
@@
 If more segments get recruited to the PB1-PB2-HA complex, the process could continue until none of the segments are able to reassort across the Victoria--Yamagata lineage boundary.
-The two alternatives to the speciation hypothesis are the `\textit{status quo}' model, where the influenza B genome continues to be homogenized via gene flow with the exception of PB1, PB2 and HA segments and the extinction scenario, whereby one of the two PB1-PB2-HA complexes goes extinct, marking the return of single-strain dynamics in the influenza B virus population.
+The two alternatives to the speciation hypothesis are the `\textit{status quo}' model, where the influenza B genome continues to be homogenized via gene flow with the exception of PB1, PB2 and HA segments and the extinction scenario, if one of the two PB1-PB2-HA complexes goes extinct, marking the return of single-strain dynamics in the influenza B virus population.
 The eventual fate of influenza B viruses will be mostly determined by the combined effects of reassortment frequency and epistatic interactions between segments over long periods of time.
@@
 To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
-In addition, by treating the relative position of each isolate within a phylogeny of one segment as a label and modelling it on phylogenies of other segments we have also developed a metric similar to `between population' diversity used in calculating $F_{ST}$, which is capable of quantifying reticulate evolution-induced loss of diversity between partitions of taxa in two or more phylogenies.
+In addition, by treating the relative position of each isolate within a phylogeny of one segment as a label and modelling it on phylogenies of other segments we have also developed a metric similar to `between population' diversity used in calculating $F_{ST}$, which is capable of quantifying reticulate evolution-induced loss of diversity between partitions of taxa in multiple phylogenies.
 
@@
 We argue that this is due to selection against viruses with mixed-lineage PB1-PB2-HA complexes.
-Given sufficient time it should become clear whether PB1-PB2-HA complexes of human influenza B viruses are simply resisting gene flow whilst the rest of the influenza B virus genome is repeatedly homogenized or whether the two PB1-PB2-HA complexes are recruiting the rest of the genomic segments into co-adapted and co-reassorting segment complexes on their way to sympatric speciation.
+Given sufficient time it should become clear whether PB1-PB2-HA complexes of human influenza B viruses are simply resisting gene flow whilst the rest of the influenza B virus genome is repeatedly homogenized or whether the two PB1-PB2-HA complexes are recruiting the rest of the genomic segments into reassortment-incompatible segment complexes on their way to sympatric speciation.
 
@@
 \section*{Methods}
-We compiled a primary dataset of 452 complete influenza B genomes from GISAID \citep{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in Supplementary information).
+We compiled a primary dataset of 452 complete influenza B genomes from GISAID \citep{GISAID} dating from 1984 to 2012.
 The longest protein coding region of each segment was extracted and used for all further analyses.
@@
 To date there has been little evidence of homologous recombination in influenza viruses \citep{chare2003,boni2008,han2010}.
-The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \citep{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \citep{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
-Each strain was thus assigned 8 lineages depending on the combination of lineages from which their genomes were derived, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as (V,V,V,V,V,V,V,Y).
+The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \citep{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment since B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \citep{lindstrom1999}.
+B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage for the NS segment.
+Every segment in each genome thus received either a Vic or a Yam lineage designation, for example the strain B/Victoria/2/87 received V-V-V-V-V-V-V-Y, since its NS segment is derived from the Yam lineage and the rest of the genome is Vic.
 
-We also collated a secondary dataset from all complete influenza B virus genomes available on Genbank as of May 5, 2014.
+We also collated a secondary dataset from all complete influenza B virus genomes available on GenBank as of May 5, 2014.
 After removing isolates that had considerable portions of any sequence missing, were isolated prior to 1980 or were suspected of having a contaminant sequence in any segment, we were left with 1603 sequences.
@@
 Temporally-calibrated phylogenies were recovered for each segment in the primary dataset using Markov chain Monte Carlo (MCMC) methods in the BEAST software package \citep{drummond2012}.
-Here, we modeled the substitution process using the HKY model of nucleotide substitution \citep{hky1985}, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \citep{obrien2009}.
+We modeled the substitution process using the HKY model of nucleotide substitution \citep{hky1985}, with separate transition models for each of the 3 codon partitions, and additionally estimated realized synonymous and non-synonymous substitution counts \citep{obrien2009}.
 We used a flexible Bayesian skyride demographic model \citep{minin2008}.
 We accounted for incomplete sampling dates for 94 sequences (of which 93 had only year and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
-A relaxed molecular clock was used, where branch lengths are drawn from a lognormal distribution \citep{drummond2006}.
+A relaxed molecular clock was used, where branch rates are drawn from a lognormal distribution \citep{drummond2006}.
 We ran 3 independent MCMC chains, each with 200 million states, sampled every 20,000 steps and discarded the first 10\% of the MCMC states as burn-in.
@@
 In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.
+Inter-phylogeny labeling approaches have been extensively used in the past to investigate reticulate evolution in influenza A viruses and HIV \citep{lycett2012,ward2013,lu2014}.
 
@@
 We had 3 replicate analyses of each segment and in order to calculate $f(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
-We subsampled our combined posterior distribution of trees to give a total of 2700 trees on which to analyze $\dtmrca$.
+We subsampled our combined posterior distribution of trees to give a total of 2,700 trees on which to analyze $\dtmrca$.
 
@@
 Our approach exploits the branch scaling used by BEAST \citep{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
-This method operates under the assumption that the segment trees capture the overall `species tree' (or `virus tree') of influenza B viruses quite well.
+In the absence of reassortment we expect the tree of every segment to recapitulate the `virus tree', a concept analogous to `species trees' in population genetics.
+Our method operates under the assumption that the segment trees capture this `virus tree' of influenza B viruses quite well.
 It is not an unreasonable assumption, given the seasonal bottlenecks influenza viruses experience.
 This makes it almost certain that influenza viruses circulating at any given time point are derived from a single genome that existed in the recent past.
-The $\dtmrca$ statistic essentially captures a facet of this influenza B `species tree' by quantifying the temporal distance between an admixture event and the original lineage that was replaced by the incoming lineage (see Figure \ref{speciesTree}).
+The $\dtmrca$ statistic essentially quantifies the temporal distance between admixture events and nodes in the `virus tree' (see Figure \ref{speciesTree}).
 Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \citep{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \citep{ruane2011}.
@@
 \subsection*{Linkage disequilibrium across the influenza B genome}
-We used the secondary Genbank dataset with 1603 complete genome sequences to estimate linkage disequilibrium (LD) between amino acid loci across the longest proteins encoded by each segment of the influenza B virus genome.
+We used the secondary GenBank dataset with 1603 complete genome sequences to estimate linkage disequilibrium (LD) between amino acid loci across the longest proteins encoded by each segment of the influenza B virus genome.
 To quantify LD we adapt the $\chiSq$ statistic from \citep{hedrick1986}:
@@
 We find that $D'$ is almost uniformly high across the influenza B virus genome and close to 1.0 for almost any pair of polymorphic loci.
-This is because most amino acid alleles in the population exist transiently, meaning that they don't get a chance to reassort and we only observe them within the backgrounds of more persistent alleles, which $D'$ quantifies as complete LD.
+This is because most amino acid alleles in the population exist transiently, meaning that they do not get a chance to reassort and we only observe them within the backgrounds of more persistent alleles, which $D'$ quantifies as complete LD.
 We think that metrics related to $r^2$, like $\chiSq$, perform much better on temporal data such as ours in finding persistent associations between alleles and are easier to interpret.
@@
 \subsection*{Data availability}
-Python scripts used to process trees and sequences, as well as their output are publicly available at:
+Python scripts used to process trees and sequences are available at:
+\\https://github.com/evogytis/fluB/tree/master/scripts.
+
+Output files from scripts, lineage designations, MCC trees, acknowledgment tables, accession numbers and redacted XML files (per GISAID Data Access Agreement) are publicly available at:
 \\https://github.com/evogytis/fluB/tree/master/data.

### 2014-07-23 fluB.tex
Move speciesTree figure to supp.
Reduces number of figures in the main text.

@@
 Although the linkage between the two polymerase segments and HA is more difficult to explain, it is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering ample time for the budding Victoria lineage to accumulate alleles causing reassortment incompatibility.
-Without more data it is difficult to tell how much of the linkage between PB1, PB2 and HA segments can be attributed to the population structure of influenza B viruses in the past.
-However, there is some evidence that the linkage between PB1 and HA might not be a phenomenon restricted to influenza B viruses.
+Without more genomic data from the past, however, it is difficult to estimate to what extent influenza B virus population structure contributed to the development of the current segment linkage.
+There is some evidence that the linkage between PB1 and HA might not be a phenomenon restricted to influenza B viruses.
 It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments derived from avian influenza A viruses \citep{kawaoka1989}.
@@
 
-\begin{figure}[h]
-	\centering	
-	\includegraphics[width=0.85\textwidth]{figures/InfB_methodSpeciesTree.pdf}
-	\caption{\textbf{Calculating $\dtmrca$ from a species tree perspective.}
-Consider an organism that has diverged into 3 taxa (A, B, C) with a genome comprised of 3 segments (blue, green and red).
-Due to reassortments taxa A and B share a slightly more recent TMRCA in the red segment, likewise for taxa B and C in the green segment.
-By comparing differences in TMRCAs between taxa A-B, A-C and B-C in blue, red and green segments we would find that the red segment has a lower `reassortment distance' (X) than the green segment (Y).
-}
-	\label{speciesTree}
-\end{figure}
-
 Calculating $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.

### 2014-07-23 fluB_supp.tex
Clean language per #38
Also add references to similar methods.
Move speciesTree figure to supp.
Reduces number of figures in the main text.

@@
 The numbers shown are the mean inferred number of trait transitions (minus one to account for the initial Vic--Yam split) in a given tree and trait combination.
-The numbers in brackets correspond to 95\% highest posterior density intervals.}
+The numbers in brackets correspond to 95\% highest posterior density intervals.
+Transitions may not be independent when more than one segment reassorts at the same time.}
 \label{Ntransitions}
@@
 \label{deltaTMRCAreplicates}
+\end{figure}
+
+\begin{figure}[h]
+	\centering	
+	\includegraphics[width=0.85\textwidth]{figures/InfB_methodSpeciesTree.pdf}
+	\caption{\textbf{Calculating $\dtmrca$ from a species tree perspective.}
+Consider an organism that has diverged into 3 taxa (A, B, C) with a genome comprised of 3 segments (blue, green and red).
+Due to reassortments taxa A and B share a slightly more recent TMRCA in the red segment, likewise for taxa B and C in the green segment.
+By comparing differences in TMRCAs between taxa A-B, A-C and B-C in blue, red and green segments we would find that the red segment has a lower `reassortment distance' (X) than the green segment (Y).
+}
+	\label{speciesTree}
 \end{figure}

### 2014-07-23 fluB.tex
Change wording slightly.
Reassortment incompatibility sounds better.

@@
 
-\subsection*{Potential mechanisms for reassortant incompatibility}
+\subsection*{Potential mechanisms for reassortment incompatibility}
 Unfortunately, the limited amount of genomic data available for the early years of the Vic--Yam split precludes any attempts of answering whether selection or drift have led to the current linkage of PB1, PB2 and HA segments.
-Although the origins of the linkage between these three segments might be difficult to explain, we can speculate on the nature of reassortant incompatibility.
+Although the origins of the linkage between these three segments might be difficult to explain, we can speculate on the nature of reassortment incompatibility.
 For example, it is intuitive for why this might be the case for PB1 and PB2: both proteins interact directly as part of the RNA-dependent RNA polymerase heterotrimer.

### 2014-07-25 fluB.tex
Revise main text, most changes to discussion.

@@
 
-Here, we extend previous methods to reveal an evolutionarily intriguing pattern of reassortment in influenza B.
-In our approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual isolates in the tree of the other segments.
+Here, we extend previous methods to reveal an intriguing pattern of reassortment in influenza B.
+In our approach, membership to either the Victoria or Yamagata lineage in one segment is used to label the individual isolates in the tree of the other segments.
 By modelling the transition between labels on a phylogenetic tree, reassortment events which result in the replacement of one segment's lineage by another show up as label changes along a branch (Figure \ref{methodFig}).
@@
 The differentiation into Vic and Yam lineages can be seen in all segments (Figure \ref{genomeGrid}) and is followed by inter-lineage reassortment events.
-In the phylogenetic trees of some segments either the Victoria or Yamagata lineage has become the `trunk' of the tree (yellow vs purple bars in Figure \ref{genomeGrid}) following reassortment.
-This pattern is apparent in the PA, NP, NA, MP and NS segments.
+In the phylogenetic trees of the PA, NP, NA, MP and NS segments either the Victoria or Yamagata lineage has become the `trunk' of the tree, with present-day viruses deriving entirely from the Victoria or Yamagata lineage (yellow vs purple bars in Figure \ref{genomeGrid}) following reassortment.
 However, the Victoria and Yamagata lineages of PB1, PB2 and HA segments continue to co-circulate to this day.
-Periodic loss of diversity in PA, NP, NA, MP and NS segments is consistent with Victoria or Yamagata lineages becoming the `trunk' of the trees of those segments, while maintenance of parallel Victoria and Yamagata lineages results in continually increasing diversity in segments PB1, PB2 and HA (Figure \ref{tmrcaOT}).
+Periodic loss of diversity in PA, NP, NA, MP and NS segments is consistent with introgression of one lineage into the other in those segments, while maintenance of parallel Victoria and Yamagata lineages results in continually increasing diversity in segments PB1, PB2 and HA (Figure \ref{tmrcaOT}).
 The PB1, PB2 and HA segments from present-day viruses maintain a common ancestor in $\sim$1983 and thus accumulate genetic diversity since the split of those segments into Vic and Yam lineages, while other segments often lose diversity with ancestors to present-day viruses appearing between $\sim$1991 and $\sim$1999.
@@
 This difference in TMRCAs, or $\dtmrca$, is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
-We normalize our $\dtmrca$ comparisons to account for alignment length induced uncertainty in tree topology stability over the course of the MCMC chain (see Methods).
+We normalize our $\dtmrca$ comparisons to account for uncertainty in tree topology (see Methods).
 Figure \ref{deltaTMRCA} shows normalized mean $\dtmrca$ values for all pairs of trees.
@@
 
-Linkage disequilibrium (LD) is a measure of non-random association between pairs of alleles at different polymorphic loci within a population.
-To estimate LD no other information, except for haplotype and allele frequencies at polymorphic amino acid or nucleotide sites, is required.
-Although not an external validation of previous phylogenetic results, we observe greater amino acid LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}) in a large secondary dataset (see Methods).
+We next sought to confirm our findings through measurement of linkage disequilibrium (LD), a measure of non-random association between polymorphic loci within a population.
+We estimated LD directly from haplotype frequencies at polymorphic amino acid sites (see Methods), thereby avoiding phylogenetic reconstruction or Vic/Yam lineage assignment.
+We observe greater amino acid LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}) in a large secondary dataset.
 This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting non-synonymous alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
@@
 
-\subsection*{The origins of the linkage remain unknown}
+\subsection*{Linkage between PB1, PB2 and HA gene segments}
+In this paper we show that the PB1, PB2 and HA segments of influenza B viruses are the only ones that have continuously maintained separate Vic and Yam lineages, while other segments have fixed either Vic or Yam lineages (Figures \ref{genomeGrid}, \ref{lineageRatiosOverTime} and \ref{railroadPlot}).
+Evidence suggests that this is a result of prolonged lack of reassortment between Vic and Yam lineages in PB1, PB2 and HA (Figure \ref{betweenDiversity}) which possess co-assorting sequences detectable as high linkage disequilibrium (Figure \ref{segmentLD}).
+The vast majority of the sampled evolutionary history of each segment of influenza B viruses since the split of Vic and Yam lineages has been spent in association with either completely Victoria or completely Yamagata lineage derived PB1-PB2-HA complexes (Figure \ref{stateTime}), suggesting that having `pure' lineage PB1-PB2-HA complexes is important for whole-genome fitness.
+We propose that this pattern of coassortment is due to the action of selection and not simply biased or rare reassortment.
+
 The origin of the strong genetic linkage between PB1, PB2 and HA segments remains unclear.
-We believe there are two alternative, but similar explanations for the origins of the strong genetic linkage between PB1, PB2 and HA segments: mutation-driven co-evolution \citep{presgraves2010} and Dobzhansky-Muller incompatibility \citep{dobzhansky1937,muller1942}.
+We believe there are two alternative, but similar explanations for the origins of the strong genetic linkage between these segments: mutation-driven co-evolution \citep{presgraves2010} and Dobzhansky-Muller incompatibility \citep{dobzhansky1937,muller1942}.
 Mutation-driven co-evolution \citep{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \citep{lee2008}, and evolves as a by-product of adaptation.
@@
 Dobzhansky-Muller incompatibility operates in a similar way, but the main difference from the scenario described earlier is that the incompatible alleles are neutral or nearly neutral in their native background and become deleterious or lethal when combined with non-native backgrounds.
+Emergence of DM incompatibility is aided by geographic isolation.
+Interestingly, the Victoria lineage of HA was restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering ample time for the budding Victoria lineage to accumulate alleles causing reassortment incompatibility.
+However, without more genomic data from the past, it is difficult to estimate to what extent influenza B virus population structure contributed to the development of the current segment linkage.
 
@@
 
-Although the linkage between the two polymerase segments and HA is more difficult to explain, it is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering ample time for the budding Victoria lineage to accumulate alleles causing reassortment incompatibility.
-Without more genomic data from the past, however, it is difficult to estimate to what extent influenza B virus population structure contributed to the development of the current segment linkage.
 There is some evidence that the linkage between PB1 and HA might not be a phenomenon restricted to influenza B viruses.
@@
 Recent experiments have found that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \citep{cobbin2013}.
-Bu we would also like to point out that there have been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \citep{gregory2002} and H1N1/09 in 2009 \citep{smith2009}.
+However, there have been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \citep{gregory2002} and H1N1/09 in 2009 \citep{smith2009}.
 
@@
 In this context, we see three potential paths of evolution for influenza B viruses.
-If more segments get recruited to the PB1-PB2-HA complex, the process could continue until none of the segments are able to reassort across the Victoria--Yamagata lineage boundary.
-The two alternatives to the speciation hypothesis are the `\textit{status quo}' model, where the influenza B genome continues to be homogenized via gene flow with the exception of PB1, PB2 and HA segments and the extinction scenario, if one of the two PB1-PB2-HA complexes goes extinct, marking the return of single-strain dynamics in the influenza B virus population.
-The eventual fate of influenza B viruses will be mostly determined by the combined effects of reassortment frequency and epistatic interactions between segments over long periods of time.
+If more segments get recruited to the PB1-PB2-HA complex, the process could continue until `speciation' occurs in which none of the segments are able to reassort across the Victoria--Yamagata lineage boundary.
+Alternatively, the influenza B genome could continue to be homogenized via gene flow with the exception of PB1, PB2 and HA segments or one of the two PB1-PB2-HA complexes could go extinct, marking the return of single-strain dynamics in the influenza B virus population.
+The eventual fate of influenza B viruses will likely be determined by the combined effects of reassortment frequency and the strength of epistatic interactions between segments.
 
-\section*{Conclusion}
-We have used the $\dtmrca$ statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
-We believe that patristic distance methods such as this, though themselves far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies.
-One of many useful applications of the $\dtmrca$ method would be identifying clades or taxa that are products of reticulate evolution.
-$\dtmrca$ measures between independent analyses of the same alignment (`within-alignment') could be used as a cutoff to detect outlier taxa with greater than expected `between-alignment' $\dtmrca$ values.
-To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
-In addition, by treating the relative position of each isolate within a phylogeny of one segment as a label and modelling it on phylogenies of other segments we have also developed a metric similar to `between population' diversity used in calculating $F_{ST}$, which is capable of quantifying reticulate evolution-induced loss of diversity between partitions of taxa in multiple phylogenies.
+%\section*{Conclusion}
+%We have used the $\dtmrca$ statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
+%We believe that patristic distance methods such as this, though themselves far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies.
+%One of many useful applications of the $\dtmrca$ method would be identifying clades or taxa that are products of reticulate evolution.
+%$\dtmrca$ measures between independent analyses of the same alignment (`within-alignment') could be used as a cutoff to detect outlier taxa with greater than expected `between-alignment' $\dtmrca$ values.
+%To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
+%In addition, by treating the relative position of each isolate within a phylogeny of one segment as a label and modelling it on phylogenies of other segments we have also developed a metric similar to `between population' diversity used in calculating $F_{ST}$, which is capable of quantifying reticulate evolution-induced loss of diversity between partitions of taxa in multiple phylogenies.
 
-In this paper we apply a novel combination of population genetics and phylogenetic methods to full genome sequences in order to describe and quantify reassortment patterns in influenza B viruses circulating in humans from 1980 to the present day.
-Our main finding is that in influenza B viruses only PB1, PB2 and HA segments maintain both Victoria and Yamagata lineages which associate with segments of their own lineage, yielding two co-circulating PB1-PB2-HA complexes: one entirely derived from Victoria and one from Yamagata lineage segments.
-We argue that this is due to selection against viruses with mixed-lineage PB1-PB2-HA complexes.
-Given sufficient time it should become clear whether PB1-PB2-HA complexes of human influenza B viruses are simply resisting gene flow whilst the rest of the influenza B virus genome is repeatedly homogenized or whether the two PB1-PB2-HA complexes are recruiting the rest of the genomic segments into reassortment-incompatible segment complexes on their way to sympatric speciation.
+%In this paper we apply a novel combination of population genetics and phylogenetic methods to full genome sequences in order to describe and quantify reassortment patterns in influenza B viruses circulating in humans from 1980 to the present day.
+%Our main finding is that in influenza B viruses only PB1, PB2 and HA segments maintain both Victoria and Yamagata lineages which associate with segments of their own lineage, yielding two co-circulating PB1-PB2-HA complexes: one entirely derived from Victoria and one from Yamagata lineage segments.
+%We argue that this is due to selection against viruses with mixed-lineage PB1-PB2-HA complexes.
+%Given sufficient time it should become clear whether PB1-PB2-HA complexes of human influenza B viruses are simply resisting gene flow whilst the rest of the influenza B virus genome is repeatedly homogenized or whether the two PB1-PB2-HA complexes are recruiting the rest of the genomic segments into reassortment-incompatible segment complexes on their way to sympatric speciation.
 

### 2014-08-07 fluB.tex
Fix #40, #39
Clarify deltaTMRCA confusion. I've decided that deltaTMRCA is best
thought of as the mean of absolute deviations in TMRCAs between 2
trees. We get normalized deltaTMRCA by quantifying deltaTMRCAs within
(=phylogenetic noise) and between (=phylogenetic noise + signal) trees.
The result of this is a measure of how similar the signal is to noise,
which in our case is the desired outcome. Hopefully that makes sense.
Also move all the LD figures to supplementary. It was done as a
confirmation of results and also addresses the figure number problem.

@@
 Although we do not find evidence of differences in total number of reassortments between segments (see Supplementary information), we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
-This difference in TMRCAs, or $\dtmrca$, is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
+The summary statistic we use that reflects this difference in TMRCAs, $\dtmrca$, is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
 We normalize our $\dtmrca$ comparisons to account for uncertainty in tree topology (see Methods).
-Figure \ref{deltaTMRCA} shows normalized mean $\dtmrca$ values for all pairs of trees.
+Figure \ref{deltaTMRCA} shows normalized $\dtmrca$ values for all pairs of trees.
 Most segment pairs show very low values for this statistic with $\dtmrca \approx 0.1$, indicating that $\dtmrca$ measurements between replicate posterior samples from the same segment are up to 10 times smaller than $\dtmrca$ values between posterior samples from different segments.
@@
 	\label{deltaTMRCA}
-\end{figure}
-
-We next sought to confirm our findings through measurement of linkage disequilibrium (LD), a measure of non-random association between polymorphic loci within a population.
-We estimated LD directly from haplotype frequencies at polymorphic amino acid sites (see Methods), thereby avoiding phylogenetic reconstruction or Vic/Yam lineage assignment.
-We observe greater amino acid LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}) in a large secondary dataset.
-This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting non-synonymous alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
-We conclude that Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions.
-Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 4 sites on PB1, 4 on PB2 and 4 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
-These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
-In addition, there are sites on PB1, PB2, HA and NA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
-
-\begin{figure}[h]
-	\centering	
-	\includegraphics[width=0.65\textwidth]{figures/InfB_LDcircle.1600.aa.ChiSqdf.minorCutoff1percent.png}
-	\caption{\textbf{LD comparison between influenza B proteins.}
-Pairwise comparisons of linkage disequilibrium between amino acid sites on influenza B proteins in the secondary dataset.
-Many polymorphic amino acid sites on PB1, PB2 and HA proteins exhibit high LD between themselves, followed by the NA protein.
-This is evidence of a considerable number of co-assorting alleles within these proteins.}
-	\label{segmentLD}
 \end{figure}
@@
 \begin{equation}
-\dtmrca(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
+normalized \dtmrca(A_i, B_i) = \frac{\dtmrca(A_i, A_i') + \dtmrca(B_i, B_i')}{2 \, \dtmrca(A_i, B_i)},
 \end{equation}
-where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $n$ is the total number of pairwise comparisons available between sets of tips.
+where $\dtmrca(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $n$ is the total number of pairwise comparisons available between sets of tips.
 Thus, $g(A_{ij},B_{ij})$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree $A$ and the \textit{i}th posterior sample of tree $B$.
-Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), to control for variability in tree topology stability over the course of the MCMC chain.
-We had 3 replicate analyses of each segment and in order to calculate $f(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
-We subsampled our combined posterior distribution of trees to give a total of 2,700 trees on which to analyze $\dtmrca$.
+Additionally, $\dtmrca(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), which is used in the normalization procedure to control for variability in tree topology stability over the course of the MCMC chain (see Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates}).
+We had 3 replicate analyses of each segment and in order to calculate $\dtmrca(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
+We subsampled our combined posterior distribution of trees to give a total of 2,700 trees on which to analyze normalized $\dtmrca$.
 
-Calculating $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
+Calculating normalized $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
 Our approach exploits the branch scaling used by BEAST \citep{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.

### 2014-08-07 fluB_supp.tex
Fix #40, #39
Clarify deltaTMRCA confusion. I've decided that deltaTMRCA is best
thought of as the mean of absolute deviations in TMRCAs between 2
trees. We get normalized deltaTMRCA by quantifying deltaTMRCAs within
(=phylogenetic noise) and between (=phylogenetic noise + signal) trees.
The result of this is a measure of how similar the signal is to noise,
which in our case is the desired outcome. Hopefully that makes sense.
Also move all the LD figures to supplementary. It was done as a
confirmation of results and also addresses the figure number problem.

@@
 \renewcommand{\thetable}{S\arabic{table}}
+
+\subsection*{Confirmation of primary findings}
+We sought to confirm our findings through measurement of linkage disequilibrium (LD), a measure of non-random association between polymorphic loci within a population.
+We estimated LD directly from haplotype frequencies at polymorphic amino acid sites (see Methods) in the secondary dataset, thereby avoiding phylogenetic reconstruction or Vic/Yam lineage assignment.
+We observe greater amino acid LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}) in a large secondary dataset.
+This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting non-synonymous alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
+We conclude that Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions.
+Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 4 sites on PB1, 4 on PB2 and 4 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
+These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
+In addition, there are sites on PB1, PB2, HA and NA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.
+
+\begin{figure}[h]
+	\centering	
+	\includegraphics[width=0.85\textwidth]{supp_figures/InfB_LDcircle.1600.aa.ChiSqdf.minorCutoff1percent.png}
+	\caption{\textbf{LD comparison between influenza B proteins.}
+Pairwise comparisons of linkage disequilibrium between amino acid sites on influenza B proteins in the secondary dataset.
+Many polymorphic amino acid sites on PB1, PB2 and HA proteins exhibit high LD between themselves, followed by the NA protein.
+This is evidence of a considerable number of co-assorting alleles within these proteins.}
+	\label{segmentLD}
+\end{figure}
+
+\begin{figure}
+\centering  
+\includegraphics[width=0.85\textwidth]  {supp_figures/InfB_LDheatmap.PB1_PB2_HA.1600.aa.ChiSqdf.minorCutoff1percent.png}
+\caption{\textbf{Heatmap of linkage disequilibrium ($\chiSq$) between amino acid sites on PB1, PB2 and HA proteins.}
+Numbers next to each row and column correspond to amino acid site number within a given protein starting from methionine.
+Amino acid sites exhibiting reciprocally high LD between PB1, PB2 and HA proteins are: 176, 211, 213, 508 (PB1), 115, 399, 470, 642 (PB2) and 96, 163, 520 and 570 (HA).
+Sites 211 and 213 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
+Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211 and 213 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
+Previous research \citep{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
+Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 position 211 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at position 211 and threonine (T) at position 213.
+It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
+Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \citep{fodor2004}.
+Similarly, site 399 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \citep{guilligay2008}.}
+\label{ChiCore}
+\end{figure}
 
@@
 \label{ChiGenome}
-\end{figure}
-
-\begin{figure}
-\centering  
-\includegraphics[width=0.85\textwidth]  {supp_figures/InfB_LDheatmap.PB1_PB2_HA.1600.aa.ChiSqdf.minorCutoff1percent.png}
-\caption{\textbf{Heatmap of linkage disequilibrium ($\chiSq$) between amino acid sites on PB1, PB2 and HA proteins.}
-Numbers next to each row and column correspond to amino acid site number within a given protein starting from methionine.
-Amino acid sites exhibiting reciprocally high LD between PB1, PB2 and HA proteins are: 176, 211, 213, 508 (PB1), 115, 399, 470, 642 (PB2) and 96, 163, 520 and 570 (HA).
-Sites 211 and 213 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
-Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211 and 213 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
-Previous research \citep{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
-Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 position 211 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at position 211 and threonine (T) at position 213.
-It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
-Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \citep{fodor2004}.
-Similarly, site 399 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \citep{guilligay2008}.}
-\label{ChiCore}
 \end{figure}

### 2014-08-07 fluB_supp.tex
Add more dTMRCA explanation to supp.

@@
 By comparing differences in TMRCAs between taxa A-B, A-C and B-C in blue, red and green segments we would find that the red segment has a lower `reassortment distance' (X) than the green segment (Y).
+In the absence of reassortment we expect every segment in the genome to have the same tree, \textit{i.e.} the tree of every segment should recapitulate the `virus' tree (analogous to `species' trees in diploid population genetics), including the dates of nodes.
+Due to population bottlenecks influenza viruses go through each year we expect strains isolated at any given time to have descended from a single recent virus genome.
+This descent from a single genome should therefore be reflected in the TMRCAs of all segments, the only exception being reassortment, which will dramatically alter the TMRCAs of the reassorted segment tree with respect to the background onto which it is reassorting.
 }

### 2014-08-08 fluB.tex
Fix #40
Dtmrca now refers to the normalized measure and dtmrca refers to the
mean of TMRCA deviations.
Should now be unambiguous across main and supplementary texts. Also fix
a few figures.

@@
 \newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}}
+\newcommand{\undtmrca}{\delta_\mathrm{TMRCA}}
 \newcommand{\dspr}{d_\mathrm{SPR}}
@@
 Although we do not find evidence of differences in total number of reassortments between segments (see Supplementary information), we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
-The summary statistic we use that reflects this difference in TMRCAs, $\dtmrca$, is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
-We normalize our $\dtmrca$ comparisons to account for uncertainty in tree topology (see Methods).
-Figure \ref{deltaTMRCA} shows normalized $\dtmrca$ values for all pairs of trees.
+The summary statistic we use that reflects this difference in TMRCAs, $\undtmrca$, is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
+We normalize our $\undtmrca$ comparisons to arrive at $\dtmrca$, which accounts for uncertainty in tree topology (see Methods).
+Figure \ref{deltaTMRCA} shows $\dtmrca$ values for all pairs of trees.
 Most segment pairs show very low values for this statistic with $\dtmrca \approx 0.1$, indicating that $\dtmrca$ measurements between replicate posterior samples from the same segment are up to 10 times smaller than $\dtmrca$ values between posterior samples from different segments.
-PB1, PB2 and HA trees, on the other hand, exhibit normalized $\dtmrca$ values that are much higher.
+PB1, PB2 and HA trees, on the other hand, exhibit $\dtmrca$ values that are much higher.
 This shows that TMRCA differences between trees of PB1, PB2 and HA segments are, though noisy, occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
@@
 	\includegraphics[width=0.65\textwidth]{figures/InfB_normalizedMuDeltaTMRCA.png}
-	\caption{\textbf{Normalized mean $\dtmrca$ statistics for different segment pairs.}
+	\caption{\textbf{$\dtmrca$ statistics for different segment pairs.}
 PB1, PB2 and HA trees exhibit reciprocally highly similar TMRCAs, unlike most other pairwise comparisons.
@@
 \subsection*{Tree to tree similarities}
-We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ for a particular posterior sample $i$, following
+We express the normalized distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ for a particular posterior sample $i$, following
 \begin{equation}
-normalized \dtmrca(A_i, B_i) = \frac{\dtmrca(A_i, A_i') + \dtmrca(B_i, B_i')}{2 \, \dtmrca(A_i, B_i)},
+\dtmrca(A_i, B_i) = \frac{\undtmrca(A_i, A_i') + \undtmrca(B_i, B_i')}{2 \, \undtmrca(A_i, B_i)},
 \end{equation}
-where $\dtmrca(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $n$ is the total number of pairwise comparisons available between sets of tips.
+where $\undtmrca(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $n$ is the total number of pairwise comparisons available between sets of tips.
 Thus, $g(A_{ij},B_{ij})$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree $A$ and the \textit{i}th posterior sample of tree $B$.
-Additionally, $\dtmrca(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), which is used in the normalization procedure to control for variability in tree topology stability over the course of the MCMC chain (see Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates}).
-We had 3 replicate analyses of each segment and in order to calculate $\dtmrca(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
-We subsampled our combined posterior distribution of trees to give a total of 2,700 trees on which to analyze normalized $\dtmrca$.
+Additionally, $\undtmrca(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), which is used in the normalization procedure to control for variability in tree topology stability over the course of the MCMC chain (see Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates}).
+We had 3 replicate analyses of each segment and in order to calculate $\undtmrca(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
+We subsampled our combined posterior distribution of trees to give a total of 2,700 trees on which to analyze $\dtmrca$.
 
-Calculating normalized $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
+Calculating the normalized $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
 Our approach exploits the branch scaling used by BEAST \citep{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
@@
 This makes it almost certain that influenza viruses circulating at any given time point are derived from a single genome that existed in the recent past.
-The $\dtmrca$ statistic essentially quantifies the temporal distance between admixture events and nodes in the `virus tree' (see Figure \ref{speciesTree}).
-Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \citep{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \citep{ruane2011}.
+The $\undtmrca$ statistic essentially quantifies the temporal distance between admixture events and nodes in the `virus tree' (see Figure \ref{speciesTree}).
+We normalize $\undtmrca$ values to get $\dtmrca$, a measure which quantifies the extent to which the similarity of two independent trees resembles phylogenetic noise.
+The $\undtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \citep{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \citep{ruane2011}.
 

### 2014-08-08 fluB_supp.tex
Fix #40
Dtmrca now refers to the normalized measure and dtmrca refers to the
mean of TMRCA deviations.
Should now be unambiguous across main and supplementary texts. Also fix
a few figures.

@@
 
+\begin{figure}
+\centering  
+\includegraphics[width=0.95\textwidth]  {supp_figures/InfB_LDheatmap.1600.aa.ChiSqdf.minorCutoff1percent.png}
+\caption{\textbf{Heatmap of genome-wide linkage disequilibrium ($\chiSq$) between polymorphic amino acid sites.}
+Patterns of LD across the genome suggest a network of reciprocally linked amino acid sites on PB1, PB2, HA and, to some extent NA, proteins.
+Proximity of sites on heatmaps might not correspond to proximity of sites within proteins.}
+\label{ChiGenome}
+\end{figure}
+
 \subsection*{Analysis of within-lineage reassortment patterns}
@@
 We calculated approximate SPR distances \citep{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
-Approximate SPR distances were normalized using the procedure used to normalize $\dspr$ (see Methods):
+Normalized approximate SPR distances,$\dspr$, were recovered using (see Methods):
 \begin{equation}
@@
 \centering  
-\includegraphics[width=0.95\textwidth]  {supp_figures/InfB_LDheatmap.1600.aa.ChiSqdf.minorCutoff1percent.png}
-\caption{\textbf{Heatmap of genome-wide linkage disequilibrium ($\chiSq$) between polymorphic amino acid sites.}
-Patterns of LD across the genome suggest a network of reciprocally linked amino acid sites on PB1, PB2, HA and, to some extent NA, proteins.
-Proximity of sites on heatmaps might not correspond to proximity of sites within proteins.}
-\label{ChiGenome}
-\end{figure}
-
-\begin{figure}
-\centering  
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_robustCounting.png}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
-\caption{\textbf{Mean $\dtmrca$ between all pairs of trees of segments.}
-Mean $\dtmrca$ between trees of segments reveal that tip pairs in PB1, PB2 and HA trees have very similar TMRCAs.
-The upper tail of the 95\% HPD (HPDs are represented as vertical lines) interval of mean $\dtmrca$ values for PB1-PB2-HA and MP-NA trees do not exceed 3 years.}
+\caption{\textbf{$\undtmrca$ between all pairs of trees of segments.}
+$\undtmrca$ between trees of segments reveal that tip pairs in PB1, PB2 and HA trees have very similar TMRCAs.
+The upper tail of the 95\% HPD (HPDs are represented as vertical lines) interval of $\undtmrca$ values for PB1-PB2-HA and MP-NA trees do not exceed 3 years.}
 \label{deltaTMRCAtrees}
@@
 \includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
-\caption{\textbf{Mean $\dtmrca$ between replicate trees of each segment.}
-Mean $\dtmrca$ values between independent analyses of each segment show that mean $\dtmrca$ values rarely exceed 1 year.}
+\caption{\textbf{$\undtmrca$ between replicate trees of each segment.}
+$\undtmrca$ values between independent analyses of each segment show that mean $\undtmrca$ values rarely exceed 1 year.}
 \label{deltaTMRCAreplicates}
@@
 	\includegraphics[width=0.85\textwidth]{figures/InfB_methodSpeciesTree.pdf}
-	\caption{\textbf{Calculating $\dtmrca$ from a species tree perspective.}
+	\caption{\textbf{Calculating $\undtmrca$ from a species tree perspective.}
 Consider an organism that has diverged into 3 taxa (A, B, C) with a genome comprised of 3 segments (blue, green and red).

### 2014-08-08 fluB.tex
A few extra details.

@@
 Figure \ref{deltaTMRCA} shows $\dtmrca$ values for all pairs of trees.
-Most segment pairs show very low values for this statistic with $\dtmrca \approx 0.1$, indicating that $\dtmrca$ measurements between replicate posterior samples from the same segment are up to 10 times smaller than $\dtmrca$ values between posterior samples from different segments.
+Most segment pairs show very low values for this statistic with $\dtmrca \approx 0.1$, indicating that $\undtmrca$ measurements between replicate posterior samples from the same segment are up to 10 times smaller than $\undtmrca$ values between posterior samples from different segments.
 PB1, PB2 and HA trees, on the other hand, exhibit $\dtmrca$ values that are much higher.
@@
 \section*{Acknowledgements}
-We would like to thank Darren Obbard and Paul Wikramaratna for helpful discussions.
+We would like to thank Darren Obbard and Paul Wikramaratna for helpful discussions and anonymous reviewers for comments and suggestions.
 GD was supported by a Natural Environment Research Council studentship D76739X.

### 2014-09-02 correspondence/responseLetter.tex
Add response letter.
Using Trevor's response style, obviously.

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color} 
\usepackage{parskip}
\usepackage{float}
\usepackage{hyperref}

% text layout
\usepackage{geometry}
\geometry{textwidth=15.25cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace} 
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

% notation
\usepackage{amsmath}
\usepackage{amssymb}

\begin{document}

\newgeometry{top=4cm}

Dear MBE editorial board,

Please find attached our extensively revised manuscript entitled ``Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex''.  This is a resubmission of manuscript number MBE-14-0575. The editorial assessment stated that:

\begin{quote}
The reviewers made many suggestions for the improvement of this paper, and I agree that these improvements are needed. 
The value of this manuscript lies primarily in the analytical techniques presented. 
The result is simple but of interest.

As both reviewers noted, the discussion consists mostly of rambling speculation. 
I found it hard to follow (particularly the part about balancing selection) and neither convincing nor very interesting. 
There is nothing wrong with just admitting that you don't understand the process behind this pattern and suggesting some experiments. 
The writing needs to be improved throughout.

The word fixation is often used in a confusing manner. 
For example, see line 28 page 5: "We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage." 
See also the sentence on lines 48-49 of that page.

The first line of the abstract and the first paragraph of the introduction make it sound like influenza B is an increasingly worrisome public health problem. 
Influenza B does not need to be hyped to be interesting, and this type of "journalism" can makes the reader wonder about the validity of everything that follows.
\end{quote}

We would like to thank the reviewers' and the editor's thorough and insightful comments.
We appreciate the time they have taken to review the manuscript and the acknowledgment of its value.
We agree that the discussion, as it was, contained mostly irrelevant information and speculation.
The discussion has been considerably slimmed down to the most essential information and reworded in places.
We have additionally altered the terminology that caused confusion, e.g. the word ``fixation'' is now explicitly refered to as a population genetics term and used only where applicable.
We have also toned down the wording with respect to the importance of influenza B viruses.

Point-by-point review responses are attached, as well as a diffed PDF detailing all the changes that have been made since the initial submission.

Sincerely,\\
Gytis Dudas

\restoregeometry

\newpage

\section*{Reviewer responses}

Original reviewer criticisms are in plain text.  Our responses follow in \textbf{bold}.

%%% REVIEWER 1 %%%
\section*{Reviewer 1}
1. "Could the authors provide an estimate of how many reassortment events have occurred in these trees in total, based on the no. of transitions?"

\textbf{We have added a new supplementary figure (Figure S10), which shows the mean number of transitions for each trait/label as well as their 95\% HPDs in each tree.
We advise to interpret the new figure with caution - many successful reassortment events we have observed involve more than one segment (see Figure 6) and thus these counts are not independent.}

2. "It would be useful, as comparison, to include a Supplementary figure similar to Fig 2, but with branches colored to show reassortment events involving the other segments (PA, NP, NA, MP, NS -- obviously this is a lot of combinations and a lot of trees, so maybe just show a representative subset)."

\textbf{We believe that although this is an interesting point, it is outside the scope of the current manuscript.
Additionally, the complete reassortment history of all segments is available in Figure 6, which can be combined with Figure 2 to reveal the phylogenetic position of all reassorting lineages.
To avoid making excessive amounts of figures with all possible combinations of trees and traits we have made the MCC trees (including trait mapping) publicly available at: \url{https://github.com/evogytis/fluB/tree/master/data/mcc\%20trees}.}

3. "I didn't find any accession numbers or description of the data set in the Supplement.  
It would be useful to know how geographically representative the data set is.  
It certainly is possible that in a spatially biased data set certain variants (as in those with mixed PB1-PB2-HA complexes) could be under-represented."

\textbf{Apologies for this. We have a separate file listing all sequence accession numbers from primary (GISAID) and secondary (GenBank) datasets but forgot to include it in the initial submission.
These files are available at \url{https://github.com/evogytis/fluB/tree/master/data/acknowledgement\%20tables}.
To make it easier to judge how representative the sequence sampling was we have included two new supplementary figures (Figures S8 and S9) which show the geographic distribution of sequences in the primary and secondary datasets.}

4. "It seems that most of the mixed PB1-PB2-HA complexes are found earlier in the tree, closer to the root, and become less frequent as time goes on.  Is this worth expanding upon?"

\textbf{The reviewer raises a very interesting point. We have addressed this on page 12: 
``Indeed, we observe that PB1--PB2 reassortants are the rarest and least persistent among mixed-lineage PB1-PB2-HA strains and have not been isolated in great numbers.
In fact most reassortants breaking the PB1-PB2-HA complex apart have occured in the past, close to the split of Vic and Yam lineages and have become very rare since.''
Unfortunately due to the low numbers of reassortments this is no more than speculation that we would rather not dwell on too much.}

5. "The authors have focused on major reassortment events between Yam and Vic lineages.  
If the definition of reassortment is expanded to include sub-clades within the Yam and Vic lineages (as appear quite clearly on the PB1 and PB2 trees), does the pattern still hold?  
Or is this only occurring at the larger scale?"

\textbf{This is a very insightful observation. The unusual pattern of PB1-PB2-HA co-reassortment could potentially be caused via physical means (e.g. packaging signals) and thus should be observable at the sub-clade level.
The alternative is linkage caused by selection, where sub-clade reassortants occur, but are not selected against as long as ``pure'' PB1-PB2-HA complexes are maintained.
We attempted to address this in supplementary information by using SPR distances (Figure S4), but because sub-clades are necessarily more closely related to each other we find that there is too much phylogenetic noise to investigate this in a satisfactory way using the current data.
It is our opinion that addressing sub-clade reassortment would require the development of new methods and a more suitable dataset.}

6. "p3, line 3: one of these should be PB2"

\textbf{This has been fixed. Thank you for spotting the mistake.}

7. "p12, line 20-21: 'do not appear to function well' sounds too virological; as we really have no idea about functional interactions, it's probably safer to stick with population dynamics and say 'do not appear to be fit'"

\textbf{We agree. It has been changed per reviewer's suggestion.}

%%% REVIEWER 2 %%%
\section*{Reviewer 2}

1.1. "Importance of influenza B: There is no need to underline this in the first two paragraphs.
Influenza B is important, just not as clinically relevant as influenza A.
The authors state in the introduction that influenza B has increased in prevalence recently, but this is just the last influenza season, and the majority of influenza cases and influenza seasons are still dominated by influenza A."

\textbf{This was pointed out by the editor too.
It has been addressed by toning down the language in the abstract and the introduction.}

2.1. "Figure 5: If I understand this figure correctly, the authors are using the BEAST MCC tree of each segment, looking at a particular point in the past, and counting the number of extant lineages that would be associated either with the original 1983 Yamagata virus or the original 1983 Victoria virus."

\textbf{Figure 5 shows the ratio at which Victoria and Yamagata sequences were isolated over time and is based on simply counting how many sequences within any given temporal bin and segment are derived from Victoria and Yamagata lineages.
It has caused confusion with reviewer 2 before, which we believe is due to our inter-changeable use of the word ``lineage'' when we actually mean ``sequence''.
This has been corrected in the text.}

2.2 "I understand what the `recent past' part of this chart means.
It means, for example, that the PA, NP, NA, and MP segments circulating today are all derived from the Yamagata lineage.
But what does the 1984 to 1987 part of this chart mean?"

\textbf{The 1984-1987 part of the chart does not exist, because we have omitted influenza seasons 1983/1984 to 1986/1987 from the figure after reviewer 2 made this comment during the previous review.
This was done because we only had single complete genomes for seasons 1983/1984 to 1986/1987.
However, we are certain that both Victoria and Yamagata lineage segments were still co-circulating in the 1994/1995 flu season.}

"Is it really true that at this time the majority or all of these segments were Victoria-type?
If yes, does this mean that when Vic and Yam split the initial diversification was in PB1-PB2-HA and the PA, NP, NA, and MP segments were largely conserved during this time?
If yes, should we really call these Victoria-type in 1984, or are they simply the ancestral type before the split occurred?"

\textbf{We believe these questions were made in ensuing misinterpretation of Figure 5.}

2.4. "Does sampling bias affect the results of Figure 5?"

\textbf{The apparent fixation of one lineage or the other will be completely robust to sampling.
In order to alleviate concerns about geographic sampling bias we have now included two additional supplementary Figures (Figures S8 and S9) which show the geographic distribution of the sequences in both primary and secondary datasets.}

2.5. "I believe we have enough influenza B sequences to reconstruct the post-2005 part of this chart."

\textbf{We are not sure what the reviewer meant with this question.
Figure 5 tracks the ratio of Victoria to Yamagata lineage sequences up to the 2011/2012 influenza season.}

2.6. "How much confidence do we have in 1995?
1985?"

\textbf{The confidence in the ratios displayed in Figure 5 is completely dependent on the numbers of genomes sequenced every year.
We provide the genome counts for each year in each segment at the top of Figure 5.
Before 1997, there are fewer than 10 genomes per season, suggesting caution in interpreting these frequencies.
Post 1997, we have more confidence.}

3.1. "Figure 8:  I could not understand this (granted, I am not a phylogeneticist).
But, something seems wrong here.
How can we compare PB1-PB1 TMRCAs to PB1-PB2 TMRCAs when PB1 and PB2 are not homologous?
I have probably misunderstood something here."

\textbf{Figure 8 has caused confusion during previous review, which we could not address directly earlier, but have included an additional figure (supplementary Figure S17) and elaborated on the $\Delta$TMRCA method in response.
Briefly, in the absence of reassortment we expect every segment in the genome to have the same tree, i.e. the tree of every segment should recapitulate the ``virus'' tree (analogous to ``species'' trees in diploid population genetics), including the dates of nodes.
Due to population bottlenecks influenza viruses go through each year we expect strains isolated at any given time to have descended from a single recent virus genome.
This descent from a single genome should therefore be reflected in the TMRCAs of all segments, the only exception being reassortment, which will dramatically alter the TMRCAs of the reassorted segment tree with respect to the background onto which it is reassorting.
We understand that this is a rather complicated and non-intuitive concept to grasp and we are willing to elaborate further on the underlying principle of the method in the text if necessary.}

3.2. "Are these TMRCAs of partial virus genomes?"

\textbf{We are not sure what the reviewer meant with this question.
We hope, however, that our description above and Figure S17 help to clarify this issue.}

3.3. "Should the caption read "Normalized mean TMRCA statistics for different segment pairs"?"

\textbf{We agree, the wording in the caption for Figure 8 is slightly misleading. It has been changed accordingly.}

4.1. "The discussion is long and somewhat disorganized.
I think the authors should spend some time improving the writing here, and shortening it.
The central theme of the paper is this PB1-PB2-HA complex, and I would structure the discussion to address the main questions surrounding this complex.  
Something like:

1.  Why are these segments so tightly linked at the genetic level?

1a.  Co-adaptation?
Possibly.
There is good functional evidence that PB1 and PB2 should be co-adapted.
There is some evidence, but much less, that HA should be co-adapted to PB1-PB2.

1b.  Simple divergence?
Possibly, especially if epistatic interactions cause an adapted PB1-PB2-HA complex to synergistically interact with the genotype defined by the other five segments.
But, we do see reassortment among the other five segments, so the evidence here is mixed."

\textbf{It is difficult to determine whether divergence or co-adaptation have caused linkage between PB1, PB2 and HA segments, and thus we have decided to combine these two scenarios into a single paragraph under the ``Linkage between PB1, PB2 and HA gene segments'' section in the discussion.
Our speculation on what might be the cause of the linkage is contained in the ``Potential mechanisms for reassortment incompatibility'' section that follows.}

4.2. "2. Is frequency-dependent selection allowing Flu B to maintain two HA types (in the same way that Flu A maintains two circulating subtypes in the population)?"

\textbf{We have decided to exclude superfluous speculation about hitchhiking and frequency-dependent selection on the HA segment in the discussion.}

4.3. "3. Are Flu B population sizes big or small?"

\textbf{We are not sure whether the reviewer refers to 'census' or effective population sizes.
It is widely accepted that influenza B viruses have a larger effective population size than influenza A viruses, but it would difficult to find experts who would hazard a guess as to the 'census' population size for influenza B viruses.}

4.4. "Could stochastic events have played a role in Flu B's population structure?
Does the lack of sequence data in the 1980s and early 1990s affect our conclusions?"

\textbf{The reviewer makes a very good point - population structure could indeed result in stochastic linkage between segments.
We have made note of this on page 12:
``Although the linkage between the two polymerase segments and HA is more difficult to explain, it is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000, offering ample time for the budding Victoria lineage to accumulate alleles causing reassortment incompatibility.
Without more genomic data from the past, however, it is difficult to estimate to what extent influenza B virus population structure contributed to the development of the current segment linkage.''}

4.5. "4. Will influenza B speciate?"

\textbf{This paragraph was moved to its own section titled ``Will influenza B viruses speciate?'' and has been made more concise and focused.}

4.6. "( Throughout the discussion "balancing selection" is used in a way I don't understand.
On page 15, if Vic and Yam are antigenically dissimilar, then it is frequency-dependent selection -- not balancing selection -- that prevents one genotype from driving the other to extinction. )"

\textbf{Apologies for the confusion, all references to balancing selection have been removed as irrelevant to the current manuscript.
We think that the maintenance of two subtypes in the influenza B population is an interesting question deserving of a separate discussion.}

\end{document}

### 2014-09-02 correspondence/responseLetter.tex
Fix quotations.

@@
 The word fixation is often used in a confusing manner. 
-For example, see line 28 page 5: "We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage." 
+For example, see line 28 page 5: ``We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.'' 
 See also the sentence on lines 48-49 of that page.
@@
 The first line of the abstract and the first paragraph of the introduction make it sound like influenza B is an increasingly worrisome public health problem. 
-Influenza B does not need to be hyped to be interesting, and this type of "journalism" can makes the reader wonder about the validity of everything that follows.
+Influenza B does not need to be hyped to be interesting, and this type of ``journalism'' can makes the reader wonder about the validity of everything that follows.
 \end{quote}
@@
 \section*{Reviewer 1}
-1. "Could the authors provide an estimate of how many reassortment events have occurred in these trees in total, based on the no. of transitions?"
+1. ``Could the authors provide an estimate of how many reassortment events have occurred in these trees in total, based on the no. of transitions?''
 
@@
 2.
-"It would be useful, as comparison, to include a Supplementary figure similar to Fig 2, but with branches colored to show reassortment events involving the other segments (PA, NP, NA, MP, NS -- obviously this is a lot of combinations and a lot of trees, so maybe just show a representative subset)."
+``It would be useful, as comparison, to include a Supplementary figure similar to Fig 2, but with branches colored to show reassortment events involving the other segments (PA, NP, NA, MP, NS -- obviously this is a lot of combinations and a lot of trees, so maybe just show a representative subset).''
 
@@
 
-3. "I didn't find any accession numbers or description of the data set in the Supplement.  
+3. ``I didn't find any accession numbers or description of the data set in the Supplement.  
 It would be useful to know how geographically representative the data set is.  
-It certainly is possible that in a spatially biased data set certain variants (as in those with mixed PB1-PB2-HA complexes) could be under-represented."
+It certainly is possible that in a spatially biased data set certain variants (as in those with mixed PB1-PB2-HA complexes) could be under-represented.''
 
@@
 
-4. "It seems that most of the mixed PB1-PB2-HA complexes are found earlier in the tree, closer to the root, and become less frequent as time goes on.  Is this worth expanding upon?"
+4. ``It seems that most of the mixed PB1-PB2-HA complexes are found earlier in the tree, closer to the root, and become less frequent as time goes on.  Is this worth expanding upon?''
 
@@
 
-5. "The authors have focused on major reassortment events between Yam and Vic lineages.  
+5. ``The authors have focused on major reassortment events between Yam and Vic lineages.  
 If the definition of reassortment is expanded to include sub-clades within the Yam and Vic lineages (as appear quite clearly on the PB1 and PB2 trees), does the pattern still hold?  
-Or is this only occurring at the larger scale?"
+Or is this only occurring at the larger scale?''
 
@@
 
-6. "p3, line 3: one of these should be PB2"
+6. ``p3, line 3: one of these should be PB2''
 
@@
 7.
-"p12, line 20-21: 'do not appear to function well' sounds too virological; as we really have no idea about functional interactions, it's probably safer to stick with population dynamics and say 'do not appear to be fit'"
+``p12, line 20-21: `do not appear to function well' sounds too virological; as we really have no idea about functional interactions, it's probably safer to stick with population dynamics and say 'do not appear to be fit'''
 
@@
 
-1.1. "Importance of influenza B: There is no need to underline this in the first two paragraphs.
+1.1. ``Importance of influenza B: There is no need to underline this in the first two paragraphs.
 Influenza B is important, just not as clinically relevant as influenza A.
-The authors state in the introduction that influenza B has increased in prevalence recently, but this is just the last influenza season, and the majority of influenza cases and influenza seasons are still dominated by influenza A."
+The authors state in the introduction that influenza B has increased in prevalence recently, but this is just the last influenza season, and the majority of influenza cases and influenza seasons are still dominated by influenza A.''
 
@@
 2.1.
-"Figure 5: If I understand this figure correctly, the authors are using the BEAST MCC tree of each segment, looking at a particular point in the past, and counting the number of extant lineages that would be associated either with the original 1983 Yamagata virus or the original 1983 Victoria virus."
+``Figure 5: If I understand this figure correctly, the authors are using the BEAST MCC tree of each segment, looking at a particular point in the past, and counting the number of extant lineages that would be associated either with the original 1983 Yamagata virus or the original 1983 Victoria virus.''
 
@@
 
-2.2 "I understand what the `recent past' part of this chart means.
+2.2 ``I understand what the `recent past' part of this chart means.
 It means, for example, that the PA, NP, NA, and MP segments circulating today are all derived from the Yamagata lineage.
-But what does the 1984 to 1987 part of this chart mean?"
+But what does the 1984 to 1987 part of this chart mean?''
 
@@
 
-"Is it really true that at this time the majority or all of these segments were Victoria-type?
+``Is it really true that at this time the majority or all of these segments were Victoria-type?
 If yes, does this mean that when Vic and Yam split the initial diversification was in PB1-PB2-HA and the PA, NP, NA, and MP segments were largely conserved during this time?
-If yes, should we really call these Victoria-type in 1984, or are they simply the ancestral type before the split occurred?"
+If yes, should we really call these Victoria-type in 1984, or are they simply the ancestral type before the split occurred?''
 
@@
 
-2.4. "Does sampling bias affect the results of Figure 5?"
+2.4. ``Does sampling bias affect the results of Figure 5?''
 
@@
 
-2.5. "I believe we have enough influenza B sequences to reconstruct the post-2005 part of this chart."
+2.5. ``I believe we have enough influenza B sequences to reconstruct the post-2005 part of this chart.''
 
@@
 
-2.6. "How much confidence do we have in 1995?
-1985?"
+2.6. ``How much confidence do we have in 1995? 1985?''
 
@@
 
-3.1. "Figure 8:  I could not understand this (granted, I am not a phylogeneticist).
+3.1. ``Figure 8:  I could not understand this (granted, I am not a phylogeneticist).
 But, something seems wrong here.
 How can we compare PB1-PB1 TMRCAs to PB1-PB2 TMRCAs when PB1 and PB2 are not homologous?
-I have probably misunderstood something here."
+I have probably misunderstood something here.''
 
@@
 
-3.2. "Are these TMRCAs of partial virus genomes?"
+3.2. ``Are these TMRCAs of partial virus genomes?''
 
@@
 
-3.3. "Should the caption read "Normalized mean TMRCA statistics for different segment pairs"?"
+3.3. ``Should the caption read `Normalized mean TMRCA statistics for different segment pairs'?''
 
@@
 
-4.1. "The discussion is long and somewhat disorganized.
+4.1. ``The discussion is long and somewhat disorganized.
 I think the authors should spend some time improving the writing here, and shortening it.
@@
 Possibly, especially if epistatic interactions cause an adapted PB1-PB2-HA complex to synergistically interact with the genotype defined by the other five segments.
-But, we do see reassortment among the other five segments, so the evidence here is mixed."
+But, we do see reassortment among the other five segments, so the evidence here is mixed.''
 
@@
 
-4.2. "2. Is frequency-dependent selection allowing Flu B to maintain two HA types (in the same way that Flu A maintains two circulating subtypes in the population)?"
+4.2. ``2. Is frequency-dependent selection allowing Flu B to maintain two HA types (in the same way that Flu A maintains two circulating subtypes in the population)?''
 
@@
 
-4.3. "3. Are Flu B population sizes big or small?"
+4.3. ``3. Are Flu B population sizes big or small?''
 
@@
 
-4.4. "Could stochastic events have played a role in Flu B's population structure?
-Does the lack of sequence data in the 1980s and early 1990s affect our conclusions?"
+4.4. ``Could stochastic events have played a role in Flu B's population structure?
+Does the lack of sequence data in the 1980s and early 1990s affect our conclusions?''
 
@@
 
-4.5. "4. Will influenza B speciate?"
+4.5. ``4. Will influenza B speciate?''
 
@@
 
-4.6. "( Throughout the discussion "balancing selection" is used in a way I don't understand.
-On page 15, if Vic and Yam are antigenically dissimilar, then it is frequency-dependent selection -- not balancing selection -- that prevents one genotype from driving the other to extinction. )"
+4.6. ``( Throughout the discussion `balancing selection' is used in a way I don't understand.
+On page 15, if Vic and Yam are antigenically dissimilar, then it is frequency-dependent selection -- not balancing selection -- that prevents one genotype from driving the other to extinction. )''
 

### 2014-09-02 correspondence/responseLetter.tex
Minor revisions to response.

@@
 
-\textbf{The 1984-1987 part of the chart does not exist, because we have omitted influenza seasons 1983/1984 to 1986/1987 from the figure after reviewer 2 made this comment during the previous review.
-This was done because we only had single complete genomes for seasons 1983/1984 to 1986/1987.
-However, we are certain that both Victoria and Yamagata lineage segments were still co-circulating in the 1994/1995 flu season.}
+\textbf{The 1984-1987 part of the chart was removed after reviewer 2 made this comment during the previous review.
+We have omitted influenza seasons 1983/1984 to 1986/1987 from our analysis because we only had single complete genomes for these seasons and we consequently lack information on relative Vic/Yam frequencies.}
 
@@
 
-2.5. ``I believe we have enough influenza B sequences to reconstruct the post-2005 part of this chart.''
-
-\textbf{We are not sure what the reviewer meant with this question.
-Figure 5 tracks the ratio of Victoria to Yamagata lineage sequences up to the 2011/2012 influenza season.}
-
-2.6. ``How much confidence do we have in 1995? 1985?''
+2.5. ``I believe we have enough influenza B sequences to reconstruct the post-2005 part of this chart. How much confidence do we have in 1995? 1985?''
 
@@
 
-\textbf{Figure 8 has caused confusion during previous review, which we could not address directly earlier, but have included an additional figure (supplementary Figure S17) and elaborated on the $\Delta$TMRCA method in response.
-Briefly, in the absence of reassortment we expect every segment in the genome to have the same tree, i.e.
-the tree of every segment should recapitulate the ``virus'' tree (analogous to ``species'' trees in diploid population genetics), including the dates of nodes.
-Due to population bottlenecks influenza viruses go through each year we expect strains isolated at any given time to have descended from a single recent virus genome.
+\textbf{Figure 8 has caused confusion during previous review, which we could not address directly earlier, but we have now included an additional figure (supplementary Figure S17) and elaborated on the $\Delta$TMRCA method in the text.
+Briefly, in the absence of reassortment we expect every segment in the genome to have the same tree, i.e.\ the tree of every segment should recapitulate the `virus' tree (analogous to `species' trees in diploid population genetics), including the dates of nodes.
+Different segments may evolve at different rates, but we should capture this through estimating separate molecular clocks across segments, so that dates of nodes should be consistent.
+Due to yearly population bottlenecks, we expect strains isolated at any given time to have descended from a single recent (within ~5-8 years) virus genome.
 This descent from a single genome should therefore be reflected in the TMRCAs of all segments, the only exception being reassortment, which will dramatically alter the TMRCAs of the reassorted segment tree with respect to the background onto which it is reassorting.
+The paper ``The genomic and epidemiological dynamics of human influenza A virus'' (Rambaut et al. 2008) uses a similar metric in the analysis of segment similarity in influenza H3N2.
 We understand that this is a rather complicated and non-intuitive concept to grasp and we are willing to elaborate further on the underlying principle of the method in the text if necessary.}
@@
 
-\textbf{We are not sure whether the reviewer refers to 'census' or effective population sizes.
-It is widely accepted that influenza B viruses have a larger effective population size than influenza A viruses, but it would difficult to find experts who would hazard a guess as to the 'census' population size for influenza B viruses.}
+\textbf{We are not sure whether the reviewer refers to `census' or effective population sizes.
+It is widely accepted that influenza B viruses have a larger effective population size than influenza A viruses, but it would difficult to find experts who would hazard a guess as to the `census' population size for influenza B viruses.}
 

### 2014-09-02 correspondence/responseLetter.tex
Fix typo.

@@
 \textbf{We are not sure whether the reviewer refers to `census' or effective population sizes.
-It is widely accepted that influenza B viruses have a larger effective population size than influenza A viruses, but it would difficult to find experts who would hazard a guess as to the `census' population size for influenza B viruses.}
+It is widely accepted that influenza B viruses have a larger effective population size than influenza A viruses, but it would be difficult to find experts who would hazard a guess as to the `census' population size for influenza B viruses.}
 

### 2014-09-02 correspondence/responseLetter.tex
A few edits and comments for the response letter.

@@
 We appreciate the time they have taken to review the manuscript and the acknowledgment of its value.
-We agree that the discussion, as it was, contained mostly irrelevant information and speculation.
-The discussion has been considerably slimmed down to the most essential information and reworded in places.
+We agree that the discussion, as it was, contained superfluous and speculative text.
+The discussion has been considerably slimmed down to the core information and reworded in places.
 We have additionally altered the terminology that caused confusion, e.g. the word ``fixation'' is now explicitly refered to as a population genetics term and used only where applicable.
@@
 
-Point-by-point review responses are attached, as well as a diffed PDF detailing all the changes that have been made since the initial submission.
+Point-by-point review responses are attached, as well as a 'diffed' PDF detailing all the changes that have been made since the initial submission.
 
@@
 \textbf{We have added a new supplementary figure (Figure S10), which shows the mean number of transitions for each trait/label as well as their 95\% HPDs in each tree.
-We advise to interpret the new figure with caution - many successful reassortment events we have observed involve more than one segment (see Figure 6) and thus these counts are not independent.}
+We advise interpreting the new figure with caution - many successful reassortment events we have observed involve more than one segment (see Figure 6) and thus these counts are not independent.}
 
@@
 
-\textbf{We believe that although this is an interesting point, it is outside the scope of the current manuscript.
+\textbf{We believe that, although this is an interesting point, it is beyond the scope of the current manuscript.
 Additionally, the complete reassortment history of all segments is available in Figure 6, which can be combined with Figure 2 to reveal the phylogenetic position of all reassorting lineages.
@@
 
-\textbf{Apologies for this. We have a separate file listing all sequence accession numbers from primary (GISAID) and secondary (GenBank) datasets but forgot to include it in the initial submission.
+\textbf{Apologies for this. We have a separate file listing all sequence accession numbers from GISAID and GenBank datasets but forgot to include it in the initial submission.
 These files are available at \url{https://github.com/evogytis/fluB/tree/master/data/acknowledgement\%20tables}.
@@
 We attempted to address this in supplementary information by using SPR distances (Figure S4), but because sub-clades are necessarily more closely related to each other we find that there is too much phylogenetic noise to investigate this in a satisfactory way using the current data.
-It is our opinion that addressing sub-clade reassortment would require the development of new methods and a more suitable dataset.}
+It is our opinion that addressing sub-clade reassortment would require the development of new methods and a more fine-scaled dataset.}
 
@@
 This has been corrected in the text.}
+%% AR - what do you mean with 'caused confusion with reviewer 2 before'?
 
@@
 \textbf{We believe these questions were made in ensuing misinterpretation of Figure 5.}
+%% AR - is this referring to 2.1, above?
 
@@
 
-\textbf{The apparent fixation of one lineage or the other will be completely robust to sampling.
+\textbf{The apparent fixation of one lineage or the other will be robust to sampling. %% AR - want to expand on this?
 In order to alleviate concerns about geographic sampling bias we have now included two additional supplementary Figures (Figures S8 and S9) which show the geographic distribution of the sequences in both primary and secondary datasets.}
@@
 The paper ``The genomic and epidemiological dynamics of human influenza A virus'' (Rambaut et al. 2008) uses a similar metric in the analysis of segment similarity in influenza H3N2.
-We understand that this is a rather complicated and non-intuitive concept to grasp and we are willing to elaborate further on the underlying principle of the method in the text if necessary.}
+%We understand that this is a rather complicated and non-intuitive concept and we are willing to elaborate further on the underlying principle of the method in the text if necessary.
+%% AR Don't offer to do something conditional on being responded to by the referee - work on the basis that this will be accepted without revision, i.e., persuade the editor that 
+%% you have completely handled the referees' concerns.
+}
 
@@
 \textbf{We are not sure what the reviewer meant with this question.
+%% AR Should we just say are TMRCAs of individual segments? Don't say we don't know what the reviewer means.
 We hope, however, that our description above and Figure S17 help to clarify this issue.}

### 2014-09-03 correspondence/responseLetter.tex
Address @rambaut comments.

@@
 \textbf{Figure 5 shows the ratio at which Victoria and Yamagata sequences were isolated over time and is based on simply counting how many sequences within any given temporal bin and segment are derived from Victoria and Yamagata lineages.
-It has caused confusion with reviewer 2 before, which we believe is due to our inter-changeable use of the word ``lineage'' when we actually mean ``sequence''.
+Reviewer 2 has had difficulties understanding Figure 5 before, which we believe is due to our inter-changeable use of the word ``lineage'' when we actually mean ``sequence''.
 This has been corrected in the text.}
-%% AR - what do you mean with 'caused confusion with reviewer 2 before'?
 
@@
 
-``Is it really true that at this time the majority or all of these segments were Victoria-type?
+2.3 ``Is it really true that at this time the majority or all of these segments were Victoria-type?
 If yes, does this mean that when Vic and Yam split the initial diversification was in PB1-PB2-HA and the PA, NP, NA, and MP segments were largely conserved during this time?
@@
 %% AR - is this referring to 2.1, above? 
+%% GD - reviewer 2's question 2 was a fairly large wall of text regarding Figure 5, so I split it into manageable chunks.
 
@@
 
-\textbf{The apparent fixation of one lineage or the other will be robust to sampling. %% AR - want to expand on this?
+\textbf{The apparent fixation of one lineage or the other will be robust to sampling.
+All influenza B viruses isolated over the last 13 years have had PA, NP, NA and MP segments derived from Yamagata lineage and NS segments derived from Victoria lineage.
+We find it highly unlikely that the sister lineages of those segments are still circulating in the human population.
 In order to alleviate concerns about geographic sampling bias we have now included two additional supplementary Figures (Figures S8 and S9) which show the geographic distribution of the sequences in both primary and secondary datasets.}
@@
 This descent from a single genome should therefore be reflected in the TMRCAs of all segments, the only exception being reassortment, which will dramatically alter the TMRCAs of the reassorted segment tree with respect to the background onto which it is reassorting.
-The paper ``The genomic and epidemiological dynamics of human influenza A virus'' (Rambaut et al. 2008) uses a similar metric in the analysis of segment similarity in influenza H3N2.
-%We understand that this is a rather complicated and non-intuitive concept and we are willing to elaborate further on the underlying principle of the method in the text if necessary.
-%% AR Don't offer to do something conditional on being responded to by the referee - work on the basis that this will be accepted without revision, i.e., persuade the editor that 
-%% you have completely handled the referees' concerns.
-}
+The paper ``The genomic and epidemiological dynamics of human influenza A virus'' (Rambaut et al. 2008) uses a similar metric in the analysis of segment similarity in influenza H3N2.}
 
@@
 
-\textbf{We are not sure what the reviewer meant with this question.
-%% AR Should we just say are TMRCAs of individual segments? Don't say we don't know what the reviewer means.
-We hope, however, that our description above and Figure S17 help to clarify this issue.}
+\textbf{The TMRCAs are of isolate pairs in individual segments.
+We hope that our description above and Figure S17 have clarified this issue.}
 

### 2014-09-03 correspondence/responseLetter.tex
Alter the wording a bit.

@@
 \textbf{Figure 5 shows the ratio at which Victoria and Yamagata sequences were isolated over time and is based on simply counting how many sequences within any given temporal bin and segment are derived from Victoria and Yamagata lineages.
-Reviewer 2 has had difficulties understanding Figure 5 before, which we believe is due to our inter-changeable use of the word ``lineage'' when we actually mean ``sequence''.
+There has been some confusion regarding Figure 5, which we believe is due to our inter-changeable use of the word ``lineage'' when we actually mean ``sequence''.
 This has been corrected in the text.}
